---
title: "The Baldwin Effect: within-lifetime learning accelerating evolution"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Baldwin Effect: within-lifetime learning accelerating evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Background

The Baldwin Effect (Baldwin 1896; Simpson 1953) is a neo-Darwinian mechanism by
which within-lifetime learning can accelerate genetic evolution without
Lamarckian inheritance. The logic has four steps:

1. In a novel or variable environment, genetically fixed behaviours are
   suboptimal. Individuals that can adjust behaviour within their lifetime
   survive long enough to reproduce.
2. Because learners reproduce more, genetic variants that predispose faster or
   deeper learning spread — learning ability itself becomes selected.
3. Over generations, the fitness benefit of learning shrinks as the genome
   gradually encodes the once-learned behaviour directly.
4. The endpoint is *genetic assimilation* (Waddington 1942): what was once
   learned is now innate. The organism no longer needs to learn it.

In `clade`, the BNN brain type (`brain_type = "bnn"`) makes this process
directly observable. Each agent's neural network carries a prior standard
deviation σ over its synaptic weights. **High σ = exploration** (wide
Thompson sampling, broad behavioural flexibility). **Low σ = canalization**
(narrow prior, behaviour largely determined by the genome). The population
mean `mean_prior_sigma` is therefore a continuous readout of where evolution
has placed the population on the exploration–assimilation axis.

## Demonstration

Two conditions, 1000 ticks, 150 agents on a 25×25 grid (`grass_rate = 0.10`).
Two readouts: `mean_prior_sigma` (the assimilation signal — high σ = exploration,
low σ = canalization) and `genetic_diversity` (mean pairwise genome distance).

```{r setup}
library(clade)
library(ggplot2)
library(dplyr)
library(patchwork)

base <- default_specs()
base$brain_type    <- "bnn"
base$n_agents_init <- 150L
base$max_ticks     <- 1000L
base$grid_rows     <- 25L
base$grid_cols     <- 25L
base$grass_rate    <- 0.10
base$random_seed   <- 42L
```

```{r run-demo}
# Condition 1: BNN baseline — genetic evolution only
env_base <- run_alife(base)

# Condition 2: BNN + epigenetics (TEI)
# Within-lifetime learning narrows the BNN prior; TEI transmits the
# narrowed priors to offspring. No stable landscape required.
s_epi <- base
s_epi$epigenetics            <- TRUE
s_epi$epigenetic_inheritance <- 0.5
s_epi$methylation_rate       <- 0.2
env_epi <- run_alife(s_epi)
```

```{r plot-demo, fig.width=8, fig.height=6}
df <- bind_rows(
  transform(get_run_data(env_base)$ticks, condition = "BNN baseline"),
  transform(get_run_data(env_epi)$ticks,  condition = "BNN + epigenetics")
)
df$condition <- factor(df$condition,
  levels = c("BNN baseline", "BNN + epigenetics"))

pal <- c("BNN baseline"      = "#377eb8",
         "BNN + epigenetics" = "#e6550d")
lty <- c("BNN baseline"      = "dashed",
         "BNN + epigenetics" = "solid")

pA <- ggplot(df, aes(t, mean_prior_sigma, colour = condition, linetype = condition)) +
  geom_hline(yintercept = 0.5, colour = "grey70", linetype = "dotted", linewidth = 0.5) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_linetype_manual(values = lty, name = NULL) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(limits = c(0, 0.55)) +
  labs(x = NULL, y = expression("Mean prior " * sigma)) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom", legend.key.width = unit(1.4, "cm"))

pB <- ggplot(df, aes(t, genetic_diversity, colour = condition, linetype = condition)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_linetype_manual(values = lty, name = NULL) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Tick", y = "Genetic diversity\n(mean pairwise genome distance)") +
  theme_classic(base_size = 11) + theme(legend.position = "none")

pA / pB + plot_layout(heights = c(2, 1))
```

```{r fig-canalization-demo, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "Top: mean BNN prior sigma over 1000 ticks", fig.cap="Top: mean BNN prior sigma over 1000 ticks. Baseline (blue dashed): sigma rises to the 0.5 exploration ceiling — no canalization in a dynamic foraging world. Epigenetics (orange): sigma declines toward zero as TEI propagates within-lifetime learning across generations. Bottom: genetic diversity (mean pairwise genome distance) rises in both conditions — epigenetic canalization of sigma does not cause a selective sweep of the genome; different genotypes can encode the same canalized behaviour."}
FIG("bnn_canalization_demo")
```

**Baseline:** σ rises to its ceiling (0.5) and stays there. The foraging landscape
shifts continuously — grass grows stochastically, population density creates local
competition, and no single fixed policy consistently outcompetes broad sampling.
Exploration is the evolutionarily stable strategy. The Baldwin Effect does *not*
occur in this default dynamic world.

**Epigenetics:** σ declines from 0.5 toward zero over 1000 ticks. The mechanism is
a two-step chain: (1) within-lifetime learning contracts each agent's weight
posterior during its lifetime, narrowing σ as the agent forages; (2) transgenerational
epigenetic inheritance (TEI) passes these narrowed priors to offspring. Each
generation starts more canalized than the last, and selection reinforces the
direction. **This is genetic assimilation via the Baldwin Effect** — but achieved
through a Lamarckian-adjacent transmission mechanism (TEI), not pure Darwinian
selection on a stable fitness peak.

**Genetic diversity:** Both conditions show *rising* diversity over 1000 ticks.
This is a notable result: epigenetic canalization of σ does not sweep the genome.
Different weight matrices can all encode the same low-σ (canalized) foraging
policy, so genetic variation is maintained even as the phenotypic trait (σ)
converges.

The five systematic experiments below ask: *when* does canalization occur, and
what disrupts it?

---

### The Hinton & Nowlan (1987) setup: `fixed_patch`

The original computational demonstration of the Baldwin Effect (Hinton & Nowlan
1987) used a needle-in-a-haystack fitness landscape: one correct genotype gave
high fitness, all others gave near-zero fitness, and learning (random search)
provided an intermediate fitness signal. Canalization occurred because agents
with *more* correct alleles needed less learning to find the needle — σ declined
as correct alleles accumulated.

`clade` provides a `fixed_patch` module that implements the closest spatial
equivalent: a permanent high-value resource cell that never moves. Agents that
genetically encode navigation toward the patch need less within-lifetime
exploration, creating selection pressure against high σ.

```{r fixed-patch-setup}
# Hinton & Nowlan (1987) analogue: sparse background, large fixed patch
s_hn <- default_specs()
s_hn$brain_type         <- "bnn"
s_hn$grass_rate         <- 0.02   # sparse — patch becomes essential
s_hn$fixed_patch        <- TRUE
s_hn$fixed_patch_value  <- 5.0
s_hn$fixed_patch_radius <- 3L     # 7×7 = 49 cells at grid centre
s_hn$fixed_patch_x      <- 13L
s_hn$fixed_patch_y      <- 13L
s_hn$max_ticks          <- 2000L
env_hn <- run_alife(s_hn)
```

The carrying capacity of the environment must match the patch size: with 49
patch cells and a sparse background, a starting population of 50–80 agents
allows selection to act without crashing to extinction. Parameter tuning
(patch radius, `grass_rate`, `n_agents_init`) determines whether the
Hinton–Nowlan result (σ declines) or a population crash occurs — reflecting
the sensitivity of genetic assimilation to the fitness landscape geometry.

The five systematic experiments below used `fixed_patch = FALSE` (dynamic
world only) to map the boundary conditions of canalization in clade's default
environment.

## Key parameters

| Parameter                | Default        | Role                                              |
|--------------------------|----------------|---------------------------------------------------|
| `brain_type`             | `"bnn"`        | Enables σ as an evolvable exploration parameter   |
| `rl_mode`                | `"none"`       | `"actor_critic"` enables within-lifetime REINFORCE|
| `rl_update_freq`         | `1L`           | RL update every N ticks                           |
| `epigenetics`            | `FALSE`        | Enables methylation marks + TEI                   |
| `epigenetic_inheritance` | `0.5`          | Probability each methylation mark is transmitted  |
| `methylation_rate`       | `0.1`          | Probability a reward event drives methylation     |
| `demethylation_rate`     | `0.02`         | Background demethylation rate per tick            |
| `social_learning`        | `FALSE`        | Prestige-biased copying of output-layer weights   |
| `fixed_patch`            | `FALSE`        | Enable a permanent high-value resource cell (H&N 1987 setup) |
| `fixed_patch_value`      | `5.0`          | Grass value maintained at patch cells each tick   |
| `fixed_patch_x`          | `NA`           | Column index of patch centre (`NA` = grid centre) |
| `fixed_patch_y`          | `NA`           | Row index of patch centre (`NA` = grid centre)    |
| `fixed_patch_radius`     | `0L`           | Chebyshev radius: 0 = 1 cell, 1 = 3×3, 2 = 5×5  |

## Systematic experiments: when does canalization occur?

The five experiments below ask a sharper question than the demonstration above:
*under what conditions does canalization occur, and what prevents it?*

The BNN brain is uniquely suited to this question because its prior standard
deviation σ (`mean_prior_sigma` in the tick log) is a direct, continuous readout
of the population's position on the exploration–assimilation axis. When σ
declines over evolutionary time, agents are becoming more certain about their
action-selection policy — their learned behaviour is being absorbed into the
genome. When σ rises or plateaus, selection is actively maintaining cognitive
flexibility: exploration is the evolutionarily stable strategy (ESS).

**A σ slope of zero is not neutral.** In a world where exploration is costly
(Thompson sampling with high σ burns energy on uncertain actions), zero slope
means selection is actively maintaining a costly trait. It does not mean the
population is indifferent. The key comparison is always against the prediction
that σ *declines* — i.e., that canalization occurs.

The five experiments manipulate environmental predictability, run length, brain
architecture, and social transmission mechanisms in turn. The designs are 3–45
runs each; all results shown are from pre-computed data (no Julia session needed).

---

### Experiment 1 — Environmental stability gradient

**Question.** Mayley (1996) showed that the Baldwin Effect operates fastest when
the fitness landscape has a single, stable peak. Does varying resource abundance
and temporal variability produce any condition where canalization occurs?

**Design.** 3 × 3 factorial: `grass_rate` ∈ {0.05, 0.10, 0.20} ×
`seasonal_amplitude` ∈ {0, 0.4, 0.8}; 5 seeds × 1000 ticks = 45 runs.
For each condition and seed, a linear regression of `mean_prior_sigma ~ tick`
was fitted; the slope is the measure of canalization (negative = assimilation,
positive = exploration selected).

**What we found.** Every one of the 9 conditions produced a positive σ slope —
no canalization in any cell. All 9 converged to the ceiling (σ = 0.5) by
tick 1000.

| grass_rate | seasonal_amp | mean slope | final σ |
|---|---|---|---|
| 0.05 | 0.0 | +2.51 × 10⁻⁴ | 0.50 |
| 0.05 | 0.4 | +5.54 × 10⁻⁴ ‡ | 0.40 |
| 0.05 | 0.8 | +2.45 × 10⁻⁴ | 0.50 |
| 0.10 | 0.0 | +1.98 × 10⁻⁴ | 0.50 |
| 0.10 | 0.4 | +1.92 × 10⁻⁴ | 0.50 |
| 0.10 | 0.8 | +1.83 × 10⁻⁴ | 0.50 |
| 0.20 | 0.0 | +1.49 × 10⁻⁴ | 0.50 |
| 0.20 | 0.4 | +1.43 × 10⁻⁴ | 0.50 |
| **0.20** | **0.8** | **+1.38 × 10⁻⁴** | **0.50** |

‡ Population-crash artefact (high variance across seeds), not canalization.

Crucially, the ordering is the **inverse** of the Baldwin Effect prediction:
resource scarcity (grass = 0.05) produced the *steepest* positive slopes, while
abundance (grass = 0.20) produced the shallowest — yet still positive. When
foraging is difficult, agents with high σ that can explore broadly have a strong
survival advantage; tight priors are fatal. The bold row is the most
canalization-favourable cell tested; it still reaches the ceiling.

```{r fig-exp1, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "Phase diagram of mean_prior_sigma slope (×10⁻⁴) across 9 resource-abundance × seasonality conditions", fig.cap="Phase diagram of mean_prior_sigma slope (×10⁻⁴) across 9 resource-abundance × seasonality conditions. All cells are positive (red). The predicted canalization signal (green) does not appear in any cell."}
FIG("bnn_exp1_stability")
```

**Interpretation.** In a competitive foraging world with spatial heterogeneity
and population-density feedbacks, no single foraging policy is universally
optimal. The fitness landscape shifts continuously, so there is no stable peak
for canalization to track. Selection maintains the ability to learn (high σ) as
the ESS across all resource and seasonality levels tested.

---

### Experiment 2 — Is the sigma ceiling an artefact of run length?

**Question.** σ plateaus at 0.5 within 300 ticks in the baseline. Could
canalization emerge given substantially longer evolutionary time — 2000 or
5000 ticks?

**Design.** The most canalization-favourable condition from Experiment 1 (grass
= 0.20, seasonal = 0.8): 5 seeds each at 1000, 2000, and 5000 ticks = 15 runs.

**What we found.** The σ slope decelerates markedly with run length:

| Run length | Mean slope | Final σ |
|---|---|---|
| 1000 ticks | +1.38 × 10⁻⁴ | 0.50 (ceiling) |
| 2000 ticks | +3.84 × 10⁻⁵ | 0.50 (ceiling) |
| 5000 ticks | +6.54 × 10⁻⁶ | 0.50 (ceiling) |

The slope shrinks 21-fold from 1000 to 5000 ticks. But `final_sigma` remains at
0.50 in all cases. This deceleration is **ceiling saturation**: σ presses against
its upper bound (0.5), so the observable slope must approach zero as the ceiling
is reached — regardless of whether canalization is occurring. If the ceiling
parameter (`prior_sigma_max`) were raised, σ would continue rising.

```{r fig-exp2, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "Sigma trajectories at 1000, 2000, and 5000 ticks (mean ± SE across 5 seeds). All runs reach the", fig.cap="Sigma trajectories at 1000, 2000, and 5000 ticks (mean ± SE across 5 seeds). All runs reach the 0.5 ceiling; the slope decelerates 21-fold but never reverses. The dashed line marks the ceiling."}
FIG("bnn_exp2_runlength")
```

**Interpretation.** Run length is not the limiting factor. The near-zero slope
at 5000 ticks looks superficially like the onset of canalization but is a
measurement artefact of the ceiling constraint. The biological conclusion is
unchanged: no canalization occurs even with five times the evolutionary time of
the baseline.

---

### Experiment 3 — Brain architecture comparison

**Question.** Does the Baldwin Effect require the BNN specifically, or would any
within-lifetime adaptation mechanism (here, REINFORCE RL) produce the same σ
dynamics?

**Design.** 4 brain conditions × 5 seeds × 1000 ticks = 20 runs, in the most
canalization-favourable environment:

| Condition | `brain_type` | `rl_mode` | Learning mechanism |
|---|---|---|---|
| BNN only | `"bnn"` | `"none"` | Thompson sampling (σ-driven) |
| BNN + RL | `"bnn"` | `"actor_critic"` | Thompson sampling + REINFORCE |
| ANN + RL | `"ann"` | `"actor_critic"` | REINFORCE only |
| ANN null | `"ann"` | `"none"` | None (genetic evolution only) |

**What we found.** Three results:

1. **RL has no effect on σ dynamics.** BNN-only and BNN+RL produce
   indistinguishable σ trajectories — the same slope, the same ceiling arrival
   tick. Adding REINFORCE on top of Thompson sampling does not alter how
   exploration evolves.

2. **RL provides no energy benefit at 1000 ticks.** ANN+RL and ANN-null
   both end at 154.97 mean energy — identical to 5 significant figures.
   REINFORCE gradient updates offer no measurable foraging advantage over
   genetic evolution of ANN weights at this timescale.

3. **BNN pays a 17% energy penalty for exploration, yet exploration is still the
   ESS.** BNN agents end at 128 energy versus ANN agents at 155. The metabolic
   cost of broad Thompson sampling is real and large — but selection still drives
   σ to ceiling. This is the clearest evidence that exploration is genuinely
   the ESS in this world: it is favoured even when it is costly.

```{r fig-exp3, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "Left: sigma trajectories for BNN-only and BNN+RL are indistinguishable — RL does not affect sigma", fig.cap="Left: sigma trajectories for BNN-only and BNN+RL are indistinguishable — RL does not affect sigma dynamics. Right: final mean energy by condition — BNN pays a 17% energy cost relative to ANN, yet sigma rises to ceiling in both BNN conditions."}
FIG("bnn_exp3_brains")
```

**Interpretation.** σ is purely a BNN property: ANN agents have no σ to
evolve. But within BNN conditions, neither RL nor social information changes the
evolutionary trajectory of σ. The driving force is the foraging fitness
differential between high-σ and low-σ agents — a differential that REINFORCE
does not alter because it acts on policy weights, not on prior width.

---

### Experiment 4 — Social modifiers: can they break the ceiling?

**Question.** If genetic evolution and within-lifetime RL both fail to produce
canalization, can social transmission mechanisms do so? Three candidates:
kin-based altruism (energy transfers to relatives), prestige-biased social
learning (copying successful neighbours' policy weights), and epigenetic
inheritance (transgenerational transmission of within-lifetime weight
contractions).

**Design.** 4 modules × 2 environments × 5 seeds × 1000 ticks = 40 runs.
The two environments are stable-abundant (grass = 0.20, seasonal = 0.0) and
scarce-stable (grass = 0.05, seasonal = 0.0).

**What we found.**

*Stable abundant (grass = 0.20, seasonal = 0):*

| Module | σ slope | Final σ |
|---|---|---|
| BNN + epigenetics | +6.6 × 10⁻⁵ | **0.342** |
| BNN + kin selection | +1.5 × 10⁻⁴ | 0.500 (ceiling) |
| BNN + social learning | +1.7 × 10⁻⁴ | 0.500 (ceiling) |
| BNN baseline | +1.7 × 10⁻⁴ | 0.500 (ceiling) |

*Scarce stable (grass = 0.05, seasonal = 0):*

| Module | σ slope | Final σ |
|---|---|---|
| BNN + epigenetics | +5.5 × 10⁻⁴ | **0.140** |
| BNN + social learning | +6.2 × 10⁻⁴ | 0.300 |
| BNN baseline | +4.5 × 10⁻⁴ | 0.400 |
| BNN + kin selection | +5.6 × 10⁻⁴ | 0.400 |

**Epigenetic inheritance is the only mechanism that substantially reduces σ.**
In the stable-abundant environment, it reduces final σ from 0.500 to 0.342 —
32% below ceiling. In the scarce-stable environment the effect is stronger: σ
reaches only 0.140, a 72% reduction from the 0.5 ceiling and 65% below the
baseline final value of 0.400.

```{r fig-exp4, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "Final mean prior sigma by social module and environment", fig.cap="Final mean prior sigma by social module and environment. Epigenetics (red dot) is the only mechanism that substantially reduces sigma below the 0.5 ceiling (dashed line). All other modules are indistinguishable from the baseline in the stable environment and provide only moderate reductions in the scarce environment."}
FIG("bnn_exp4_social")
```

**Interpretation.** The epigenetic mechanism here is transgenerational epigenetic
inheritance (TEI): when an agent contracts its weight posterior during its
lifetime (σ narrows as it learns), offspring inherit partially-contracted priors
with probability `tei_prob`. This is a Lamarckian shortcut — within-lifetime
narrowing is transmitted directly to descendants, bypassing the slow route that
requires mutations to fix via selection over many generations. In contrast:

- **Social learning** had no significant effect in the stable environment
  because copying output-layer weights from successful neighbours propagates
  *what* to do, not *how certain to be about it*. σ is a prior width, not a
  policy weight, and social copying does not transmit it.
- **Kin selection** was predicted to retard canalization (energy subsidies
  buffer the cost of exploration, maintaining high σ). The effect was
  negligible in the stable environment; kin altruism had no detectable effect
  on σ dynamics at these parameter values.

The prediction was reversed: epigenetics produced the largest effect across all
four experiments. Social learning, predicted to accelerate assimilation via rapid
spread of optimal policies, had no effect on σ.

---

### Experiment 5 — MAP-Elites: mapping the canalization-compatible parameter space

**Question.** The experiments above all used the same BNN baseline parameters.
Are there *any* parameter combinations — across the full space that `clade` can
reach — that produce low-σ solutions? And if so, do those solutions coexist with
high genetic diversity (the Baldwin Effect proper) or with low genetic diversity
(drift-driven pseudo-canalization)?

**Design.** `search_map_elites()` with a 11 × 11 archive over σ × genetic
diversity space (both 0–0.5, step 0.05 = 121 cells), 150 iterations, 500-tick
runs per evaluation. The archive records the highest-scoring parameter
configuration found for each (σ, gd) combination.

**What we found.** 20 of 121 cells were filled (16.5%). Low-σ solutions were
found — σ as low as 0.028 — but every low-σ cell had very low genetic diversity
(gd < 0.15 for all cells with σ < 0.3). The σ–gd correlation across the archive
was +0.593. The five highest-scoring cells (highest genetic diversity) all had σ
> 0.40.

**No low-σ + high-gd cell was found anywhere in the archive.**

```{r fig-exp5, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "MAP-Elites archive: each point is a filled cell (20 of 121 total), coloured by genetic diversity", fig.cap="MAP-Elites archive: each point is a filled cell (20 of 121 total), coloured by genetic diversity. Low sigma is only found at low genetic diversity (bottom-left region). The bottom-right quadrant — low sigma + high diversity, required for the Baldwin Effect proper — is empty."}
FIG("bnn_exp5_mapelites")
```

**Interpretation.** The Baldwin Effect *proper* requires low σ coexisting with
high genetic diversity: the genome has converged on specific adaptive weight
configurations (canalization at key loci) while maintaining diversity at others.
MAP-Elites found no such solution in the accessible parameter space.

Low σ in the archive is entirely explained by **genetic drift to fixation**:
when a small or bottlenecked population becomes genetically homogeneous (low gd),
heterozygosity collapses everywhere, driving σ toward zero as a byproduct of
fixation, not of selection for canalized behaviour. This is *pseudo-canalization*
via drift — it looks like genetic assimilation from the σ readout alone, but the
mechanism is demographic, not adaptive.

The MAP-Elites result closes the argument. Epigenetics (Experiment 4) provides
the only route to substantially reduced σ in viable, genetically diverse
populations. Without it, the competitive foraging world in `clade` consistently
selects for maintained exploration.

---

## Addendum: a calibrated regime where canalization emerges

After the five original experiments, we ran clade's built-in CMA-ES
search (`clade::search_cmaes`, driven by the harness under
[`dev/audit/calibration/`](https://github.com/itchyshin/clade/tree/main/dev/audit/calibration))
against the objective "negative slope of `mean_prior_sigma`" — i.e.
reward any parameter regime that makes σ *narrow*. Full writeup at
[[`dev/audit/calibration/RESULTS.md`](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md)](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md).

The search found a regime that does produce canalization:

```{r baldwin-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES.
s <- default_specs()
s$brain_type              <- "bnn"
s$n_agents_init           <- 80L
s$max_ticks               <- 400L
s$grass_rate              <- 0.027   # default 0.05 — scarcer resources
s$learning_rate_init_mean <- 0.007   # default 0.01 — slower RL update
# env <- run_alife(s)   # uncomment to run the calibrated regime
```

The resulting σ trajectory has a small but consistently **negative**
slope (the observable fitness ratio is 1.2× over the default baseline
in the calibration report, a sign-flip from the exploration-dominated
regime). Canalization is weaker than Hinton & Nowlan's prediction
because even at these specs, the clade foraging world does not have a
truly stable peak — it has a shifting optimum driven by population
competition. But the result confirms that canalization **is**
accessible; it just requires scarcer resources and slower within-
lifetime learning than the displayed defaults.

Biological reading: in environments where foraging is costly and
learning is slow, selection favours agents whose genome already
encodes a close-to-optimal policy. Agents with wider priors (high σ)
waste energy sampling low-reward actions and are outcompeted. In the
abundant / fast-learning regime, high σ is protective because
exploration finds good cells before starvation. The sign of selection
on σ therefore hinges on the ratio of exploration cost to exploration
benefit — exactly the Mayley (1996) framing.

The five original experiments establish that at *commonly used*
parameter defaults, canalization does not occur. The addendum
establishes that canalization is not mechanistically impossible in
clade — it just sits in a narrower parameter neighbourhood than the
broader exploration regime.

---

## References

Baldwin, J.M. (1896) A new factor in evolution. *American Naturalist*
30(354):441–451.

Henrich, J. & Gil-White, F.J. (2001) The evolution of prestige: freely
conferred deference as a mechanism for enhancing the benefits of cultural
transmission. *Evolution and Human Behavior* 22(3):165–196.

Jablonka, E. & Lamb, M.J. (2005) *Evolution in Four Dimensions: Genetic,
Epigenetic, Behavioral, and Symbolic Variation in the History of Life.*
MIT Press, Cambridge MA.

Laland, K.N. (2004) Social learning strategies. *Learning and Behavior*
32(1):4–14.

Rendell, L. et al. (2010) Why copy others? Insights from the social learning
strategies tournament. *Science* 328(5975):208–213.

Simpson, G.G. (1953) The Baldwin Effect. *Evolution* 7(2):110–117.

Sutton, R.S. & Barto, A.G. (2018) *Reinforcement Learning: An Introduction.*
2nd ed. MIT Press, Cambridge MA.

Waddington, C.H. (1942) Canalization of development and the inheritance of
acquired characters. *Nature* 150:563–565.

Williams, R.J. (1992) Simple statistical gradient-following algorithms for
connectionist reinforcement learning. *Machine Learning* 8(3–4):229–256.
