---
title: "The Baldwin Effect: within-lifetime learning accelerating evolution"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Baldwin Effect: within-lifetime learning accelerating evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Background

The Baldwin Effect (Baldwin 1896; Simpson 1953) is a neo-Darwinian mechanism by
which phenotypic plasticity — the ability of an organism to adjust its behaviour
within its lifetime — can accelerate genetic evolution without Lamarckian
inheritance.  The logic is as follows:

1. In a novel environment, genetically fixed behaviours are suboptimal.
2. Individuals that can learn (or copy successful neighbours) survive long
   enough to reproduce, even when their genome does not yet encode the optimal
   behaviour.
3. Because learners reproduce more, genetic variants that predispose faster or
   deeper learning spread in the population.
4. Over generations, the benefit of learning decreases as the genome gradually
   encodes the once-learned behaviour directly — a process called *genetic
   assimilation* (Waddington 1942).

`clade` provides two within-lifetime learning modules that can be used
independently or together to study these dynamics:

- **Within-lifetime RL** (`rl_mode = "actor_critic"`): REINFORCE with
  baseline updates the action-selection layer of an agent's neural network
  each tick, driven by energy-gain reward.
- **Social learning** (`social_learning = TRUE`): prestige-biased copying —
  each agent blends a fraction of the highest-energy neighbour's output-layer
  weights into its own policy, following the strategy of Henrich & Gil-White
  (2001).

## Minimal demonstration

The experiment below compares three conditions over 300 ticks on a 30×30 grid:

| Condition | `rl_mode`       | `social_learning` | Expected outcome             |
|-----------|-----------------|-------------------|------------------------------|
| Baseline  | `"none"`        | `FALSE`           | Slowest adaptation           |
| RL only   | `"actor_critic"`| `FALSE`           | Faster individual adaptation |
| RL + SL   | `"actor_critic"`| `TRUE`            | Fastest cultural spread      |

```{r setup}
library(clade)
library(ggplot2)
library(patchwork)
```

```{r base-specs}
base <- default_specs()
base$grid_rows     <- 30L
base$grid_cols     <- 30L
base$n_agents_init <- 80L
base$max_agents    <- 500L
base$max_ticks     <- 300L
base$random_seed   <- 42L
# Start in a low-resource environment to create selection pressure for learning
base$grass_rate    <- 0.03
base$grass_init_prob <- 0.3
```

```{r run-conditions}
# Condition 1: no within-lifetime learning
s_base <- base
s_base$rl_mode        <- "none"
s_base$social_learning <- FALSE
env_base <- run_alife(s_base)

# Condition 2: individual RL only
s_rl <- base
s_rl$rl_mode        <- "actor_critic"
s_rl$rl_update_freq <- 1L
s_rl$social_learning <- FALSE
env_rl <- run_alife(s_rl)

# Condition 3: RL + prestige-biased social learning (full Baldwin pathway)
s_both <- base
s_both$rl_mode             <- "actor_critic"
s_both$rl_update_freq      <- 1L
s_both$social_learning      <- TRUE
s_both$social_learning_freq <- 5L
s_both$social_learning_rate <- 0.1
env_both <- run_alife(s_both)
```

## Comparing population trajectories

```{r trajectories}
d_base <- get_run_data(env_base)$ticks
d_rl   <- get_run_data(env_rl)$ticks
d_both <- get_run_data(env_both)$ticks

# Combine into a single tidy data frame
d <- rbind(
  transform(d_base, condition = "Baseline"),
  transform(d_rl,   condition = "RL only"),
  transform(d_both, condition = "RL + SL")
)
d$condition <- factor(d$condition, levels = c("Baseline", "RL only", "RL + SL"))
```

```{r plot-pop, fig.width=7, fig.height=4.5}
p_agents <- ggplot(d, aes(t, n_agents, colour = condition)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Tick", y = "Population size",
       colour = NULL, title = "Population size") +
  theme_minimal(base_size = 12)

p_energy <- ggplot(d, aes(t, mean_energy, colour = condition)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Tick", y = "Mean energy",
       colour = NULL, title = "Mean energy per agent") +
  theme_minimal(base_size = 12)

p_agents / p_energy + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The RL + SL condition typically reaches higher mean energy sooner than either
the baseline or RL-only condition, reflecting the two-step Baldwin pathway:
individual reinforcement discovers locally good policies, while social copying
spreads successful policies faster than genetic transmission alone.

## Tracing genetic assimilation

Genetic assimilation leaves a signature in the population's genetic diversity:
early in the run, diversity is maintained because the genome does not yet encode
the optimal policy; later, as the learned behaviour becomes genetically fixed,
diversity decreases as a small number of well-adapted genotypes dominate.

```{r plot-diversity, fig.width=7, fig.height=3}
ggplot(d, aes(t, genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Tick", y = "Mean pairwise genome distance",
       colour = NULL,
       title = "Genetic diversity — signature of genetic assimilation") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

A steeper late-run decline in the RL + SL condition relative to the baseline
is consistent with accelerated assimilation: the learning buffer reduces
selection variance early, allowing more genetic variants to persist, but the
benefit of learning narrows once the genome catches up, driving a sharper
selective sweep.

## Combining with epigenetics

The `epigenetics = TRUE` module adds a second non-genetic inheritance channel:
methylation marks accumulated during an agent's lifetime are transmitted to
offspring with probability `epigenetic_inheritance`. This can accelerate the
Baldwin pathway further because an offspring inherits not only its parent's
genome but also a prior on which synaptic weights to canalize.

```{r run-epigenetics}
s_epi <- s_both
s_epi$epigenetics            <- TRUE
s_epi$epigenetic_inheritance <- 0.5
s_epi$methylation_rate       <- 0.2
s_epi$demethylation_rate     <- 0.05
env_epi <- run_alife(s_epi)
```

```{r plot-epi, fig.width=7, fig.height=3}
d_epi <- get_run_data(env_epi)$ticks
d2 <- rbind(
  transform(d_both, condition = "RL + SL"),
  transform(d_epi,  condition = "RL + SL + Epigenetics")
)
ggplot(d2, aes(t, mean_prior_sigma, colour = condition)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Tick", y = "Mean BNN prior sigma",
       colour = NULL,
       title = "BNN uncertainty — canalization via epigenetics") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

`mean_prior_sigma` (the mean standard deviation of BNN weight distributions)
measures how certain agents are about their action-selection policy. Lower
values indicate more canalized — and potentially more heritably fixed — neural
circuits. The epigenetic condition should show faster sigma reduction because
offspring inherit methylation marks that bias sigma contraction towards the
parent's effective policy.

## Quantifying the effect

`estimate_heritability()` provides a rough proxy for the heritability of the
learned component:

```{r heritability}
h2_base <- estimate_heritability(env_base, trait = "mean_energy")
h2_rl   <- estimate_heritability(env_rl,   trait = "mean_energy")
h2_both <- estimate_heritability(env_both, trait = "mean_energy")

cat(sprintf("Heritability proxy (lag-1 autocorrelation):\n"))
cat(sprintf("  Baseline:  %.3f\n", h2_base$h2))
cat(sprintf("  RL only:   %.3f\n", h2_rl$h2))
cat(sprintf("  RL + SL:   %.3f\n", h2_both$h2))
```

Higher autocorrelation in the RL + SL condition reflects a more stable
(genetically assimilated) population trajectory: once the optimal policy is
encoded in the genome, mean energy varies less from tick to tick.

## Key parameters

| Parameter               | Default       | Role in Baldwin pathway                   |
|-------------------------|---------------|-------------------------------------------|
| `rl_mode`               | `"none"`      | `"actor_critic"` enables REINFORCE        |
| `rl_update_freq`        | `1L`          | RL update every N ticks                   |
| `social_learning`       | `FALSE`       | Enables prestige-biased copying           |
| `social_learning_freq`  | `10L`         | Copy every N ticks                        |
| `social_learning_rate`  | `0.1`         | Fraction of model weights copied          |
| `epigenetics`           | `FALSE`       | Enables methylation + TEI                 |
| `epigenetic_inheritance`| `0.5`         | Probability each mark is transmitted      |
| `methylation_rate`      | `0.1`         | Probability reward drives methylation     |
| `demethylation_rate`    | `0.02`        | Background demethylation rate             |

## Pre-computed experimental results

The five figures below summarise a systematic study of the Baldwin Effect using
the BNN brain type (`brain_type = "bnn"`), where the prior standard deviation σ
is the observable marker of canalization: declining σ indicates genetic
assimilation of a learned behaviour; rising σ indicates selection for maintained
flexibility. All results are from runs stored in `Rdata/` — no Julia session
required to view them.

### Experiment 1 — Environmental stability gradient

3 × 3 factorial design (grass_rate × seasonal_amplitude), 5 seeds × 1000 ticks
(45 runs). All 9 conditions produced a positive σ slope: exploration is selected
everywhere. The shallowest slope (grass = 0.20, seasonal = 0.8, bold row) still
reaches the σ ceiling.

```{r fig-exp1, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Phase diagram of mean_prior_sigma slope across 9 resource-abundance × seasonality conditions. All cells positive (red) — no canalization in any environment tested."}
FIG("bnn_exp1_stability")
```

### Experiment 2 — Is the sigma ceiling an artefact of run length?

Most canalization-favourable condition (grass = 0.20, seasonal = 0.8) extended
to 5000 ticks. The σ slope decelerates 21-fold from 1000 to 5000 ticks, but
final σ remains at 0.50 in all cases. This is ceiling saturation, not the onset
of canalization.

```{r fig-exp2, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Sigma trajectories at 1000, 2000, and 5000 ticks (mean +/- SE, 5 seeds). All runs hit the 0.5 ceiling; the slope decelerates but never reverses."}
FIG("bnn_exp2_runlength")
```

### Experiment 3 — Brain architecture comparison

BNN-only, ANN+RL, BNN+RL, ANN-null, 5 seeds × 1000 ticks. RL has no effect on
σ dynamics: BNN-only and BNN+RL produce identical σ trajectories. BNN agents
pay a 17% energy penalty for broad Thompson sampling yet the exploration ESS
persists.

```{r fig-exp3, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Left: sigma trajectories for BNN-only vs BNN+RL — indistinguishable. Right: final mean energy by condition — BNN pays 17% relative to ANN but exploration remains the ESS."}
FIG("bnn_exp3_brains")
```

### Experiment 4 — Social modifiers

4 modules × 2 environments × 5 seeds × 1000 ticks. Epigenetic inheritance
(`epigenetics = TRUE`) is the only mechanism that substantially reduces σ —
32% below ceiling in the stable-abundant environment and 72% below ceiling in
the scarce-stable environment. Social learning and kin selection have negligible
effects.

```{r fig-exp4, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Final mean prior sigma by module and environment. Epigenetics (red) is the only social mechanism to substantially reduce sigma. All others remain at or near the 0.5 ceiling (dashed line)."}
FIG("bnn_exp4_social")
```

### Experiment 5 — MAP-Elites on σ × genetic diversity space

150 MAP-Elites iterations on the BNN baseline, 500-tick runs, 121-cell archive
(σ × genetic diversity, both 0–0.5, step 0.05): 20 cells filled. Low σ was
found (as low as 0.028), but every low-σ cell also had low genetic diversity
(gd < 0.15). The σ–gd correlation across the archive was +0.59. No low-σ +
high-gd cells exist — ruling out the Baldwin Effect proper and confirming that
low σ arises via genetic drift to fixation, not via genetic assimilation of a
learned strategy.

```{r fig-exp5, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="MAP-Elites archive (20/121 cells filled). Low sigma only coexists with low genetic diversity — drift pseudo-canalization, not the Baldwin Effect. No solutions occupy the low-sigma + high-diversity quadrant."}
FIG("bnn_exp5_mapelites")
```

---

## References

Baldwin, J.M. (1896) A new factor in evolution. *American Naturalist*
30(354):441–451.

Henrich, J. & Gil-White, F.J. (2001) The evolution of prestige: freely
conferred deference as a mechanism for enhancing the benefits of cultural
transmission. *Evolution and Human Behavior* 22(3):165–196.

Jablonka, E. & Lamb, M.J. (2005) *Evolution in Four Dimensions: Genetic,
Epigenetic, Behavioral, and Symbolic Variation in the History of Life.*
MIT Press, Cambridge MA.

Laland, K.N. (2004) Social learning strategies. *Learning and Behavior*
32(1):4–14.

Rendell, L. et al. (2010) Why copy others? Insights from the social learning
strategies tournament. *Science* 328(5975):208–213.

Simpson, G.G. (1953) The Baldwin Effect. *Evolution* 7(2):110–117.

Sutton, R.S. & Barto, A.G. (2018) *Reinforcement Learning: An Introduction.*
2nd ed. MIT Press, Cambridge MA.

Waddington, C.H. (1942) Canalization of development and the inheritance of
acquired characters. *Nature* 150:563–565.

Williams, R.J. (1992) Simple statistical gradient-following algorithms for
connectionist reinforcement learning. *Machine Learning* 8(3–4):229–256.
