---
title: "Basics — five minutes to a clade run"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics — five minutes to a clade run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## What this vignette is for

The shortest path from a fresh install to a clade run you can read.
Five short sections, each anchored on one of the four core functions
(`default_specs()`, `run_alife()`, `get_run_data()`, `plot_run()`) and
a closing pointer to the deeper material. If you want narrative motivation,
the architectural picture, or the lab vision behind clade, read
`vignette("getting-started")` first; this vignette assumes you just
want a working run.

> This vignette is built incrementally across the Phase A foundation
> walk (`~/.claude/plans/purring-honking-dove.md`, items 1–4). Each
> Tier-A0 audit contributes one section; the file becomes complete
> when item 4 (`plot_run()`) lands.

## 1. Start from `default_specs()`

`default_specs()` is the single source of truth for every parameter
clade understands. It takes no arguments and returns a named list of
~300 fields with sensible defaults — sparse 30 × 30 grid, 50 starting
agents, 500-tick run, BNN brain, diploid sex, no optional modules
(no disease, no kin selection, no predators, etc.). The discipline
across the codebase is *always* copy from `default_specs()` and
modify the copy; never hand-build a specs list from scratch, because
new modules add fields over time and you would miss them.

```{r}
library(clade)

specs <- default_specs()       # the canonical starting list
specs$n_agents_init <- 100L    # one common change
specs$max_ticks     <- 200L    # another
specs$random_seed   <- 42L     # reproducibility
```

To see what changed from defaults, use `print_specs(specs, diff_only
= TRUE)`. To see every parameter grouped by biological theme, use
`print_specs(specs)` (no arguments needed). To see what biological
modules are off-by-default (the opt-in design), look at
`vignette("parameter-reference")` or read the field-by-field roxygen
in `?default_specs`.

## 2. Run the simulation with `run_alife()`

`run_alife(specs)` is the one call you need. Internally it starts a
Julia session on the first invocation (one-time ~60-90 s JIT cost),
sends `specs` across the R-Julia boundary once, runs every tick in
Julia, and returns an `env` list containing the surviving agents,
per-tick statistics, per-death records, and (if you set
`specs$log_genomes = TRUE`) per-tick genome snapshots. After the run,
`env$viability` carries the verdict from `viability_report()` and the
function emits a `warning()` if the population crashed — so silent
collapses can't slip past you. Pass `verbose = FALSE` for batch /
test usage; the viability warning still fires.

```{r}
env <- run_alife(specs)        # one boundary crossing, one return
env$t                          # the final tick (== specs$max_ticks)
env$viability$verdict          # "viable", "weak", or "crashed"
```

`run_clade()` is an alias for `run_alife()` if you prefer the
package name in your scripts. For multi-seed replicates use
`batch_seeds(specs, seeds = 1:5)`; for a list of different specs use
`batch_alife(list_of_specs, n_cores = 4L)` (each worker pays the
Julia startup cost separately, so parallelism is worth it from
roughly 20 scenarios upward).

## 3. Extract the run with `get_run_data()`

*(Filled by Phase A item 3.)*

## 4. Plot the run with `plot_run()`

*(Filled by Phase A item 4.)*

## 5. What next?

*(Filled by Phase A item 4.)*
