---
title: "Basics — five minutes to a clade run"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics — five minutes to a clade run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## What this vignette is for

The shortest path from a fresh install to a clade run you can read.
Five short sections, each anchored on one of the four core functions
(`default_specs()`, `run_alife()`, `get_run_data()`, `plot_run()`) and
a closing pointer to the deeper material. If you want narrative motivation,
the architectural picture, or the lab vision behind clade, read
`vignette("getting-started")` first; this vignette assumes you just
want a working run.

## 1. Start from `default_specs()`

`default_specs()` is the single source of truth for every parameter
clade understands. It takes no arguments and returns a named list of
~300 fields with sensible defaults — sparse 30 × 30 grid, 50 starting
agents, 500-tick run, BNN brain, diploid sex, no optional modules
(no disease, no kin selection, no predators, etc.). The discipline
across the codebase is *always* copy from `default_specs()` and
modify the copy; never hand-build a specs list from scratch, because
new modules add fields over time and you would miss them.

```{r}
library(clade)

specs <- default_specs()       # the canonical starting list
specs$n_agents_init <- 100L    # one common change
specs$max_ticks     <- 200L    # another
specs$random_seed   <- 42L     # reproducibility
```

To see what changed from defaults, use `print_specs(specs, diff_only
= TRUE)`. To see every parameter grouped by biological theme, use
`print_specs(specs)` (no arguments needed). To see what biological
modules are off-by-default (the opt-in design), look at
`vignette("parameter-reference")` or read the field-by-field roxygen
in `?default_specs`.

## 2. Run the simulation with `run_alife()`

`run_alife(specs)` is the one call you need. Internally it starts a
Julia session on the first invocation (one-time ~60-90 s JIT cost),
sends `specs` across the R-Julia boundary once, runs every tick in
Julia, and returns an `env` list containing the surviving agents,
per-tick statistics, per-death records, and (if you set
`specs$log_genomes = TRUE`) per-tick genome snapshots. After the run,
`env$viability` carries the verdict from `viability_report()` and the
function emits a `warning()` if the population crashed — so silent
collapses can't slip past you. Pass `verbose = FALSE` for batch /
test usage; the viability warning still fires.

```{r}
env <- run_alife(specs)        # one boundary crossing, one return
env$t                          # the final tick (== specs$max_ticks)
env$viability$verdict          # "viable", "weak", or "crashed"
```

`run_clade()` is an alias for `run_alife()` if you prefer the
package name in your scripts. For multi-seed replicates use
`batch_seeds(specs, seeds = 1:5)`; for a list of different specs use
`batch_alife(list_of_specs, n_cores = 4L)` (each worker pays the
Julia startup cost separately, so parallelism is worth it from
roughly 20 scenarios upward).

## 3. Extract the run with `get_run_data()`

`get_run_data(env)` reshapes the run into two tidy data frames
(`$ticks` and `$deaths`) plus a third (`$genomes`) that is `NULL`
unless you set `specs$log_genomes = TRUE`. `$ticks` has one row per
logged tick and ~60 population columns — the always-present core
(`t`, `n_agents`, `n_births`, `n_deaths`, `mean_energy`, `mean_age`,
`mean_body_size`, `genetic_diversity`, `grass_coverage`, ...) plus
module-specific columns that are zero when the corresponding module
is off, so the shape is stable across specs. `$deaths` has one row
per agent death with seven columns (`id`, `t`, `age`, `energy`,
`cause`, `body_size`, `num_offspring`); inspecting `$deaths$cause`
is the fastest way to see whether a run was starvation-dominated,
age-dominated, or (when predators are on) predation-dominated.

```{r}
rd <- get_run_data(env)
head(rd$ticks)
tail(rd$ticks[, c("t", "n_agents", "mean_energy", "genetic_diversity")])
table(rd$deaths$cause)
```

For the full column list of a specific run, call
`colnames(rd$ticks)` — the canonical schema lives in
[`inst/julia/src/logging.jl`](https://github.com/itchyshin/clade/blob/main/inst/julia/src/logging.jl)
(`_init_progress`). For genome snapshots, use `get_genome_data(env)`
(same `$genomes` long-format data frame, plus reserved placeholders
for heterozygosity and FST).

## 4. Plot the run with `plot_run()`

`plot_run(rd)` returns a 2 × 3 dashboard (a `patchwork` object) of
six time-series diagnostics: population size, mean energy ± 1 SD,
genetic diversity, births vs deaths per tick, grass coverage, and a
sixth panel that switches between *BNN prior sigma* (when the BNN
brain's plasticity is varying — the Baldwin Effect panel after
Baldwin 1896 / Hinton & Nowlan 1987) and *mean body size* (when
sigma is flat or absent). It's the fastest way to read whether a
run is biologically sensible: a healthy run shows population
holding above a viability floor, energy stable or oscillating,
diversity not collapsing to zero, births roughly tracking deaths
after burn-in, and grass not stripped to bare ground. If any of
those panels look pathological (population to zero, energy
collapsing, diversity to zero in <50 ticks), the run probably
won't support trait-mean interpretations — and `env$viability$verdict`
will already say so.

```{r}
plot_run(rd)
```

`plot_run()` is intentionally robust to the unhappy paths:
it errors with a clear message if `rd$ticks` is missing, returns a
ggplot placeholder (rather than crashing) if `$ticks` has zero
rows, and quietly truncates the timeline when `run_alife()` crashed
mid-run (the post-crash zero-padded rows are filtered by `t > 0`).
So you can pipe a viability-flagged crashed run straight into
`plot_run()` and read the partial story.

## 5. What next?

Once you can read `plot_run()` you have most of clade's surface.
Where to go from here, in order of investment:

- **Multi-seed replicates.** Every conclusion in clade should be
  defended at ≥ 5 seeds; one run is a story, five is data.
  `batch_seeds(specs, seeds = 1:5)` runs the same specs across
  seeds; pair with `viability_report()` per replicate and keep only
  the viable ones.
- **Scenario vignettes.** `vignette(package = "clade")` lists ~40
  worked scenarios under `s-*` (scenarios) and `paper-*` (paper
  reproductions). Start with `vignette("s-baseline")` for the
  shortest fully-narrated run, then `vignette("scenarios")` for the
  full catalogue.
- **Paper reproductions.** If you have a behavioural-ecology paper
  with a quantitative prediction, the `paper-template.Rmd` vignette
  shows how to turn it into a clade experiment. The published
  reproductions (Wolf 2007, Trivers 1971, Réale 2010, Griesser
  2023, Ryan 1990, ...) each follow the same template and report
  what reproduces vs what doesn't.
- **Parameter-space search.** When you want the best (or most
  diverse) configurations rather than a single run,
  `search_map_elites()` / `search_cmaes()` / `search_viability()`
  run optimisation over the spec list. `vignette("ps-introduction")`
  is the starting point.
- **Narrative depth.** For the architectural picture, the four
  brain types, the R-Julia split, the lab values, and the vision:
  `vignette("getting-started")` and
  [`dev/design/00-vision.md`](https://github.com/itchyshin/clade/blob/main/dev/design/00-vision.md).
