---
title: "Genetic Diversity Search in clade"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetic Diversity Search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## 1. Why search for diversity?

Genetic diversity quantifies the breadth of heritable variation in a simulated
population. It is a proxy for evolutionary potential (a genetically uniform
population has little raw material for selection to act on), resilience to
environmental change (diverse populations are less likely to be uniformly
maladapted to a novel challenge), and the richness of phenotypic exploration
(high diversity increases the probability that at least one lineage will
colonise a new niche).

In conservation biology, the loss of genetic diversity is a leading indicator of
extinction risk. In artificial life research, diversity is the substrate on which
open-ended evolution depends. Finding which parameter combinations sustain high
diversity — and understanding the trade-offs they entail — is therefore a central
design problem in both fields.

`clade` provides four complementary tools for this search:

| Function | Strategy | When to use |
|---|---|---|
| `search_random()` | Stochastic sweep | Fast initial screen |
| `search_map_elites()` | Quality-diversity (MAP-Elites) | Illuminate the achievable space |
| `search_cmaes()` | CMA-ES optimisation | Find a single global optimum |
| `search_gradient()` | Finite-difference gradient ascent | Local refinement |

The workflow below moves from coarse to fine: screen with `search_random()`,
characterise trade-offs with `search_map_elites()`, and maximise with
`search_cmaes()`.

---

## 2. Quick stochastic screen with `search_random()`

`search_random()` draws `n_samples` independent parameter configurations from
the supplied ranges and evaluates each. It is fast enough to run interactively
and is the right entry point when you have no prior knowledge of which regions
of parameter space are promising.

The example below screens three parameters simultaneously. Each sample draws
`mutation_sd`, `grass_rate`, and `n_agents_init` independently and uniformly
from the specified ranges, runs a full simulation, and records mean genetic
diversity.

```{r random-screen}
library(clade)

result <- search_random(
  specs_base    = default_specs(),
  search_params = list(
    mutation_sd   = c(0.01, 0.5),
    grass_rate    = c(0.05, 0.8),
    n_agents_init = c(20L, 150L)
  ),
  n_samples  = 30L,
  objective  = "genetic_diversity"
)

# Top configurations by genetic diversity
head(result)

# Partial-dependence plot for mutation_sd
plot(result$mutation_sd, result$score,
     xlab = "mutation_sd", ylab = "mean genetic diversity",
     pch  = 19, col = "steelblue")
```

The `result` data frame is sorted by descending score. Each row corresponds to
one simulation; the `score` column holds mean genetic diversity across ticks.
The highest-diversity configurations found by random sampling serve as starting
points for the targeted searches below.

The `specs_list` attribute stores the full specs for each row, ordered to match
the data frame. Retrieve the best configuration with:

```{r random-retrieve}
best_specs <- attr(result, "specs_list")[[1]]
cat("Best mutation_sd:", best_specs$mutation_sd, "\n")
cat("Best grass_rate:", best_specs$grass_rate, "\n")
```

---

## 3. MAP-Elites for the diversity–population trade-off

The MAP-Elites algorithm (Mouret & Clune 2015) fills a grid of niches in
behavioural space. Unlike a standard optimiser, it does not converge to a single
point. Instead, every cell of the archive holds the parameter set that achieves
that particular (diversity, population) combination with the highest diversity
score. Empty cells indicate combinations that no simulated configuration
achieved.

This illumination of the achievable space is the key advantage over random
search: you can read off not just "what maximises diversity" but "what diversity
is achievable at each population size" — a biologically meaningful trade-off.

```{r map-elites}
result <- search_map_elites(
  specs_base   = default_specs(),
  archive_dims = list(
    genetic_diversity = seq(0, 1, by = 0.1),
    n_agents          = seq(0, 200, by = 20)
  ),
  n_iterations = 300L,
  objective    = "genetic_diversity",
  verbose      = FALSE
)

# Archive heatmap: x = genetic_diversity niche, y = n_agents niche,
# fill = objective score (NA = empty cell)
result$map
```

Interpreting the heatmap:

- **Filled cells** show achievable (diversity, population) pairs. The colour
  encodes the objective score: brighter means higher diversity within that niche.
- **Empty cells** (shown in grey) are combinations that no parameter set
  produced. If high-diversity, large-population cells are empty, that combination
  may be intrinsically unattainable under the chosen model.
- **The archive** itself (`result$archive`) is a list of length equal to the
  number of cells. Non-null elements contain `$specs`, `$score`, and
  `$behavioural_descriptor`.

```{r map-elites-inspect}
# Extract all non-empty cells and compare their mutation_sd
non_empty <- Filter(Negate(is.null), result$archive)
cat("Filled cells:", length(non_empty), "\n")
cat("mutation_sd range in archive:",
    range(sapply(non_empty, function(x) x$specs$mutation_sd)), "\n")
```

---

## 4. CMA-ES for maximum diversity

`search_cmaes()` wraps a CMA-ES-style real-valued evolutionary strategy (via the
`GA` package) to find the single parameter set that maximises the objective. Use
it when you want one best answer rather than a map of trade-offs.

```{r cmaes}
result <- search_cmaes(
  specs_base = default_specs(),
  params     = c("mutation_sd", "grass_rate", "crossover_rate"),
  objective  = "genetic_diversity",
  n_iterations = 20L,
  popsize    = 10L
)

cat("Best diversity:", result$score, "\n")
cat("Best params:\n")
print(result$specs[c("mutation_sd", "grass_rate", "crossover_rate")])
```

`search_cmaes()` requires the `GA` package:

```{r cmaes-install, eval = FALSE}
install.packages("GA")
```

**When to use CMA-ES vs MAP-Elites:**

- Use `search_cmaes()` when you want the highest possible diversity and do not
  care about the population-size cost. It converges to a single point.
- Use `search_map_elites()` when you want to understand how diversity scales
  with other outcomes (population size, speciation, cooperation level). It
  reveals the full Pareto-like frontier of achievable outcomes.
- CMA-ES is faster per run for a single-objective answer; MAP-Elites is more
  informative per compute budget for multi-objective insight.

---

## 5. Custom diversity objective function

All four search functions accept a custom `objective` function in place of a
column name. The function receives the `env` object returned by `run_alife()` and
must return a numeric scalar.

The example below defines an objective that rewards both diversity and sustained
population size. Runs that go extinct (mean population below 10) receive a score
of zero regardless of their transient diversity, preventing the optimiser from
finding parameters that briefly achieve high diversity before collapse.

```{r custom-objective}
custom_obj <- function(env) {
  d   <- get_run_data(env)$ticks
  div <- mean(d$genetic_diversity, na.rm = TRUE)
  pop <- mean(d$n_agents,          na.rm = TRUE)
  if (pop < 10) return(0)      # discard runs that go near-extinct
  div * log1p(pop / 50)        # reward diversity weighted by log-population
}

result <- search_random(
  specs_base    = default_specs(),
  search_params = list(
    mutation_sd = c(0.01, 0.4),
    grass_rate  = c(0.1,  0.7)
  ),
  n_samples = 20L,
  objective = custom_obj
)

head(result)
```

The same `custom_obj` function can be passed to `search_map_elites()`,
`search_cmaes()`, or `search_gradient()` without modification.

---

## References

Mouret, J.-B. & Clune, J. (2015) Illuminating search spaces by mapping elites.
*arXiv:1504.04909*.

Hansen, N. & Ostermeier, A. (2001) Completely derandomized self-adaptation in
evolution strategies. *Evolutionary Computation* 9(2):159–195.

Frankham, R., Ballou, J.D. & Briscoe, D.A. (2010) *Introduction to Conservation
Genetics*, 2nd edn. Cambridge University Press.
