---
title: "Getting Started with clade"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with clade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

`clade` is an agent-based evolutionary simulator. Digital organisms carry
heritable neural-network genomes and forage on a renewable resource grid.
Natural selection acts on brain weights, life-history traits, and — with
optional modules — body size, dispersal tendency, wing morphology, helper
behaviour, and more.

The simulation runs entirely in Julia (via
[JuliaConnectoR](https://cran.r-project.org/package=JuliaConnectoR)), so
the R session only crosses the R–Julia boundary once per call: `run_alife()`
sends specs in and receives the full environment back.

---

## 1. Installation

### R package

```{r install-clade, eval=FALSE}
# From GitHub (development version)
remotes::install_github("shinichi-nakagawa/clade")

# Or from CRAN when released
# install.packages("clade")
```

### Julia

clade requires Julia >= 1.9. Download it from
[julialang.org](https://julialang.org/downloads/) or use
[juliaup](https://github.com/JuliaLang/juliaup):

```bash
curl -fsSL https://install.julialang.org | sh
```

On the **first** call to `run_alife()`, Julia compiles the simulation kernel.
This takes 60–90 seconds and is cached for all subsequent runs in the same
Julia environment.

---

## 2. Check that Julia is ready

```{r julia-check}
library(clade)

if (!julia_is_ready()) {
  stop("Julia not found. See ?julia_is_ready for setup instructions.")
}
cat("Julia", julia_version(), "ready.\n")
```

If Julia is not found, the most common fixes are:

- Add Julia to `PATH` (run `export PATH="$PATH:~/.juliaup/bin"` in your shell).
- Set `JULIA_BINDIR` in your `.Renviron` file.
- Reinstall Julia via `juliaup`.

---

## 3. Your first run

`default_specs()` returns the full parameter list with sensible defaults.
Modify only what you need, then pass to `run_alife()`.

```{r first-run}
library(clade)

specs <- default_specs()
specs$grid_rows     <- 20L
specs$grid_cols     <- 20L
specs$n_agents_init <- 40L
specs$max_ticks     <- 200L
specs$random_seed   <- 1L

env <- run_alife(specs)

# S3 print: one-line summary
print(env)
```

```
#> clade_env  [200 ticks | 47 agents | genetic_diversity: 0.312]
```

`env` is a `clade_env` object. Its key fields are:

| Field | Contents |
|---|---|
| `env$agents` | List of surviving agents (one list per agent) |
| `env$progress` | Data frame of per-tick population statistics |
| `env$deaths` | Data frame of per-death records |
| `env$specs` | The specs list used for this run |
| `env$grass` | Final grass coverage matrix |

---

## 4. Extracting and plotting results

```{r plot-run}
data <- get_run_data(env)

# data$ticks  — per-tick population statistics
# data$deaths — one row per agent death

plot_run(data)   # multi-panel dashboard: population, energy, diversity
```

```{r fig-run-dashboard, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="plot_run() produces a six-panel dashboard: population size, mean energy (±SD ribbon), genetic diversity, births and deaths per tick, grass coverage, and BNN prior sigma."}
FIG("01_run_dashboard")
```

`data$ticks` contains columns including `t`, `n_agents`, `mean_energy`,
`genetic_diversity`, `n_births`, `n_deaths`, and columns for every active
module (e.g. `mean_wing_size` when `complex_landscape = TRUE`).

---

## 5. Inspecting parameters

`print_specs()` prints every parameter grouped by biological theme. Use
`diff_only = TRUE` to show only what you changed:

```{r print-specs}
specs2 <- default_specs()
specs2$kin_selection    <- TRUE
specs2$complex_landscape <- TRUE
specs2$max_ticks        <- 400L

print_specs(specs2, diff_only = TRUE)
```

```
#> -- clade specs (88 parameters) [diff only] --
#>
#>   Grid & population
#>     max_ticks                              400 *
#>
#>   Kin selection
#>     kin_selection                         TRUE *
#>
#>   Complex landscape
#>     complex_landscape                     TRUE *
```


---

## 6. Key parameters at a glance

```{r params-table, echo=FALSE, eval=TRUE}
params <- data.frame(
  Parameter = c(
    "grid_rows / grid_cols", "n_agents_init", "max_agents",
    "max_ticks", "random_seed",
    "brain_type", "ploidy",
    "grass_rate", "mutation_sd",
    "kin_selection", "disease", "complex_landscape",
    "dispersal_evolution", "spatial_sorting",
    "iffolk_selection", "cooperative_breeding",
    "body_size_evolution", "niche_construction",
    "social_learning", "parental_care"
  ),
  Default = c(
    "20 x 20", "50", "300", "300", "42",
    "\"bnn\"", "2",
    "0.10", "0.05",
    "FALSE", "FALSE", "FALSE",
    "FALSE", "FALSE",
    "FALSE", "FALSE",
    "FALSE", "FALSE",
    "FALSE", "FALSE"
  ),
  Role = c(
    "World size", "Starting population", "Population cap",
    "Simulation length", "Reproducibility",
    "Neural architecture (bnn/ann/ctrnn/grn/transformer/synthesis)",
    "Haploid (1) or diploid (2)",
    "Grass regrowth probability per tick", "Genome mutation rate",
    "Enable kin altruism", "Enable SIR disease", "Enable 3-layer habitat",
    "Allow dispersal to evolve", "Enable invasion-front assortment",
    "Enable inclusive fitness transfers", "Enable alloparental helpers",
    "Enable body-size scaling", "Enable shelter building",
    "Enable copying neighbours' brain weights",
    "Enable offspring carried until graduation"
  )
)
knitr::kable(params, caption = "Selected `default_specs()` parameters.")
```

---

## 7. Running multiple scenarios in parallel

`batch_alife()` runs a list of specs across R worker processes:

```{r batch}
specs_list <- lapply(c(0.05, 0.10, 0.20), function(gr) {
  s <- default_specs()
  s$grass_rate <- gr
  s$max_ticks  <- 200L
  s
})

results <- batch_alife(specs_list, n_cores = 3L)

# Compare mean genetic diversity across grass rates
sapply(results, function(env) {
  mean(get_run_data(env)$ticks$genetic_diversity, na.rm = TRUE)
})
```

---

## 8. Next steps

- **Biological scenarios**: see `vignette("scenarios")` for a tour of all
  modules with code and expected outputs.
- **Parameter search**: see `vignette("diversity-search")` for CMA-ES,
  MAP-Elites, and the viability mapping workflow.
- **Auto-calibration** (0.3.0): each scenario has a per-vignette CMA-ES
  harness at `dev/audit/calibration/` that discovers the regime where
  its claimed biology emerges. Run with
  `bash dev/audit/calibration/run_all.sh`; results in
  `dev/audit/calibration/RESULTS.md`.
- **Custom modules**: see `vignette("custom-modules")` to write your own
  per-tick hooks.
- **Full showcase**: see `vignette("showcase")` for figures from all 36
  scenarios.
- **Audit harness** (0.3.0): `dev/audit/run_audit.R` runs all 35
  vignette scenarios against the live kernel and confirms each
  produces the signal its "What we found" prose claims.
