---
title: "Getting started with clade: agent-based evolution in R"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clade: agent-based evolution in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## What clade is

`clade` is an R package for agent-based simulation of evolution and foraging
ecology. Digital organisms carry heritable neural-network genomes (haploid or
diploid) and one of several brain types, and they evolve over generations
through natural selection on behaviour, life history, and cognition. The
package is designed for evolutionary biologists who want to run controlled
*in silico* experiments without writing low-level simulation code.

`clade` is the successor to the experimental `alifeR` package. Where `alifeR`
embedded the simulation hot path in Rcpp, `clade` runs the entire simulation
in Julia (via [JuliaConnectoR](https://CRAN.R-project.org/package=JuliaConnectoR)).
The R–Julia boundary is crossed *once* per `run_alife()` call, not once per
tick. For 200 agents and 1,000 ticks this is roughly 25× faster than the
Rcpp-based alifeR backend, and it makes runs of 100,000 agents practical.
The Julia backend also enables MAP-Elites quality-diversity search and
derivative-free optimisation (CMA-ES), neither of which is feasible at this
scale from an Rcpp simulator.

This vignette is an orientation, not a reference. It introduces the workflow,
the parameter list, the brain types, and the biological modules that can be
toggled on or off. All code chunks below are shown but not evaluated, because
the Julia backend is not available during package check or vignette build.

## Installation

`clade` is currently distributed from GitHub:

```{r install}
# install.packages("devtools")
devtools::install_github("itchyshin/clade")
```

You will also need a working Julia installation (≥ 1.9). On the first call
to `run_alife()`, `clade` starts a Julia session, installs its own Julia
dependencies into a private environment, and precompiles the simulation
package. This first start-up takes a minute or two; subsequent runs in the
same R session are immediate.

## A minimal run

The default configuration is a small grid world with a single brain type
and the core ecological loop. Three calls are enough to produce a tidy
result and a diagnostic plot:

```{r minimal}
library(clade)

specs <- default_specs()

# Modules are off by default; enable any combination independently, e.g.:
# specs$disease <- TRUE
# specs$body_size_evolution <- TRUE

env  <- run_alife(specs)
data <- get_run_data(env)

plot_run(data)
```

`get_run_data()` returns a list with two tidy data frames:

- `data$ticks` — one row per logged tick, with population-level summaries
  (mean energy, mean body size, genetic diversity, grass cover, and so on).
- `data$deaths` — one row per agent death, with cause, age, energy, and
  parental output.

These are the canonical inputs to every analysis function in the package.

## Parameter overview

`default_specs()` returns a long named list. The parameters that most
strongly shape the run are summarised below. All others are documented in
`?default_specs`.

| Parameter              | Default | What it controls                                                              |
|------------------------|---------|-------------------------------------------------------------------------------|
| `brain_type`           | `"bnn"` | The cognitive architecture used by every agent (see next section).            |
| `ploidy`               | `2L`    | Diploid (`2L`) or haploid (`1L`) genomes.                                     |
| `max_ticks`            | `500L`  | Length of the simulation in time steps.                                       |
| `n_agents_init`        | `50L`   | Number of founders.                                                           |
| `disease`              | `FALSE` | Toggles the SIR disease module.                                               |
| `kin_selection`        | `FALSE` | Toggles pedigree-based altruism toward Moore-neighbourhood relatives.         |
| `cooperation_evolution`| `FALSE` | Toggles the evolution of an inheritable cooperation propensity.               |
| `niche_construction`   | `FALSE` | Toggles shelter-building and its protective and grass-suppressing effects.    |
| `social_learning`      | `FALSE` | Toggles output-layer copying from successful neighbours.                      |
| `rl_mode`              | `"none"`| Within-lifetime reinforcement learning (`"actor_critic"` or `"none"`).        |

Each module can be toggled independently, and the simulation will warn if
two modules with conflicting assumptions are active.

## Brain types

Every agent runs the same architecture, set by `specs$brain_type`. The
choice changes both how agents make decisions and what kind of plasticity
is available.

| Brain type      | Description                                                                                                                |
|-----------------|----------------------------------------------------------------------------------------------------------------------------|
| `"bnn"`         | A Bayesian neural network with a Gaussian prior over weights; `prior_sigma` evolves under selection (Neal 1996).            |
| `"ann"`         | A standard multilayer perceptron with feed-forward computation and weight inheritance.                                     |
| `"ctrnn"`       | A continuous-time recurrent network with leaky integrator neurons; supports dynamical behaviour (Beer 1995).                |
| `"grn"`         | A gene regulatory network whose nodes are genes and whose interactions evolve via mutation of the regulatory matrix.       |
| `"transformer"` | A small attention-based architecture with a fixed history window, useful for sequence-dependent foraging.                  |
| `"synthesis"`   | Symbolic rule synthesis: agents carry an evolved program of if-then rules rather than weights.                             |
| `"random"`      | Uniformly random action selection; useful as a null model baseline.                                                        |

`"bnn"` is the default because it admits an evolvable prior width and so
captures Baldwin-style learning–evolution interactions (Baldwin 1896,
Mouret & Clune 2015) cleanly.

## Biological modules

Modules are toggled through `specs`. They are all off by default, so the
core simulation is a clean baseline of foraging, reproduction, and death.

| Module                    | Spec field                        | One-line description                                                                             |
|---------------------------|-----------------------------------|--------------------------------------------------------------------------------------------------|
| Body size evolution       | `body_size_evolution`             | Heritable body size scales metabolic cost, foraging gain, and energy storage (Kleiber 1947).     |
| Natal dispersal           | `dispersal_evolution`             | Heritable tendency to move away from birthplace each tick, reducing kin competition.             |
| Group defense             | `group_defense`                   | Agents in the same cell share an anti-predator benefit that scales with local group size.        |
| Habitat preference        | `habitat_preference_evolution`    | Heritable preference for particular terrain or grass-density zones.                              |
| Seasonal dynamics         | `seasonal_amplitude`              | Grass productivity cycles sinusoidally, imposing periodic resource bottlenecks.                  |
| Signals and mate choice   | `signal_dims`                     | Agents evolve a multi-dimensional signal; receivers apply a heritable preference at mating.      |
| Speciation                | `speciation`                      | Reproductive isolation emerges when mate-preference divergence exceeds a threshold.              |
| Predators                 | `n_predators_init`                | Evolving predator agents pursue prey; predator numbers are set at initialisation.                |
| Parental care             | `parental_care`                   | Offspring are carried until graduation; parent pays a per-tick care cost.                        |
| Cooperative breeding      | `cooperative_breeding`            | Non-breeding helpers contribute energy to raise offspring in the same cell.                      |
| Mimicry and toxicity      | `mimicry` (+ `batesian_mimicry`)  | Heritable toxicity deters predators via Rescorla–Wagner avoidance learning. With `batesian_mimicry = TRUE`, palatable mimics exploit the learned aversion (Bates 1862) with predator-betrayal decay. |
| Phenotypic plasticity     | `phenotypic_plasticity`           | Within-lifetime weight adjustment in response to experienced environments.                       |
| Niche construction        | `niche_construction` (+ `shelter_occupancy_bonus`) | Agents build shelters that protect against predators and slow grass regrowth on the cell. With `shelter_occupancy_bonus > 0`: occupants of sheltered cells receive an energy subsidy proportional to shelter depth — heritable niche effect (Odling-Smee et al. 2003). |
| Scavenging                | `scavenging`                      | Agents may consume carrion left behind by recently dead conspecifics.                            |
| Disease (SIR)             | `disease`                         | Susceptible–infected–recovered transmission with energy costs and recovered immunity.            |
| Kin selection             | `kin_selection`                   | Pedigree-based altruistic energy transfer to related neighbours (Hamilton 1964).                 |
| Cooperation (PGG)         | `cooperation_evolution`           | An inheritable cooperation trait that scales the propensity to share with conspecifics.          |
| Within-lifetime RL        | `rl_mode`                         | REINFORCE-with-baseline updates to the output layer at runtime.                                  |
| Social learning           | `social_learning`                 | Periodic copying of output-layer weights from successful Moore-neighbourhood teachers.           |
| Epigenetic inheritance    | `epigenetics`                     | Lamarckian-style transmission of within-lifetime weight changes (Jablonka & Lamb 2005).         |
| Clutch size evolution     | `clutch_size_evolution`           | Heritable clutch size evolves under the life-history trade-off between offspring number and cost.|
| Parental investment       | `parental_investment_evolution`   | Heritable per-offspring energy allocation evolves jointly with clutch size.                      |
| Stress hypermutation      | `stress_hypermutation`            | Mutation rate increases when agent energy falls below a threshold, accelerating escape from traps.|

A typical experimental design fixes the brain type and toggles one module
at a time, comparing the resulting `get_run_data()` outputs across
conditions.

## Example: body size evolution

When `body_size_evolution = TRUE`, agents evolve a heritable body size trait.
Larger agents eat more per grass cell but pay a higher metabolic surcharge;
smaller agents eat less but are cheaper to run. Reference size 1.0 has no
metabolic correction. The trait is inherited via the genome and bounded by
`body_size_min` (default 0.3) and `body_size_max` (default 3.0).

```{r body-size}
specs                       <- default_specs()
specs$body_size_evolution   <- TRUE
specs$body_size_init_mean   <- 1.0
specs$body_size_mutation_sd <- 0.08

env  <- run_alife(specs)
data <- get_run_data(env)

# Time series of mean and SD of body size
head(data$ticks[, c("t", "n_agents", "mean_body_size", "sd_body_size")])
```

## Example: natal dispersal

When `dispersal_evolution = TRUE`, agents evolve a heritable probability of
taking one step away from their birthplace each tick. Dispersal reduces
inbreeding and kin competition at the cost of `dispersal_cost` energy.

```{r dispersal}
specs                    <- default_specs()
specs$dispersal_evolution <- TRUE
specs$dispersal_init_mean <- 0.2

env  <- run_alife(specs)
data <- get_run_data(env)

# Total dispersal events
sum(data$ticks$n_dispersal_events)
```

## Example: disease dynamics

To produce an SIR-style epidemic in the population, set the `disease`
flag and use the dedicated plotting helper:

```{r disease}
specs          <- default_specs()
specs$disease  <- TRUE
env            <- run_alife(specs)
data           <- get_run_data(env)

plot_disease_dynamics(data)
```

`plot_disease_dynamics()` reads the `n_infected` and `n_new_infections`
columns from `data$ticks` and overlays the susceptible, infected, and
recovered classes through time. The transmission probability, infectious
period, and immune duration are all controlled through `default_specs()`.

## Performance note

`clade` crosses the R–Julia boundary exactly **once per `run_alife()`
call**, regardless of the number of ticks or the number of agents. The
entire tick loop, sensing, neural network forward pass, action selection,
movement, eating, reproduction, and death runs inside a single Julia
process. For 200 agents and 1,000 ticks this is roughly 25× faster than
the Rcpp-based `alifeR` backend, and it scales smoothly to populations of
~100,000 agents on a workstation.

The first call in an R session pays a one-off start-up cost (~30–60 s)
for Julia compilation. Subsequent calls reuse the warm Julia session and
return results immediately.

## Where to go next

For a richer set of experiments — combining multiple modules, comparing
brain types, and reading the outputs with the analysis helpers — see the
[scenario discovery guide](scenarios.html). Each of the 35 scenarios has
its own page with runnable code, expected dynamics, and discovery
experiments.

For full parameter documentation, see `?default_specs`. Each module's
parameters are grouped in that help page under the module name.

## References

- Baldwin, J.M. (1896) A new factor in evolution. *American Naturalist*
  30(354): 441–451.
- Beer, R.D. (1995) On the dynamics of small continuous-time recurrent
  neural networks. *Adaptive Behavior* 3(4): 469–509.
- Falconer, D.S. & Mackay, T.F.C. (1996) *Introduction to Quantitative
  Genetics*, 4th ed. Longman, Harlow.
- Hamilton, W.D. (1964) The genetical evolution of social behaviour.
  *Journal of Theoretical Biology* 7(1): 1–52.
- Jablonka, E. & Lamb, M.J. (2005) *Evolution in Four Dimensions*. MIT
  Press, Cambridge, MA.
- Kermack, W.O. & McKendrick, A.G. (1927) A contribution to the
  mathematical theory of epidemics. *Proceedings of the Royal Society A*
  115(772): 700–721.
- Mouret, J.-B. & Clune, J. (2015) Illuminating search spaces by mapping
  elites. *arXiv:1504.04909*.
- Neal, R.M. (1996) *Bayesian Learning for Neural Networks*. Springer,
  New York.
