---
title: "Getting started with clade: agent-based evolution in R"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with clade: agent-based evolution in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## What clade is

`clade` is an R package for agent-based simulation of evolution and foraging
ecology. Digital organisms carry heritable neural-network genomes (haploid or
diploid) and one of several brain types, and they evolve over generations
through natural selection on behaviour, life history, and cognition. The
package is designed for evolutionary biologists who want to run controlled
*in silico* experiments without writing low-level simulation code.

`clade` is the successor to the experimental `alifeR` package. Where `alifeR`
embedded the simulation hot path in Rcpp, `clade` runs the entire simulation
in Julia (via [JuliaConnectoR](https://CRAN.R-project.org/package=JuliaConnectoR)).
The R–Julia boundary is crossed *once* per `run_alife()` call, not once per
tick. For 200 agents and 1,000 ticks this is roughly 25× faster than the
Rcpp-based alifeR backend, and it makes runs of 100,000 agents practical.
The Julia backend also enables MAP-Elites quality-diversity search and
gradient-through-simulation parameter tuning (using `Zygote.jl`), neither of
which is feasible from an Rcpp simulator.

This vignette is an orientation, not a reference. It introduces the workflow,
the parameter list, the brain types, and the biological modules that can be
toggled on or off. All code chunks below are shown but not evaluated, because
the Julia backend is not available during package check or vignette build.

## Installation

`clade` is currently distributed from GitHub:

```{r install}
# install.packages("devtools")
devtools::install_github("shinichi-nakagawa/clade")
```

You will also need a working Julia installation (≥ 1.9). On the first call
to `run_alife()`, `clade` starts a Julia session, installs its own Julia
dependencies into a private environment, and precompiles the simulation
package. This first start-up takes a minute or two; subsequent runs in the
same R session are immediate.

## A minimal run

The default configuration is a small grid world with a single brain type
and the core ecological loop. Three calls are enough to produce a tidy
result and a diagnostic plot:

```{r minimal}
library(clade)

specs <- default_specs()
env   <- run_alife(specs)
data  <- get_run_data(env)

plot_run(data)
```

`get_run_data()` returns a list with two tidy data frames:

- `data$ticks` — one row per logged tick, with population-level summaries
  (mean energy, mean body size, genetic diversity, grass cover, and so on).
- `data$deaths` — one row per agent death, with cause, age, energy, and
  parental output.

These are the canonical inputs to every analysis function in the package.

## Parameter overview

`default_specs()` returns a long named list. The parameters that most
strongly shape the run are summarised below. All others are documented in
`?default_specs`.

| Parameter              | Default | What it controls                                                              |
|------------------------|---------|-------------------------------------------------------------------------------|
| `brain_type`           | `"bnn"` | The cognitive architecture used by every agent (see next section).            |
| `ploidy`               | `1L`    | Haploid (`1L`) or diploid (`2L`) genomes.                                     |
| `max_ticks`            | `500L`  | Length of the simulation in time steps.                                       |
| `n_agents_init`        | `50L`   | Number of founders.                                                           |
| `disease`              | `FALSE` | Toggles the SIR disease module.                                               |
| `kin_selection`        | `FALSE` | Toggles pedigree-based altruism toward Moore-neighbourhood relatives.         |
| `cooperation_evolution`| `FALSE` | Toggles the evolution of an inheritable cooperation propensity.               |
| `niche_construction`   | `FALSE` | Toggles shelter-building and its protective and grass-suppressing effects.    |
| `social_learning`      | `FALSE` | Toggles output-layer copying from successful neighbours.                      |
| `rl_mode`              | `"none"`| Within-lifetime reinforcement learning (`"actor_critic"` or `"none"`).        |

Each module can be toggled independently, and the simulation will warn if
two modules with conflicting assumptions are active.

## Brain types

Every agent runs the same architecture, set by `specs$brain_type`. The
choice changes both how agents make decisions and what kind of plasticity
is available.

| Brain type      | Description                                                                                                                |
|-----------------|----------------------------------------------------------------------------------------------------------------------------|
| `"bnn"`         | A Bayesian neural network with a Gaussian prior over weights; `prior_sigma` evolves under selection (Neal 1996).            |
| `"ann"`         | A standard multilayer perceptron with feed-forward computation and weight inheritance.                                     |
| `"ctrnn"`       | A continuous-time recurrent network with leaky integrator neurons; supports dynamical behaviour (Beer 1995).                |
| `"grn"`         | A gene regulatory network whose nodes are genes and whose interactions evolve via mutation of the regulatory matrix.       |
| `"transformer"` | A small attention-based architecture with a fixed history window, useful for sequence-dependent foraging.                  |
| `"synthesis"`   | Symbolic rule synthesis: agents carry an evolved program of if-then rules rather than weights.                             |

`"bnn"` is the default because it admits an evolvable prior width and so
captures Baldwin-style learning–evolution interactions (Baldwin 1896,
Mouret & Clune 2015) cleanly.

## Biological modules

Modules are toggled through `specs`. They are all off by default, so the
core simulation is a clean baseline of foraging, reproduction, and death.

| Module               | Spec field             | One-line description                                                                       |
|----------------------|------------------------|--------------------------------------------------------------------------------------------|
| Disease (SIR)        | `disease`              | Susceptible–infected–recovered transmission with energy costs and recovered immunity.       |
| Kin selection        | `kin_selection`        | Pedigree-based altruistic energy transfer to related neighbours (Hamilton 1964).            |
| Cooperation          | `cooperation_evolution`| An inheritable cooperation trait that scales the propensity to share with conspecifics.     |
| Scavenging           | `scavenging`           | Agents may consume carrion left behind by recently dead conspecifics.                       |
| Niche construction   | `niche_construction`   | Agents build shelters that protect against predators and slow grass regrowth on the cell.  |
| Social learning      | `social_learning`      | Periodic copying of output-layer weights from successful Moore-neighbourhood teachers.      |
| Within-lifetime RL   | `rl_mode`              | REINFORCE-with-baseline updates to the output layer at runtime (epigenetic plasticity).     |
| Phenotypic plasticity| `epigenetics`          | Lamarckian-style transmission of within-lifetime weight changes (Jablonka & Lamb 2005).     |

A typical experimental design fixes the brain type and toggles one module
at a time, comparing the resulting `get_run_data()` outputs across
conditions.

## Example: disease dynamics

To produce an SIR-style epidemic in the population, set the `disease`
flag and use the dedicated plotting helper:

```{r disease}
specs          <- default_specs()
specs$disease  <- TRUE
env            <- run_alife(specs)
data           <- get_run_data(env)

plot_disease_dynamics(data)
```

`plot_disease_dynamics()` reads the `n_infected` and `n_new_infections`
columns from `data$ticks` and overlays the susceptible, infected, and
recovered classes through time. The transmission probability, infectious
period, and immune duration are all controlled through `default_specs()`.

## Performance note

`clade` crosses the R–Julia boundary exactly **once per `run_alife()`
call**, regardless of the number of ticks or the number of agents. The
entire tick loop, sensing, neural network forward pass, action selection,
movement, eating, reproduction, and death runs inside a single Julia
process. For 200 agents and 1,000 ticks this is roughly 25× faster than
the Rcpp-based `alifeR` backend, and it scales smoothly to populations of
~100,000 agents on a workstation.

The first call in an R session pays a one-off start-up cost (~30–60 s)
for Julia compilation. Subsequent calls reuse the warm Julia session and
return results immediately.

## References

- Baldwin, J.M. (1896) A new factor in evolution. *American Naturalist*
  30(354): 441–451.
- Beer, R.D. (1995) On the dynamics of small continuous-time recurrent
  neural networks. *Adaptive Behavior* 3(4): 469–509.
- Falconer, D.S. & Mackay, T.F.C. (1996) *Introduction to Quantitative
  Genetics*, 4th ed. Longman, Harlow.
- Hamilton, W.D. (1964) The genetical evolution of social behaviour.
  *Journal of Theoretical Biology* 7(1): 1–52.
- Jablonka, E. & Lamb, M.J. (2005) *Evolution in Four Dimensions*. MIT
  Press, Cambridge, MA.
- Kermack, W.O. & McKendrick, A.G. (1927) A contribution to the
  mathematical theory of epidemics. *Proceedings of the Royal Society A*
  115(772): 700–721.
- Mouret, J.-B. & Clune, J. (2015) Illuminating search spaces by mapping
  elites. *arXiv:1504.04909*.
- Neal, R.M. (1996) *Bayesian Learning for Neural Networks*. Springer,
  New York.
