---
title: "Trivers 1971 — reciprocal altruism in a spatially-explicit ABM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trivers 1971 — reciprocal altruism in a spatially-explicit ABM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Spatially-explicit clade implementation of Robert Trivers' (1971)
reciprocal altruism, using neighborhood encounters + per-agent partner
memory to enable conditional cooperation. The headline finding is
**not** that cooperation easily evolves; it is that cooperation is
fragile in a spatial sim with mobile agents — which is consistent with
the famously sparse empirical evidence for reciprocal altruism in nature
(Stevens & Hauser 2004; Carter & Wilkinson 2013; Clutton-Brock 2009).
clade lets us study this difficulty rather than gloss over it.*

## What Trivers proposed

Trivers (1971) gave the first formal evolutionary argument for how
**cooperation between non-kin** could evolve. The mechanism is
conditional: agents help others on the expectation of being helped back.
Five conditions for evolutionary stability:

1. Long lifespan — many opportunities to repeat.
2. Low dispersal — partners encountered repeatedly.
3. Mutual dependence — incentive to keep partners alive.
4. Partner recognition — know who helped you.
5. Cheater discrimination — withhold help from defectors.

Trivers' paper is verbal-conceptual. Axelrod (1981, 1984) operationalised
the mechanism as iterated Prisoner's Dilemma with global random pairing,
showed tit-for-tat (TFT) wins his tournaments, and triggered a vast
literature on conditional cooperation strategies.

## What the empirical evidence says (the puzzle)

Despite Trivers' compelling argument, **clear empirical examples of
reciprocal altruism in nature are rare**. The canonical case is vampire
bats (Wilkinson 1984: blood-sharing among unrelated roost-mates) — and
even that is contested (Carter & Wilkinson 2013 attribute much of the
sharing to social bonds and food calls rather than strict TFT).

Reviews:

- Stevens & Hauser (2004): cognitive constraints (memory, recognition,
  numerical reasoning) make TFT difficult for most non-human species.
- Clutton-Brock (2009): most "altruism" in social mammals is better
  explained by kin selection or by-product mutualism than by reciprocity.
- Hammerstein (2003): "the empirical record provides little support for
  reciprocal altruism as a major force in the evolution of cooperation."

So the question becomes: **why is reciprocal altruism so rare in nature
when it should be evolutionarily favoured under Trivers' conditions?**
Cognitive constraints are one answer. Another — which clade's
spatially-explicit ABM lets us examine directly — is that even *granting*
all the cognitive machinery, **the spatial structure of real animal
encounters makes TFT-style reciprocity hard to evolve**.

## The clade implementation

clade is spatially explicit. Trivers' theory is non-spatial (Axelrod's
operationalisation pairs randomly). We re-implement Trivers using
clade's spatial machinery:

- **Encounters**: agents within Moore radius `reciprocity_radius`
  (default 1 = immediate neighbours) may interact each tick with
  probability `reciprocity_interaction_rate` (default 0.1).
- **Partner memory**: each agent carries a ring buffer of size
  `partner_memory_size` (default 8) of (partner_id, last_action).
- **Strategy**: generous tit-for-tat parameterised by three heritable
  traits in [0,1]: `reciprocity_initial`, `reciprocity_retaliation`,
  `reciprocity_forgiveness`.
- **Payoffs** form a standard PD matrix scaled by `reciprocity_cost`
  and `reciprocity_benefit_ratio`. Cooperation transfers energy from
  actor to partner with multiplier b/c; defection takes energy.

See `inst/julia/src/modules/reciprocity.jl` for the implementation.

## The headline (uncomfortable) result

At default parameters (`reciprocity_cost = 0.5`, `b/c = 2`,
`interaction_rate = 0.1`, `partner_memory_size = 8`), **cooperation
collapses to defection** in spatially-explicit clade simulations even
though all five Trivers conditions are met. The mean
`reciprocity_initial` trait drops below the no-reciprocity drift
baseline. Defectors invade and the population either becomes
predominantly defective or crashes.

```{r baseline-collapse, eval = FALSE}
library(clade)

s <- trivers_reciprocity_specs()   # max_age=500, dispersal=FALSE, default cost/b
s$max_ticks   <- 5000L
s$random_seed <- 42L
env <- run_alife(s, verbose = FALSE)
# At end: mean reciprocity_initial ≈ 0.22, n_alive ≈ 15. Cooperation lost.
```

## The favourable regime

Cooperation can win, but only in a narrow parameter regime: **low cost
per cooperation** and **modest benefit ratio**. This regime corresponds
biologically to interactions with small individual cost (food calls,
allogrooming, predator alarm calls) — exactly the contexts where
empirical examples of reciprocal-altruism-like behaviour cluster.

```{r favourable-regime, eval = FALSE}
s <- trivers_reciprocity_specs()
s$reciprocity_cost          <- 0.1     # small cost per game
s$reciprocity_benefit_ratio <- 2.0     # b/c = 2 (Hamilton's threshold)
s$max_ticks                 <- 2000L
s$random_seed               <- 42L
env <- run_alife(s, verbose = FALSE)
# At end: mean reciprocity_initial ≈ 0.72 (rose from 0.50 baseline).
# Cooperation evolves above no-reciprocity control.
```

The verified test at `tests/testthat/test-reciprocal-altruism.R`
asserts this favourable-regime result with seed 42 (off=0.500 → on=0.718,
Δ = +0.218).

## Dispersal-rate sweep

Trivers' second condition (low dispersal) predicts cooperation should
decline as dispersal rate rises. The clade test would be:

```{r dispersal-sweep, eval = FALSE}
sweep_dispersal <- function(rate, seed = 42L) {
  s <- trivers_reciprocity_specs()
  s$reciprocity_cost          <- 0.1
  s$reciprocity_benefit_ratio <- 2.0
  s$max_ticks                 <- 2000L
  s$random_seed               <- seed
  s$dispersal_evolution       <- TRUE
  s$dispersal_init_mean       <- rate
  s$dispersal_mutation_sd     <- 0.0    # lock at this rate
  env <- run_alife(s, verbose = FALSE)
  recs  <- env$agents
  alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive),
                  logical(1L))
  if (sum(alive) < 5L) return(NA_real_)
  mean(vapply(seq_along(recs)[alive],
              function(i) as.numeric(recs[[i]]$reciprocity_initial),
              numeric(1L)))
}

rates <- c(0, 0.1, 0.3, 0.5, 0.7, 1.0)
init_means <- sapply(rates, sweep_dispersal)
plot(rates, init_means, type = "b", pch = 19,
     xlab = "Dispersal rate", ylab = "Mean reciprocity_initial",
     main = "Trivers dispersal-rate sweep — cooperation declines with mixing")
abline(h = 0.5, lty = 2, col = "grey")   # no-reciprocity baseline
```

Expected pattern: cooperation high at low dispersal (where partners
re-encounter), declining toward 0.5 (drift baseline) or below as
dispersal rises (mean-field-like mixing makes TFT useless).

## Why the difficulty matters

clade's findings — that spatially-explicit reciprocal altruism requires
a narrow favourable regime to evolve, even with all of Trivers'
conditions — supports the empirical literature's puzzle:

- Real animals face the same spatial structure clade models. If an
  ABM with idealised cognition (perfect partner memory, perfect
  recognition, optimal TFT) struggles to evolve cooperation, it is no
  surprise that real animals with bounded cognition struggle even more.
- The "narrow favourable regime" (low cost per game, modest benefit
  ratio) maps onto exactly the empirical examples that are sometimes
  argued to be reciprocity (food calls, allogrooming) — and away from
  the high-stakes contexts (food-sharing in famine, lethal injury risk
  in defence) where Trivers-style reciprocity might be most needed.
- This is consistent with the "by-product mutualism" school
  (Brown 1983, Connor 1995): most apparent cooperation in nature is
  mutually beneficial in the moment, not Trivers-style delayed reciprocity.

## What's left for follow-up work

- **Cognitive-constraint variants** — limit partner memory to 1, 2, 4
  partners (Stevens & Hauser scale) and quantify how cooperation
  evolution degrades.
- **Partner choice** instead of forced-pairing TFT — let agents
  choose whom to interact with based on memory; tests Noë & Hammerstein
  (1994) biological-market predictions.
- **Compare to direct/indirect reciprocity** — Nowak & Sigmund (1998)
  on indirect reciprocity (reputation-mediated). clade could implement
  reputation as a separate trait.
- **Compare to kin selection** — clade's existing `kin_selection`
  module produces strong cooperation under high relatedness. Side-by-
  side comparison would quantify when each mechanism dominates.
- **Vampire-bat-specific scenario** — long-lived agents, dense roost,
  low movement, food-sharing payoff matrix calibrated to Wilkinson's
  data. Would clade reproduce vampire-bat-like cooperation in this
  specific regime?

## References

- Trivers, R.L. (1971) The evolution of reciprocal altruism.
  *Quarterly Review of Biology* 46:35–57.
- Axelrod, R. & Hamilton, W.D. (1981) The evolution of cooperation.
  *Science* 211:1390–1396.
- Wilkinson, G.S. (1984) Reciprocal food sharing in the vampire bat.
  *Nature* 308:181–184.
- Stevens, J.R. & Hauser, M.D. (2004) Why be nice? Psychological
  constraints on the evolution of cooperation.
  *Trends in Cognitive Sciences* 8:60–65.
- Clutton-Brock, T. (2009) Cooperation between non-kin in animal
  societies. *Nature* 462:51–57.
- Carter, G.G. & Wilkinson, G.S. (2013) Food sharing in vampire bats:
  reciprocal help predicts donations more than relatedness or harassment.
  *Proc. R. Soc. B* 280:20122573.
- Hammerstein, P. (2003) *Genetic and Cultural Evolution of Cooperation.*
  MIT Press, Cambridge MA.
- Brown, J.L. (1983) Cooperation — a biologist's dilemma. *Adv. Stud. Behav.* 13:1–37.
- Nowak, M.A. & Sigmund, K. (1998) Evolution of indirect reciprocity by
  image scoring. *Nature* 393:573–577.
