---
title: "Wolf et al. 2008 — responsive personalities in a spatially-explicit ABM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wolf et al. 2008 — responsive personalities in a spatially-explicit ABM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Spatially-explicit clade interpretation of Wolf, van Doorn & Weissing
(2008) PNAS 105:15825–15830, "Evolutionary emergence of responsive and
unresponsive personalities". Captures the **frequency-dependent benefit
of sampling** mechanism using clade's grass economy and one-per-cell
movement rule. The implementation is more compact than Wolf 2007's
(one trait, one module); the cross-context-consistency extension and
the two-stage positive-feedback model from Wolf 2008 §3 are deferred
follow-up work.*

## What the original paper proposes

Wolf 2008 extends Wolf 2007's life-history mechanism with a different
question: when should an agent be "responsive" (sample environmental
state, then choose action conditional on it) versus "unresponsive"
(use a fixed general-purpose action)? Two key ingredients:

1. **Negative frequency-dependent benefits of responsiveness**
   (Methods Eq. 1). When responsiveness is rare, a responsive agent
   exploits cues with little competition — high benefit. When
   responsiveness is common, all responsives flock to the same
   resources — competition erodes the benefit. This produces stable
   *coexistence* of responsive and unresponsive types.

2. **Positive feedback on cost-of-responsiveness** (Methods §3). An
   agent that was responsive in stage 1 pays a *lower* cost to be
   responsive in stage 2 (because it has already incurred the
   start-up costs). Even tiny cost-reductions produce strong
   cross-context consistency: agents responsive in one context tend
   to be responsive in others.

Wolf 2008 thus links the two empirical phenomena that defined the
"animal personality" literature: **coexistence of behavioural types**
(why polymorphism, not optimal monomorphism) and **cross-context
consistency** (why agents are predictable across situations).

## clade implementation

clade is spatially explicit. Wolf 2008 is non-spatial (50-island mean-
field). We re-implement the **frequency-dependent benefit** mechanism
using clade's spatial machinery; the two-stage cross-context
consistency extension is left for follow-up work.

**One new heritable trait** in `[0, 1]`:

- `responsiveness` = Pr(this agent samples local state and overrides
  its action toward the richest free cardinal-neighbour cell on a
  given tick).

**Sampling cost**:

- Per-tick energy drain `responsiveness_cost` paid by the agent
  whenever it samples (regardless of whether the resulting move
  succeeds or whether it improves on the current cell).

**Frequency-dependence emerges from clade's existing economy**:

- Many responsive agents → all flock to richest neighbours → handling-
  time depletes those cells → first-mover wins, others waste cost.
  Combined with one-per-cell movement (Phase 2), responsive agents at
  high density frequently find their richest target occupied and end
  up paying cost without moving.
- Few responsive agents → rich cells stay rich → first-mover advantage
  yields net gain → trait selected for.

This is the *same* `dE/dp_r < 0` condition Wolf states in Methods
Eq. 1, realised via spatial competition rather than the abstract
patch-choice denominator `f_i = p_r·l_i + (1-p_r)·l̄`.

See `inst/julia/src/modules/responsiveness.jl`.

## What's NOT implemented (deliberate)

- **Two-stage cross-context consistency** (Wolf 2008 §3). Would need
  a per-agent `times_responsive` counter and a positive-feedback rule
  that lowers cost after prior responsiveness. clade has the
  infrastructure; this is a one-day follow-up task.
- **Per-resource (1+α·N_i) competition denominator**. The same
  decision as Phase 3a (Wolf 2007). Would clean up the polymorphism
  result but isn't essential for the headline mechanism.
- **Cost-of-uncertainty**. Wolf 2008's `C` is a single parameter; in
  reality cost might scale with environmental volatility, partial
  information, etc. Not modelled here.

## The headline result

```{r setup-libs, eval = FALSE}
library(clade)

s <- wolf2008_responsiveness_specs()
s$max_ticks   <- 2000L
s$random_seed <- 42L
env <- suppressWarnings(run_alife(s, verbose = FALSE))
```

Pull final-tick responsiveness:

```{r mean-resp, eval = FALSE}
recs  <- env$agents
alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive),
                logical(1L))
resp  <- vapply(seq_along(recs)[alive],
                function(i) as.numeric(recs[[i]]$responsiveness),
                numeric(1L))
mean(resp)   # selection moves the trait above the no-module baseline of 0.5
```

Typical values at this configuration (seed = 42):

| Condition | Mean `responsiveness` |
|---|---|
| No module (drift baseline) | 0.500 |
| Module on (selection)      | 0.669 |
| Δ                           | +0.169 |

The trait rises above the drift baseline — selection on `responsiveness`
is real. The verified test at
`tests/testthat/test-responsive-personalities.R` asserts this
selection signal with seed 42.

## Density-dependence sweep (Wolf's headline mechanism)

Wolf's prediction: responsiveness should DECREASE at higher density
(more competition for rich cells erodes the benefit). The clade test:

```{r density-sweep, eval = FALSE}
sweep_density <- function(n_init, seed = 42L) {
  s <- wolf2008_responsiveness_specs()
  s$n_agents_init <- n_init
  s$max_agents    <- max(800L, 2L * n_init)
  s$max_ticks     <- 2000L
  s$random_seed   <- seed
  env <- suppressWarnings(run_alife(s, verbose = FALSE))
  recs  <- env$agents
  alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive),
                  logical(1L))
  if (sum(alive) < 5L) return(NA_real_)
  mean(vapply(seq_along(recs)[alive],
              function(i) as.numeric(recs[[i]]$responsiveness),
              numeric(1L)))
}

n_inits <- c(50L, 100L, 200L, 350L, 500L)
resp_means <- sapply(n_inits, sweep_density)
plot(n_inits, resp_means, type = "b", pch = 19,
     xlab = "Initial population density",
     ylab = "Mean responsiveness at tick 2000",
     main = "Wolf 2008 density sweep — clade implementation")
abline(h = 0.5, lty = 2, col = "grey")
```

Expected pattern (per Wolf): monotonically decreasing. **Observed in
clade**: non-monotonic, with strongest selection-for-responsiveness at
intermediate density. At very high density, populations crash before
the trait equilibrates. This is the gap between Wolf's mean-field model
and clade's spatial implementation — see "honest discussion" below.

## Honest discussion (continuing the Phase 4 theme)

Wolf 2008 reports clean coexistence of responsive + unresponsive types
at intermediate `C` (Fig. 2c). clade's implementation produces selection
on the responsiveness trait but does NOT cleanly reproduce Wolf's
polymorphic equilibrium. Two reasons:

1. **No per-resource competition denominator** (Wolf 2008 only — Wolf
   2007's denominator was added in Phase 6 of the 0.7.0 work; the
   responsiveness module here does not yet have an analogue). Wolf's
   `(1+α·N_i)` in the patch-choice payoff is the mathematical machinery
   that makes per-agent benefit decline smoothly with the fraction of
   responders. clade's spatial competition (handling time + one-per-cell)
   is more discrete: the first responsive agent to a rich cell gets all
   the benefit; later movers get nothing. This produces *winner-take-all*
   rather than *smoothly frequency-dependent*.

2. **Population dynamics**. clade's grass-economy regulator interacts
   with the responsiveness trait in ways Wolf's model doesn't see. At
   high density, the population crashes before the trait equilibrates
   — and trait-mean interpretations on a crashing population are
   unreliable (`viability_report()` flags this).

Like Phase 4 (Trivers): clade's implementation captures the mechanism
but the spatial setting changes the dynamics in ways that are
themselves biologically interesting.

## What's left for follow-up

- **Two-context cross-consistency test** (Wolf 2008 §3): pair the
  responsiveness module with the existing kin-altruism or hawk-dove
  modules; quantify whether agents responsive in foraging are also
  responsive in social contexts.
- **Positive-feedback cost reduction**: implement `times_responsive`
  counter; cost in stage 2 = `C × (1 − k × times_responsive)`.
- **Tunable per-resource denominator**: optional `(1+α·N_i)` payoff
  scaling driven by per-cell agent density (cheap to compute via
  `cell_to_agents`-style enumeration that we already designed during
  Phase 2 even if Policy A wasn't ultimately taken).
- **Compare with Wolf 2007**: pair the responsiveness module with
  Wolf 2007 personality syndrome — does responsiveness correlate
  with the boldness-aggressiveness syndrome under both selection
  pressures?

## References

- Wolf, M., van Doorn, G.S. & Weissing, F.J. (2008) Evolutionary
  emergence of responsive and unresponsive personalities.
  *Proc. Natl. Acad. Sci. USA* 105:15825–15830. doi:10.1073/pnas.0805473105.
- DeWitt, T.J., Sih, A. & Wilson, D.S. (1998) Costs and limits of
  phenotypic plasticity. *Trends in Ecology & Evolution* 13:77–81.
- Sih, A., Bell, A.M., Johnson, J.C. & Ziemba, R.E. (2004) Behavioral
  syndromes: an integrative overview. *Q. Rev. Biol.* 79:241–277.
- Bell, A.M. (2007) Future directions in behavioural syndromes
  research. *Proc. R. Soc. B* 82:755–761.
