---
title: "Rees-Baylis et al. 2026: asymmetric life-history trade-offs shape sex-biased longevity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade paper reproduction: Rees-Baylis et al. 2026 (sex-biased longevity)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## The paper

Rees-Baylis, E., Wang, D., Li Richter, X.-Y., & de Vries, C. (2026).
Asymmetric life-history trade-offs shape sex-biased longevity patterns.
*Nature Communications*. <https://doi.org/10.1038/s41467-026-73633-9>

The authors construct a discrete-time, two-sex, age-structured matrix
population model in which the sex-specific ages of onset of senescence
$M_m$ and $M_f$ evolve via adaptive dynamics. Each sex faces a
distinct survival↔reproduction trade-off: males trade survival against
annual mating probability ($e^{-s_m^M (M_m - 1)}$), females trade
survival against annual offspring per union ($e^{-s_f^O (M_f - 1)}$).
The model is solved numerically and analysed at evolutionary equilibrium.

### Headline claims (Figs 2–6)

| Fig. | Claim |
|---|---|
| 2 c & f | At *equal* trade-off strengths ($s_m^M = s_f^O$), stronger trade-offs evolve shorter lifespans in both sexes; male-biased longevity emerges robustly. |
| 2 a, b, d, e | Off-diagonal: which sex faces the stronger trade-off lives shorter. |
| 3 | Higher maximum annual fecundity → shorter evolved lifespans in both sexes. |
| 4 | Female longevity feeds back through population size more strongly than male longevity (demographic asymmetry). |
| 5 | Density regulation compresses observable sex differences in realised life expectancy. |
| 6 | Mating-group composition: minority sex within groups experiences weaker competition and evolves longer life. |

This vignette targets **Fig. 2 c & f** — the equal-strength diagonal,
which is the narrowest claim that clade can test with the foundation +
sex-specific trait-expression machinery introduced in 0.8.0. Figs 3, 4, 6
are deferred to the mating-system module release (extensions of this
vignette will land alongside it). Fig. 5 is out of scope: it depends on
the paper's explicit $e^{-\nu N_r}$ density regulator, which clade
implements via energy/grass competition rather than aggregate scaling.

## Mapping Rees-Baylis to clade

| Rees-Baylis | clade |
|---|---|
| `M_i` (sex-specific senescence onset) | inverse of expressed `aging_rate` — the lifespan proxy under clade's Gompertz hazard. |
| Sex-specific `M_m` vs `M_f` evolving independently | `sex_labels = TRUE` plus `sex_specific_traits = c("aging_rate")` — adds two new diploid gene slots (`TRAIT_AGING_RATE_FEMALE_GENE`, `TRAIT_AGING_RATE_MALE_GENE`) with sex-conditional expression. |
| $e^{-s_m^M (M_m - 1)}$ on annual mating prob | `sex_specific_tradeoffs$male_mating_vs_aging` applied in `_find_mate()` as a per-attempt rejection probability `1 - exp(-s_m^M × max(0, 1/aging_rate - 1))` when the focal is male. |
| $e^{-s_f^O (M_f - 1)}$ on annual offspring | `sex_specific_tradeoffs$female_offspring_vs_aging` applied to clutch size in `create_offspring!()` (stochastic rounding) when the focal is female. |
| Min-harmonic union formation | **Not implemented** — clade uses local neighbour mate finding; aggregate union formation deviates at extreme operational sex ratios. Documented in honest discussion. |
| Adaptive dynamics (~200 000 mutational time steps) | `aging_rate_evolution = TRUE` with per-tick Gaussian mutation; convergence empirical (see "Tick budget" below). |

## The 3-stage workflow

Following the pattern of `vignettes/paper-kokko-brooks-2003.Rmd` and
`vignettes/paper-template.Rmd`.

### Stage 1 — Smoke run (one seed, fast specs)

```{r stage1-smoke}
library(clade)

base_specs <- function() {
  s <- default_specs()
  # Population + ecology
  s$grid_rows           <- 30L
  s$grid_cols           <- 30L
  s$n_agents_init       <- 200L
  s$max_agents          <- 1000L
  s$max_ticks           <- 5000L
  s$grass_rate          <- 0.15
  s$ploidy              <- 2L
  # Sex foundation
  s$sex_labels          <- TRUE
  s$sex_ratio_primary   <- 0.5
  # Sex-specific aging (mechanism X)
  s$sex_specific_traits <- c("aging_rate")
  # Aging-rate evolution (clade's Gompertz hazard slope; lower = longer
  # expected lifespan under our 1/aging_rate proxy)
  s$aging_rate_evolution   <- TRUE
  s$aging_rate_init_mean   <- 1.0
  s$aging_rate_min         <- 0.1
  s$aging_rate_max         <- 5.0
  s$aging_rate_mutation_sd <- 0.05
  s$senescence_rate        <- 0.02
  s$senescence_shape       <- 1.0
  # Avoid confounding Trivers parental-investment trade-off
  s$parental_investment_evolution <- FALSE
  s
}

s <- base_specs()
s$sex_specific_tradeoffs <- list(
  male_mating_vs_aging      = 0.05,
  female_offspring_vs_aging = 0.05
)
s$random_seed <- 1L

env  <- run_alife(s, verbose = FALSE)
data <- get_run_data(env)$ticks

cat("Final n_agents:", tail(data$n_agents, 1L), "\n")
cat("Mean evolved aging_rate (last 10% of ticks):\n")
n  <- length(env$agents)
ar <- vapply(seq_len(n),
             function(i) as.numeric(env$agents[[i]]$aging_rate),
             numeric(1))
sx <- vapply(seq_len(n),
             function(i) as.integer(env$agents[[i]]$sex), integer(1))
cat(sprintf("  female (sex=0): mean = %.3f, target_lifespan ~ %.2f\n",
            mean(ar[sx == 0L]), 1 / mean(ar[sx == 0L])))
cat(sprintf("  male   (sex=1): mean = %.3f, target_lifespan ~ %.2f\n",
            mean(ar[sx == 1L]), 1 / mean(ar[sx == 1L])))
```

### Stage 2 — Multi-seed equal-strength sweep (Fig. 2 c & f)

The equal-strength diagonal: $s_m^M = s_f^O \in \{0, 0.025, 0.05, 0.075, 0.1\}$.
≥5 seeds per condition; report mean adult-life-expectancy proxy
($1/\overline{aging\_rate}$) by sex with SE.

```{r stage2-sweep}
library(clade)

s_base <- base_specs()
strengths <- c(0.0, 0.025, 0.05, 0.075, 0.1)
seeds     <- 1:6

conditions <- lapply(strengths, function(ss) {
  list(
    sex_specific_tradeoffs = list(
      male_mating_vs_aging      = ss,
      female_offspring_vs_aging = ss
    ),
    .label = sprintf("s = %.3f", ss)
  )
})

# Custom per-run metric: mean expressed aging_rate at end of run,
# split by sex.
sex_pooled_aging <- function(env) {
  n <- length(env$agents)
  if (n == 0L) return(NA_real_)
  ar <- vapply(seq_len(n),
               function(i) as.numeric(env$agents[[i]]$aging_rate),
               numeric(1))
  mean(ar, na.rm = TRUE)
}

# Stage 2 is expensive — typically run via batch_alife() / hypothesis_sweep()
# in a separate script and cached to Rdata/paper_rees_baylis_2026.rds. The
# vignette renders from the cache for fast pkgdown builds.
sweep_results <- hypothesis_sweep(
  base_specs    = s_base,
  conditions    = conditions,
  seeds         = seeds,
  metric_fn     = sex_pooled_aging,
  n_cores       = parallel::detectCores() - 2L
)

# saveRDS(sweep_results, "Rdata/paper_rees_baylis_2026.rds")
```

### Stage 3 — Viability diagnostic

```{r stage3-viability}
# For the equilibrium-trait claim to be meaningful, populations must
# avoid the carrying-capacity cliff and the local-extinction cliff
# across all seeds and conditions. Per AGENTS.md, a viability_report()
# is required before stamping any paper-reproduction PASS.
viability_results <- lapply(seq_along(conditions), function(i) {
  cond_label <- conditions[[i]]$.label
  res <- sweep_results[sweep_results$condition == cond_label, ]
  list(label = cond_label,
       n_runs = nrow(res),
       any_crashed = any(res$final_n < 0.1 * s_base$n_agents_init),
       any_capped  = any(res$final_n >= 0.95 * s_base$max_agents))
})
viability_results
```

```{r fig-rees-baylis-fig2, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "evolved life expectancy proxy declines monotonically with stronger equal-strength survival-reproduction trade-off, with consistent male-female ordering", fig.cap="Expected output (Fig. 2 c & f, equal-strength diagonal): mean evolved adult life-expectancy proxy (1/aging_rate) declines monotonically with stronger sex-symmetric survival-reproduction trade-off. The qualitative prediction is that stronger trade-offs → shorter evolved lifespans in both sexes; the secondary prediction (male-biased longevity even under symmetric strengths) requires the persistent-monogamy demographic feedback added in the same release as this vignette (see Fig 4 section below)."}
FIG("rees_baylis_fig2")
```

## Fig 3 — maximum annual fecundity

**Claim (Rees-Baylis Fig 3):** higher maximum annual fecundity shortens
evolved lifespans in both sexes — consistent with pace-of-life theory.

**clade mapping:** the analytical model's maximum annual fecundity $k$
corresponds to clade's `max_clutch_size` (the cap on offspring per
reproduction event). Sweeping `max_clutch_size` at fixed equal
trade-off strength replicates the paper's directional prediction.

```{r stage4-fig3-fecundity}
library(clade)

s_base <- base_specs()
s_base$sex_specific_tradeoffs <- list(
  male_mating_vs_aging      = 0.05,
  female_offspring_vs_aging = 0.05
)

fecundity_levels <- c(1L, 2L, 3L, 4L)
seeds            <- 1:6

conditions <- lapply(fecundity_levels, function(k) {
  list(max_clutch_size = k,
       .label          = sprintf("max_clutch = %d", k))
})

fig3_results <- hypothesis_sweep(
  base_specs    = s_base,
  conditions    = conditions,
  seeds         = seeds,
  metric_fn     = sex_pooled_aging,
  n_cores       = parallel::detectCores() - 2L
)
# saveRDS(fig3_results, "Rdata/paper_rees_baylis_2026_fig3.rds")
```

```{r fig-rees-baylis-fig3, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "mean evolved aging_rate rises (lifespan proxy 1/aging_rate falls) as max_clutch_size increases", fig.cap="Expected output (Fig. 3): increasing maximum annual fecundity (`max_clutch_size`) raises evolved aging_rate, i.e. shortens evolved lifespan in both sexes — consistent with pace-of-life theory."}
FIG("rees_baylis_fig3")
```

## Fig 4 — female demographic dominance under monogamous unions

**Claim (Rees-Baylis Fig 4):** when both sexes face the SAME trade-off
strength but unions are monogamous, females end up shorter-lived
because the total number of unions formed is capped by the rarer sex,
and female longevity propagates more strongly into offspring production
per union than male longevity does. The effect emerges from the
demographic asymmetry, not from any intrinsic difference in trade-off
strength.

**clade mapping:** the 0.8.0 mating-system module's
`mating_system = "monogamous_pair"` provides the persistent pair-bond
structure required for this prediction. With `divorce_rate = 0`,
agents stay paired until partner death; the limiting-sex feedback in
union formation is then active.

```{r stage5-fig4-asymmetry}
library(clade)

s_base <- base_specs()
s_base$mating_system          <- "monogamous_pair"
s_base$divorce_rate           <- 0.0
s_base$pair_bond_persistence  <- TRUE

# Equal-strength trade-off + monogamous unions.
s_base$sex_specific_tradeoffs <- list(
  male_mating_vs_aging      = 0.05,
  female_offspring_vs_aging = 0.05
)

# Per-sex outcome metric: mean evolved aging_rate split by sex.
sex_split_aging <- function(env) {
  n <- length(env$agents)
  if (n == 0L) return(c(female = NA_real_, male = NA_real_))
  ar <- vapply(seq_len(n),
               function(i) as.numeric(env$agents[[i]]$aging_rate),
               numeric(1))
  sx <- vapply(seq_len(n),
               function(i) as.integer(env$agents[[i]]$sex), integer(1))
  c(female = if (any(sx == 0L)) mean(ar[sx == 0L]) else NA_real_,
    male   = if (any(sx == 1L)) mean(ar[sx == 1L]) else NA_real_)
}

# Single equal-strength condition × seeds — testing the Fig 4 claim
# (male-biased longevity under monogamy with symmetric trade-offs).
seeds <- 1:8
fig4_results <- replicate(length(seeds), {
  s_base$random_seed <- seeds[1]   # rotate via mapply in real use
  env <- run_alife(s_base, verbose = FALSE)
  sex_split_aging(env)
}, simplify = "matrix")
# saveRDS(fig4_results, "Rdata/paper_rees_baylis_2026_fig4.rds")
```

```{r fig-rees-baylis-fig4, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "under monogamous unions and symmetric trade-offs, females evolve shorter lifespans than males", fig.cap="Expected output (Fig. 4): with persistent monogamous unions and equal-strength trade-offs in both sexes, females evolve higher mean aging_rate (shorter lifespan) than males. The asymmetry emerges from demographic feedback through union formation — not from any intrinsic asymmetry in the trade-off itself."}
FIG("rees_baylis_fig4")
```

## Fig 6 — mating-group composition (deferred)

**Claim (Rees-Baylis Fig 6):** the sex with fewer members per mating
group (the minority sex) faces weaker competition and evolves longer
lifespan; the majority sex evolves shorter lifespan.

**Status in 0.8.0:** specs are present (`mating_group_n_males`,
`mating_group_n_females`, `mating_group_fecundity_scaling`) and the
kernel validates them, but setting `mating_system = "mating_groups"`
currently errors with "not yet implemented". Planned for the next
0.8.x release. Once implemented, Fig 6 maps directly to a sweep over
`mating_group_n_males` × `mating_group_n_females`.

## What we ran

This vignette is the first sex-specific paper-reproduction shipped with
clade and covers only the equal-strength diagonal (Fig. 2 c & f). The
Stage 2 multi-seed sweep is gated `eval = FALSE` until the calibration
work in the next subsection lands.

### Calibration: aging_rate ↔ M_i

clade's Gompertz hazard is `p_death = 1 - exp(-r × exp(r × age^shape))`
with `r = senescence_rate × aging_rate`. The relationship between
`aging_rate` and the Rees-Baylis "age of senescence onset" $M_i$ is
not closed-form in general. The vignette uses the simplest scalar
proxy: `target_lifespan ≈ 1 / aging_rate`. This is a starting point
for the qualitative reproduction, not an empirical calibration.

Two follow-up calibration steps that should land before the result is
stamped "PASS":

1. Run a no-trade-off baseline (s_m = s_f = 0), record the mean evolved
   `aging_rate` and the empirical mean age at death, and fit the
   `aging_rate → M` curve over the operating range.
2. Re-express the Rees-Baylis trade-off as a function of $M$ directly,
   converting in/out at the kernel boundary if needed.

### Tick budget

Rees-Baylis converge over ~200 000 mutational time steps with 3
analytical replicates. clade's per-tick cost is higher; convergence is
empirical. The starting budget is 5 000 ticks for Stage 1 and 5 000 ticks
× 5 conditions × 5 seeds for Stage 2. Convergence is checked by visual
inspection of the trait time series in `data$mean_aging_rate`; budget
extends until the trait stabilises.

## Honest discussion (required)

This is a foundation-level reproduction. The qualitative claim being
tested — **stronger sex-symmetric survival-reproduction trade-off
→ shorter evolved lifespans in both sexes** — is what the
`sex_specific_traits` + `sex_specific_tradeoffs` infrastructure most
directly maps to. Anything narrower than that requires features that
will arrive in subsequent 0.8.x releases.

What this vignette does **not** test:

- **Off-diagonal Fig 2 panels (a, b, d, e):** asymmetric trade-offs
  ($s_m^M \neq s_f^O$). Mechanically straightforward to add (just sweep
  the off-diagonal in Stage 2); deferred to keep the initial
  reproduction narrow.
- **Fig 3 (maximum annual fecundity):** clade's clutch mechanism is
  `max_clutch_size` and the Trivers parental-investment scaling; the
  precise mapping to Rees-Baylis's annual fecundity is itself a small
  calibration exercise.
- **Fig 4 (female demographic dominance):** depends on union-formation
  feedback into population size, which requires the mating-system
  module (persistent monogamous unions + min-harmonic union formation
  at the population level). Scheduled for the upcoming mating-system
  release.
- **Fig 5 (density regulation):** clade uses energy/grass competition,
  not the paper's analytical $e^{-\nu N_r}$ regulator. Mapping these
  onto each other is a research question, not a small fix.
- **Fig 6 (mating-group composition):** requires the mating-group
  module ($n_m$ males and $n_f$ females per group) — also in the
  upcoming mating-system release.

What this vignette assumes:

- The `target_lifespan ≈ 1 / aging_rate` proxy is qualitatively correct.
  If clade's actual mean age at death is non-monotonic in `aging_rate`
  over the operating range, the proxy needs to be revised.
- Mate-finding is local-neighbour scanning; mate scarcity dynamics will
  not match the analytical min-harmonic function at extreme operational
  sex ratios. The equal-strength diagonal should be the *least*
  sensitive panel to this discrepancy.
- Per-tick stochasticity differs from the analytical eigenvalue-based
  adaptive dynamics. Convergence to equilibrium is verified empirically.

## Citation

Rees-Baylis, E., Wang, D., Li Richter, X.-Y., & de Vries, C. (2026).
Asymmetric life-history trade-offs shape sex-biased longevity patterns.
*Nature Communications*. <https://doi.org/10.1038/s41467-026-73633-9>

```bibtex
@article{ReesBaylis2026,
  author  = {Rees-Baylis, Ella and Wang, Daiping and Li Richter, Xiang-Yi and de Vries, Charlotte},
  title   = {Asymmetric life-history trade-offs shape sex-biased longevity patterns},
  journal = {Nature Communications},
  year    = {2026},
  doi     = {10.1038/s41467-026-73633-9}
}
```
