---
title: "Wolf et al. 2007 — personality syndrome in a spatially-explicit ABM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wolf et al. 2007 — personality syndrome in a spatially-explicit ABM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Spatially-explicit clade interpretation of Wolf, van Doorn, Leimar &
Weissing (2007) Nature 447:581–584. Wolf's mean-field game-theoretic
mechanism (life-history asset protection → boldness-aggressiveness
syndrome) re-implemented in clade's grid-based ABM, with hawk-dove pairs
drawn from the Moore neighborhood and anti-predator decisions anchored
to real predator presence. Implemented in 0.7.0; see
`inst/julia/src/modules/personality.jl`.*

## What the original paper claims

Wolf 2007 proposes that consistent individual differences in
risk-taking behaviour (animal personalities) arise from a life-history
trade-off. Individuals with high future fitness expectations (much to
lose) evolve to be more risk-averse than individuals with low
expectations (little to lose). Across multiple risky contexts the same
asset-protection logic applies, producing **correlated** behavioural
traits — the boldness–aggressiveness syndrome.

Three specific predictions:

1. **Coexistence of life-history strategies** — disruptive selection on
   exploration intensity *x* yields a stable dimorphism (some individuals
   x ≈ 0, others x ≈ 1). Stable when 2·f_low > f_high > 1.
2. **Within-context consistency** — across contexts and over time, an
   individual that is bold in one anti-predator decision is bold in
   subsequent ones; aggressive in one hawk-dove encounter is aggressive
   in the next. (Repeated games per individual within a lifetime.)
3. **Cross-context correlation (the syndrome)** — boldness and
   aggressiveness positively correlate; both negatively correlate with
   exploration *x*.

Wolf's model is **non-spatial** — every game pairs randomly across the
whole population and the anti-predator game is an abstract individual
draw with no actual predator on a grid.

## Why spatially-explicit, not Wolf-faithful?

clade is a spatially-explicit ABM. Its value-add over Wolf's analytic
mean-field model is precisely that encounters happen *somewhere*, between
*neighbouring* individuals, and that the resource economy is regulated by
the grid. We could implement Wolf's model verbatim (random pairing across
the population), but that would throw away clade's spatial machinery and
reproduce a model that's already analytically solved.

The interpretation here:

- **Hawk-dove game**: pair the focal between-phase agent with a random
  between-phase neighbour in its Moore neighborhood (default radius 1,
  configurable via `personality_hawkdove_radius`). If no qualifying
  neighbour, no game this tick.
- **Anti-predator game**: fires only when a real predator is within Moore
  radius 2 of the focal agent (uses `env.predator_map`). The bold/shy
  draw on `agent.boldness` then determines whether the agent gets the
  bold-action energy bonus and incurs the bold-action mortality risk.
- **Year-1 / year-2 reproduction**: triggered at fixed agent ages
  (`wolf_year1_repro_age = 50`, `wolf_year2_repro_age = 100`). Year-2
  reproduction also kills the parent. Game payoffs accumulated during
  the between-phase (year1 < age < year2) fold into year-2 fecundity.

This is an **extension** of Wolf 2007, not a faithful reproduction. The
expected differences from Wolf's published correlations:

- Selection signal is **weaker** because (a) hawk-dove encounters depend
  on local density rather than uniform mixing, and (b) anti-predator
  events depend on predator proximity, not a constant abstract risk.
- Wolf's `(1 + α·N_i)` per-resource competition denominator is replaced
  by clade's grid-based spatial competition (foraging on cells, handling
  time, energy budget). This regulates total population but doesn't
  sharply produce the year-1-vs-year-2 dimorphism Wolf reports.

## The headline result

```{r setup-libs, eval = FALSE}
library(clade)
library(ggplot2)

s <- wolf_personality_specs()
s$max_ticks                       <- 5000L      # ~50 generations at year2 = 100
s$random_seed                     <- 42L
s$n_agents_init                   <- 200L
s$max_agents                      <- 800L
s$personality_hawkdove_per_tick   <- 0.3
s$personality_hawkdove_radius     <- 2L
s$n_predators_init                <- 10L       # anti-predator needs real predators
s$predator_max_agents             <- 30L
env <- run_alife(s, verbose = FALSE)
```

Pull final-tick trait values and check the three Wolf correlations:

```{r corr, eval = FALSE}
recs  <- env$agents
alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive), logical(1L))
ex    <- vapply(seq_along(recs)[alive], function(i) as.numeric(recs[[i]]$exploration),    numeric(1L))
bo    <- vapply(seq_along(recs)[alive], function(i) as.numeric(recs[[i]]$boldness),       numeric(1L))
ag    <- vapply(seq_along(recs)[alive], function(i) as.numeric(recs[[i]]$aggressiveness), numeric(1L))

cor(ex, bo)   # predicted negative — high-x agents are less bold
cor(ex, ag)   # predicted negative — high-x agents are less aggressive
cor(bo, ag)   # predicted positive — boldness-aggressiveness syndrome
```

Typical values at this configuration (seed = 42 only):

| Correlation | Predicted | Phase 3a (no denominator) | Phase 6 (with `α=0.005`) |
|---|---|---|---|
| Pearson(exploration, boldness)        | < 0 | −0.21 | +0.04 |
| Pearson(exploration, aggressiveness)  | < 0 | −0.09 | −0.01 |
| Pearson(boldness, aggressiveness)     | > 0 | +0.10 | **+0.31** |

This single-seed observation suggested the Phase 6 `(1+α·N_i)`
denominator strengthened the bold-aggro syndrome from +0.10 to
**+0.31** — a threefold improvement. **But that read does not survive
multi-seed scrutiny.** The next section reports the rigorous 8-seed
verdict, which reclassifies all three correlations as null.

## Multi-seed verification (2026-05-16)

A multi-seed regression baseline at the default `wolf_personality_specs()`
configuration — 8 seeds × 5000 ticks, single condition — was generated
by `dev/audit/fidelity/paper_wolf2007.R` and saved to
`dev/audit/fidelity/paper_wolf2007.rds`. Per-seed correlations:

| seed | cor(bold, aggro) | cor(exp, bold) | cor(exp, aggro) | n_alive |
|---:|---:|---:|---:|---:|
| 1 | **−0.370** | −0.017 | −0.023 | 402 |
| 2 | +0.248 | +0.013 | −0.115 | 401 |
| 3 | −0.201 | +0.017 | −0.226 | 424 |
| 4 | +0.201 | +0.068 | +0.058 | 402 |
| 5 | −0.331 | −0.142 | +0.262 | 422 |
| 6 | +0.052 | +0.032 | −0.084 | 430 |
| 7 | +0.193 | −0.034 | +0.014 | 419 |
| 8 | −0.251 | +0.064 | −0.163 | 418 |

Aggregate across 8 seeds, with 2σ verdict ladder:

| Correlation | Predicted | Mean ± SE | t | Verdict |
|---|---|---:|---:|:---:|
| cor(bold, aggro) | > 0 (syndrome) | **−0.057 ± 0.091** | −0.63 | **null** |
| cor(exp, bold) | < 0 (asset-protection) | +0.000 ± 0.024 | +0.01 | **null** |
| cor(exp, aggro) | < 0 (asset-protection) | −0.034 ± 0.054 | −0.64 | **null** |

**The boldness-aggressiveness syndrome is null at 8 seeds.** Per-seed
correlations range from **−0.37** (seed 1) to **+0.25** (seed 2) with
no systematic direction. The single-seed observation reported above
(+0.31 at seed 42, which is *not* one of the 8 seeds in this rerun
— the rerun used seeds 1–8 to match clade's `batch_seeds()` default)
falls inside the +0.21 to +0.25 upper tail of the 8-seed distribution
but isn't representative.

**Honest reclassification:** in clade's current kernel, the spatially-
explicit interpretation of Wolf 2007 does NOT robustly reproduce the
personality syndrome at the 5000-tick / 8-seed scale tested here.
This is consistent with the "Honest discussion of magnitudes" below
(weaker selection signal from local-density encounters) but **more
severe** than the single-seed result suggested — the verdict is null,
not a small but reliable positive.

Possible explanations for the gap, in decreasing order of likelihood:

- **Seed variance dominates the signal at 5000 ticks.** The SE on
  bold-aggro (0.091) exceeds the absolute mean of every single-seed
  observation except 1 and 5. Longer runs (15 000+ ticks) might
  resolve an equilibrium, but that's also the regime where Wolf's
  syndrome decays per McElreath 2007 — see "Critique-aware framing"
  below.
- **Spatial-pairing structure breaks the mean-field signal.** Wolf's
  result depends on every agent pairing with every other in the
  between-phase; clade's Moore-neighborhood pairing produces a
  fundamentally different game structure. The per-strategy
  `(1+α·N_i)` denominator (Phase 6) is the only kernel knob that
  partially bridges the gap; tweaking the snapshot frequency or
  hawk-dove radius might push the signal up but should be tested
  explicitly.
- **Implementation issue in `inst/julia/src/modules/personality.jl`.**
  Lower-likelihood but worth keeping on the list. The deterministic
  kernel module has been reviewed but a multi-seed null at this scale
  is a candidate finding for Sergio's v0.8-core review.

The seed-42 +0.31 observation should be read as a noise-envelope
landing, not a robust syndrome signal. Phase 6's
`(1+α·N_i)` denominator may still be doing useful work (none of the
seeds produce the +0.30+ magnitude under the pre-Phase-6 codepath
either) — but its benefit has to be re-evaluated against the multi-
seed baseline, not the seed-42 single-shot.

## Visualising the trait distribution

```{r plot-traits, eval = FALSE}
trait_df <- data.frame(
  exploration    = ex,
  boldness       = bo,
  aggressiveness = ag
)

ggplot(trait_df, aes(x = exploration, y = boldness)) +
  geom_point(alpha = 0.5, colour = "#377eb8") +
  geom_smooth(method = "lm", se = TRUE, colour = "#e41a1c") +
  labs(title = "Wolf 2007 prediction: high exploration → low boldness",
       subtitle = sprintf("n = %d  Pearson r = %+.3f", length(ex), cor(ex, bo)),
       x = "Exploration trait (x in Wolf's notation)",
       y = "Boldness") +
  theme_classic(base_size = 12)

ggplot(trait_df, aes(x = boldness, y = aggressiveness)) +
  geom_point(alpha = 0.5, colour = "#984ea3") +
  geom_smooth(method = "lm", se = TRUE, colour = "#e41a1c") +
  labs(title = "Wolf 2007 syndrome: boldness and aggressiveness positively correlate",
       subtitle = sprintf("n = %d  Pearson r = %+.3f", length(ex), cor(bo, ag)),
       x = "Boldness", y = "Aggressiveness") +
  theme_classic(base_size = 12)
```

## Honest discussion of magnitudes

Wolf 2007 reports correlations typically > 0.5 in his Fig 4 quantitative-
genetics extension. clade's spatially-explicit interpretation produces
correlations of order 0.1. Why the difference?

1. **Encounter rate**. Wolf's mean-field model has every agent paired
   with another every game tick. clade's neighborhood pairing depends on
   local density. At 200 initial agents on 30×30 = 900 cells (~22%
   density), an agent's Moore-radius-2 neighborhood (24 cells) contains
   ~5 other agents on average — but only some of those will be in the
   between-phase, so actual game encounters are sparser than Wolf's mean-
   field rate.

2. **The `(1 + α·N_i)` competition denominator (Phase 6 refinement)**.
   clade now applies Wolf's per-strategy denominator with
   `N_high ≈ N·mean(x)` and `N_low ≈ N·mean(1-x)`, snapshotted once per
   tick. This strengthened the bold-aggro syndrome from +0.10 to +0.31,
   but did not produce Wolf's year-1-vs-year-2 dimorphism: the population
   still locks at x≈0 because the snapshot-once-per-tick approximation
   plus clade's spatial competition is coarser than Wolf's continuous
   per-individual evaluation. The asset-protection signs (exp ↔ bold,
   exp ↔ aggro) are weaker / sign-flipped in the locked equilibrium.

3. **Population dynamics**. clade's energy economy regulates carrying
   capacity in addition to the (1+α·N) denominator. The two interact:
   the energy economy already prevents runaway growth, so the
   denominator's regulating role is partly redundant and the resulting
   selection gradient on *x* is weaker than in Wolf's pure mean-field
   model.

These are real interpretation choices, not bugs. clade's value is that
it embeds Wolf's mechanism in a richer biological setting; the trade-off
is weaker / sign-flipped asset-protection signal even though the
defining bold-aggro syndrome is preserved.

## Diploid extension (Wolf 2007 Fig 4) — clade does NOT recover the syndrome

Wolf 2007 §"Genetic architecture" reports that the diploid quantitative-
genetics version of his model yields **stronger** correlations
(r ≈ 0.65) than the haploid basic model. The conventional explanation is
that diploidy + Mendelian inheritance buffers selection and allows
genetic variation to accumulate at the asset-protection equilibrium.

clade supports diploidy out of the box — the `genome.jl` meiosis logic
handles ploidy = 2 — so testing this is one line in the preset:

```{r diploid, eval = FALSE}
s <- wolf_personality_specs()
s$ploidy <- 2L            # diploid; Wolf Fig 4 setting
env <- run_alife(s, verbose = FALSE)
```

**Observed result at seed 42 (5000 ticks, otherwise identical
configuration)**:

| Correlation | Predicted | Haploid (Phase 6) | Diploid |
|---|---|---:|---:|
| Pearson(exploration, boldness)        | < 0 | +0.044 | **−0.075** |
| Pearson(exploration, aggressiveness)  | < 0 | −0.005 | **−0.068** |
| Pearson(boldness, aggressiveness)     | > 0 | **+0.307** | −0.003 |

The diploid extension produces the **opposite** trade-off from haploid
Phase 6:

- The **bold-aggro syndrome disappears** (+0.307 → −0.003). Recombination
  decouples the two trait expressions across generations, so the
  shared-asset-protection mechanism that links them in haploid mode
  no longer produces a correlation.
- The **asset-protection signs flip toward the right direction**
  (−0.075, −0.068). Diploidy seems to push the population into a
  different equilibrium structure where exploration *does* trade off
  against risk-taking, in line with Wolf's prediction.

Neither pattern is what Wolf reports for his analytic diploid model.
Wolf's *all three* correlations strengthen together; clade's diploid
extension produces a *swap* — losing one and gaining the others.

The most likely explanation is the same as the haploid Phase 6 trade-off:
clade's grid-based competition and one-per-cell offspring placement
constrain the equilibrium population structure in ways Wolf's mean-field
model doesn't see. Diploid recombination interacts with this
spatially-imposed structure to produce a new attractor.

The test at `tests/testthat/test-personality-syndrome.R` asserts only
that the diploid scenario completes without crashing and produces a
viable population; the specific correlation magnitudes are documented
here as honest findings rather than enforced.

## What's left for follow-up work

- **Per-individual (not snapshot) frequency evaluation** — switch to
  per-agent re-computation of `(1+α·N_high)` / `(1+α·N_low)` to recover
  the smoothness Wolf's continuous model assumes. Restructures
  `apply_lifehistory_tradeoff!`.
- **Spatial vs random pairing comparison** — run side-by-side with
  `personality_hawkdove_radius = 1` (clade-native) vs an alternative
  global-pairing implementation (Wolf-faithful) to quantify the cost of
  spatial restriction.
- **Sensitivity to game frequency** — the
  `personality_hawkdove_per_tick` and `personality_antipred_per_tick`
  parameters control encounter rate; sweeping these would map the
  encounter-rate-to-syndrome-strength relationship.

## Critique-aware framing

Two published critiques of Wolf 2007 are directly testable in clade.
Both have now been audited with multi-seed regression baselines
(2026-05-16) — and both single-seed claims that *appeared* to
confirm the critiques **are themselves contradicted at 8 seeds**.

- **Massol & Crochet (2008)** Nature 451:doi:10.1038/nature06744 —
  argue the dimorphism depends critically on β (the trade-off
  shape). Tested in `vignette("paper-massol-crochet-2008")`. The
  single-seed sweep suggested syndrome peaks at Wolf's published
  β = 1.25 (+0.307). At 8 seeds × 5 β values, **β = 1.25 is the
  trough** (−0.077, marginal-negative); no β value produces a
  robust positive syndrome. Range across β shrinks from single-seed
  0.39 to multi-seed 0.13. The β-sensitivity claim survives; the
  "peak at Wolf's β" claim is contradicted.
- **McElreath, Luttbeg, Fogarty, Brodin & Sih (2007)** Nature
  450:doi:10.1038/nature06326 — argue asset protection is a
  *negative-feedback* mechanism that erodes individual differences
  over time. Tested in `vignette("paper-mcelreath-2007")`. The
  single-seed trajectory "+0.307 at 5 000 ticks decaying to +0.032
  at 15 000" was a noise-envelope landing. At 8 seeds × 3 horizons,
  bold-aggro is null at every horizon (−0.075, −0.077, +0.034) —
  there's no peak to decay from. Surprise positive finding:
  `cor(exp, aggro)` at 2 000 ticks **PASSes negative** (t = −4.12),
  opposite to McElreath's "later-peak" framing.

> **Cumulative multi-seed verdict** across all three Wolf 2007
> reproductions in clade (`paper-wolf2007`,
> `paper-massol-crochet-2008`, `paper-mcelreath-2007`): the
> boldness-aggressiveness syndrome does NOT robustly emerge at any
> (β, time-horizon) combination tested. All seed-42 +0.307 single-
> seed observations were noise-envelope landings on the positive
> tail. Both critique vignettes' *apparent confirmations* dissolved
> with the rds upgrade. **One robust positive finding survives**
> (confirmed at 16 seeds, PR #147): in the McElreath horizon sweep,
> `cor(exp, aggro)` at 2 000 ticks PASSes-negative — the
> asset-protection signal *does* appear in clade's spatial kernel,
> but in the asset-protection correlation rather than the syndrome
> correlation, and at the *earliest* horizon (opposite to McElreath's
> "later-peak" prediction). Both the syndrome null and the
> early-horizon asset-protection finding are candidates for the
> v0.8-core review.

## References

- Wolf, M., van Doorn, G.S., Leimar, O. & Weissing, F.J. (2007)
  Life-history trade-offs favour the evolution of animal personalities.
  *Nature* 447:581–584. doi:10.1038/nature05835.
- Massol, F. & Crochet, P. (2008) Do animal personalities emerge?
  *Nature* 451 (Brief Communications Arising). doi:10.1038/nature06744.
- McElreath, R., Luttbeg, B., Fogarty, S.P., Brodin, T. & Sih, A. (2007)
  Evolution of animal personalities. *Nature* 450 (Brief Communications
  Arising). doi:10.1038/nature06326.
- Clark, C.W. (1994) Antipredator behavior and the asset-protection
  principle. *Behavioral Ecology* 5:159–170.
