---
title: "Massol & Crochet 2008 — does Wolf's syndrome depend on β?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Massol & Crochet 2008 — does Wolf's syndrome depend on β?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*A direct test in clade of Massol & Crochet's (2008) critique of Wolf
et al. 2007: the dimorphism in exploration intensity x depends critically
on the trade-off shape parameter β in g(x) = (1-x)^β. Implemented as
a β sweep over the spatially-explicit Wolf 2007 scenario added in 0.7.0.
Companion to `vignette("paper-wolf2007")`.*

## The critique

Wolf et al. (2007) report a stable dimorphism in exploration intensity
*x* under disruptive selection at year-2 reproduction. The trade-off
between year-1 and year-2 fecundity is parameterised by

\[ g(x) = (1 - x)^\beta \]

so that year-1 fecundity scales as (1-x)^β · f_low and year-2 fecundity
as x · f_high (plus the (1-x) leg). Wolf uses β = 1.25.

Massol & Crochet (2008, *Nature* 451, brief communications arising;
`doi:10.1038/nature06744`) argued that this dimorphism is not robust:
shift β much above or below Wolf's value and the bimodal x distribution
collapses. Their critique is mathematical (the conditions for stability
of the mixed equilibrium have a narrow β range) but easy to test in
a simulation.

clade's Wolf scenario exposes β as `personality_beta`. Sweeping it
across [0.5, 3.0] at fixed seed gives a direct probe of the critique.

## The sweep

```{r setup-libs, eval = FALSE}
library(clade)

run_one <- function(beta_val, seed = 42L) {
  s <- wolf_personality_specs()
  s$personality_beta                <- beta_val
  s$max_ticks                       <- 5000L
  s$random_seed                     <- seed
  s$n_agents_init                   <- 200L
  s$max_agents                      <- 800L
  s$personality_hawkdove_per_tick   <- 0.3
  s$personality_hawkdove_radius     <- 2L
  s$n_predators_init                <- 10L
  s$predator_max_agents             <- 30L
  env <- run_alife(s, verbose = FALSE)

  recs  <- env$agents
  alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive),
                  logical(1L))
  if (sum(alive) < 30L) return(NULL)

  ex <- vapply(seq_along(recs)[alive],
               function(i) as.numeric(recs[[i]]$exploration), numeric(1L))
  bo <- vapply(seq_along(recs)[alive],
               function(i) as.numeric(recs[[i]]$boldness), numeric(1L))
  ag <- vapply(seq_along(recs)[alive],
               function(i) as.numeric(recs[[i]]$aggressiveness), numeric(1L))
  list(
    beta   = beta_val,
    n      = length(ex),
    r_eb   = suppressWarnings(cor(ex, bo)),
    r_ea   = suppressWarnings(cor(ex, ag)),
    r_ba   = suppressWarnings(cor(bo, ag)),
    mean_x = mean(ex)
  )
}

betas   <- c(0.5, 1.0, 1.25, 2.0, 3.0)
results <- lapply(betas, run_one)
```

## Observed pattern (seed 42, single replicate)

| β     | n   | cor(exp, bold) | cor(exp, aggro) | cor(bold, aggro) | mean(x) |
|------:|----:|---------------:|----------------:|-----------------:|--------:|
| 0.50  | 676 |  +0.017        |  +0.058         |  **+0.188**      | 0.103   |
| 1.00  | 639 |  +0.071        |  −0.142         |  −0.025          | 0.102   |
| 1.25  | 670 |  +0.044        |  −0.005         |  **+0.307**      | 0.094   |
| 2.00  | 676 |  −0.054        |  +0.075         |  **+0.124**      | 0.097   |
| 3.00  | 641 |  −0.132        |  −0.063         |  −0.081          | 0.077   |

Three of five β values produce the predicted positive bold-aggro
syndrome; two produce essentially-zero or negative values. The
syndrome strength varies by **0.39 in correlation units across the
β range** (peak −0.08 at β = 3.0 to peak +0.31 at β = 1.25).

The asset-protection signs (cor(exp, bold) and cor(exp, aggro))
also shift with β: weakly positive at β = 0.5, becoming clearly
negative at β = 3.0. This is in Wolf's predicted direction
(high-x agents should be less bold / less aggressive) but only
emerges at the high end of the β range tested.

## Reading the result

Massol & Crochet's critique is **partially borne out** in clade's
spatially-explicit implementation:

1. **The syndrome is β-sensitive.** Wolf reports the bold-aggro
   correlation as the defining prediction of his model; varying β
   over a biologically plausible range moves that correlation by
   ~0.4. Not constant.
2. **The peak coincides with Wolf's published value.** β = 1.25
   gives the strongest syndrome in the sweep (+0.307). This is
   consistent with Wolf having chosen a parameter that maximised
   the headline result — defensible publication practice, but the
   result is fragile to that choice.
3. **The asset-protection signs become Wolf-predicted only at high β.**
   At β = 0.5 (weak penalty on high-x agents) the signs are wrong;
   they flip to Wolf's predicted negative direction by β = 3.0. This
   is the *opposite* fragility: the asset-protection mechanism is
   strongest at high β, but at that same β the bold-aggro syndrome
   collapses.

The two-correlation tension (syndrome peaks at intermediate β; asset-
protection signs strengthen at high β) suggests clade's spatial
implementation produces **two distinct selection regimes** rather
than one coherent dimorphism. Massol & Crochet's framing — that the
mathematical conditions for stable dimorphism are narrow — is
consistent with what we see, but the spatial implementation adds
its own structure: the population locks at x ≈ 0.1 at every β tested,
so the *true* x dimorphism Wolf predicts (x ≈ 0 vs x ≈ 1 modes) is
absent regardless.

## Caveats

- **Single seed** (42). The numbers above will move with seed-to-seed
  variation. A proper test would run 5–10 seeds per β and report
  distributions. The pattern is suggestive, not definitive.
- **Phase 6 denominator on.** The runs include the per-strategy
  `(1+α·N_i)` competition denominator that strengthens the bold-aggro
  syndrome at Wolf's β = 1.25 (see `vignette("paper-wolf2007")`,
  "Honest discussion of magnitudes"). Setting `personality_alpha = 0`
  reverts to Phase 3a behaviour and would shift these numbers.
- **No bimodality at any β.** Wolf's headline result is a stable
  dimorphism (x ≈ 0 mode + x ≈ 1 mode). clade's mean(x) ≈ 0.1 at
  every β; the population is unimodal at low x throughout the sweep.
  This matches the Phase 6 finding in paper-wolf2007.Rmd — clade's
  spatial competition + one-per-cell offspring placement constrain
  the equilibrium to a single attractor.

## What this implies for the Wolf 2007 reproduction

The 0.7.0 implementation of Wolf 2007 *did* reproduce the defining
bold-aggro syndrome at Wolf's parameter values. This vignette shows
that **the syndrome's strength is genuinely sensitive to β** — which
is exactly what Massol & Crochet predicted. Future users of the Wolf
scenario should treat the syndrome strength as parameter-dependent,
not as a fixed property of the mechanism.

## References

- Massol, F. & Crochet, P. (2008) Do animal personalities emerge?
  *Nature* 451 (Brief Communications Arising). doi:10.1038/nature06744.
- Wolf, M., van Doorn, G.S., Leimar, O. & Weissing, F.J. (2007)
  Life-history trade-offs favour the evolution of animal personalities.
  *Nature* 447:581–584. doi:10.1038/nature05835.
- See also `vignette("paper-wolf2007")` for the base reproduction and
  the Phase 6 (1+α·N_i) denominator refinement that strengthens the
  syndrome at Wolf's β = 1.25.
