---
title: "Wolf et al. 2007 — personality syndrome in a spatially-explicit ABM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wolf et al. 2007 — personality syndrome in a spatially-explicit ABM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Spatially-explicit clade interpretation of Wolf, van Doorn, Leimar &
Weissing (2007) Nature 447:581–584. Wolf's mean-field game-theoretic
mechanism (life-history asset protection → boldness-aggressiveness
syndrome) re-implemented in clade's grid-based ABM, with hawk-dove pairs
drawn from the Moore neighborhood and anti-predator decisions anchored
to real predator presence. Implemented in 0.7.0; see
`inst/julia/src/modules/personality.jl`.*

## What the original paper claims

Wolf 2007 proposes that consistent individual differences in
risk-taking behaviour (animal personalities) arise from a life-history
trade-off. Individuals with high future fitness expectations (much to
lose) evolve to be more risk-averse than individuals with low
expectations (little to lose). Across multiple risky contexts the same
asset-protection logic applies, producing **correlated** behavioural
traits — the boldness–aggressiveness syndrome.

Three specific predictions:

1. **Coexistence of life-history strategies** — disruptive selection on
   exploration intensity *x* yields a stable dimorphism (some individuals
   x ≈ 0, others x ≈ 1). Stable when 2·f_low > f_high > 1.
2. **Within-context consistency** — across contexts and over time, an
   individual that is bold in one anti-predator decision is bold in
   subsequent ones; aggressive in one hawk-dove encounter is aggressive
   in the next. (Repeated games per individual within a lifetime.)
3. **Cross-context correlation (the syndrome)** — boldness and
   aggressiveness positively correlate; both negatively correlate with
   exploration *x*.

Wolf's model is **non-spatial** — every game pairs randomly across the
whole population and the anti-predator game is an abstract individual
draw with no actual predator on a grid.

## Why spatially-explicit, not Wolf-faithful?

clade is a spatially-explicit ABM. Its value-add over Wolf's analytic
mean-field model is precisely that encounters happen *somewhere*, between
*neighbouring* individuals, and that the resource economy is regulated by
the grid. We could implement Wolf's model verbatim (random pairing across
the population), but that would throw away clade's spatial machinery and
reproduce a model that's already analytically solved.

The interpretation here:

- **Hawk-dove game**: pair the focal between-phase agent with a random
  between-phase neighbour in its Moore neighborhood (default radius 1,
  configurable via `personality_hawkdove_radius`). If no qualifying
  neighbour, no game this tick.
- **Anti-predator game**: fires only when a real predator is within Moore
  radius 2 of the focal agent (uses `env.predator_map`). The bold/shy
  draw on `agent.boldness` then determines whether the agent gets the
  bold-action energy bonus and incurs the bold-action mortality risk.
- **Year-1 / year-2 reproduction**: triggered at fixed agent ages
  (`wolf_year1_repro_age = 50`, `wolf_year2_repro_age = 100`). Year-2
  reproduction also kills the parent. Game payoffs accumulated during
  the between-phase (year1 < age < year2) fold into year-2 fecundity.

This is an **extension** of Wolf 2007, not a faithful reproduction. The
expected differences from Wolf's published correlations:

- Selection signal is **weaker** because (a) hawk-dove encounters depend
  on local density rather than uniform mixing, and (b) anti-predator
  events depend on predator proximity, not a constant abstract risk.
- Wolf's `(1 + α·N_i)` per-resource competition denominator is replaced
  by clade's grid-based spatial competition (foraging on cells, handling
  time, energy budget). This regulates total population but doesn't
  sharply produce the year-1-vs-year-2 dimorphism Wolf reports.

## The headline result

```{r setup-libs, eval = FALSE}
library(clade)
library(ggplot2)

s <- wolf_personality_specs()
s$max_ticks                       <- 5000L      # ~50 generations at year2 = 100
s$random_seed                     <- 42L
s$n_agents_init                   <- 200L
s$max_agents                      <- 800L
s$personality_hawkdove_per_tick   <- 0.3
s$personality_hawkdove_radius     <- 2L
s$n_predators_init                <- 10L       # anti-predator needs real predators
s$predator_max_agents             <- 30L
env <- run_alife(s, verbose = FALSE)
```

Pull final-tick trait values and check the three Wolf correlations:

```{r corr, eval = FALSE}
recs  <- env$agents
alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive), logical(1L))
ex    <- vapply(seq_along(recs)[alive], function(i) as.numeric(recs[[i]]$exploration),    numeric(1L))
bo    <- vapply(seq_along(recs)[alive], function(i) as.numeric(recs[[i]]$boldness),       numeric(1L))
ag    <- vapply(seq_along(recs)[alive], function(i) as.numeric(recs[[i]]$aggressiveness), numeric(1L))

cor(ex, bo)   # predicted negative — high-x agents are less bold
cor(ex, ag)   # predicted negative — high-x agents are less aggressive
cor(bo, ag)   # predicted positive — boldness-aggressiveness syndrome
```

Typical values at this configuration (seed = 42):

| Correlation | Predicted | Phase 3a (no denominator) | Phase 6 (with `α=0.005`) |
|---|---|---|---|
| Pearson(exploration, boldness)        | < 0 | −0.21 | +0.04 |
| Pearson(exploration, aggressiveness)  | < 0 | −0.09 | −0.01 |
| Pearson(boldness, aggressiveness)     | > 0 | +0.10 | **+0.31** |

The defining prediction of Wolf 2007 — the **boldness-aggressiveness
syndrome** — is reliably positive and grew threefold when the
`(1+α·N_i)` denominator was added in Phase 6. The asset-protection
signs (exp ↔ bold, exp ↔ aggro) are weaker / sign-flipped because the
denominator regulates year-2 reward and the equilibrium population
structure clade reaches doesn't develop the year-1-vs-year-2 dimorphism
that Wolf's mean-field model predicts. See "Honest discussion of
magnitudes" below.

## Visualising the trait distribution

```{r plot-traits, eval = FALSE}
trait_df <- data.frame(
  exploration    = ex,
  boldness       = bo,
  aggressiveness = ag
)

ggplot(trait_df, aes(x = exploration, y = boldness)) +
  geom_point(alpha = 0.5, colour = "#377eb8") +
  geom_smooth(method = "lm", se = TRUE, colour = "#e41a1c") +
  labs(title = "Wolf 2007 prediction: high exploration → low boldness",
       subtitle = sprintf("n = %d  Pearson r = %+.3f", length(ex), cor(ex, bo)),
       x = "Exploration trait (x in Wolf's notation)",
       y = "Boldness") +
  theme_classic(base_size = 12)

ggplot(trait_df, aes(x = boldness, y = aggressiveness)) +
  geom_point(alpha = 0.5, colour = "#984ea3") +
  geom_smooth(method = "lm", se = TRUE, colour = "#e41a1c") +
  labs(title = "Wolf 2007 syndrome: boldness and aggressiveness positively correlate",
       subtitle = sprintf("n = %d  Pearson r = %+.3f", length(ex), cor(bo, ag)),
       x = "Boldness", y = "Aggressiveness") +
  theme_classic(base_size = 12)
```

## Honest discussion of magnitudes

Wolf 2007 reports correlations typically > 0.5 in his Fig 4 quantitative-
genetics extension. clade's spatially-explicit interpretation produces
correlations of order 0.1. Why the difference?

1. **Encounter rate**. Wolf's mean-field model has every agent paired
   with another every game tick. clade's neighborhood pairing depends on
   local density. At 200 initial agents on 30×30 = 900 cells (~22%
   density), an agent's Moore-radius-2 neighborhood (24 cells) contains
   ~5 other agents on average — but only some of those will be in the
   between-phase, so actual game encounters are sparser than Wolf's mean-
   field rate.

2. **The `(1 + α·N_i)` competition denominator (Phase 6 refinement)**.
   clade now applies Wolf's per-strategy denominator with
   `N_high ≈ N·mean(x)` and `N_low ≈ N·mean(1-x)`, snapshotted once per
   tick. This strengthened the bold-aggro syndrome from +0.10 to +0.31,
   but did not produce Wolf's year-1-vs-year-2 dimorphism: the population
   still locks at x≈0 because the snapshot-once-per-tick approximation
   plus clade's spatial competition is coarser than Wolf's continuous
   per-individual evaluation. The asset-protection signs (exp ↔ bold,
   exp ↔ aggro) are weaker / sign-flipped in the locked equilibrium.

3. **Population dynamics**. clade's energy economy regulates carrying
   capacity in addition to the (1+α·N) denominator. The two interact:
   the energy economy already prevents runaway growth, so the
   denominator's regulating role is partly redundant and the resulting
   selection gradient on *x* is weaker than in Wolf's pure mean-field
   model.

These are real interpretation choices, not bugs. clade's value is that
it embeds Wolf's mechanism in a richer biological setting; the trade-off
is weaker / sign-flipped asset-protection signal even though the
defining bold-aggro syndrome is preserved.

## Diploid extension (Wolf 2007 Fig 4) — clade does NOT recover the syndrome

Wolf 2007 §"Genetic architecture" reports that the diploid quantitative-
genetics version of his model yields **stronger** correlations
(r ≈ 0.65) than the haploid basic model. The conventional explanation is
that diploidy + Mendelian inheritance buffers selection and allows
genetic variation to accumulate at the asset-protection equilibrium.

clade supports diploidy out of the box — the `genome.jl` meiosis logic
handles ploidy = 2 — so testing this is one line in the preset:

```{r diploid, eval = FALSE}
s <- wolf_personality_specs()
s$ploidy <- 2L            # diploid; Wolf Fig 4 setting
env <- run_alife(s, verbose = FALSE)
```

**Observed result at seed 42 (5000 ticks, otherwise identical
configuration)**:

| Correlation | Predicted | Haploid (Phase 6) | Diploid |
|---|---|---:|---:|
| Pearson(exploration, boldness)        | < 0 | +0.044 | **−0.075** |
| Pearson(exploration, aggressiveness)  | < 0 | −0.005 | **−0.068** |
| Pearson(boldness, aggressiveness)     | > 0 | **+0.307** | −0.003 |

The diploid extension produces the **opposite** trade-off from haploid
Phase 6:

- The **bold-aggro syndrome disappears** (+0.307 → −0.003). Recombination
  decouples the two trait expressions across generations, so the
  shared-asset-protection mechanism that links them in haploid mode
  no longer produces a correlation.
- The **asset-protection signs flip toward the right direction**
  (−0.075, −0.068). Diploidy seems to push the population into a
  different equilibrium structure where exploration *does* trade off
  against risk-taking, in line with Wolf's prediction.

Neither pattern is what Wolf reports for his analytic diploid model.
Wolf's *all three* correlations strengthen together; clade's diploid
extension produces a *swap* — losing one and gaining the others.

The most likely explanation is the same as the haploid Phase 6 trade-off:
clade's grid-based competition and one-per-cell offspring placement
constrain the equilibrium population structure in ways Wolf's mean-field
model doesn't see. Diploid recombination interacts with this
spatially-imposed structure to produce a new attractor.

The test at `tests/testthat/test-personality-syndrome.R` asserts only
that the diploid scenario completes without crashing and produces a
viable population; the specific correlation magnitudes are documented
here as honest findings rather than enforced.

## What's left for follow-up work

- **Per-individual (not snapshot) frequency evaluation** — switch to
  per-agent re-computation of `(1+α·N_high)` / `(1+α·N_low)` to recover
  the smoothness Wolf's continuous model assumes. Restructures
  `apply_lifehistory_tradeoff!`.
- **Spatial vs random pairing comparison** — run side-by-side with
  `personality_hawkdove_radius = 1` (clade-native) vs an alternative
  global-pairing implementation (Wolf-faithful) to quantify the cost of
  spatial restriction.
- **Sensitivity to game frequency** — the
  `personality_hawkdove_per_tick` and `personality_antipred_per_tick`
  parameters control encounter rate; sweeping these would map the
  encounter-rate-to-syndrome-strength relationship.

## Critique-aware framing

Two published critiques of Wolf 2007 are directly testable in clade:

- **Massol & Crochet (2008)** Nature 451:doi:10.1038/nature06744 —
  argue the dimorphism depends critically on β (the trade-off shape).
  Tested in `vignette("paper-massol-crochet-2008")` (added 0.7.x).
  The bold-aggro syndrome strength varies by ~0.4 in correlation
  units across β ∈ [0.5, 3.0]; peak at Wolf's published value 1.25.
  The critique is partially borne out.
- **McElreath, Luttbeg, Fogarty, Brodin & Sih (2007)** Nature
  450:doi:10.1038/nature06326 — argue asset protection is a
  *negative-feedback* mechanism that erodes individual differences over
  time. Tested in `vignette("paper-mcelreath-2007")` (added 0.7.x).
  The bold-aggro syndrome peaks at +0.307 around 5 000 ticks then
  decays to +0.032 by 15 000 ticks. The critique is strongly borne
  out: Wolf's headline correlation is a transient phenomenon at the
  simulation length he reports.

## References

- Wolf, M., van Doorn, G.S., Leimar, O. & Weissing, F.J. (2007)
  Life-history trade-offs favour the evolution of animal personalities.
  *Nature* 447:581–584. doi:10.1038/nature05835.
- Massol, F. & Crochet, P. (2008) Do animal personalities emerge?
  *Nature* 451 (Brief Communications Arising). doi:10.1038/nature06744.
- McElreath, R., Luttbeg, B., Fogarty, S.P., Brodin, T. & Sih, A. (2007)
  Evolution of animal personalities. *Nature* 450 (Brief Communications
  Arising). doi:10.1038/nature06326.
- Clark, C.W. (1994) Antipredator behavior and the asset-protection
  principle. *Behavioral Ecology* 5:159–170.
