---
title: "McElreath et al. 2007 — does Wolf's syndrome erode over time?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{McElreath et al. 2007 — does Wolf's syndrome erode over time?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*A direct test in clade of McElreath, Luttbeg, Fogarty, Brodin & Sih's
(2007) critique of Wolf et al. 2007: asset protection is a
**negative-feedback** mechanism that erodes individual differences
over time, so the bold-aggro syndrome should be transient rather than
stable. Implemented as a time-horizon sweep over the spatially-explicit
Wolf 2007 scenario added in 0.7.0. Companion to
`vignette("paper-wolf2007")` and `vignette("paper-massol-crochet-2008")`.*

## The critique

Wolf et al. (2007) report a stable bold-aggro syndrome at the
end of their simulations. McElreath, Luttbeg, Fogarty, Brodin & Sih
(2007, *Nature* 450, brief communications arising;
`doi:10.1038/nature06326`) argued that this stability is misleading:

> The mechanism Wolf et al. identify is a **negative-feedback loop**.
> Agents with high assets become risk-averse; risk-aversion preserves
> assets; assets become more uniform across individuals; and the
> trait correlations that defined the syndrome erode.

Their prediction: run the simulation longer than Wolf did and the
syndrome will fade. The headline syndrome strength is a transient
artifact of having stopped at an early time-point.

clade exposes the time horizon directly via `max_ticks`. Sweeping it
gives a clean test.

## The sweep

```{r setup-libs, eval = FALSE}
library(clade)

run_one <- function(max_ticks_val, seed = 42L) {
  s <- wolf_personality_specs()
  s$max_ticks                       <- as.integer(max_ticks_val)
  s$random_seed                     <- seed
  s$n_agents_init                   <- 200L
  s$max_agents                      <- 800L
  s$personality_hawkdove_per_tick   <- 0.3
  s$personality_hawkdove_radius     <- 2L
  s$n_predators_init                <- 10L
  s$predator_max_agents             <- 30L
  env <- run_alife(s, verbose = FALSE)

  recs  <- env$agents
  alive <- vapply(seq_along(recs), function(i) as.logical(recs[[i]]$alive),
                  logical(1L))
  if (sum(alive) < 30L) return(NULL)

  ex <- vapply(seq_along(recs)[alive],
               function(i) as.numeric(recs[[i]]$exploration), numeric(1L))
  bo <- vapply(seq_along(recs)[alive],
               function(i) as.numeric(recs[[i]]$boldness), numeric(1L))
  ag <- vapply(seq_along(recs)[alive],
               function(i) as.numeric(recs[[i]]$aggressiveness), numeric(1L))
  list(
    max_ticks = max_ticks_val,
    n         = length(ex),
    r_eb      = suppressWarnings(cor(ex, bo)),
    r_ea      = suppressWarnings(cor(ex, ag)),
    r_ba      = suppressWarnings(cor(bo, ag)),
    mean_x    = mean(ex),
    sd_x      = sd(ex)
  )
}

horizons <- c(2000L, 5000L, 15000L)
results  <- lapply(horizons, run_one)
```

## Observed pattern (seed 42, single replicate)

| Ticks  | n   | mean(x) | sd(x) | cor(exp, bold) | cor(exp, aggro) | cor(bold, aggro) |
|-------:|----:|--------:|------:|---------------:|----------------:|-----------------:|
|  2 000 | 585 | 0.143   | 0.127 | −0.021         | −0.099          |  −0.004          |
|  5 000 | 670 | 0.094   | 0.096 | +0.044         | −0.005          | **+0.307**       |
| 15 000 | 665 | 0.094   | 0.100 | +0.044         | −0.058          |  +0.032          |

The defining bold-aggro syndrome (single seed = 42 only):

- **2 000 ticks** — not yet established (r ≈ 0). Population still
  approaching equilibrium.
- **5 000 ticks** — peak at r = +0.307. This is the Wolf 2007
  reproduction value reported in `vignette("paper-wolf2007")` and
  the Phase 6 commit.
- **15 000 ticks** — decayed to r = +0.032, an order of magnitude
  weaker. Essentially gone.

The asset-protection signs (cor(exp, bold), cor(exp, aggro)) stay
near zero across all horizons.

The single-seed picture above looked like a clean confirmation of
McElreath's transience prediction. **But that read does not survive
multi-seed scrutiny** — see the next section.

## Multi-seed verification (2026-05-16)

A multi-seed regression baseline was generated by
`dev/audit/fidelity/paper_mcelreath_2007.R` (3 horizons × 8 seeds =
24 runs, ~24 min wall-clock). Saved to
`dev/audit/fidelity/paper_mcelreath_2007.rds`. Headline finding:
**the peak-then-decay trajectory disappears at 8 seeds.**

### cor(bold, aggro) by horizon (8 seeds per cell)

| Horizon (ticks) | mean ± SE |
|---:|---:|
|  2 000 | **−0.075 ± 0.045** |
|  5 000 | **−0.077 ± 0.049** |
| 15 000 | +0.034 ± 0.049 |

- The single-seed +0.307 at 5 000 ticks is a noise-envelope landing
  on the positive tail; the multi-seed mean is **−0.077** —
  opposite sign.
- The 15 000-tick "decay to +0.032" is also single-seed noise; mean
  is +0.034.
- The multi-seed pattern is **null at every horizon** (none reach
  2σ PASS). There's no peak to decay from.

### Asset-protection correlations across horizons

| Horizon | cor(exp, bold) | cor(exp, aggro) |
|---:|---:|---:|
|  2 000 | +0.035 ± 0.029 | **−0.172 ± 0.042**, t = −4.12 **PASS-negative** |
|  5 000 | −0.069 ± 0.035 | −0.020 ± 0.044 |
| 15 000 | +0.015 ± 0.027 | −0.063 ± 0.042 |

**Surprise positive finding**: at the *shortest* horizon (2 000
ticks), `cor(exp, aggro)` is **−0.172 ± 0.042, t = −4.12 → PASS-negative**.
This is the only multi-seed cell across all three Wolf 2007
reproductions in clade (Wolf 2007 + Massol-Crochet + McElreath) that
produces a 2σ PASS in the predicted direction. Notably, this is
*opposite to McElreath's prediction*: the asset-protection signal is
strongest at the *earliest* time-point, not later. If
asset-protection is a real mechanism in clade's spatial kernel, it
appears early and is washed out by 5 000 ticks — but the signal is
in `cor(exp, aggro)` specifically, not in the bold-aggro syndrome.

### Trait-mean stability check

`mean(exploration)` across horizons: 0.119 → 0.112 → 0.113. The
population *is* in a stable equilibrium structure at all three
horizons. So the disappearing syndrome cannot be blamed on
equilibrium drift — the equilibrium is there; the syndrome simply
doesn't emerge across seeds.

## Reading the result — honest revision

The original single-seed reading was: "McElreath's critique is
**strongly borne out** — the syndrome peaks at 5 000 ticks then
decays." At 8 seeds, **neither McElreath's transience prediction nor
Wolf's syndrome prediction is robust**: the bold-aggro syndrome is
null at every horizon tested. Both authors' opposing claims about
that syndrome are *equally untestable* at this sample size — there's
no signal to be transient or persistent.

The one robust multi-seed finding is **opposite to McElreath**:
`cor(exp, aggro)` shows asset-protection at the *shortest* horizon
(2 000 ticks, PASS-negative), not the longest. The mechanism may be
real but presents in the asset-protection correlation rather than
the cross-context syndrome, and on the early-decay side of the
time-axis rather than the late-decay side.

Recommended interpretation: clade's spatial Wolf 2007 reproduction
is null on the syndrome metric at 8 seeds, so the McElreath transience
critique cannot be tested rigorously without first strengthening the
underlying signal (longer runs, larger population, kernel changes —
or perhaps the cor(exp, aggro) is the actual robust asset-protection
signal and the syndrome metric was a red herring). Documented as a
candidate finding for the v0.8-core review.

## Implications for the Wolf 2007 reproduction

The base reproduction in `vignette("paper-wolf2007")` ran to 5 000
ticks because that's roughly where Wolf 2007 stops. At that horizon
clade matches Wolf's qualitative pattern (positive bold-aggro
syndrome). But at 15 000 ticks — three times longer — the syndrome
is gone. The asset-protection mechanism is real, but its visible
effect on trait correlations is short-lived.

Users running the Wolf scenario in clade should be aware of this:
the syndrome's strength is *time-window-dependent*. Quoting a
"final" correlation without specifying the simulation length is
misleading.

## Two critiques together: a coherent multi-seed picture

Reading `vignette("paper-massol-crochet-2008")` and this vignette
together at the multi-seed level (8 seeds each, both rds saved):

- **Massol & Crochet 2008**: the single-seed claim "peak at β = 1.25"
  *flips at 8 seeds* — β = 1.25 is the trough (mean = −0.077,
  marginal-negative), and no β value produces a robust positive
  syndrome.
- **McElreath et al. 2007** (this vignette): the single-seed claim
  "peak at 5 000 ticks decaying to near-zero by 15 000" flips at
  8 seeds — bold-aggro is null at every horizon; there's no peak
  to decay from.

Both critiques' apparent confirmations from single seeds were noise-
envelope landings, not real signals. The cumulative multi-seed
picture across all three Wolf 2007 reproductions in clade
(`paper-wolf2007`, `paper-massol-crochet-2008`, `paper-mcelreath-2007`)
is that **the boldness-aggressiveness syndrome does not robustly
emerge** in clade's spatially-explicit kernel at any (β, horizon)
combination tested. Documented as a candidate finding for the
v0.8-core review.

## Caveats

- **Single seed** (42). The qualitative pattern (peak then decay) is
  visible at this seed; a multi-seed sweep would firm up the timing
  of the peak and the half-life of the decay. Documented as a
  limitation.
- **Phase 6 (1+α·N_i) denominator on**. Setting `personality_alpha = 0`
  reverts to Phase 3a behaviour and might shift the timing.
- **Spatial implementation**. clade's grid + one-per-cell offspring +
  Moore-neighbourhood hawk-dove pairing produces a different
  equilibrium structure than Wolf's mean-field model. McElreath's
  argument is mathematical and applies to any negative-feedback
  setting; the time-scale of decay in clade may differ from what
  pure Wolf would show.
- **Computational cost**. A 15 000-tick run takes ~130 s on a typical
  M1 laptop. Doing the same sweep at 5–10 seeds per horizon takes
  10–20 minutes. Worth doing for any quantitative claim that
  generalises beyond seed 42.

## What's left for follow-up work

- **Multi-seed verification** of the timing of peak and decay.
- **Per-tick correlation tracking** — using the `log_genomes` feature
  (wired in 0.7.x) one could compute the bold-aggro correlation at
  every logged tick rather than just at three endpoints, producing a
  full time-course rather than three snapshots.
- **Negative-feedback rate as a function of population size**.
  McElreath's theory predicts faster decay in larger populations
  (more statistical averaging). The current sweep is at a single
  population size.

## References

- McElreath, R., Luttbeg, B., Fogarty, S.P., Brodin, T. & Sih, A.
  (2007) Evolution of animal personalities. *Nature* 450 (Brief
  Communications Arising). doi:10.1038/nature06326.
- Wolf, M., van Doorn, G.S., Leimar, O. & Weissing, F.J. (2007)
  Life-history trade-offs favour the evolution of animal personalities.
  *Nature* 447:581–584. doi:10.1038/nature05835.
- See also `vignette("paper-wolf2007")` for the base reproduction and
  `vignette("paper-massol-crochet-2008")` for the parallel β-sweep
  test of the same Wolf 2007 mechanism.
