---
title: "Reproducing YOUR paper — a template"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing YOUR paper: a template for the 3-stage workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Skeleton for reproducing a behavioural-ecology paper in clade.
Copy this file, replace the TODO blocks with your paper's
specifics, run, interpret. The
[five worked examples](paper-kokko-brooks-2003.html) use exactly
this pattern.*

---

## Before you start

### Pick a paper that's in-scope

clade is a behavioural / cognitive / social-evolution testbed
with a neural-brained, energy-fitness kernel. Papers **in scope**:

- **Selection on a heritable trait clade already models** (brain
  size, body size, signals, helper tendency, metabolic rate,
  plasticity, etc. — see [scenarios](scenarios.html) for the full
  list).
- **Population-level predictions expressible as sweeps**
  ("condition X → more of Y").
- **Evolutionary-ABM predictions** — directional evolution,
  equilibrium densities, cluster formation.

Papers **out of scope** (use a different tool):

- Genome-scale pop-gen (linkage, sweeps, demographic inference) —
  [SLiM](https://messerlab.org/slim/)
- Coalescent / tree-sequence inference —
  [msprime](https://tskit.dev/msprime/)
- Specific-species management models — use empirical-data ABMs
  (Murphy 2025 framework)

See the [landing page](../index.html) fit-table for the full
"when clade fits / when it doesn't" summary.

### Identify the paper's *quantitative* prediction

The sharpest reproductions come from predictions that translate
to a directional claim about a clade-measurable outcome. Examples
from the 5 worked reproductions:

| Paper | The directional claim |
|---|---|
| Kokko & Brooks 2003 | signals_effect becomes more negative as grass drops |
| Griesser 2023 | `care_duration ↑ → brain_size ↑` (Spearman > 0) |
| Dieckmann & Doebeli 1999 | `isolation_threshold ↓ → n_species ↑` (Spearman < 0) |
| Réale 2010 | `metabolic_rate ↑ → mean_age ↓` (Spearman < 0) |
| Emlen 1982 | `resource scarcity ↑ → helping rate ↑` |

If you can't phrase your paper's claim as `X → Y at rate ρ`, the
paper may be qualitative-only and not a clean clade fit.

## The 3-stage workflow

All five showcase vignettes follow the same structure:

1. **Grid-search** a 2-D parameter slice to find which regime
   makes the signal visible. Tools: `grid_specs()` + `batch_alife()`.
2. **Multi-seed validation** at the chosen regime, computing
   contrast statistics. Tools: `hypothesis_sweep()` +
   `hypothesis_report()`.
3. **Viability diagnostic** confirming populations aren't sitting
   at a fitness cliff that would invalidate the Δ statistics.
   Tool: `viability_report()`.

Don't skip Stage 1. Naive parameter choices often give null
results (see Griesser 2023); grid search identifies the regime
where the predicted mechanism has room to express.

## Template

Copy and edit the blocks marked TODO.

### Setup

```{r setup-specs}
library(clade)

# TODO: pick a base regime appropriate to your paper's scale
base <- default_specs()          # or fast_specs() / realistic_specs()
base$grid_rows       <- 40L
base$grid_cols       <- 40L
base$n_agents_init   <- 120L
base$max_agents      <- 500L
base$max_ticks       <- 2000L
base$grass_rate      <- 0.15
base$n_predators_init <- 0L

# TODO: enable the modules your paper's mechanism requires.
# Examples from the 5 showcase vignettes:
#
#   Brain-size evolution (Griesser, s-brain-size):
#     base$brain_size_evolution         <- TRUE
#     base$brain_size_cost_scale        <- 2.0
#     base$neonatal_foraging_deficit    <- 0.4
#     base$parental_care                <- TRUE
#
#   Sexual selection (K&B):
#     base$signal_dims           <- 3L
#     base$signal_cost           <- 0.2
#     base$mate_choice_mode      <- "preference"
#
#   Speciation (D&D):
#     base$speciation                  <- TRUE
#
#   Life-history syndrome (Réale):
#     base$metabolic_rate_evolution       <- TRUE
#     base$max_age_scales_with_metabolism <- TRUE
#
#   Cooperative breeding (Emlen):
#     base$cooperative_breeding          <- TRUE
#     base$parental_care                 <- TRUE
#
# See ?default_specs for the full parameter catalogue.
```

### Stage 1 — grid search for the right regime

```{r stage1-grid}
# TODO: pick two parameters to vary — usually
#   (a) one that controls the MECHANISM strength (cost, rate,
#       threshold, etc.)
#   (b) one that controls the ENVIRONMENT your paper varies
#       (resource, stress, duration, etc.)

specs_list <- grid_specs(
  base,
  param_a = c(0.5, 1.0, 1.5, 2.0),   # TODO: your parameter A
  param_b = c(0.05, 0.10, 0.15, 0.20), # TODO: your parameter B
  seed_from = 1L
)

results <- batch_alife(specs_list, n_cores = 16L)

grid_tbl <- do.call(rbind, mapply(function(env, s) {
  d <- get_run_data(env)$ticks
  data.frame(
    param_a = s$param_a,
    param_b = s$param_b,
    # TODO: the outcome your paper predicts
    outcome = mean(tail(d$YOUR_OUTCOME_COLUMN, 500), na.rm = TRUE),
    final_n = mean(tail(d$n_agents, 500), na.rm = TRUE)
  )
}, results, specs_list, SIMPLIFY = FALSE))

# Inspect: where is the signal? Which cells are viable?
print(grid_tbl)
```

**Common pitfall**: if your grid returns almost all `final_n` near
zero, the regime is extinction-prone — back off on whatever
makes selection harsh (cost scales, mutation rates, scarcity).
See the Griesser vignette for a concrete debugging walkthrough.

### Stage 2 — multi-seed validation

```{r stage2-sweep}
# Pick the cell from Stage 1 where the signal is strongest AND
# populations are viable.
final_base <- base
final_base$param_a <- 1.5   # TODO: from Stage 1 winner

sweep <- hypothesis_sweep(
  base_specs = final_base,
  conditions = list(
    # TODO: one condition per level of the treatment you're sweeping
    low    = list(param_b = 0.05),
    medium = list(param_b = 0.10),
    high   = list(param_b = 0.15),
    very_high = list(param_b = 0.20)
  ),
  seeds = 1:8,
  metrics = list(
    # TODO: the outcome measure. Match units to your paper!
    #   - Per-capita rate if the paper talks per-individual
    #   - Aggregate count if the paper talks population-level
    #   - Peak value, final value, or 500-tick mean — choose
    #     based on what the paper reports
    outcome = function(ticks) mean(tail(ticks$YOUR_OUTCOME_COLUMN, 500),
                                   na.rm = TRUE),
    final_n = function(ticks) mean(tail(ticks$n_agents, 500), na.rm = TRUE),
    crashed = function(ticks) tail(ticks$n_agents, 1) < 10
  ),
  n_cores = 32L
)
print(sweep)

# Pairwise contrasts — the direction the paper predicts
hypothesis_report(
  sweep,
  contrasts = list(
    # TODO: name the comparisons your paper makes.
    # Convention: c(reference, test) — delta is test − reference.
    medium_vs_low    = c("low", "medium"),
    high_vs_low      = c("low", "high"),
    very_high_vs_low = c("low", "very_high")
  ),
  metric = "outcome"
)
```

### Stage 3 — viability diagnostic

```{r stage3-viability}
# Pick the most-stressed cell (usually the extreme of your
# treatment sweep). Confirm populations aren't crashing.
s_check <- final_base
s_check$param_b     <- 0.20   # TODO: most-extreme treatment
s_check$random_seed <- 1L
env_check <- run_alife(s_check, verbose = FALSE)
viability_report(get_run_data(env_check))
```

If `viability_report()` reports `"viable"`, your Stage 2
statistics are interpretable. If it reports `"crashed"` or
`"near-extinction"`, the conclusion about that condition is
"mechanism triggers collapse at these parameters", not "mechanism
produces Δ = X".

## Interpretation checklist

When writing up your result, check each of these **before**
claiming the paper reproduces (or doesn't):

- [ ] **Direction** — does Spearman / contrast sign match the
      paper's prediction?
- [ ] **Magnitude** — does any contrast pass 2σ (`|t| ≥ 2`)? Note
      the difference between "direction-correct sub-threshold"
      (marginal) and "direction-correct decisive" (PASS).
- [ ] **Units** — is your metric per-individual or aggregate?
      Raw population-level counts can invert per-capita rates
      (the Réale and Emlen vignettes document this trap
      explicitly).
- [ ] **Viability** — did populations survive? A crashed
      condition is an observation, not a failure — but it's a
      different observation from the paper's one.
- [ ] **Mechanism match** — does clade's kernel encode the same
      causal structure as the paper's theoretical framework?
      When they differ (e.g., linear per-tick cost in clade vs
      stress-multiplicative cost in K&B's theory), the
      interaction direction can reverse. Document this.
- [ ] **Kernel limits** — if the reproduction is a null, does
      the kernel have all the mechanisms the paper's theory
      needs? The s-kin invasion honest-null (no kin-weighted
      fitness in the cooperative_breeding module) is one example.

## Common outcomes and what they mean

Based on the 5 showcase reproductions:

| Outcome | What it looks like | Interpretation |
|---|---|---|
| **Clean ✅** | Contrast t ≥ 2, Spearman strong | Kernel mechanism matches theory; paper's prediction robust. D&D 1999 is this pattern. |
| **Conditional ✅** | Direction-correct only in specific parameter regime | Prediction works in the regime the theory is about, not at arbitrary defaults. s-mimicry, s-niche, s-parental-care are this pattern. |
| **Direction-correct sub-threshold** | Spearman correct sign, t < 2 at tested seed count | Mechanism present but signal below noise floor. Griesser 2023 is this pattern; could strengthen with more seeds or larger populations. |
| **Inverts on raw, recovers on per-capita** | Aggregate counts go wrong direction | Measurement-unit mismatch. Normalise by `n_agents` and re-test. Réale 2010 (per-tick births) and Emlen 1982 (raw helping counts) are this pattern. |
| **Contradiction** | Direction opposite to paper | Mechanism-level mismatch between kernel and theory. Document honestly; look for what cost structure, feedback loop, or selection mode differs. K&B 2003 is this pattern. |
| **Honest null** | No signal in any tested regime | Kernel may be missing a mechanism the paper requires. s-kin invasion dynamics is this pattern (needs kin-weighted reproduction). |

All six outcomes are scientifically legitimate. The audit
methodology treats null and contradiction as equally informative
as a clean ✅ — they reveal what kernel or theory needs to
change.

## Research-paper integration

When you publish a reproduction, cite:

```bibtex
@article{YOUR_PAPER_KEY,
  author  = {YOUR AUTHOR},
  title   = {YOUR TITLE},
  journal = {YOUR JOURNAL},
  year    = {YEAR},
  doi     = {YOUR DOI}
}

@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package, version 0.6.0+},
  url     = {https://github.com/itchyshin/clade}
}
```

A typical results-paragraph in a published reproduction might
read:

> *"To verify the Author et al. (year) prediction in a
> mechanistic evolutionary ABM, we reproduced the main
> experimental contrast in clade (Nakagawa 2026). Across a
> hypothesis_sweep with 8 seeds per condition, the direction of
> the predicted effect reproduced with Δ = X ± SE (t = Y). A
> grid search over (param_a × param_b) identified the
> [concrete regime] as the parameter cell where the effect is
> strongest. Full reproduction protocol is in [supplementary
> script]."*

## Going further

Once you have a working reproduction:

- Add seeds (16, 32, 64) to shrink the noise floor.
- Add `realistic_specs()` scale (60×60, 2000 ticks, predator
  guild) to stress-test the result.
- Run the **inverse prediction**: show clade *doesn't* reproduce
  the claim at parameter regimes outside the paper's
  applicability, demonstrating where the theory actually
  depends.
- Publish your reproduction as a standalone vignette; consider
  submitting it as a clade-documented contribution.

## Related vignettes

- [Kokko & Brooks 2003](paper-kokko-brooks-2003.html) — contradictions
- [Griesser et al. 2023](paper-griesser-2023.html) — direction-correct sub-threshold
- [Dieckmann & Doebeli 1999](paper-dieckmann-doebeli-1999.html) — clean ✅
- [Réale et al. 2010](paper-reale-2010.html) — unit-sensitivity
- [Emlen 1982](paper-emlen-1982.html) — aggregate vs per-capita

Each demonstrates the same 3-stage workflow on a different
mechanism. Cross-reference them to see how the template maps
onto real papers.
