---
title: "Reproducing a paper — Kokko & Brooks 2003"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing a paper: Kokko & Brooks 2003 (Sexy to die for?)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Worked example: take a published behavioural-ecology paper, turn
its core prediction into a clade experiment, report what
reproduces and what doesn't. Uses the `hypothesis_sweep()` +
`hypothesis_report()` helpers to keep the whole workflow inside
clade's vocabulary — no separate scripts, no bespoke plumbing.*

---

## The paper

**Kokko, H. & Brooks, R. (2003).** *Sexy to die for? Sexual
selection and the risk of extinction.* *Annales Zoologici
Fennici* 40, 207–219.

Core claim: **evolutionary "suicide" from costly sexual traits is
*unlikely* in stable environments, but becomes possible under
environmental stress**. The interaction — how the fitness effect
of signals scales with environmental pressure — is the paper's
distinctive empirical prediction.

The prediction is qualitative in the original: populations carrying
costly sexual signals should face a steeper extinction-risk gradient
as environments deteriorate. For an in-silico test we need a
quantitative surrogate. The natural one is:

> At each resource level, measure `Δn = n(signals on) − n(signals off)`.
> Plot Δn as a function of resource level. K&B predict the slope is
> *negative*: as resources deteriorate, signals become an increasingly
> costly liability.

This is a clean 2 × k factorial (signals × resource level), and a
clean *interaction test* on the slope of `signals_effect`.

## Experimental design

Four resource levels crossed with signals on/off, 8 seeds each:
64 runs at 40 × 40 grid, 2000 ticks, no predators (to isolate
resource-driven stress from external mortality).

**Signals-on arm**: `signal_dims = 3L`, `signal_cost = 0.2` (above
the drift-scale noise documented in s-signals.Rmd),
`mate_choice_mode = "preference"`, `mate_choice_strength = 0.7`.
**Signals-off arm**: `signal_dims = 0L`, `mate_choice_mode =
"random"`.

**Resource levels**: `grass_rate ∈ {0.20, 0.12, 0.08, 0.05}`
spanning abundant → very_scarce.

```{r kb-sweep}
library(clade)

base <- default_specs()
base$grid_rows       <- 40L
base$grid_cols       <- 40L
base$n_agents_init   <- 120L
base$max_agents      <- 500L
base$max_ticks       <- 2000L
base$n_predators_init <- 0L   # isolate resource stress

SIGNAL_COST  <- 0.2
SIGNAL_DIMS  <- 3L
GRASS_LEVELS <- c(abundant = 0.20, mid = 0.12,
                  scarce   = 0.08, very_scarce = 0.05)

make_cond <- function(grass, signals_on) {
  if (signals_on) list(
      signal_dims            = SIGNAL_DIMS,
      signal_cost            = SIGNAL_COST,
      signal_evolution_drift = TRUE,
      signal_drift_sd        = 0.01,
      mate_choice_mode       = "preference",
      mate_choice_strength   = 0.7,
      grass_rate             = grass
    )
  else list(
      signal_dims      = 0L,
      signal_cost      = 0.0,
      mate_choice_mode = "random",
      grass_rate       = grass
    )
}

conds <- list()
for (ln in names(GRASS_LEVELS)) {
  g <- GRASS_LEVELS[[ln]]
  conds[[paste0(ln, "_no_signals")]]   <- make_cond(g, FALSE)
  conds[[paste0(ln, "_with_signals")]] <- make_cond(g, TRUE)
}

sweep <- hypothesis_sweep(
  base_specs = base,
  conditions = conds,
  seeds = 1:8,
  metrics = list(
    final_n     = function(ticks) mean(tail(ticks$n_agents, 500), na.rm = TRUE),
    crashed     = function(ticks) tail(ticks$n_agents, 1) < 10,
    mean_energy = function(ticks) mean(tail(ticks$mean_energy, 500), na.rm = TRUE)
  ),
  n_cores = 32L
)
print(sweep)

# Contrasts: one signals-effect test per resource level
signals_contrasts <- setNames(
  lapply(names(GRASS_LEVELS), function(ln) {
    c(paste0(ln, "_no_signals"), paste0(ln, "_with_signals"))
  }),
  paste0("signals_effect_", names(GRASS_LEVELS))
)
rpt <- hypothesis_report(sweep, signals_contrasts, metric = "final_n")
print(rpt)
```

## Results

### Per-condition equilibrium populations (8 seeds each)

| grass_rate | signals OFF | signals ON |
|---|---|---|
| abundant (0.20) | 237.7 ± 5.3 | 210.8 ± 3.7 |
| mid (0.12) | 157.8 ± 3.2 | 150.4 ± 3.9 |
| scarce (0.08) | 120.6 ± 2.3 | 109.9 ± 1.6 |
| very_scarce (0.05) | 70.2 ± 4.9 | 67.1 ± 2.6 |

Main effect of resource: massive. Going from abundant to
very_scarce cuts equilibrium population by ~70% regardless of
signals.

### Signals effect per resource level

| resource level | Δ (signals − no signals) ± SE | t | verdict |
|---|---|---|---|
| abundant | **−26.90 ± 6.46** | **−4.17** | **PASS** |
| mid | −7.36 ± 5.06 | −1.46 | null |
| scarce | **−10.77 ± 2.82** | **−3.82** | **PASS** |
| very_scarce | −3.11 ± 5.49 | −0.57 | null |

### The K&B interaction

K&B predict: `signals_effect @ very_scarce` should be *more
negative* than `signals_effect @ abundant` — the sexual-selection
drag gets worse under stress.

clade shows the opposite:

```
signals_effect @ abundant    = -26.90 ± 6.46
signals_effect @ very_scarce =  -3.11 ± 5.49
Δ(very_scarce - abundant)    = +23.79 ± 8.48, t = +2.81  [PASS]
```

**The interaction is positive and statistically significant.** In
clade, costly signals are a *larger* absolute demographic drag on
abundant populations than on stressed ones — the opposite of the
K&B prediction.

### Extinction rates

Zero extinctions across all 64 runs. clade's 2000-tick ceiling +
500-agent cap + resource-limited dynamics produce stable
equilibria at every tested condition; K&B's primary outcome
(extinction risk) is not expressible in this regime.

## What this tells us — honest interpretation

Three observations stand apart:

1. **The direction of the main effect matches K&B qualitatively:**
   at the signals-on arm, the population is smaller than the
   control. Zahavi-style handicap costs measurable in clade.

2. **The interaction direction does NOT match.** The sexual-
   selection cost is *diluted* by resource scarcity in clade: when
   the population is already energy-limited (low equilibrium),
   each agent's per-tick `signal_cost = 0.2` removes less total
   energy from the system than at the high equilibrium. The
   signals-effect curve is concave, not steepening.

3. **No extinctions.** K&B's primary outcome is not observed
   because clade's dynamics produce stable low-density equilibria
   rather than demographic collapse even at the most extreme
   tested resource level.

### Why the mismatch?

clade's signal cost is a **per-tick per-agent linear energy
drain** on the agent's own budget. K&B's theoretical framework
(built on allele-dynamic models) assumes costs that interact
multiplicatively with stress — for example, signal-maintenance
costs that reduce an individual's fasting tolerance under
scarcity. These are different mechanistic claims dressed in the
same surface-level "costly trait" vocabulary.

This is **scientifically informative, not a failure**. The audit
methodology surfaces exactly this kind of mechanism-level
mismatch. A behavioural-ecology researcher reading K&B's paper and
asking "would my empirical system show this?" can use clade to
find out whether their modelled mechanism carries the claim
through or not.

### Paths to a clade-side K&B reproduction

Three extensions would likely reproduce the K&B pattern:

- **Stress-multiplicative costs**: replace `signal_cost` with a
  mechanism where signal-bearing agents face higher starvation
  probability at low energy (not just a flat energy tax). Requires
  a kernel change in `tick.jl` — signal load → fasting-tolerance
  penalty.

- **Longer runs + harder stress**: 8000 ticks × `grass_rate =
  0.03` might produce extinctions; K&B's outcome is extinction
  probability, not equilibrium Δ. Pure parameter-space change, no
  kernel work. Deferred; not attempted here.

- **Sexual dimorphism**: clade does not currently implement
  sex-specific mortality. K&B predict that sexual dimorphism can
  *raise* carrying capacity when high male mortality relaxes
  density-dependence on females. Implementing sex-specific
  survival would let clade test both the extinction-risk and
  dimorphism-paradox halves of the paper's argument.

## Takeaway for empirical researchers

The workflow this vignette demonstrates is the common shape of
an in-silico verification study:

1. **Read the paper** carefully enough to extract a *quantitative*
   prediction (not a paraphrase). K&B's is "signals_effect slope
   is negative with resource deterioration".
2. **Translate to clade terms**: which parameters encode the
   mechanism, which encode the environment, which outcome
   measures the prediction.
3. **Use `hypothesis_sweep()`** to cross the relevant parameters
   across enough seeds to distinguish signal from drift.
4. **Use `hypothesis_report()`** to compute the specific contrasts
   your prediction requires, including interactions.
5. **Report the result honestly** — including when the kernel's
   mechanistic realisation of the cited theory differs from the
   paper's own framework, producing a different prediction
   direction.

Zero code was written outside of spec lists + the two helpers.
Total compute: ~30 s on 32 PSOCK cores for 64 runs × 2000 ticks.

## Citation

If you use this scenario or workflow in published work, cite both
the paper and clade:

```bibtex
@article{kokko2003sexy,
  author  = {Kokko, Hanna and Brooks, Robert},
  title   = {Sexy to die for? Sexual selection and the risk of extinction},
  journal = {Annales Zoologici Fennici},
  year    = {2003},
  volume  = {40},
  pages   = {207--219}
}

@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```

Audit protocol and raw outputs:
[dev/audit/fidelity/kokko_brooks_2003.R](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/kokko_brooks_2003.R)
and `kokko_brooks_2003.rds`.
