---
title: "Parameter Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, collapse = TRUE, comment = "#>",
                       echo = FALSE)
library(clade)
.print_tbl <- function(grp) {
  df <- clade:::.param_table(grp)
  if (nrow(df) == 0L) {
    cat("_(no fields registered in this group)_\n")
    return(invisible(NULL))
  }
  knitr::kable(df, format = "markdown",
               col.names = c("Parameter", "Default", "Type"),
               row.names = FALSE)
}
```

This article lists every parameter in [`default_specs()`](../reference/default_specs.html),
grouped by biological theme. For each parameter you'll see the default value
and its type.

**The tables on this page are generated at vignette-build time** from
`default_specs()` itself, so they cannot drift from what the code actually
ships. The full description of each parameter — including units, valid
ranges, references, and module-interaction notes — lives in the
[`default_specs()` reference page](../reference/default_specs.html). This
page is a quick lookup; that page is the authoritative documentation.

For an interactive view of a customised spec list, use `print_specs()`:

```{r print-specs-example, eval = FALSE, echo = TRUE}
s <- default_specs()
s$kin_selection      <- TRUE
s$complex_landscape  <- TRUE
print_specs(s, diff_only = TRUE)
```

---

## Grid and population

The toroidal grid, founder population, simulation horizon, and the two
0.7.0 kernel-discipline switches.

```{r grid-population}
.print_tbl("Grid & population")
```

`random_tick_order = TRUE` (default) enables random asynchronous agent
scheduling per Grimm & Railsback (2005); setting `FALSE` recovers the
legacy fixed-order behaviour. `max_agents_per_cell = 1L` (default)
enforces one-per-cell movement; `0L` allows unbounded co-occupancy
(legacy). See `NEWS.md` for the 0.7.0 audit that surfaced these.

---

## Energy and metabolism

Per-agent energy budget: starting energy, costs of moving / waiting /
reproducing, and the threshold below which agents starve.

```{r energy}
.print_tbl("Energy & metabolism")
```

`repro_cost_mode` and `offspring_energy_mode` control whether the cost
and birth-energy are fixed amounts or fractions of parental state — see
`default_specs()` for the full Smith-Fretwell quality-quantity options.

---

## Grass dynamics

The base food supply: how cells initialise, how often they regrow, and
the per-cell saturation.

```{r grass}
.print_tbl("Grass dynamics")
```

---

## Brain architecture

The four implemented brain types plus a random null-model baseline.
`"transformer"` and `"synthesis"` are reserved names — the kernel
errors if either is requested at present.

```{r brain}
.print_tbl("Brain architecture")
```

`bnn_sigma_source` is required for s-baldwin / s-plasticity scenarios
(0.4.0 Tier 5A). `bnn_sample_freq = 5` lets REINFORCE gradients
accumulate (required for the 0.4.1 s-rl ✅ verdict).

---

## Brain metabolic cost

How brains cost energy: a fixed-per-weight floor, an activity-scaled
component, BNN-uncertainty information cost, and the Kleiber-style
size exponent.

```{r brain-cost}
.print_tbl("Brain metabolic cost")
```

---

## Reproduction and genetics

Ploidy, mate-choice, recombination, and reproduction-age controls.

```{r reproduction}
.print_tbl("Reproduction & sex")
```

`mate_choice_mode = "preference"` enables signal-preference assortative
mating; `"random"` reverts to baseline. `mate_search_radius` is the
Moore-neighbourhood radius for partner search.

---

## Life history

Maximum lifespan, the Gompertz senescence parameters (rate × shape),
and the optional pace-of-life lifespan-scaling coupling.

```{r life-history}
.print_tbl("Life history")
```

The Gompertz hazard is
\(p = 1 - \exp(-r \cdot \exp(r \cdot \mathrm{age}^{\mathrm{shape}}))\)
with \(r = \mathrm{senescence\_rate} \times \mathrm{aging\_rate}\).
`senescence_shape = 1` (default) is the classic Gompertz; > 1 accelerates
late-life mortality; < 1 gives a late-life plateau (Vaupel et al. 1998).

---

## Body size evolution

Enable with `body_size_evolution = TRUE`.

```{r body-size}
.print_tbl("Body size")
```

---

## Brain size evolution

Enable with `brain_size_evolution = TRUE`. Cognitive bonus is balanced
against the metabolic cost set in "Brain metabolic cost" above.

```{r brain-size}
.print_tbl("Brain size evolution")
```

---

## Dispersal evolution

Enable with `dispersal_evolution = TRUE`. Combine with
`spatial_sorting = TRUE` for invasion-front dynamics.

```{r dispersal}
.print_tbl("Dispersal")
```

---

## Habitat preference

Enable with `habitat_preference_evolution = TRUE`.

```{r habitat}
.print_tbl("Habitat preference")
```

---

## Kin selection

Enable with `kin_selection = TRUE`. Pedigree-based relatedness; see the
`s-kin` vignette for Hamilton's-rule examples.

```{r kin}
.print_tbl("Kin selection")
```

---

## Cooperation (PGG)

Enable with `cooperation_evolution = TRUE`. Public-goods games on the
neighbourhood; see `s-cooperation`.

```{r cooperation}
.print_tbl("Cooperation (PGG)")
```

---

## Disease (SIR dynamics)

Enable with `disease = TRUE`.

```{r disease}
.print_tbl("Disease (SIR)")
```

---

## Predators

Enable with `n_predators_init > 0`. Predators are co-evolving agents
with their own 15-input sensory brain.

```{r predators}
.print_tbl("Predators")
```

`predator_sense_graded = TRUE` (0.4.2 default): prey's predator
sensory input at distance `d` is `1/(d+1)` (graded threat-level).
`FALSE` falls back to pre-0.4.2 binary presence.

---

## Group defense

Enable with `group_defense = TRUE`. Coordinated anti-predator
behaviour at high local density.

```{r group-defense}
.print_tbl("Group defense")
```

---

## Niche construction

Enable with `niche_construction = TRUE`. Shelters modify the local
selection environment.

```{r niche}
.print_tbl("Niche construction")
```

`shelter_occupancy_bonus > 0` activates the heritable-niche effect
(Odling-Smee, Laland & Feldman 2003): agents on a shelter cell receive
`bonus × depth` energy per tick.

---

## Within-lifetime reinforcement learning

Enable with `rl_mode = "actor_critic"`. `plasticity_cost` charges a
per-tick metabolic cost proportional to `learning_rate` (Laughlin et
al. 1998 *Nature Neuroscience*).

```{r rl}
.print_tbl("Within-lifetime RL")
```

`lamarckian = TRUE` writes within-lifetime learned weights back to the
genome before meiosis. Off by default.

---

## Social learning

Enable with `social_learning = TRUE`.

```{r social-learning}
.print_tbl("Social learning")
```

---

## Signals and mate choice

Sexual-selection machinery. Enable signals by setting `signal_dims > 0`.

```{r signals}
.print_tbl("Signals & mate choice")
```

`preference_bias_target` + `preference_bias_strength` implement the
β_N leg of Fuller, Houle & Travis (2005); see `paper-ryan-1990`.

---

## Mimicry and toxicity

Enable with `mimicry = TRUE`. Müllerian by default; set
`batesian_mimicry = TRUE` for Batesian dynamics. As of 0.4.4 the
predator memory is a `signal_dims`-length vector updated via the
Widrow-Hoff delta rule.

```{r mimicry}
.print_tbl("Mimicry & toxicity")
```

---

## Coevolving parasites (Hamilton 1980 Red Queen)

Enable with `coevolving_parasites = TRUE` AND either `signal_dims > 0`
(continuous mode) or `n_parasite_loci > 0` (discrete mode).

```{r coevolving-parasites}
.print_tbl("Coevolving parasites")
```

Discrete mode (`n_parasite_loci >= 8`) is the canonical Hamilton
mechanism and is the regime where clade can show sex > asex.

---

## Parental care

Enable with `parental_care = TRUE`.

```{r parental-care}
.print_tbl("Parental care")
```

`neonatal_foraging_deficit > 0` creates the expensive-brain selection
pressure for parental provisioning (Aiello & Wheeler 1995; Isler & van
Schaik 2009).

---

## Cooperative breeding

Enable with `cooperative_breeding = TRUE`. Usually combined with
`iffolk_selection = TRUE`. See `paper-emlen-1982`.

```{r coop-breeding}
.print_tbl("Cooperative breeding")
```

---

## Parental investment (Trivers 1972)

Enable with `parental_investment_evolution = TRUE`. The female bears
`female_investment` of per-offspring cost; offspring birth energy
scales by `2 × female_investment`.

```{r parental-investment}
.print_tbl("Parental investment")
```

---

## Pace of life (Réale et al. 2010)

Heritable metabolic rate and aging rate; see `paper-reale-2010` and
`s-pace-of-life`.

```{r pace}
.print_tbl("Pace of life")
```

---

## Mutation-rate evolution

Enable with `mutation_rate_evolution = TRUE`. Per-agent heritable
`mutation_sd`.

```{r mutrate}
.print_tbl("Mutation-rate evol.")
```

---

## Phenotypic plasticity

Enable with `phenotypic_plasticity = TRUE`. The `plasticity` trait
scales how much an agent's effective reproductive threshold depends
on local resource density.

```{r plasticity}
.print_tbl("Phenotypic plasticity")
```

---

## Stress hypermutation

Enable with `stress_hypermutation = TRUE`. SOS-style mutation-rate
spike below `stress_threshold` energy.

```{r stress}
.print_tbl("Stress hypermutation")
```

---

## Clutch size (r/K)

Enable with `clutch_size_evolution = TRUE`. r/K trade-off between
clutch count and per-offspring investment.

```{r clutch}
.print_tbl("Clutch size (r/K)")
```

---

## Spatial sorting

Enable with `spatial_sorting = TRUE`. Requires `dispersal_evolution = TRUE`
and a bounded grid (`toroidal = FALSE`).

```{r spatial-sorting}
.print_tbl("Spatial sorting")
```

---

## IFfolk inclusive fitness

Enable with `iffolk_selection = TRUE`. Haig 2000 / Fromhage & Jennions
2019 intragenomic-conflict + parliament suppression.

```{r iffolk}
.print_tbl("IFfolk incl. fitness")
```

---

## Complex landscape (forest world)

Enable with `complex_landscape = TRUE`. Three-layer foraging:
ground grass + shrubs + canopy. Wing-size evolves for canopy access.

```{r complex-landscape}
.print_tbl("Complex landscape")
```

---

## Fixed patch

A single fixed-rich-cell helper for testing local-resource scenarios.

```{r fixed-patch}
.print_tbl("Fixed patch")
```

---

## Epigenetics

Enable with `epigenetics = TRUE`. Methylation inheritance on BNN sigma
(Jablonka & Lamb 2005).

```{r epigenetics}
.print_tbl("Epigenetics")
```

---

## Speciation

Enable with `speciation = TRUE`. Genome-distance clustering + reproductive
isolation. See `paper-dieckmann-doebeli-1999`.

```{r speciation}
.print_tbl("Speciation")
```

---

## Seasonal dynamics

```{r seasonal}
.print_tbl("Seasonal dynamics")
```

`seasonal_spatial_bias = TRUE` (0.5.18) flips the spatial grass
distribution between seasons — required for the Baldwin / plasticity
scenarios to demonstrate fluctuating selection.

---

## Scavenging

Enable with `scavenging = TRUE`. Carcasses decay over time and can
transmit disease.

```{r scavenging}
.print_tbl("Scavenging")
```

---

## ANN regularisation

Optional weight regularisation on ANN brains.

```{r ann-reg}
.print_tbl("ANN regularisation")
```

---

## Wolf 2007 personality syndrome (0.7.0)

Enable with `personality_syndrome = TRUE`. See
`paper-wolf2007` for the boldness-aggressiveness syndrome reproduction
and the Massol-Crochet / McElreath critiques.

```{r personality}
.print_tbl("Wolf 2007 personality")
```

---

## Trivers 1971 reciprocal altruism (0.7.0)

Enable with `reciprocal_altruism = TRUE`. Generous-TFT via three
heritable traits (initial / retaliation / forgiveness). See
`paper-trivers1971`.

```{r reciprocity}
.print_tbl("Trivers 1971 reciprocity")
```

---

## Wolf 2008 responsive personalities (0.7.0)

Enable with `responsive_personalities = TRUE`. Frequency-dependent
sampling: high-responsiveness agents override toward richer cells at
a per-tick metabolic cost. See `paper-wolf2008`.

```{r responsive}
.print_tbl("Wolf 2008 responsive")
```

---

## Logging and output

```{r logging}
.print_tbl("Logging & search")
```

`log_genomes = TRUE` (wired in 0.7.x) captures every agent's trait
vector each `log_freq` ticks — consumed by `plot_tsne_genomes()` for
population-genetic structure visualisations.

---

## Inspecting defaults programmatically

```{r inspect-defaults, eval = TRUE, echo = TRUE}
defs <- default_specs()

# All parameter names
length(defs)

# First 10 parameters with their values
str(defs[1:10], max.level = 1, give.attr = FALSE)
```

For an interactive view of all parameters at once, use
[`print_specs()`](../reference/print_specs.html).
