---
title: "Parameter Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, collapse = TRUE, comment = "#>")
library(clade)
```

This article lists every parameter in `default_specs()`, grouped by biological
theme. For each parameter: the default value, type, and what it controls.

Use `print_specs(diff_only = TRUE)` to see which parameters differ from
defaults in a given specs list.

```{r print-specs-example, eval=FALSE}
s <- default_specs()
s$kin_selection      <- TRUE
s$complex_landscape  <- TRUE
print_specs(s, diff_only = TRUE)
```

---

## Grid and population

| Parameter | Default | Type | Description |
|---|---|---|---|
| `grid_rows` | 20 | integer | Number of rows in the toroidal grid |
| `grid_cols` | 20 | integer | Number of columns |
| `n_agents_init` | 50 | integer | Starting population size |
| `max_agents` | 300 | integer | Hard population cap (no births above this) |
| `max_ticks` | 300 | integer | Simulation length |
| `random_seed` | 42 | integer | Random seed (set to reproduce runs exactly) |

---

## Energy and metabolism

| Parameter | Default | Type | Description |
|---|---|---|---|
| `energy_init` | 30.0 | numeric | Starting energy for each agent |
| `energy_max` | 200.0 | numeric | Energy cap per agent |
| `move_cost` | 1.0 | numeric | Energy cost per move action |
| `idle_cost` | 0.5 | numeric | Energy cost per tick of inaction |
| `eat_gain` | 20.0 | numeric | Energy gained from eating grass |
| `min_repro_energy` | 80.0 | numeric | Energy threshold to reproduce |
| `repro_cost` | 20.0 | numeric | Energy cost to parent per offspring |
| `offspring_energy` | 20.0 | numeric | Starting energy for offspring |
| `starvation_threshold` | 0.0 | numeric | Die if energy falls to or below this |

---

## Grass dynamics

| Parameter | Default | Type | Description |
|---|---|---|---|
| `grass_init_prob` | 0.5 | numeric \[0,1\] | Fraction of cells with grass at tick 1 |
| `grass_rate` | 0.10 | numeric \[0,1\] | Per-tick regrowth probability on empty cells |
| `grass_max` | 1.0 | numeric | Maximum grass per cell |

---

## Neural architecture

| Parameter | Default | Options | Description |
|---|---|---|---|
| `brain_type` | `"bnn"` | bnn, ann, ctrnn, grn, transformer, synthesis | Neural network architecture |
| `hidden_layers` | `c(16, 8)` | integer vector | Hidden layer sizes |
| `n_inputs` | auto | integer | Sensory input vector size (set by `create_environment()`) |
| `n_outputs` | 4 | integer | Action output size (N/E/S/W) |

The six brain types differ in expressive power and computational cost:

- **bnn** (Bayesian neural network): default; learns a distribution over weights.
- **ann**: standard multilayer perceptron.
- **ctrnn**: continuous-time recurrent network; suitable for temporally extended tasks.
- **grn**: gene regulatory network topology; sparse, biologically motivated.
- **transformer**: self-attention architecture; highest capacity, slowest.
- **synthesis**: symbolic rule extraction from evolved weights.

---

## Reproduction and genetics

| Parameter | Default | Type | Description |
|---|---|---|---|
| `ploidy` | 1 | 1 or 2 | Haploid or diploid |
| `sex_ratio` | 0.5 | numeric \[0,1\] | Fraction of agents that are female |
| `mating_system` | `"random"` | character | Random, monogamy, polygyny, polyandry |
| `mutation_sd` | 0.05 | numeric | Gaussian noise on genome weights |
| `crossover_rate` | 0.1 | numeric \[0,1\] | Per-locus crossover probability (diploid) |
| `min_repro_age` | 0 | integer | Minimum age to reproduce |
| `max_repro_age` | Inf | integer | Maximum age to reproduce (Inf = no limit) |

---

## Life history

| Parameter | Default | Options | Description |
|---|---|---|---|
| `max_age` | 500 | integer | Die if age exceeds this |
| `life_history` | `"iteroparous"` | iteroparous, semelparous | Semelparous: reproduce once then die |
| `senescence_rate` | 0.0 | numeric | Gompertz mortality scaling with age |
| `repro_senescence` | 0.0 | numeric | Age-related fertility decline rate |

---

## Body size evolution

Enable with `body_size_evolution = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `body_size_evolution` | FALSE | Enable |
| `body_size_init_mean` | 1.0 | Starting value (relative units) |
| `body_size_mutation_sd` | 0.05 | Mutation noise |
| `body_size_min` | 0.1 | Lower clamp |
| `body_size_max` | 5.0 | Upper clamp |

---

## Dispersal evolution

Enable with `dispersal_evolution = TRUE`. Combine with `spatial_sorting = TRUE`
for invasion-front dynamics.

| Parameter | Default | Description |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation noise |
| `dispersal_min` | 0.0 | Minimum clamp |
| `dispersal_max` | 1.0 | Maximum clamp |

---

## Kin selection

Enable with `kin_selection = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `kin_selection` | FALSE | Enable kin altruism |
| `kin_altruism_r_min` | 0.25 | Minimum pedigree relatedness to donate |
| `kin_altruism_cost` | 2.0 | Energy transferred per altruistic act |
| `kin_altruism_min_donor_energy` | 50.0 | Donor energy floor |

---

## SIR disease

Enable with `disease = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `disease` | FALSE | Enable SIR dynamics |
| `transmission_prob` | 0.15 | Per-tick transmission probability |
| `disease_energy_cost` | 5.0 | Energy cost per tick while infected |
| `disease_duration` | 10 | Ticks until recovery |
| `immune_duration` | 20 | Ticks of post-recovery immunity |
| `disease_death_prob` | 0.01 | Additional per-tick mortality while infected |
| `disease_seed_prob` | 0.02 | Fraction of agents infected at tick 1 |

---

## Predators

Enable with `n_predators_init > 0`.

| Parameter | Default | Description |
|---|---|---|
| `n_predators_init` | 0 | Starting predator count |
| `predator_energy_init` | 60.0 | Starting energy for predators |
| `predator_attack_strength` | 10.0 | Damage dealt to prey per attack |
| `predator_min_repro_energy` | 100.0 | Predator reproduction threshold |
| `predator_max_agents` | 50 | Predator population cap |

---

## Niche construction

Enable with `niche_construction = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `niche_construction` | FALSE | Enable shelter building |
| `shelter_build_prob` | 0.1 | Per-tick build probability (when energy sufficient) |
| `shelter_min_energy` | 40.0 | Minimum donor energy to build |
| `shelter_max_depth` | 5 | Maximum shelter units per cell |
| `shelter_decay_rate` | 0.02 | Fraction of shelters lost per tick |

---

## Within-lifetime RL

Enable with `rl_mode = "actor_critic"`.

| Parameter | Default | Description |
|---|---|---|
| `rl_mode` | `"none"` | `"actor_critic"` or `"hebbian"` |
| `learning_rate` | 0.01 | Output-layer update step size |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve genetically |
| `rl_update_freq` | 5 | Ticks between RL updates |

---

## Social learning

Enable with `social_learning = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `social_learning` | FALSE | Enable copying successful neighbours |
| `social_learning_freq` | 10 | Ticks between learning events |
| `social_learning_radius` | 3 | Neighbourhood radius for model selection |

---

## Signals and mate choice

| Parameter | Default | Description |
|---|---|---|
| `signal_dims` | 2 | Dimensionality of heritable signal vector |
| `signal_mutation_sd` | 0.05 | Signal mutation noise |
| `preference_mutation_sd` | 0.05 | Mate-preference mutation noise |

---

## Mimicry and toxicity

Enable with `mimicry = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `mimicry` | FALSE | Enable toxicity and predator learning |
| `toxicity_init_mean` | 0.0 | Starting toxicity |
| `toxin_dose` | 2.0 | Damage per toxicity unit per attack |
| `signal_memory` | 20 | Predator memory window (Rescorla-Wagner) |

---

## Parental care

Enable with `parental_care = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `parental_care` | FALSE | Enable carried offspring |
| `care_duration` | 5 | Ticks offspring are carried |
| `care_energy_cost` | 2.0 | Parent energy cost per juvenile per tick |
| `juvenile_energy_gain` | 3.0 | Energy transferred to juvenile per tick |

---

## Cooperative breeding

Enable with `cooperative_breeding = TRUE`. Usually combined with
`iffolk_selection = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `cooperative_breeding` | FALSE | Enable `helper_tendency` trait |
| `helper_tendency_init_mean` | 0.2 | Starting helper tendency |
| `helper_tendency_mutation_sd` | 0.05 | Mutation noise |

---

## Phenotypic plasticity

Enable with `phenotypic_plasticity = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.5 | Starting plasticity |
| `plasticity_mutation_sd` | 0.05 | Mutation noise |

---

## Spatial sorting

Enable with `spatial_sorting = TRUE`. Requires `dispersal_evolution = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `spatial_sorting` | FALSE | Enable invasion-front mating assortment |
| `sorting_front_threshold` | 0.75 | Fraction from front to define "front zone" |
| `sorting_mating_boost` | 3.0 | Mating encounter fold-increase at front |

---

## IFfolk inclusive fitness

Enable with `iffolk_selection = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `iffolk_selection` | FALSE | Enable energy transfers to relatives |
| `iffolk_r_min` | 0.125 | Minimum relatedness (cousins and closer) |
| `iffolk_radius` | 5 | Neighbourhood radius for kin search |
| `iffolk_transfer` | 3.0 | Energy transferred per act |
| `iffolk_min_energy` | 60.0 | Donor energy floor |
| `parliament_suppression` | FALSE | Penalise defectors among cooperators |
| `parliament_cost` | 0.5 | Energy cost for defectors per cooperator neighbour |

---

## Complex landscape (forest world)

Enable with `complex_landscape = TRUE`.

| Parameter | Default | Description |
|---|---|---|
| `complex_landscape` | FALSE | Enable 3-layer habitat |
| `shrub_density` | 0.35 | Fraction of cells with shrubs |
| `shrub_energy` | 25.0 | Energy gained from shrubs |
| `shrub_growth_rate` | 0.04 | Shrub regrowth rate per tick |
| `canopy_density` | 0.15 | Fraction of cells with canopy |
| `canopy_energy` | 55.0 | Energy gained from canopy |
| `canopy_growth_rate` | 0.005 | Canopy regrowth rate |
| `canopy_threshold` | 0.6 | Minimum wing_size for canopy access |
| `wing_size_init_mean` | 0.1 | Starting wing size |
| `wing_size_mutation_sd` | 0.05 | Wing size mutation noise |
| `wing_size_min` | 0.0 | Lower clamp |
| `wing_size_max` | 1.0 | Upper clamp |

---

## Logging and output

| Parameter | Default | Description |
|---|---|---|
| `log_genomes` | FALSE | Log full genome matrices (large output) |
| `log_deaths` | TRUE | Record every agent death |
| `log_freq` | 1 | Ticks between progress log entries |
| `verbose_julia` | FALSE | Print Julia-side debug messages |

---

## Inspecting defaults programmatically

```{r inspect-defaults}
defs <- default_specs()
# All parameter names
names(defs)
# All parameters with their values
str(defs, max.level = 1, give.attr = FALSE)
```
