---
title: "Parameter search — algorithms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter search — algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This is the reference for the search algorithms themselves — what each
does, when each shines, and how the code looks. For the experimental
question ("when should I search at all?") see
[ps-introduction.html](ps-introduction.html). For which parameters to
search, see [ps-agent-parameters.html](ps-agent-parameters.html) and
[ps-environment-parameters.html](ps-environment-parameters.html).

`clade` ships five search functions. All of them take the same core
arguments — `specs_base`, `search_params` (or `params`), `objective` —
and return results with the same structure, so you can swap algorithms
without changing the surrounding code.

---

## Random search {#random-search}

**When.** A cheap initial screen before committing to a more expensive
algorithm. No priors about where the good region is needed.

**What it does.** Samples `n_samples` points uniformly from the
hyper-rectangle you specify, runs a simulation at each, scores them,
returns them sorted by descending score.

**Cost.** 30 simulations × 5–30 s each = 2–15 minutes at default
settings.

```{r random}
library(clade)

result <- search_random(
  specs_base    = default_specs(),
  search_params = list(
    mutation_sd   = c(0.01, 0.5),
    grass_rate    = c(0.05, 0.8),
    n_agents_init = c(20L, 150L)
  ),
  n_samples  = 30L,
  objective  = "genetic_diversity"
)

head(result)
```

```
#>   mutation_sd grass_rate n_agents_init score
#> 1       0.312      0.621           127 0.847
#> 2       0.287      0.583           119 0.831
#> 3       0.341      0.548           143 0.819
```

Retrieve the full specs of the best row:

```{r random-retrieve}
best_specs <- attr(result, "specs_list")[[1]]
cat("Best mutation_sd:", best_specs$mutation_sd, "\n")
cat("Best grass_rate:", best_specs$grass_rate, "\n")
```

**Strengths.** Embarrassingly parallel, needs no priors, easy to reason
about, gives partial-dependence plots as a free by-product.

**Weaknesses.** Scales badly with dimension (coverage drops fast beyond
5 parameters). Doesn't exploit structure; if two parameters interact,
random search won't exploit it.

---

## MAP-Elites {#map-elites}

**When.** You want to understand the *achievable set* of outcomes, not
just the single best point. MAP-Elites fills a grid of niches in
behavioural space and reports the best parameter set for each niche.

**What it does.** Divides a behavioural space (two or more dimensions of
`get_run_data()$ticks` columns) into bins. Starts with random parameter
sets, and at each iteration picks an existing archive member, perturbs
it, runs a simulation, and places the result in whichever bin it falls
into — *iff* that bin was empty or the new score beats the incumbent.

**Cost.** 300 simulations = 20–90 minutes.

```{r map-elites}
result <- search_map_elites(
  specs_base   = default_specs(),
  archive_dims = list(
    genetic_diversity = seq(0, 1, by = 0.1),
    n_agents          = seq(0, 200, by = 20)
  ),
  n_iterations = 300L,
  objective    = "genetic_diversity",
  verbose      = FALSE
)

# Archive heatmap: x = genetic_diversity niche, y = n_agents niche,
# fill = objective score (NA = empty cell).
result$map
```

**Interpreting the heatmap.**

- *Filled cells* show achievable (diversity, population) pairs. The
  colour encodes the objective score.
- *Empty cells* (grey) are combinations no simulated parameter set
  produced. If high-diversity, large-population cells are empty, that
  combination may be intrinsically unattainable.
- The raw archive (`result$archive`) is a list of length equal to the
  number of cells. Non-null entries have `$specs`, `$score`, and
  `$behavioural_descriptor`.

Inspect the archive directly:

```{r map-elites-inspect}
non_empty <- Filter(Negate(is.null), result$archive)
cat("Filled cells:", length(non_empty), "\n")
cat("mutation_sd range in archive:",
    range(sapply(non_empty, function(x) x$specs$mutation_sd)), "\n")
```

**Strengths.** Illumination — reveals Pareto-like frontiers and
impossible regions that single-objective methods silently skip over.
Robust to local optima.

**Weaknesses.** Expensive per cell filled. Needs meaningful bin
definitions; if your archive dims don't carve the space usefully, the
result is uninformative.

**Known issue (0.4.0 backlog).** At default mutation, MAP-Elites
sometimes fills only one cell — the candidate mutations produce near-
identical behavioural descriptors. See the
[s-map-elites audit](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/map_elites.md)
for the workaround (wider mutation + broader archive bins).

---

## CMA-ES {#cma-es}

**When.** You want the single best parameter combination for one
scalar objective, and the parameters are continuous (not discrete
categorical choices like `brain_type`).

**What it does.** Maintains a multivariate Gaussian over the parameter
space. Each generation samples `popsize` candidates, runs them, and uses
the ranked top half to update the Gaussian's mean and covariance. The
step size (sigma) shrinks as the search converges. `clade`'s
`search_cmaes()` is a pure-R implementation (Hansen 2006) — no external
packages required.

**Cost.** `n_iterations × popsize` simulations = 200–500, i.e. 15–90
minutes.

```{r cmaes}
result <- search_cmaes(
  specs_base = default_specs(),
  params     = c("mutation_sd", "grass_rate", "crossover_rate"),
  objective  = "genetic_diversity",
  n_iterations = 20L,
  popsize    = 10L
)

cat("Best diversity:", result$score, "\n")
cat("Best params:\n")
print(result$specs[c("mutation_sd", "grass_rate", "crossover_rate")])
```

**Strengths.** State of the art for continuous derivative-free
optimisation. Converges faster than random search, handles
ill-conditioned objectives. Treats `run_alife()` as a black box —
discrete events (death, birth) are no obstacle.

**Weaknesses.** Returns a single point, not a trade-off map. Can get
stuck in local optima (mitigated by running multiple restarts with
different seeds). Doesn't handle categorical parameters
(`brain_type`, `life_history`).

**Rule of thumb for choosing between CMA-ES and MAP-Elites:** use
CMA-ES when you care about *where the optimum is*; use MAP-Elites when
you care about *what the landscape looks like around the optimum*.
CMA-ES is faster per run for a single-objective answer; MAP-Elites is
more informative per compute budget for multi-objective insight.

### Scenario auto-calibration harness

A per-scenario CMA-ES harness ships under
[`dev/audit/calibration/`](https://github.com/itchyshin/clade/tree/main/dev/audit/calibration).
Each `s-*.Rmd` scenario has a fitness function encoding its biological
claim (e.g. "negative slope of `mean_prior_sigma`" for the Baldwin
Effect, "late `n_altruistic_acts` per tick" for kin selection). Running
`bash dev/audit/calibration/run_all.sh` launches all 31 in parallel
(xargs -P 16) and writes per-scenario JSON with the best-discovered
specs. Results are summarised in
[[`dev/audit/calibration/RESULTS.md`](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md)](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md).

Several scenarios retracted their prior "Calibrated regime" claims
after the 0.4.0 fidelity audit found the CMA-ES had optimised a generic
fitness rather than the stated theoretical signal — see
[ps-introduction.html](ps-introduction.html#designing-an-objective-function)
for the discipline this requires.

---

## Viability search {#viability-search}

**When.** Before CMA-ES in any region where you're not sure the
population survives at all. Also useful as a standalone sanity-check
when setting up a new scenario.

**What it does.** Runs a rectangular 2-parameter grid with `n_reps`
replicates per cell. Reports a ggplot heatmap coloured by viability
(green = population survives; red = extinction) and a tidy data frame.

**Cost.** `length(values_x) × length(values_y) × n_reps` simulations.
For 5 × 5 × 3 = 75 runs, 5–15 minutes.

```{r viability-map}
library(clade)

s <- default_specs()
s$complex_landscape <- TRUE

vmap <- search_viability(
  s,
  param_x = "shrub_density",  values_x = seq(0.1, 0.6, by = 0.1),
  param_y = "canopy_density", values_y = seq(0.05, 0.35, by = 0.1),
  n_reps  = 3L
)

vmap$map     # ggplot2 heatmap
vmap$data    # data frame with viability, mean_final_pop per cell
```

**Strengths.** Clearest possible picture of where the system is alive.
Runs fast because extinction-prone cells finish early. Cheap enough to
run every time you set up a new scenario.

**Weaknesses.** Only 2 parameters at a time (by design — heatmap output).
Doesn't optimise; just shows you where it's safe to optimise.

### Pre-packaged tuners

For the two scenarios where viability mapping is particularly worthwhile,
`clade` ships domain-specific tuners that combine viability mapping with
a CMA-ES pass inside the viable region:

- `tune_complex_landscape()` — five landscape parameters + custom
  objective measuring joint increase in wing size, niche diversity, and
  survival.
- `tune_iffolk()` — IFfolk transfer rate and parliament cost + custom
  objective rewarding upward trends in `mean_helper_tendency`.

---

## Gradient ascent {#gradient-ascent}

**When.** Final-tick refinement when the objective surface is smooth
near a known good point — and you want fractional improvement on a
parameter you already trust.

**What it does.** Estimates the gradient of the objective with respect to
each parameter by finite differences, then takes a step in the positive
direction. Repeats until the step size drops below a threshold.

**Cost.** `n_iterations × (n_params × 2 + 1)` simulations per iteration
(two finite-difference evaluations per parameter plus one at the
midpoint). Rarely the most efficient option.

**Strengths.** Simple. Good for the last 1–5% of optimisation after
CMA-ES has found the basin.

**Weaknesses.** Gets stuck in local optima. Sensitive to step size.
Doesn't handle noisy objectives well (and clade's simulations are
stochastic, so objectives *are* noisy).

Usage mirrors the CMA-ES signature:

```{r gradient, eval=FALSE}
result <- search_gradient(
  specs_base   = default_specs(),
  params       = c("mutation_sd", "grass_rate"),
  objective    = "genetic_diversity",
  n_iterations = 20L,
  step_size    = 0.02
)
```

---

## Why not Julia autodiff?

Julia has mature automatic differentiation (Zygote.jl, Enzyme.jl). It
would be natural to ask whether we could differentiate `run_alife()`
end-to-end and use gradient descent. We cannot. `run_alife()` contains
several operations that are not differentiable:

- **Agent death** is a threshold event: `energy < starvation_threshold`.
- **Reproduction** is conditional on `energy >= min_repro_energy`.
- **Action selection** uses `argmax` over brain logits.
- **Mating** is a categorical sample.

Autodiff through any of these either fails at trace time or silently
produces a zero gradient. The correct class of optimisers for stochastic
discrete-event simulations is derivative-free: CMA-ES, MAP-Elites,
random sampling. They treat the simulator as a black box and estimate
curvature from population statistics.

---

## Compute parallelism

All five search functions accept an `n_cores` argument (default 1). Set
it to the number of R workers you want clade to fan the inner
simulations across. For `n_iterations × popsize = 200` simulations, a
4-core machine can cut wall time from ~40 minutes to ~10 minutes.

Julia caches its precompiled simulation package per worker, so the first
call in each worker pays the ~60-second compile cost. For batches of
10 or more simulations per worker, this overhead is amortised; for
smaller batches, consider running on fewer cores or using
`batch_alife()` to warm the Julia sessions first.

---

## Choosing the right algorithm — a decision tree

```
Is the population surviving at all at the starting point?
├── No  → search_viability() first.
└── Yes → Do you want one best answer or a landscape?
    ├── One best answer
    │   ├── Parameters continuous?    → search_cmaes()
    │   └── Parameters categorical?   → loop over conditions
    │                                   with batch_alife()
    └── A landscape
        ├── 2D trade-offs known in advance  → search_map_elites()
        └── Exploring without priors        → search_random()
                                              (then iterate)
```

---

## See also

- **[ps-introduction.html](ps-introduction.html)** — when to search and
  how to design objectives.
- **[ps-agent-parameters.html](ps-agent-parameters.html)** — per-organism
  parameters to search.
- **[ps-environment-parameters.html](ps-environment-parameters.html)** —
  world-level parameters to search.
- **`?search_random`, `?search_map_elites`, `?search_cmaes`,
  `?search_viability`, `?search_gradient`** — full function reference.

---

## References

- Hansen, N. & Ostermeier, A. (2001) Completely derandomized
  self-adaptation in evolution strategies. *Evolutionary Computation*
  9(2):159–195.
- Hansen, N. (2006) The CMA Evolution Strategy: A Comparing Review.
  *Towards a New Evolutionary Computation*, Springer: 75–102.
- Mouret, J.-B. & Clune, J. (2015) Illuminating search spaces by
  mapping elites. *arXiv:1504.04909*.
