---
title: "Parameter search — introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter search — introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

## Why search the parameter space at all?

`clade` exposes roughly 90 parameters (see `vignette("parameter-reference")`).
Even if most of them are at a biologically defensible default, any serious
experiment asks a question of the form *which parameter combination produces
outcome X?* — and "outcome X" is rarely a single parameter turned up.

Examples that motivate searching:

- **Discovery experiments.** "What grass-renewal rate × metabolic-rate
  combination produces the largest sustained population?"
- **Scenario calibration.** "At what `toxin_dose × avoid_threshold` does
  Müllerian aposematism actually evolve, rather than drift?"
- **Regime mapping.** "Where is the viable region for the complex-landscape
  module — which `shrub_density × canopy_density` combinations avoid
  population collapse?"
- **Trade-off visualisation.** "What is the achievable set of (genetic
  diversity, population size) pairs? Is high diversity at low population
  size the only option, or can both be large simultaneously?"

Each of these questions maps to a different search algorithm. A single "tune
this" button would be the wrong tool for most of them.

---

## The four tools at a glance

```{r tools-table, echo=FALSE, eval=TRUE}
knitr::kable(
  data.frame(
    Function = c("search_random()",
                 "search_map_elites()",
                 "search_cmaes()",
                 "search_viability()",
                 "search_gradient()"),
    Strategy = c("Stochastic sweep",
                 "Quality-diversity archive (Mouret & Clune 2015)",
                 "CMA-ES optimisation (Hansen 2006)",
                 "Rectangular viability map",
                 "Finite-difference gradient ascent"),
    When = c("Fast initial screen — before you know what to tune",
             "Illuminate the achievable behavioural space, not a single optimum",
             "Find the single best parameter combination for one scalar objective",
             "Identify which parameter region keeps the population alive at all",
             "Local refinement of an already-good parameter set")
  ),
  caption = "`clade` parameter search functions at a glance."
)
```

A typical workflow moves from coarse to fine:

1. **`search_viability()`** — if you are not sure the default region even
   sustains life, map it first.
2. **`search_random()`** — a cheap 30-run sweep to find roughly productive
   regions of parameter space.
3. **`search_map_elites()`** — illuminate the trade-offs among the top
   candidates from step 2.
4. **`search_cmaes()`** — zoom into the cell you care about most and find
   the single best point.
5. **`search_gradient()`** — optional final-tick refinement if the objective
   surface is smooth near the optimum.

Most users will live in steps 2 and 3. The others are specialist tools.

---

## What gets searched: agent vs environment parameters

A parameter in `default_specs()` is usually one of two kinds, and the two
kinds imply different experimental designs and different choices of search
method. We dedicate a vignette to each:

- **[Agent-level parameters](ps-agent-parameters.html)** — anything that
  varies per organism or per species: heritable traits (`body_size_*`,
  `metabolic_rate_*`, `brain_size_*`), brain architecture (`brain_type`,
  `bnn_sigma_*`, `rl_mode`), life-history (`max_age`, `senescence_rate`,
  `life_history`), reproduction (`repro_cost_*`, `offspring_energy_*`,
  `clutch_size_*`, `female_investment`), mutation (`mutation_sd`,
  `stress_mutation_multiplier`), signals and mate choice (`signal_dims`,
  `mate_choice_*`), social behaviours (`kin_altruism_*`, `iffolk_*`,
  `parliament_*`, `cooperation_*`).
- **[Environment-level parameters](ps-environment-parameters.html)** —
  anything about the world: grid and population bookkeeping (`grid_rows`,
  `grid_cols`, `toroidal`, `n_agents_init`, `max_agents`, `max_ticks`),
  resources (`grass_rate`, `grass_max`, `eat_gain`, `max_bite`), temporal
  and spatial structure (`seasonal_amplitude`, `season_length`,
  `complex_landscape`, `fixed_patch`), external forces (`n_predators_init`
  + predator params, `disease` + disease params).

The two categories overlap at their edges: predators, disease, and niche
construction all sit where agents and world meet. In those cases we flag
the fuzzy parameter on *both* pages with a cross-link rather than pick a
side.

---

## Designing an objective function

Every search function accepts either a column name (from
`get_run_data()$ticks`) or a custom function that takes the run
environment and returns a scalar. The custom function form is the
important one: it is what separates a valid scientific search from a
fitness-hacking exercise.

Good objective functions for clade typically have three properties.

**1. They measure the biology you actually claim.** If the scenario's
biological claim is "Hamilton's rule predicts population rises with *rB/C*",
the objective should be the Spearman correlation between `rB/C` and
`mean_n_agents` across a *grid* of `B` and `C` values — not the population
size at a single parameter point. This guards against Goodhart's law, which
the earlier CMA-ES-calibrated regimes in some scenario vignettes fell
victim to.

**2. They penalise degenerate outcomes.** A run that goes extinct on tick
50 should not be rewarded for its transient high diversity. The canonical
pattern is:

```{r custom-objective}
custom_obj <- function(env) {
  d   <- get_run_data(env)$ticks
  div <- mean(d$genetic_diversity, na.rm = TRUE)
  pop <- mean(d$n_agents,          na.rm = TRUE)
  if (pop < 10) return(0)      # discard runs that go near-extinct
  div * log1p(pop / 50)        # reward diversity weighted by log-population
}
```

This penalises extinction, smooths the score surface, and keeps the
optimiser out of the "high-diversity-just-before-collapse" trap.

**3. They are insensitive to run length when it should be.** If you want
"equilibrium diversity", average over the second half of the run rather
than the whole run — the first half is dominated by initial-condition
transients.

The same `custom_obj` can be passed unchanged to
[`search_random()`](ps-algorithms.html#random-search),
[`search_map_elites()`](ps-algorithms.html#map-elites),
[`search_cmaes()`](ps-algorithms.html#cma-es), or
[`search_gradient()`](ps-algorithms.html#gradient-ascent).

---

## Why not autodiff?

Julia has excellent automatic differentiation (Zygote.jl, Enzyme.jl). It is
tempting to ask whether `clade`'s simulator could be differentiated end-to-end
and optimised with gradient descent. It cannot, for good biological reasons.
`run_alife()` contains several operations that are not differentiable:

- **Agent death** is a threshold event: `energy < starvation_threshold` triggers
  removal. The gradient of a Bernoulli draw w.r.t. energy is zero almost
  everywhere.
- **Reproduction** is conditional on `energy >= min_repro_energy` — another
  threshold.
- **Action selection** uses `argmax` over the brain's output logits. `argmax`
  has zero gradient w.r.t. the logits almost everywhere.
- **Mating** is a categorical sample, again a zero-gradient operation.

Autodiff through these either fails at trace time or silently produces a zero
gradient. The correct optimisers here are **derivative-free** — CMA-ES,
MAP-Elites, random sampling — which treat `run_alife()` as a black box and
estimate curvature from population statistics alone.

---

## Compute budget

A practical constraint: each `run_alife()` call takes roughly 5–30 seconds on
a modern machine (default grid, 500 ticks). A typical search therefore
involves ~300 runs and costs tens of minutes to an hour. Two ways to
multiply throughput:

- Use `n_cores > 1` in any of the search functions. clade will fan the
  inner simulations across R workers.
- Use a smaller `max_ticks` during search (300–500) and re-run the winning
  candidate at full length (1,000+) to confirm the signal is not
  transient. This typically doubles effective throughput.

Avoid running CMA-ES or MAP-Elites at `max_ticks > 1000` until you know
the winning region; the coarse-to-fine workflow above is cheaper.

---

## Where to go next

- **[Agent parameters](ps-agent-parameters.html)** — canonical list,
  search examples per module.
- **[Environment parameters](ps-environment-parameters.html)** — grid,
  grass, seasonality, landscape, external forces.
- **[Algorithms](ps-algorithms.html)** — CMA-ES, MAP-Elites, viability,
  gradient: how each works and when each shines, with full working code.
- **Scenario auto-calibration harness.** A per-scenario CMA-ES harness
  ships under
  [`dev/audit/calibration/`](https://github.com/itchyshin/clade/tree/main/dev/audit/calibration).
  See [`RESULTS.md`](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md)
  for what was discovered and what was retracted after the 0.4.0
  fidelity audit.

---

## References

- Hansen, N. & Ostermeier, A. (2001) Completely derandomized
  self-adaptation in evolution strategies. *Evolutionary Computation*
  9(2):159–195.
- Mouret, J.-B. & Clune, J. (2015) Illuminating search spaces by mapping
  elites. *arXiv:1504.04909*.
