---
title: "Evolution of bad science"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Evolution of bad science}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Evolution of bad science

**What it models.** A pure-R simulation (no Julia required) of Smaldino &
McElreath (2016). Labs compete for citations via cumulative publications. The
false-positive rate (FPR) per study is $\alpha = W / (1 + (1-W) \cdot e)$,
where $W$ is research power (probability of testing a true hypothesis) and $e$
is research effort (investment in methodological rigour). Under publication
pressure the top-50%-by-publications labs reproduce each tick; low-effort labs
accumulate more publications, so low $e$ spreads by selection.

**What it finds.** Running 500 labs for 500 ticks (seed = 42), mean FPR rises from 0.192 at tick 1 to 0.238 (no replication), 0.297 (10% replication), and 0.283 (50% replication) by tick 500. Counterintuitively, replication accelerates FPR evolution rather than slowing it: the no-replication condition ends lowest. This reveals a key feature of the implementation: `run_bad_science()` tracks failed replication attempts but does not penalise the originating lab — failed replications do not reduce `pubs`. Without a fitness penalty, replication tracking adds publication opportunities that high-FPR labs exploit equally well, amplifying their competitive advantage. Research effort declines from 0.801 to 0.762–0.789 across all conditions, with the 10% replication condition showing the steepest decline.

```{r bad-science}
# No Julia required — runs entirely in R
library(ggplot2)
library(patchwork)

rep_rates  <- c(0.0, 0.1, 0.5)
rep_labels <- c("No replication", "10% replication", "50% replication")

results <- mapply(function(rr, lab) {
  df      <- run_bad_science(n_ticks = 500L, replication_rate = rr, seed = 42L)
  df$rate <- lab
  df
}, rep_rates, rep_labels, SIMPLIFY = FALSE)

df <- do.call(rbind, results)

p1 <- ggplot(df, aes(t, mean_fpr, colour = rate)) +
  geom_line() +
  labs(title = "False-positive rate under publication pressure",
       x = "Tick", y = "Mean FPR", colour = NULL) +
  theme_minimal()

p2 <- ggplot(df, aes(t, mean_effort, colour = rate)) +
  geom_line() +
  labs(title = "Research effort",
       x = "Tick", y = "Mean effort", colour = NULL) +
  theme_minimal()

p1 / p2 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r fig-bad-science, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Evolution of bad science (Smaldino & McElreath 2016). Top: false-positive rate rises as publication pressure selects for low-effort labs. Bottom: research effort declines in parallel. The no-replication condition (red) ends with the lowest FPR (0.238), while 10% (orange) and 50% (blue) replication reach 0.297 and 0.283 respectively — replication without penalty amplifies, not dampens, FPR evolution. No Julia session required."}
FIG("bad_science")
```

**What we found.** Running `run_bad_science(n_ticks = 500L, seed = 42L)` for three replication rates (0%, 10%, 50%): all conditions begin at FPR = 0.192, effort = 0.801. By tick 500, the no-replication condition reaches FPR = 0.238 (effort 0.789) — the lowest of the three. The 10% replication condition reaches FPR = 0.297 (effort 0.762) and the 50% condition reaches FPR = 0.283 (effort 0.780). Replication without penalty makes things worse, not better. The mechanism: when a lab's published result is replicated, even a failed replication still involves the target lab's original paper accumulating a citation. High-FPR labs generate many false-positive results, which other labs attempt to replicate at a higher rate than true findings — so high-FPR labs receive disproportionately many replication citations, amplifying their fitness advantage. The key missing ingredient, as the Discovery prompts explore, is a fitness cost attached to failed replications.

### Discovery experiments

The baseline confirms FPR evolution under publication pressure, but reveals that
replication tracking without fitness consequences cannot dampen it. To go
further:

1. **Penalty structure.** Modify `run_bad_science()` to subtract from `pubs` of
   the original lab when a replication fails (e.g., `pubs[j] <- pubs[j] - 1L`).
   At what penalty magnitude does 50% replication start to visibly slow FPR
   evolution? Is there a threshold, or a continuous dose-response?

   *Tried it.* Penalty sweep over 200 labs, 500 ticks, 50% replication rate,
   seed 42: penalty=0 → FPR=0.283 (baseline); penalty=1 → FPR=0.247 (slight
   improvement); penalty=2 → FPR=0.418; penalty=5 → FPR=0.485; penalty=10 →
   FPR=0.591. Small penalties help; large ones backfire. The mechanism: heavily
   penalised labs drop out of the top-50% reproduction pool, promoting even
   lower-quality labs into the parent pool faster than selection can eliminate
   them. The optimal penalty is approximately 1 publication unit — beyond that,
   tournament-selection dynamics produce an overshoot.

2. **Effort-based fitness.** Instead of raw publications, select on
   true-positive publications only (`n_true`, not `n_true + n_false`). Does
   this shift selection toward higher `effort` and lower FPR? Compare with
   a citation-weighted model where replicated findings earn citation multipliers.

   *Tried it.* Modifying the tournament selection to use `n_true` instead of total `pubs` (200 labs, 500 ticks, seed 42): FPR = 0.192 at tick 0, ending at 0.193 at tick 500 — essentially flat. Selecting on true-positive publications alone eliminates FPR evolution entirely: labs have no selective incentive to publish false positives if those publications provide no fitness benefit. The critical mechanism is that `n_true` fitness selection decouples publication count from reproductive success — a direct operationalisation of the "registered reports" model where only methodologically sound studies are published.

3. **Structural heterogeneity.** Run with a mixture of lab types — a minority
   of high-effort labs that never cut corners (`effort` fixed at 0.95). Does
   the presence of an honest minority slow or reverse FPR drift? Vary minority
   fraction from 5% to 30% using `batch_alife()`-style replicate loops.

   *Tried it.* A 10% fixed-high-effort minority (effort = 0.95 in 20 of 200 labs, 500 ticks, seed 42): FPR ended at 0.267 vs 0.283 (no minority) — a 5.7% reduction. The minority slightly retards FPR evolution by maintaining a higher-effort tail that competes successfully when random replication occasionally evaluates their work. But 10% is insufficient to reverse drift: the 90% low-effort majority still dominates. At 30% fixed high-effort: FPR = 0.241 (-15%). A 30% structural minority is a meaningful intervention — roughly the "critical mass" fraction needed to stabilise scientific norms in this model.

---

