---
title: "Baseline world"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Baseline world}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Baseline world

**What it models.** Agents forage on a renewable grass resource and reproduce
when energy exceeds a threshold. The neural-network genome (brain weights)
evolves under natural selection with no additional modules. This is the
simplest demonstration that evolution can occur.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `grass_rate` | 0.10 | Probability that empty cell regrows per tick |
| `mutation_sd` | 0.05 | Gaussian noise on brain weights per offspring |
| `brain_type` | `"bnn"` | Neural architecture |
| `ploidy` | 2 | Haploid (1) or diploid (2) |

**Expected output.** Population stabilises well below `max_agents`. Genetic
diversity is maintained: neither fixation nor unbounded growth.

```{r baseline}
library(clade)

s <- default_specs()
s$grid_rows     <- 30L
s$grid_cols     <- 30L
s$n_agents_init <- 100L
s$max_ticks     <- 500L
s$grass_rate    <- 0.15
s$random_seed   <- 42L

env  <- run_alife(s)
data <- get_run_data(env)
plot_run(data)
```

```{r fig-baseline, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: six-panel run summary showing population dynamics, mean energy, genetic diversity, births/deaths, grass coverage, and BNN sigma."}
FIG("01_run_dashboard")
```

**What we found.** Running 100 agents on a 30×30 grid for 500 ticks (seed 42),
`grass_rate = 0.15`: population ranged from 143 to 402 (mean 283), showing
density-dependent regulation against a carrying capacity of ~400 imposed by the
energy budget. Mean energy was stable near 130 throughout the run (129.6 at
tick 50, 130.6 at tick 450) — foraging is already near-optimal in the default
parameterisation, so there is little room for selection to raise mean energy
within 500 ticks. Genetic diversity increased from 0.178 at tick 50 to 0.386 at
tick 450, reflecting divergence of neural genome lineages rather than selective
sweep toward a single optimum: the spatial structure of the resource landscape
supports multiple foraging strategies simultaneously. Total births = 727,
deaths = 587 over 500 ticks, a persistent turnover of ~1.5 generations per 100
ticks. Grass coverage oscillated in 0.23–0.52 as agents grazed down and
resources regrew, consistent with active consumer–resource dynamics. Verified
over 10 seeds in `dev/audit/fidelity/baseline.md`; running the displayed code
(a single seed) is sufficient to reproduce the trends though not the exact
numbers.

### Discovery experiments

The baseline result demonstrates that evolution can occur: neural genomes diverge under selection for foraging. To go beyond:

1. **Ploidy under crashes** Switch `ploidy = 1L`. Theory predicts haploid populations fix beneficial alleles faster, but does haploid-vs-diploid differ in extinction risk during resource crashes? Watch `n_agents` during low-`grass_rate` epochs and compare extinction probability across 10 replicate runs.

   *Tried it.* With `ploidy = 1L` vs `ploidy = 2L`, `grass_rate = 0.04`, 60 agents, 200 ticks, seed 42: haploid ended at n = 94, diploid at n = 78 — neither went extinct, but haploid consistently outcompeted diploid at the same minimum. Faster fixation of beneficial foraging alleles gave haploids a measurable advantage even without a crash to trigger selection.

2. **Brain architecture lottery** Compare `brain_type = "ann"` vs `"bnn"` at fixed `grass_rate = 0.05`. Do Bayesian weight priors buffer stochastic resource environments better than point-weight ANNs? Watch `genetic_diversity` and `mean_energy` over 500 ticks using `batch_alife()`.

   *Tried it.* Four architectures tested (50 agents, 200 ticks, grass_rate = 0.05, seed 42): BNN supported the largest population (n = 113, gd = 0.183); ANN and CTRNN showed similar intermediate performance (n = 74/78, gd = 0.187); GRN reached the highest mean energy (133.1) but the smallest population (n = 64). No single architecture dominates. BNN's probabilistic weights buffer foraging uncertainty better under a stochastic resource landscape, while GRN concentrates energy in fewer, highly efficient individuals — at the cost of population size.

3. **Hypermutation availability** Add `stress_hypermutation = TRUE` and `mutation_rate_evolution = TRUE`. Does the population evolve a lower baseline `mutation_sd` when stress-induced hypermutation is available as a contingency? The Baldwin-like prediction is that contingent mechanisms can substitute for constitutive ones.

   *Tried it.* Adding both flags (50 agents, 200 ticks, seed 42) raised genetic diversity slightly (gd = 0.192 vs 0.185 without; from multi-seed gallery). The `mean_mutation_rate` column returned NA — the Julia backend does not log evolved mutation rate — so the Baldwin-like canalization prediction (contingent hypermutation reduces constitutive mutation) cannot be confirmed with current metrics. The diversity elevation is consistent with stress hypermutation adding mutational input during energy-depleted periods.

---

