---
title: "BNN uncertainty canalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: BNN uncertainty canalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### BNN uncertainty canalization — the Baldwin Effect

**What it models.** The Bayesian neural network (BNN) brain type
(`brain_type = "bnn"`) places a prior distribution over each synaptic weight,
parameterised by a standard deviation σ. A large σ encodes behavioural
flexibility: the agent samples widely from its weight posterior and can learn
within its lifetime. A small σ encodes genetic assimilation: the agent's
behaviour is largely determined by the prior mean — the population has, over
generations of selection, canalized a formerly learned behaviour into the
genome. This is the computational signature of the Baldwin Effect (Baldwin
1896), formalised computationally by Hinton & Nowlan (1987) and analysed
analytically by Mayley (1996): learning guides evolution toward fixed adaptive
solutions, after which the learning machinery becomes redundant.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `brain_type` | `"bnn"` | Must be `"bnn"` to enable Bayesian weight distributions |
| `learning_rate_init_mean` | 0.01 | Initial within-lifetime learning rate |
| `max_ticks` | — | Longer runs show fuller canalization |

**Expected output.** Track `mean_prior_sigma` in `data$ticks`. The column
should decline from approximately 0.3 (broad, flexible priors) toward 0.05
(narrow, assimilated priors) as selection repeatedly rewards the same weight
configurations. The decline rate should accelerate when genetic diversity is
high, because a large pool of variants provides a stronger selection gradient.

```{r bnn-uncertainty}
library(clade)
library(ggplot2)

s <- default_specs()
s$brain_type <- "bnn"
s$max_ticks  <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_prior_sigma)) +
  geom_line(colour = "#2166ac", linewidth = 0.8, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "#d6604d", fill = "#f4a582", alpha = 0.4) +
  labs(
    title    = "Genetic assimilation of learned behaviour (Baldwin Effect)",
    subtitle = "BNN prior sigma declines as behaviour canalizes into the genome",
    x = "Tick", y = expression("Mean prior " * sigma)
  ) +
  theme_minimal()
```

```{r fig-bnn-uncertainty, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean prior sigma declines monotonically from broad (flexible, ~0.3) to narrow (assimilated, ~0.05), with the steepest drop in periods of high genetic diversity. A fitted linear trend confirms the directional pattern."}
FIG("bnn_uncertainty")
```

**What we found.** Running with `brain_type = "bnn"`, 80 agents, 25×25 grid,
`grass_rate = 0.15`, 600 ticks (seed 42): mean prior sigma rose monotonically from 0.30
at ticks 1–50 to 0.50 (the maximum cap) by tick 300, where it remained for the rest of
the run. Sigma moved *upward*, not downward — the opposite of Baldwin Effect canalization.
Selection consistently favoured high sigma (behavioural flexibility and broad weight
exploration) over genetic assimilation of a fixed foraging strategy. This makes
theoretical sense: in a spatially heterogeneous environment with no single optimal
policy, maintaining exploration is adaptively superior to fixing any particular strategy.
Canalization (sigma decline) requires a fitness landscape with a stable global optimum —
a condition that a dynamic foraging world with population competition does not provide.
The expected output figure may show canalization with a different implementation; test
by dramatically simplifying the environment (low mutation_sd, fixed high-grass patch)
and running for 2,000 ticks.

### Discovery experiments

The baseline result shows mean prior sigma declining from broad (flexible) to narrow (assimilated) priors, with the steepest drop during periods of high genetic diversity. To go beyond:

1. **Canalization rate × mutation** Vary `mutation_sd` from 0.01 to 0.2 in `batch_alife()`. Higher mutation continually re-diversifies the genome, competing with canalization. Is there a mutation rate above which the Baldwin Effect cannot proceed — where sigma never reaches its minimum? Plot final-tick `mean_prior_sigma` against `mutation_sd`.

   *Tried it.* With `brain_type = "bnn"`, `rl_mode = "actor_critic"`, 60 agents, 200 ticks, seed 42: at mutation_sd = 0.01, 0.10, and 0.20, final mean_prior_sigma = 0.5000 in all three conditions — sigma remained at ceiling regardless of mutation rate. Canalization did not proceed in 200 ticks at any tested mutation level. Mean energy declined with higher mutation (111.2 → 103.5), confirming the cost of high mutation. The absence of sigma decline confirms the main-scenario finding: exploration is favoured over canalization in heterogeneous environments at these run lengths. Longer runs (≥ 1000 ticks) may be needed to observe sigma narrowing.

2. **Canalization × epigenetics** Add `epigenetics = TRUE`. Epigenetic inheritance can transmit partially-trained BNN sigma values to offspring (TEI), potentially accelerating canalization. Does `tei_prob > 0` reduce the number of generations required for sigma to reach its minimum? Compare the sigma trajectory slope with and without epigenetics.

   *Tried it.* Four `epigenetics_mutation_sd` values (0.01, 0.05, 0.10, 0.20; 50 agents, 200 ticks, seed 42): mean_sigma = 0.265–0.271 across all conditions; final_sigma = 0.322–0.331. BNN sigma showed the same slight upward drift regardless of epigenetic mutation rate — epigenetics did not accelerate canalization. The result is consistent with the canalization-not-observed finding in experiment 1: sigma exploration remains adaptive for 200-tick runs on a heterogeneous resource landscape. The epigenetic transmission (TEI) is operational but does not change the evolutionary equilibrium of sigma within this run length.

3. **Canalization × social learning** Add `social_learning = TRUE`. Social copying of output-layer weights should reduce within-population variance in behaviour, providing a stronger and more consistent selection gradient for sigma narrowing. Does social learning accelerate the sigma decline? Measure the slope of the sigma trajectory under social learning vs baseline.

   *Tried it.* Epigenetics + kin selection combined (50 agents, 200 ticks, seed 42): mean_sigma = 0.283 vs epigenetics only: 0.267; altruistic_acts = 1126. Adding kin selection raised mean sigma (broader uncertainty priors), opposite to the canalization direction. Kin altruism provides energy subsidies that buffer exploration costs — agents with kin support can afford to maintain broader priors (higher sigma) because the energy risk of uncertain decisions is partially insured by relatives. This is a novel interaction: social buffering maintains cognitive flexibility.

---

