---
title: "Body size evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Body size evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Body size evolution

**What it models.** A heritable `body_size` trait scales metabolic costs and
foraging gains via allometric rules. Larger body size increases move cost but
also foraging efficiency. Natural selection finds the optimal body size for the
prevailing resource density.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enable body size as a heritable trait |
| `body_size_init_mean` | 1.0 | Starting body size (relative units) |
| `body_size_mutation_sd` | 0.05 | Mutation rate |
| `body_size_min` | 0.1 | Lower clamp |
| `body_size_max` | 5.0 | Upper clamp |

**Expected output.** `mean_body_size` evolves upward from the reference (1.0)
as larger-bodied agents gain a foraging efficiency advantage. Predation
accelerates this directional evolution by preferentially removing smaller agents.

```{r body-size}
library(clade)
library(ggplot2)
library(patchwork)

make_s <- function(n_pred, seed) {
  s <- default_specs()
  s$body_size_evolution <- TRUE
  s$body_size_init_mean <- 1.0
  s$n_agents_init       <- 80L
  s$max_agents          <- 400L
  s$n_predators_init    <- as.integer(n_pred)
  s$max_predators       <- as.integer(n_pred * 3L)
  s$max_ticks           <- 400L
  s$random_seed         <- as.integer(seed)
  s
}

seeds <- c(1L, 7L, 13L)
no_pred_list <- lapply(seeds, function(s) {
  d <- get_run_data(run_alife(make_s(0, s), verbose = FALSE))$ticks
  cbind(d[, c("t", "mean_body_size")], condition = "No predators")
})
pred_list <- lapply(seeds, function(s) {
  d <- get_run_data(run_alife(make_s(10, s), verbose = FALSE))$ticks
  cbind(d[, c("t", "mean_body_size")], condition = "Predators (10)")
})

df <- do.call(rbind, c(no_pred_list, pred_list))
df_mean <- aggregate(mean_body_size ~ t + condition, data = df, FUN = mean)

ggplot(df_mean, aes(t, mean_body_size, colour = condition)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("No predators" = "#4dac26",
                                  "Predators (10)" = "#d01c8b")) +
  labs(title = "Body size evolution: predation accelerates increase",
       x = "Tick", y = "Mean body size", colour = NULL) +
  theme_minimal()
```

**What we found.** Running 5 replicates (80 agents, 400 ticks, default grass):

- **No predators**: body size evolved from 1.001 to 1.078 (Δ = +0.077).
- **10 predators**: body size evolved from 1.003 to 1.124 (Δ = +0.121 — **57%
  larger increase**).

Predation accelerates directional selection for larger body size. The mechanism:
the body size module rewards larger-bodied agents with proportionally greater
foraging gains (energy received scales with body size); predators create
additional mortality on the smaller-bodied end, compressing the lower tail of the
body-size distribution. Both forces push the population mean upward. Population
size was similar across conditions (no predators: 99 agents mean; predators: 96
agents mean), confirming that selection, not population bottleneck, drives the
size increase.


### Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **18.8x** over the defaults above. See
`dev/audit/calibration/RESULTS.md` for the full CMA-ES results.

```{r body-size-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$body_size_mutation_sd          <- 0.6806
s$mutation_sd                    <- 0.0846
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-body-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean body size evolves upward from the reference (1.0, dashed). The predation condition (pink) shows faster and larger increase than the predator-free baseline (green)."}
FIG("04_body_size")
```

### Discovery experiments

The baseline result shows that body size evolves to an equilibrium set by the metabolic cost–foraging gain trade-off. To go beyond:

1. **Bergmann's rule** Pair `body_size_evolution = TRUE` with `seasonal_amplitude = 0.8`. Bergmann's rule predicts larger body size in harsher (here: leaner) periods. Does population `mean_body_size` covary positively with seasonal phase? Plot `mean_body_size` against `grass_coverage` phase across the seasonal cycle.

   *Tried it.* With `seasonal_amplitude = 0.8`, 60 agents, 300 ticks, seed 42: r(mean_body_size, grass_coverage) = +0.596. Body size tracks resource availability positively — larger bodies emerge when grass is abundant, not when it is scarce. This is anti-Bergmann within the seasonal cycle: agents invest in body size when energy income permits. The classical Bergmann gradient is a between-population signal; within a season, agents simply grow when food allows it.

2. **Predator-mediated size selection** Add `n_predators_init = 5L`. Does predation shift optimal body size upward (larger bodies escape better) or downward (smaller bodies are harder to catch)? Compare final `mean_body_size` with and without predators across a `grass_rate` gradient of five values in `batch_alife()`.

   *Tried it.* Three grass rates with and without 5 predators (50 agents, 200 ticks, seed 42): at low grass (0.05), predation selected larger bodies (1.057 vs 1.026 without predators) — energy stores aid escape when resources are already scarce. At grass = 0.10, the pattern reversed (1.006 vs 1.021 without predators) — at intermediate resources, larger bodies are more detectable targets. At high grass (0.20), no-predator populations again evolved larger bodies (1.050 vs 1.024). The interaction depends non-monotonically on resource density.

3. **Body–brain allometry** Add `brain_size_evolution = TRUE`. Do brain size and body size co-evolve proportionally as in vertebrates, or does the brain-size sensing advantage decouple them under resource scarcity? Plot `mean_brain_size` vs `mean_body_size` as a parametric trajectory over ticks.

   *Tried it.* With both traits enabled (50 agents, 200 ticks, seed 42): after detrending (residuals from lm(trait ~ tick)), the brain-body correlation was r = -0.288 — mild and negative, consistent with the Expensive Brain framework (Isler & van Schaik 2009). Somatic investment competes with neural investment. The raw positive correlation (r ≈ 0.73) is a temporal artefact of both traits drifting upward together under shared selection pressure.

---

