---
title: "Brain size evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Brain size evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Brain size evolution and the parental provisioning hypothesis

The **parental provisioning hypothesis** proposes that parental care is a
prerequisite for brain size evolution (van Schaik et al. 2023; Griesser et al.
2023; Song et al. 2025). The mechanism has three components:

1. **Expensive brain hypothesis**: brain tissue is metabolically costly and
   cannot be down-regulated. Each unit of brain size adds to the idle metabolic
   cost that must be paid every tick, regardless of whether the agent is
   foraging successfully.
2. **Bootstrapping problem**: large-brained newborns immediately pay the
   metabolic surcharge but forage poorly (ANN weights are random at birth).
   They starve before their cognitive advantage can offset the cost — *unless*
   a parent provides exogenous energy during infancy.
3. **Sensing quality**: larger-brained agents perceive grass resource gradients
   more clearly. Grass inputs to the neural network are scaled by
   `brain_size ^ brain_size_sensing_exponent`, amplifying the signal from
   nearby food cells. This provides a directional navigation advantage that
   complements the on-cell foraging bonus and compounds as brain size evolves
   upward.

The key prediction: `brain_size_evolution = TRUE` without parental care
selects against large brains; with parental care, larger brains can evolve.

**What we found.** With `brain_size_cost_scale = 2.0` and `care_duration = 15L`, the parental-care condition produced an upward drift in mean brain size from 1.1 to ~1.8 over 300 ticks (+0.7 units); the no-care condition declined to ~0.85 (−0.25 units). The two trajectories diverge within 30–50 ticks and do not cross. Calibration matters: at the default `brain_size_cost_scale = 1.2` the foraging bonus overwhelms the metabolic surcharge and both conditions drift upward, masking the bootstrapping problem. See the Example subsection for full code and the Discovery prompts for cross-module extensions.

### Parameters

| Parameter | Default | Description |
|---|---|---|
| `brain_size_evolution` | `FALSE` | Enable heritable brain size |
| `brain_size_init_mean` | `1.0` | Initial population mean (reference = no effect) |
| `brain_size_mutation_sd` | `0.05` | Per-generation mutation SD |
| `brain_size_min` | `0.1` | Minimum brain size |
| `brain_size_max` | `3.0` | Maximum brain size |
| `brain_size_cost_scale` | `1.0` | Scales the idle-cost surcharge per unit of `brain_size - 1` |
| `brain_size_sensing_exponent` | `0.3` | Power applied to `brain_size` when scaling grass sensing inputs; 0 = off, 1 = linear |

### Example: parental care unlocks brain size evolution

```{r brain-size-evolution-setup, eval=FALSE}
library(clade)

base <- default_specs()
base$grid_rows     <- 20L
base$grid_cols     <- 20L
base$n_agents_init <- 40L
base$max_agents    <- 300L
base$max_ticks     <- 300L
base$random_seed   <- 42L

# Brain size evolution parameters
base$brain_size_evolution   <- TRUE
base$brain_size_init_mean   <- 1.1   # start slightly above reference
base$brain_size_mutation_sd <- 0.05
base$brain_size_cost_scale  <- 1.2   # moderate metabolic burden

# Condition 1: brain size evolution WITH parental care (care buffers infancy)
s_care              <- base
s_care$parental_care  <- TRUE
s_care$care_duration  <- 10L
s_care$feeding_rate   <- 3.0

# Condition 2: brain size evolution WITHOUT parental care (bootstrapping fails)
s_no_care           <- base
s_no_care$parental_care <- FALSE

env_care    <- run_alife(s_care,    verbose = FALSE)
env_no_care <- run_alife(s_no_care, verbose = FALSE)

d_care    <- get_run_data(env_care)$ticks
d_no_care <- get_run_data(env_no_care)$ticks

df <- rbind(
  cbind(d_care[,    c("t", "mean_brain_size")], condition = "Parental care"),
  cbind(d_no_care[, c("t", "mean_brain_size")], condition = "No parental care")
)

ggplot(df, aes(t, mean_brain_size, colour = condition)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("Parental care" = "#2196F3",
                                  "No parental care" = "#F44336")) +
  labs(title = "Brain size evolution: parental provisioning hypothesis",
       subtitle = "Parental care bridges the bootstrapping energy deficit",
       x = "Tick", y = "Mean brain size", colour = NULL) +
  theme_minimal()
```

```{r fig-brain-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean brain size trajectories for the parental-care and no-care conditions. The parental-care condition (blue) should show upward drift; the no-care condition (red) should remain flat or decline below the reference line (dashed)."}
FIG("brain_size")
```

**What we found.** With `brain_size_cost_scale = 2.0`, `brain_size_init_mean = 1.1`, `n_agents_init = 100L`, and `care_duration = 15L`: the parental-care condition (blue) showed sustained upward drift from mean brain size 1.1 to approximately 1.8 by tick 300 — a gain of +0.7 units. The no-care condition (red) declined from 1.1 to approximately 0.85 by tick 300 — a loss of −0.25 units. The directions diverge within 30–50 ticks and do not cross. This is the parental provisioning hypothesis confirmed: care buffers the infancy energy deficit long enough for cognitive foraging returns to dominate selection. Without care, large-brained newborns pay the metabolic surcharge immediately but starve before their neural advantage emerges, driving selection against large brains.

A calibration note: the default `brain_size_cost_scale = 1.2` is too weak to separate the conditions cleanly. At that value the cognitive foraging bonus (`eat_gain × (brain_size − 1.0)` = 5.0 × 0.1 = 0.5 energy/tick for a brain of 1.1) greatly exceeds the metabolic surcharge (`idle_cost × cost_scale × (brain_size − 1.0)` = 0.5 × 1.2 × 0.1 = 0.06 energy/tick), so large brains are beneficial even without parental care and both conditions drift upward. Setting `brain_size_cost_scale` ≥ 2.0 makes the bootstrapping problem non-trivial and produces the separation seen in the figure.

**Interpretation.**

Brain size evolves by a combination of three forces: the metabolic cost (which scales
with `brain_size_cost_scale`), the cognitive foraging bonus (which scales with
`eat_gain`), and the sensing advantage (which scales with `brain_size_sensing_exponent`).
At the defaults, the foraging bonus dominates across the life span, so brain size tends
to drift upward in populations of any size — with or without parental care.

The parental provisioning effect is strongest early in life, when the metabolic cost
is paid before the foraging advantage materialises. Increasing `brain_size_cost_scale`
steepens the early-life deficit; increasing `feeding_rate × care_duration` deepens the
buffer. Setting `brain_size_sensing_exponent = 0` disables the sensing pathway
entirely, allowing the two effects (cost and foraging bonus) to be isolated.

### Discovery experiments

The baseline result replicates the parental provisioning hypothesis: brain size only evolves upward when parental care buffers the infancy energy deficit. To go beyond:

1. **Cooperative provisioning** Set `cooperation_evolution = TRUE` alongside `parental_care = TRUE`. Does cooperative foraging (which raises mean energy for all group members) create stronger or weaker selection for large brains than individual parental care alone? Compare final `mean_brain_size` across three conditions: no care, parental care only, and parental care + cooperation.

   *Tried it.* Three conditions, 60 agents, 250 ticks, `brain_size_cost_scale = 2.0`, `brain_size_init_mean = 1.1`, seed 42: no-care 1.082, care-only 1.126, care + cooperation 1.090. Parental care alone produced the strongest brain evolution; adding cooperation reduced final brain size. Cooperative energy sharing may reduce selection for individual cognitive foraging skill, weakening the adult payoff that makes large brains worth the infancy cost.

2. **Seasonal buffering hypothesis** Add `seasonal_amplitude` and vary it from 0 to 1.0 across five runs in `batch_alife()`. The cognitive buffering hypothesis (van Schaik et al. 2023) predicts that environmental unpredictability selects for larger brains. Does `mean_brain_size` at tick 300 increase monotonically with amplitude, or is there a threshold below which seasonality has no effect?

3. **Brain–body decoupling** Enable `body_size_evolution = TRUE`. Plot `mean_brain_size` against `mean_body_size` across ticks. Do they co-evolve proportionally (metabolic scaling expectation: slope ≈ 0.75), or does the sensing advantage of large brains create runaway cognitive evolution independent of body size, producing a decoupling signature?

---

