---
title: "Comparing brain architectures side-by-side"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Comparing brain architectures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Comparing brain architectures side-by-side

**What it models.** clade ships five working brain architectures
for evolving agents — Bayesian neural network (BNN, default),
multilayer perceptron (ANN), continuous-time recurrent network
(CTRNN), gene regulatory network (GRN), and a random-action null
baseline. Each carries a different inductive bias: BNN learns with
uncertainty, ANN is a pure feedforward policy, CTRNN retains
temporal state, GRN encodes a regulatory network, random is a
behavioural control. The canonical question is *how do they
compare when placed in the same ecology?*

Historically clade had no direct head-to-head benchmark. This
vignette is that benchmark: same grass-foraging ecology, same
population size, same seeds — only `brain_type` varies.

The setup exercises the 0.5.6 PSOCK-parallel `batch_alife()` path
at a non-trivial scale: 5 brain types × 5 seeds = **25 independent
clade simulations** dispatched across 25 R+Julia worker sessions,
summarised into a tidy table with `summarize_batch()`. See
`vignette("parameter-space-search")` for the surrounding tooling.

**Two unimplemented reserved names.** `"transformer"` and
`"synthesis"` are listed in `?default_specs` as reserved names only
— the Julia kernel errors with *"'transformer' and 'synthesis' are
reserved names, not implemented"*. This vignette covers the five
working types.

**Reproducible example.** The code below is the same
`dev/benchmarks/brain_comparison.R` runner used to produce the
figure, distilled to a readable form:

```{r brain-bench}
library(clade)
library(ggplot2); library(patchwork)

BRAIN_TYPES <- c("bnn", "ann", "ctrnn", "grn", "random")
SEEDS       <- c(1L, 7L, 13L, 19L, 25L)

base <- default_specs()
base$n_agents_init <- 80L
base$max_agents    <- 400L
base$grid_rows     <- 30L
base$grid_cols     <- 30L
base$grass_rate    <- 0.15
base$max_ticks     <- 500L

# grid_specs() expands the full 5 × 5 factorial; batch_alife()
# fans out across 25 PSOCK workers (one R + Julia session each);
# summarize_batch() pulls the per-run statistics into a tibble.
specs_list <- grid_specs(base,
                         brain_type  = BRAIN_TYPES,
                         random_seed = SEEDS)
results  <- batch_alife(specs_list, n_cores = length(specs_list))
summary  <- summarize_batch(results, specs_list,
                            param_names = c("brain_type", "random_seed"))
```

```{r fig-brain-comparison, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Brain-type benchmark (5 architectures × 5 seeds, 500 ticks each, PSOCK-parallel via batch_alife). Top row: population trajectories per brain type; thin lines are individual seeds, bold line is seed-mean. Bottom row: mean-energy trajectories. BNN runs up a large population at low per-agent energy; GRN / CTRNN maintain small populations at high per-agent energy. random baseline is the most fragile."}
FIG("brain_comparison")
```

**What we found (2026-04-17 benchmark).** Summary statistics over
the last-tick state, averaged across 5 seeds:

| brain_type | n_final   | mean_energy | diversity | viability (5 seeds) |
|---|---|---|---|---|
| **bnn** | **196 ± 5** | 125 ± 2 | 0.45 ± 0.01 | 5/5 viable |
| ann     | 87 ± 26 | 154 ± 3 | 0.40 ± 0.01 | 5/5 viable |
| ctrnn   | 70 ± 5  | 160 ± 8 | 0.47 ± 0.02 | 5/5 viable |
| grn     | 56 ± 8  | **170 ± 5** | 0.44 ± 0.01 | 5/5 viable |
| random  | 36 ± 14 | 165 ± 6 | 0.43 ± 0.04 | 2/5 viable, 1/5 crashed |

Three things are worth noting.

1. **BNN trades per-capita quality for density.** BNN maintains a
   ~3× larger population than any other type but at the lowest
   per-agent energy (125 vs 154–170). The Thompson-sampled
   weights let BNN agents *explore* more aggressively; they forage
   less efficiently per individual but populations grow faster and
   crash less. Zero crashes across 5 seeds.
2. **GRN / CTRNN trade density for quality.** The recurrent and
   gene-regulatory architectures produce the smallest populations
   (56, 70) at the highest per-agent energy (170, 160). Agents
   behave more deterministically, extracting more per tick but
   sustaining fewer individuals. Also zero crashes.
3. **ANN sits between**, as expected — it's a pure feedforward
   policy without BNN's exploration or CTRNN's memory.
4. **random is the least robust.** One of five seeds crashes to
   extinction; the others sit at ~36 agents with moderately good
   mean energy. Without evolved behaviour, survival is
   seed-sensitive.

None of these results say one architecture is *better* — they
partition the r-vs-K axis of life-history strategy in a way that
maps cleanly onto the canonical cognitive-style trade-off. The
benchmark itself is reproducible by running
`dev/benchmarks/brain_comparison.R` from a fresh clade install.

### Discovery experiments

The above benchmark is a single ecology. Fuller characterisation
would look at:

1. **Under scarcity.** Halve `grass_rate` to 0.075. Does BNN still
   dominate the density axis, or do the high-quality per-agent
   strategies (GRN, CTRNN) catch up when marginal foragers starve?
   *Not yet tried — good first follow-up.*
2. **Under predation.** Add `n_predators_init = 10`. Does CTRNN's
   temporal memory pay off for escape sequences? *Not yet tried.*
3. **Under seasonality.** `seasonal_amplitude = 0.7, season_length
   = 50`. Plastic architectures (BNN via sigma, ANN none) should
   diverge. *Not yet tried.*
4. **In mixed populations.** Would require a kernel change —
   currently `brain_type` is a single spec per simulation. Worth
   opening as an issue; it would make the classic
   Boyd-Richerson-style architecture tournament possible.

---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
