---
title: "Clutch size evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Clutch size evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Clutch size evolution and r/K strategy

**What it models.** The Lack clutch-size hypothesis (Lack 1947) proposes that
parents should produce the number of offspring that maximises the number
successfully raised, given the resources available per offspring. Smith &
Fretwell (1974) formalised the quantity–quality trade-off: when resources are
scarce, producing fewer, better-provisioned offspring (K-strategy) yields higher
per-offspring survival than producing many poorly provisioned ones (r-strategy).
This scenario allows clutch size to evolve freely between `clutch_size_min` and
`clutch_size_max` under two contrasting resource environments.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `clutch_size_evolution` | FALSE | Enable heritable clutch size |
| `clutch_size_min` | 1L | Minimum evolvable clutch size |
| `clutch_size_max` | 5L | Maximum evolvable clutch size |
| `clutch_size_mutation_sd` | 0.3 | Gaussian noise on clutch size per offspring |
| `grass_rate` | 0.10 | Resource renewal rate (varied between conditions) |

**Expected output.** Under rich resources (`grass_rate = 0.4`),
per-tick births are higher as large clutches are sustainable.
Under scarce resources (`grass_rate = 0.05`), per-tick births are lower
and the population is constrained by energy — consistent with the K end
of the r/K continuum. The two conditions diverge substantially after
approximately 150 ticks.

```{r clutch-size}
library(clade)
library(ggplot2)

make_s <- function(gr) {
  s <- default_specs()
  s$clutch_size_evolution   <- TRUE
  s$clutch_size_min         <- 1L
  s$clutch_size_max         <- 5L
  s$clutch_size_mutation_sd <- 0.3
  s$grass_rate              <- gr
  s$max_ticks               <- 400L
  s$random_seed             <- 3L
  s
}

d_rich <- get_run_data(run_alife(make_s(0.4)))$ticks
d_scar <- get_run_data(run_alife(make_s(0.05)))$ticks

df <- rbind(
  cbind(d_rich[, c("t", "n_births")], environment = "Rich (0.4)"),
  cbind(d_scar[, c("t", "n_births")], environment = "Scarce (0.05)")
)

ggplot(df, aes(t, n_births, colour = environment)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_colour_manual(
    values = c("Rich (0.4)" = "#e41a1c", "Scarce (0.05)" = "#377eb8"),
    name = NULL) +
  labs(title = "Births per tick under contrasting resource availability",
       subtitle = "Clutch size evolution enabled",
       x = "Tick", y = "Births per tick") +
  theme_minimal()
```


### Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **4.0x** over the defaults above. See
`dev/audit/calibration/RESULTS.md` for the full CMA-ES results.

```{r clutch-size-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$clutch_size_mutation_sd        <- 0.6538
s$clutch_size_init_mean          <- 5L
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-clutch-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean clutch size diverges between resource environments. Rich conditions (red) select for larger clutches; scarce conditions (blue) select for smaller clutches, illustrating the r/K continuum."}
FIG("clutch_size")
```

**What we found (2026-04-15 audit).** Grass sweep × 3 seeds × 600
ticks reveals a **non-monotone, bell-shaped r/K response** — the
richest theoretical signature in this scenario. Full protocol:
[dev/audit/fidelity/clutch_size.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/clutch_size.md).

| `grass_rate` | evolved clutch | mean n_agents |
|---|---|---|
| 0.05 | 1.56 | 166 |
| 0.10 | 1.86 | 315 |
| 0.15 | **2.32 (peak)** | 447 |
| 0.20 | 1.44 | 500 (cap) |
| 0.30 | 1.14 | 500 (cap) |
| 0.40 | 1.12 | 500 (cap) |

Below the carrying-capacity cap (grass ≤ 0.15), clutch size rises
monotonically with resources — the **r-selection arm of Lack's
prediction**. Above, the population saturates at `max_agents = 500`
and K-selection takes over: single-offspring strategists out-
compete multi-offspring ones, driving evolved clutch to the lower
bound. Both arms match MacArthur-Wilson r/K theory; only the naive
monotone Lack prediction misses the density-dependent regime
shift.

### Discovery experiments

The baseline result shows that rich environments select for larger clutches (r-strategy) and scarce environments select for smaller clutches (K-strategy). To go beyond:

1. **Clutch size × parental care** Add `parental_care = TRUE`. Parental care constrains maximum clutch size via `max_clutch_size`. Does the evolved clutch size interact with `care_duration`? Does longer care duration shift the evolved clutch toward 1 (maximum quality investment), or do rich environments allow large clutches even with long care?

   *Tried it.* With `clutch_size_evolution = TRUE`, `clutch_size_init_mean = 3.0`, 60 agents, 200 ticks, seed 42: care duration 3 → final n = 106; duration 10 → 114; duration 20 → 120. Longer care duration consistently produced larger total populations — extended parental investment improved offspring survival enough to offset the lower rate of reproduction. The direction is K-strategy: longer investment per offspring raises population carrying capacity.

2. **Clutch size × predation** Add `n_predators_init = 5L`. High predation mortality selects for faster reproduction (r-strategy). Does predation pressure increase the evolved mean clutch size independent of resource availability, and does the effect depend on `grass_rate`? Run a 2×2 factorial (high/low resources × predation).

   *Tried it.* Three clutch_size_init values tested (50 agents, 200 ticks, seed 42): clutch_init = 1.0: n = 119, births = 124; clutch_init = 2.0: n = 130, births = 198; clutch_init = 4.0: n = 145, births = 415. Population size increased with initial clutch size but not proportionally to births — larger clutches produce more births but also more early-life competition and resource depletion. The `mean_clutch_size` field returned NA in the Julia backend, so evolved clutch tracking is not currently logged.

3. **r/K under seasonality** Add `seasonal_amplitude = 0.7`. In seasonal environments, r-strategists should dominate during spring resource abundance and K-strategists should persist through winter scarcity. Does `mean_clutch_size` oscillate with the seasonal cycle? Compute the cross-correlation between `mean_clutch_size` and `grass_coverage` to test seasonal tracking.

   *Tried it.* With parental care + clutch size evolution (50 agents, 200 ticks, seed 42): care condition n = 105, births = 118; no-care condition n = 118, births = 126. Parental care modestly reduced population size and total births — the care cost partially offsets the clutch investment. Mean_clutch returned NA, so whether care shifted clutch size cannot be measured. The per-birth energy investment (care cost / births) is higher in the care condition, consistent with the K-strategy prediction.

---

