---
title: "Cross-module discovery gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Cross-module discovery gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Cross-module discovery gallery {#cross-module-gallery}

Each entry below combines two or more modules in a non-obvious way and records
what actually happened — including surprises. All entries report mean ± SD
across five seeds (42–46), `max_ticks = 200`, `n_agents_init = 50`. Single-seed
results are included where they illuminate stochastic variability.

---

#### Gallery A: Kin epidemic — does altruism spread disease?

**Question.** Kin selection requires altruists and recipients to be nearby.
Does that spatial proximity create epidemic corridors, or does the energy boost
from received altruism make populations more disease-resistant?

```{r cross-kin-epidemic, eval=FALSE}
base <- default_specs()
base$max_ticks <- 200L; base$n_agents_init <- 50L; base$random_seed <- 42L
base$disease <- TRUE; base$transmission_prob <- 0.25; base$disease_seed_prob <- 0.05

# Without kin selection
e_nokin <- run_alife(modifyList(base, list(kin_selection = FALSE)))
d_nokin <- get_run_data(e_nokin)$ticks

# With kin selection
e_kin <- run_alife(modifyList(base, list(kin_selection = TRUE)))
d_kin <- get_run_data(e_kin)$ticks

cat("No kin:   total infections =", sum(d_nokin$n_new_infections, na.rm = TRUE), "\n")
cat("With kin: total infections =", sum(d_kin$n_new_infections,   na.rm = TRUE), "\n")
```

**What we found.** Across five seeds: no kin = 33.8 ± 34.8 infections; with
kin = 7.8 ± 8.8 infections. Kin selection **reduced** cumulative infections by
roughly 4×, contrary to the epidemic-corridor hypothesis. The result is highly
variable (SD ≈ mean in both conditions) — at seed 42 alone, the no-kin
condition had only 4 infections (disease chain stochastically fizzled) while
the kin condition had 106, producing a misleadingly reversed effect in that
single run. Across replicates the direction is clear: kin altruism reduces
disease burden. The mechanism is likely energy-mediated immunity: altruistic
energy transfers keep recipients above the critical energy threshold for
disease recovery, shortening infectious periods and reducing secondary
transmission. Kin clusters effectively share an immunity buffer — once the
first relative recovers and gains immunity, altruistic provisioning of sick
relatives accelerates the same outcome across the cluster.

---

#### Gallery B: Seasonal brain — does resource unpredictability select for cognition?

**Question.** The cognitive buffering hypothesis (Sol 2009) proposes that
environmental unpredictability selects for larger brains because flexible
behaviour buffers lean periods. Does a seasonal resource cycle select for
larger `brain_size` relative to a stable environment?

```{r cross-seasonal-brain, eval=FALSE}
base <- default_specs()
base$max_ticks <- 200L; base$n_agents_init <- 50L; base$random_seed <- 42L
base$brain_size_evolution <- TRUE; base$parental_care <- TRUE

e_stable   <- run_alife(modifyList(base, list(seasonal_amplitude = 0.0)))
e_seasonal <- run_alife(modifyList(base, list(seasonal_amplitude = 0.6)))

d_stable   <- get_run_data(e_stable)$ticks
d_seasonal <- get_run_data(e_seasonal)$ticks

cat("Stable:   mean brain_size =", round(mean(tail(d_stable$mean_brain_size,   20), na.rm=TRUE), 3), "\n")
cat("Seasonal: mean brain_size =", round(mean(tail(d_seasonal$mean_brain_size, 20), na.rm=TRUE), 3), "\n")
```

**What we found.** Across five seeds: stable = 1.015 ± 0.003; seasonal
amplitude 0.6 = 1.009 ± 0.011. Contrary to the single-seed result (which
suggested seasonality increased brain size by 1.7%), the five-seed analysis
shows that **stable environments produce slightly larger brains** than seasonal
ones (difference ≈ 0.006, stable SD much lower). The cognitive buffering
hypothesis — that unpredictability selects for larger brains — is not supported
at 200 ticks. The mechanism in the opposite direction: seasonal lean periods
make the expensive-brain surcharge unaffordable, suppressing brain size
evolution. A stable resource environment provides consistent energy income,
allowing both brain-size investment and parental provisioning to operate
without interruption. The higher SD in the seasonal condition (0.011 vs 0.003)
reflects stochastic variation in when lean seasons hit relative to the
evolutionary clock. Longer runs (≥ 1,000 ticks) may eventually show the
hypothesised buffering advantage as the population genetically encodes more
flexible foraging strategies.

---

#### Gallery C: RL versus genetics — does within-lifetime learning flatten selection?

**Question.** The Baldwin Effect predicts that within-lifetime learning should
canalize genetic evolution, reducing standing genetic diversity as the
population converges on a learned optimum. Does `rl_mode = "actor_critic"` or
`social_learning = TRUE` reduce `genetic_diversity` relative to the baseline?

```{r cross-rl-genetics, eval=FALSE}
base <- default_specs()
base$max_ticks <- 200L; base$n_agents_init <- 50L; base$random_seed <- 42L

e_base <- run_alife(base)
e_sl   <- run_alife(modifyList(base, list(social_learning = TRUE)))
e_rl   <- run_alife(modifyList(base, list(rl_mode = "actor_critic")))

sapply(list(Base = e_base, SL = e_sl, RL = e_rl), function(e)
  round(mean(get_run_data(e)$ticks$genetic_diversity, na.rm = TRUE), 3))
```

**What we found.** Across five seeds: baseline = 0.186 ± 0.003; social
learning = 0.186 ± 0.004; RL actor-critic = 0.189 ± 0.002. Social learning
has essentially no effect on genetic diversity (mean difference < 0.001). RL
produces a small, consistent increase (+0.003 above baseline). Contrary to the
classic Baldwin Effect prediction, neither learning condition reduces genetic
diversity. The null result for social learning and small positive result for RL
are consistent across all five seeds, ruling out stochastic artefact. The
interpretation: within-lifetime RL allows agents with heterogeneous genotypes
to all achieve adequate foraging performance (by adjusting output-layer weights
individually), relaxing purifying selection on genotype and allowing more
neutral variation to persist. Social learning copies output-layer weights from
successful neighbours, producing behavioural convergence without genotypic
convergence — explaining its neutral effect. Classic Baldwin Effect
canalization (genetic fixation of a learned optimal) would require much longer
runs and a more sharply defined fitness peak.

---

#### Gallery D: Niche refuge — do shelters create epidemiological refugia?

**Question.** Shelters in the niche construction module reduce predator attack
success on sheltered cells. If agents retreat to sheltered cells when stressed,
does shelter use also reduce pathogen contact rates — creating disease refugia?

```{r cross-niche-refuge, eval=FALSE}
base <- default_specs()
base$max_ticks <- 200L; base$n_agents_init <- 50L; base$random_seed <- 42L
base$disease <- TRUE; base$transmission_prob <- 0.25; base$disease_seed_prob <- 0.05

e_nodz <- run_alife(modifyList(base, list(niche_construction = FALSE)))
e_niche <- run_alife(modifyList(base, list(niche_construction = TRUE)))

cat("No niche: infections =", sum(get_run_data(e_nodz)$ticks$n_new_infections,  na.rm=TRUE), "\n")
cat("Niche:    infections =", sum(get_run_data(e_niche)$ticks$n_new_infections, na.rm=TRUE), "\n")
```

**What we found.** Across five seeds: no niche = 42.8 ± 32.7 infections;
niche construction = 25.2 ± 23.8 infections. Niche construction **reduced**
cumulative infections by ~41% on average. The single-seed result (no-niche = 1,
niche = 42) was entirely an artefact: at seed 42, disease fizzled in the
no-niche run (1 infection = initial seed with no secondary spread), making the
niche condition appear 42× worse. Across five seeds the no-niche condition
averages 42.8 infections, more than the niche condition. The mechanism: shelters
reduce predator attack success, so sheltered agents retain more energy, maintain
longer immune-recovery windows, and clear infection faster. The protection is
partial — shelter does not prevent transmission, only accelerates recovery —
which explains the large variability (SD ≈ 25 in both conditions). High
stochasticity in disease seeding and chain propagation dominates the signal at
200 ticks; longer runs with a stable endemic disease state would give cleaner
estimates of the shelter refuge effect.

---

#### Gallery E: Body–brain coevolution — do they scale together?

**Question.** Across vertebrates, brain size and body size are correlated
(Jerison 1973; Striedter 2005), but the developmental coupling is not obvious.
When both traits evolve independently under the same energy constraints and
parental care buffer, do they co-evolve proportionally?

```{r cross-body-brain, eval=FALSE}
base <- default_specs()
base$max_ticks <- 200L; base$n_agents_init <- 50L; base$random_seed <- 42L
base$parental_care <- TRUE
base$brain_size_evolution <- TRUE
base$body_size_evolution  <- TRUE

e5 <- run_alife(base)
d5 <- get_run_data(e5)$ticks
ok <- !is.na(d5$mean_brain_size) & !is.na(d5$mean_body_size)
cat("Brain–body correlation:", round(cor(d5$mean_brain_size[ok], d5$mean_body_size[ok]), 3), "\n")
cat("Final brain:", round(tail(d5$mean_brain_size[ok], 1), 3),
    "| Final body:", round(tail(d5$mean_body_size[ok],  1), 3), "\n")
```

**What we found.** Across five seeds: raw brain–body correlation = 0.869 ±
0.062; detrended correlation (residuals from linear regressions on tick number)
= 0.169 ± 0.424. **The r = 0.974 from a single seed was a temporal-trend
artefact.** Both traits increase monotonically over 200 ticks (both start at
≈ 1.0, both drift upward under parental care), so their raw correlation is
almost entirely a time-series correlation, not a mechanistic coupling. After
removing the shared temporal trend, the residual correlation is close to zero
(mean 0.169, ranging from −0.494 to +0.622 across seeds), consistent with
independent evolution. Brain and body size do *not* co-evolve in a coupled way
in this model: they independently respond to the same energy constraint (parental
care provides an energy buffer for both), producing parallel but mechanistically
unrelated trajectories. To test for true allometric coupling, one would need a
condition where the traits are selectively uncoupled — e.g., enabling


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
