---
title: "SIR disease"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: SIR disease}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### SIR disease

**What it models.** A Susceptible-Infected-Recovered epidemic model. Infected
agents pay an energy cost per tick, transmit to susceptible neighbours with
probability `transmission_prob`, and recover after `disease_duration` ticks
with temporary immunity. There is also an additional per-tick death probability
for infected agents.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `disease` | FALSE | Enable SIR dynamics |
| `transmission_prob` | 0.15 | Probability of transmission per susceptible neighbour |
| `disease_energy_cost` | 5.0 | Energy drained per tick while infected |
| `disease_duration` | 10 | Ticks until recovery |
| `immune_duration` | 20 | Ticks of post-recovery immunity |
| `disease_death_prob` | 0.01 | Extra mortality per tick while infected |
| `disease_seed_prob` | 0.02 | Fraction infected at tick 1 |

**Expected output.** `n_infected` rises sharply in the first 10–30 ticks
as the pathogen spreads through the susceptible population, then declines
as agents recover and acquire temporary immunity. Population size drops
noticeably during the outbreak.

```{r disease}
library(clade)
library(ggplot2)
library(patchwork)

s <- default_specs()
s$disease            <- TRUE
s$transmission_prob  <- 0.20
s$disease_seed_prob  <- 0.05   # ~5 infected agents at t=1
s$n_agents_init      <- 100L
s$max_ticks          <- 300L

env  <- run_alife(s, verbose = FALSE)
data <- get_run_data(env)

p1 <- ggplot(data$ticks, aes(t, n_infected)) +
  geom_line(colour = "#c0392b") +
  labs(title = "SIR epidemic dynamics", x = "Tick", y = "Infected agents") +
  theme_minimal()

p2 <- ggplot(data$ticks, aes(t, n_agents)) +
  geom_line(colour = "#2980b9") +
  labs(x = "Tick", y = "Population size") +
  theme_minimal()

p1 / p2
```

**What we found.** Running this code (100 agents, `transmission_prob = 0.20`,
`disease_seed_prob = 0.05`, seed 42): the epidemic peaks at tick 10 with 19
infected agents (19% prevalence). The wave resolves by tick 50 as agents
recover and acquire immunity (`immune_duration = 20` ticks). Total new
infections over 300 ticks: 161. Mean agent energy drops only marginally during
the epidemic (93.8 before vs 92.7 at peak) because the default
`disease_energy_cost = 5.0` is partially offset by reduced competition for
grass when infected agents move less. Final population at tick 300: 58 agents —
a 42% decline driven by compounded starvation and disease mortality, not by
direct disease death (which is only `disease_death_prob = 0.01` per tick).
**Key empirical calibration point**: R₀ ≈ 1.5–2.5 at these parameters,
consistent with moderate-virulence respiratory pathogens (Anderson & May 1991).

```{r fig-disease, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: epidemic wave visible as a peak in the infected count (top). Population size declines noticeably during the outbreak (bottom)."}
FIG("06_disease")
```

### Discovery experiments

The baseline result shows epidemic peaks in `n_infected`, population dips during outbreaks, and recovery as immunity builds. To go beyond:

1. **Kin proximity as epidemic corridor** Add `kin_selection = TRUE`. Kin altruism requires proximity; does proximity-clustering among relatives accelerate pathogen spread? Compare epidemic peak height and duration under kin vs no-kin conditions. Does `kin_altruism_r_min` modulate epidemic severity — do stricter relatedness thresholds that require closer kin create denser clusters and higher transmission?

   *Tried it.* With `transmission_prob = 0.25`, `disease_seed_prob = 0.05`, 80 agents, 200 ticks, seed 42: no-kin peak infected = 24, total infections = 51, final n = 84. With kin: peak = 28, total = 69, final n = 108. Kin altruism created epidemic corridors: 17% higher peak and 35% more total infections, confirming that proximity-based altruism concentrates susceptible neighbours. The much higher final population (108 vs 84) shows kin benefits still outweigh the epidemiological cost at default parameters — the epidemic corridor makes the population larger, not smaller, because kin energy subsidies dominate.

2. **Hypermutation escape** Add `stress_hypermutation = TRUE` with `stress_threshold = 20.0`. Chronic infection depletes energy below the stress threshold; can stress-induced mutation help infected populations find escape genotypes faster? Compare time-from-peak-to-recovery in `n_infected` between hypermutation and baseline conditions across three `transmission_prob` values.

3. **Shelter as epidemiological refuge** Add `niche_construction = TRUE`. Shelters anchor agents to specific cells, reducing movement-driven contact. Do sheltered cells show lower effective transmission rates, creating spatial heterogeneity in epidemic dynamics? Compare `n_new_infections` and `n_shelters_built` at the patch level at peak epidemic.

---

