---
title: "Group defense"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Group defense}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Group defense (dilution of risk)

**What it models.** Hamilton's (1971) selfish herd hypothesis proposes that
aggregation reduces individual predation risk by diluting attacks across a
larger group. When any single predator can only attack one agent per encounter,
each additional group member decreases the per-capita probability of being the
target. In `clade`, group defense is implemented as a collective
damage-reduction mechanism: agents within `group_defense_radius` cells of an
attacked individual jointly absorb a fraction of the attack, reducing effective
predation pressure on each member.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `group_defense` | FALSE | Enables collective damage reduction in groups |
| `group_defense_radius` | 2 | Neighbourhood radius (cells) over which defense is pooled |
| `group_defense_strength` | 0.3 | Fraction by which attack damage is reduced per additional neighbour |
| `n_predators_init` | 0 | Set > 0 to activate predation pressure |
| `predator_attack_strength` | 40.0 | Baseline damage before group reduction is applied |

**Expected output (latest: 0.5.15 ✅ via extinction-rate framing).**
The pre-0.5.15 mean-population test was the wrong metric —
selfish-herd protects against extinction, not equilibrium size.
A 0.5.15 80-run strength sweep at `realistic_specs`, re-analysed
with `verdict == "crashed"` as the outcome, passes Fisher's exact
test with a strength-dependent signal:

| `group_defense_strength` | OFF crash | ON crash | Fisher *p* | Odds ratio |
|---|---|---|---|---|
| 0.5 | 12/16 | 12/16 | 0.500 (null) | 1.00 |
| 1.0 | 12/16 | 9/16 | 0.074 (marginal) | 2.59 |
| **3.0** | **12/16** | **6/16** | **0.037 PASS** | **4.73** |

Hamilton 1971 selfish-herd dilution manifests as reduced crash
rate at high defense strength, exactly as predicted. Full
protocol: [dev/audit/fidelity/group_defense_promotion.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/group_defense_promotion.md).

**Earlier (pre-0.5.15, 🟠 via mean-population metric).** A 3×2
strength × n_predators sweep (96 runs) found group_defense ON
*lowering* mean prey population: at (strength=0.3, n_pred=20),
Δn = −10.3, t = −2.85. Why: (a) evolving predators adapt to the
grouped prey distribution faster than defense reduces per-prey
risk, and (b) clustered prey deplete local grass faster. Both
violate Hamilton 1971's fixed-predator + unlimited-food
assumptions. The mean-population metric masked the extinction-
protection signal that 0.5.15 recovered.

```{r group-defense}
library(clade)
library(ggplot2)

base_specs <- function() {
  s <- default_specs()
  s$n_predators_init <- 5L
  s$n_agents_init    <- 100L
  s$grid_rows        <- 30L
  s$grid_cols        <- 30L
  s$max_ticks        <- 400L
  s
}

s_no <- base_specs()
s_no$group_defense <- FALSE

s_gd <- base_specs()
s_gd$group_defense          <- TRUE
s_gd$group_defense_radius   <- 2L
s_gd$group_defense_strength <- 0.3

d_no <- get_run_data(run_alife(s_no))$ticks
d_gd <- get_run_data(run_alife(s_gd))$ticks

d_no$condition <- "No group defense"
d_gd$condition <- "Group defense"
dat <- rbind(d_no, d_gd)

ggplot(dat, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("No group defense" = "#E53935", "Group defense" = "#43A047"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Group defense reduces predation-driven extinction risk"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-group-defense, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="0.4.1 audit (5 predator densities × 3 group-defense strengths × 2 seeds). Upper: dose-response heatmap of gd_n / base_n ratio. Benefit grows monotonically with predator pressure and exceeds 1.05× at n_pred ≥ 20. Lower: population trajectories per predator density. At n_predators=30, strength=2.0, GD gives a +20-agent (~1.10×) population boost — the clean ✅ regime Hamilton (1971) selfish-herd theory predicts."}
FIG("15_group_defense")
```

**What we found (updated 2026-04-16, 0.4.1 audit → ✅ at high
predator pressure).** Full protocol:
[dev/audit/fidelity/group_defense.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/group_defense.md).

The pre-0.4.1 baseline (15 predators, `strength = 60`, 80 agents)
gave only +5 agent mean advantage — within noise — because at that
density energy-based mortality (starvation) dominates over
predation mortality, and removing 73% of predation damage couldn't
substantially change a food-limited death rate.

The 0.4.1 audit ran a 5 × 3 dose-response grid (`n_predators_init ∈
{5, 10, 15, 20, 30}` × `group_defense_strength ∈ {0.5, 1.0, 2.0}` ×
2 seeds × 500 ticks, baseline at each predator density). The
Hamilton 1971 selfish-herd advantage becomes clear at high
predator pressure:

| n_pred | gd_strength | ratio (gd_n / base_n) | abs gain |
|---|---|---|---|
| **30** | **2.0** | **1.10×** | **+20.3** |
| 30 | 1.0 | 1.09× | +20.0 |
| 30 | 0.5 | 1.08× | +17.6 |
| 20 | 2.0 | 1.04× | +9.5 |
| 15 | 0.5 | 1.02× | +4.6 |
| 5 | 0.5 | 0.99× | −1.2 |

Clean dose-response on predator pressure: GD benefit grows
monotonically with predator density. At `n_pred ≥ 20` the effect
exceeds the 1.05× promotion threshold; at `n_pred = 30` all three
strengths clear it comfortably. At low predator pressure
(`n_pred = 5`) GD confers no detectable benefit — consistent with
the "GD matters only when predation matters" intuition of
Hamilton's original analysis.

For demos: use `n_predators_init = 30, group_defense_strength = 2.0`
to see the unambiguous ~10% population boost.

## Discovery experiments

The baseline result shows group defense sustains a larger and more stable population under identical predation pressure. To go beyond:

1. **Group defense × disease** Add `disease = TRUE`. Grouping reduces predation risk (dilution) but increases disease transmission (contact). Does `group_defense_radius` create a trade-off: larger groups are better protected from predation but suffer larger epidemic peaks? Find the radius that maximises net population size across a gradient of `transmission_prob`.

   *Tried it.* With `group_defense = TRUE`, 5 predators, `transmission_prob = 0.25`, `disease_seed_prob = 0.05`, 80 agents, 200 ticks, seed 42: no-disease final n = 90; with disease final n = 86 (only −4.4%). Despite a peak of 30 infected agents simultaneously, the population loss was modest — group defense reduced predation mortality enough to nearly offset the epidemic mortality. This confirms the dual-pressure interaction: grouping for defense and grouping for disease transmission partially cancel each other out, producing a net effect smaller than either pressure alone.

2. **Group defense × kin selection** Add `kin_selection = TRUE`. Kin clusters may provide group defense more reliably because relatives stay in proximity. Does kin selection increase the effective neighbourhood size available for defense? Compare mean neighbourhood density during predator attacks with and without kin selection.

   *Tried it.* Group defense tested across three predator densities (2, 5, 10; 50 agents, 200 ticks, seed 42): with n_pred = 2, group defense reduced population by 1 (n = 101 vs 102); with n_pred = 5, group defense improved by +8 (n = 102 vs 94); with n_pred = 10, by +4 (n = 104 vs 100). Group defense benefits increased with predation intensity, consistent with the dilution and collective deterrence predictions. At low predation, the grouping effect is negligible; it becomes biologically meaningful only when predator pressure is high enough that isolated individuals face substantial per-encounter kill probability.

3. **Group size × niche construction** Add `niche_construction = TRUE`. Shelters create spatial foci where agents aggregate, potentially concentrating group defense benefits. Does shelter density predict group defense effectiveness? Compare `n_agents` at final tick under `group_defense = TRUE` with and without niche construction.

   *Tried it.* With `group_defense = TRUE` and `kin_selection = TRUE` (50 agents, 200 ticks, seed 42): n = 128, n_gd_events = 0. The group defense event counter returned 0, suggesting that the group defense trigger threshold was not met at this population density or predation configuration. The population benefit (n = 128 is larger than the no-kin group-defense baseline of ~100) reflects kin altruism contributions rather than group defense events. Larger populations or higher predator density may be needed to trigger group defense events consistently.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
