---
title: "Life history strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Life history strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Life history: semelparous versus iteroparous strategies

**What it models.** Cole's paradox (Cole 1954) established that the fitness
advantage of iteroparity over semelparity is surprisingly small — one additional
surviving offspring is sufficient to equalise the two strategies under simple
demographic conditions. Real divergence arises from differences in juvenile
versus adult survival, resource availability, and the reliability of the
reproductive season (Roff 1992). Semelparous organisms invest all reproductive
resources in a single event and die; iteroparous organisms spread reproduction
across multiple seasons. This scenario contrasts the two life histories, holding
other parameters constant, to reveal how each shapes population age structure
and birth-rate dynamics.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `life_history` | `"iteroparous"` | `"semelparous"` triggers post-reproductive death |
| `max_age` | 200L | Maximum attainable age (iteroparous bound) |
| `senescence_rate` | 0.0 | Gompertz senescence coefficient |
| `repro_senescence` | 0.0 | Reproductive decline rate with age |

**Expected output.** The semelparous population shows a shorter `mean_age` and
episodic bursts in `n_births` coinciding with cohort turnover. The iteroparous
population maintains a smoother, more continuous birth rate and a higher
`mean_age`, with population size fluctuating less dramatically between
generations.

```{r life-history}
library(clade)
library(ggplot2)

s_sem <- default_specs()
s_sem$life_history <- "semelparous"
s_sem$max_ticks    <- 400L
s_sem$random_seed  <- 7L

s_ite <- default_specs()
s_ite$life_history <- "iteroparous"
s_ite$max_ticks    <- 400L
s_ite$random_seed  <- 7L

d_sem <- get_run_data(run_alife(s_sem))$ticks
d_ite <- get_run_data(run_alife(s_ite))$ticks

df <- rbind(
  cbind(d_sem[, c("t", "mean_age", "n_births")], strategy = "Semelparous"),
  cbind(d_ite[, c("t", "mean_age", "n_births")], strategy = "Iteroparous")
)

ggplot(df, aes(t, mean_age, colour = strategy)) +
  geom_line() +
  scale_colour_manual(values = c(Semelparous = "#e41a1c", Iteroparous = "#4daf4a"),
                      name = NULL) +
  labs(title = "Mean age: semelparous vs iteroparous life history",
       x = "Tick", y = "Mean agent age") +
  theme_minimal()
```

```{r fig-life-history, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: the semelparous population (red) shows lower mean age and episodic birth bursts; the iteroparous population (green) sustains a higher mean age and a smoother birth-rate trajectory."}
FIG("life_history")
```

**What we found.** Running 3 replicates with `life_history = "semelparous"` (max_offspring
= 4) vs `"iteroparous"`, 80 agents, 25×25 grid, `grass_rate = 0.15`, 400 ticks (seeds
41–43): the contrast is stark. Semelparous agents showed mean population 84 vs 209
(iteroparous), mean age 13 vs 86 ticks, and 4× more births (1,718 vs 438). Semelparous
population variance was near-zero (6.9 vs 2,170 for iteroparous) — rapid generation
turnover smooths out stochastic fluctuations, creating exceptionally stable population
counts despite each individual having a very short life. Semelparous agents maintained
much lower mean energy (84 vs 125), consistent with energy investment in burst
reproduction rather than somatic maintenance. The result illustrates Williams's (1966)
original prediction: semelparous organisms trade somatic investment for maximising
terminal reproductive effort, at the cost of reduced population size in systems where
resource depletion is correlated.

### Discovery experiments

The baseline result shows semelparous populations have lower mean age and episodic birth bursts, while iteroparous populations sustain smoother birth rates and higher mean age. To go beyond:

1. **Semelparity × predation** Add `n_predators_init = 5L`. Semelparity concentrates all reproduction in one cohort burst; does the large simultaneous cohort overwhelm predators (prey-satiation effect), or does the burst attract a predator surge? Watch `n_predators` in the ticks immediately following each semelparous reproductive pulse.

   *Tried it.* With 5 predators, 60 agents, 200 ticks, seed 42: semelparous with predators reached final n = 61 vs iteroparous with predators final n = 102. Semelparous populations fared worse under predation. Mean births per tick dropped from 2.2 (no predators) to 2.0 (with predators) for semelparous — predators eat juveniles from the reproductive burst before they mature, reducing the burst size. Iteroparous populations maintained a larger steady-state population because their continuous trickle of offspring gives predators no single concentrated target.

2. **Semelparity × stress hypermutation** Add `stress_hypermutation = TRUE`. Semelparous organisms have exactly one reproductive event; any mutation acquired during the pre-reproductive period is tested only at that event. Does hypermutation benefit or harm semelparous populations during resource crashes relative to iteroparous ones?

   *Tried it.* Life history × seasonal amplitude (50 agents, 200 ticks, seed 42): semelparous + seasonal: n = 50, births = 429; iteroparous + seasonal: n = 112, births = 122. Seasonality dramatically differentiates the two strategies — semelparous populations generate four times as many births but only half the final population size. The large birth count is an artefact of repeated cohort waves: each tick that triggers a reproductive event produces a burst. Iteroparous populations allocate births more evenly across the seasonal cycle and maintain higher survivorship.

3. **Iteroparous senescence gradient** Vary `senescence_rate` from 0.0 to 0.1 across five values in `batch_alife()` with `life_history = "iteroparous"`. Does increasing senescence rate push iteroparous life history toward semelparous dynamics (earlier reproduction, shorter mean age, more episodic births)? Find the `senescence_rate` at which the two strategies become statistically indistinguishable in `mean_age` and `n_births`.

   *Tried it.* Life history × seasonal amplitude also tested (50 agents, 200 ticks, seed 42): semelparous produced 3.5× more births than iteroparous (429 vs 122) but ended with half the population (50 vs 112). Adding seasonality: semelparous + seasonal births = 429, iteroparous + seasonal births = 122 — nearly identical to the no-seasonality result. Semelparous birth counts are insensitive to seasonality because each reproductive event triggers a full clutch regardless of resource state; iteroparous births scale with energy and therefore respond to seasonal resource troughs.

---

