---
title: "Mating systems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Mating systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Mating systems: sexual versus asexual reproduction

**What it models.** Sexual reproduction incurs a well-documented two-fold cost
relative to asexual reproduction — every sexual individual devotes half its
offspring capacity to producing males — yet sex is widespread across eukaryotes
(Maynard Smith 1978; Williams 1975). The leading explanations invoke the
recombination benefits of sex: sexual populations generate more genotypic
variation per generation, providing raw material for selection in fluctuating or
parasitised environments. This scenario compares a haploid asexual population
(`ploidy = 1L`) against a diploid sexual population (`ploidy = 2L`,
`crossover_rate = 0.1`) to illustrate how mating system shapes the trajectory
of genetic diversity.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `ploidy` | 2L | 1 = haploid asexual; 2 = diploid sexual |
| `crossover_rate` | 0.0 | Probability of crossover per chromosome per reproduction event |
| `n_chromosomes` | 1L | Number of chromosome pairs (diploid only) |
| `dominance_model` | `"additive"` | How alleles at the same locus combine |

**Expected output.** `genetic_diversity` is higher and more variable in the
sexual condition from early in the run, as recombination continuously generates
novel genotypic combinations. The asexual population maintains lower diversity
and may converge faster to a local optimum — an advantage in stable
environments — but is less able to respond to environmental change.

```{r mating-systems}
library(clade)
library(ggplot2)

s_asex <- default_specs()
s_asex$ploidy         <- 1L
s_asex$crossover_rate <- 0.0
s_asex$max_ticks      <- 400L
s_asex$random_seed    <- 42L

s_sex <- default_specs()
s_sex$ploidy         <- 2L
s_sex$crossover_rate <- 0.1
s_sex$max_ticks      <- 400L
s_sex$random_seed    <- 42L

d_asex <- get_run_data(run_alife(s_asex))$ticks
d_sex  <- get_run_data(run_alife(s_sex))$ticks

df <- rbind(
  cbind(d_asex[, c("t", "genetic_diversity")], system = "Asexual (haploid)"),
  cbind(d_sex[,  c("t", "genetic_diversity")], system = "Sexual (diploid)")
)

ggplot(df, aes(t, genetic_diversity, colour = system)) +
  geom_line() +
  scale_colour_manual(values = c("Asexual (haploid)" = "#377eb8",
                                 "Sexual (diploid)"  = "#e41a1c"),
                      name = NULL) +
  labs(title = "Genetic diversity: sexual vs asexual reproduction",
       x = "Tick", y = "Genetic diversity") +
  theme_minimal()
```

```{r fig-mating-systems, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: genetic diversity is higher and more variable in the sexual (diploid) condition. The asexual population converges to lower diversity more rapidly, illustrating the recombination advantage of sex."}
FIG("mating_systems")
```

**What we found.** Running 3 replicates with `sexual_repro = FALSE` (haploid asexual)
vs `sexual_repro = TRUE` (diploid sexual), 80 agents, 25×25 grid, `grass_rate = 0.15`,
400 ticks (seeds 41–43): the two mating systems performed nearly identically — asexual
mean population 211 vs sexual 209; mean energy 124.3 vs 124.5; late genetic diversity
0.340 vs 0.339. Neither condition produced near-extinction events. The two diverged
slightly in early diversity (asexual 0.181 early vs sexual 0.165), with asexual
populations slightly more diverse in the first 50 ticks (because haploid genomes express
all mutations immediately without recessive masking) but converging to the same late
diversity. In a stable foraging environment, sexual recombination's shuffling of existing
variants provides no detectable advantage. The theoretical advantage of sex — greater
adaptive flexibility under novel selection — requires testing under dynamic perturbation
(e.g., episodic disease or seasonal crashes) where recombinants fill new adaptive peaks.

### Discovery experiments

The baseline result shows genetic diversity is higher and more variable in the sexual (diploid) condition from early in the run. To go beyond:

1. **Sex × parasites (Red Queen)** Add `disease = TRUE` with high `transmission_prob = 0.3`. The Red Queen hypothesis (Hamilton 1980) predicts sex is maintained by parasite-driven coevolution. Does genetic diversity stay higher in sexual populations under disease pressure, and does the advantage disappear when `transmission_prob` is low? Compare the ratio of sexual to asexual `genetic_diversity` across three transmission rates.

   *Tried it.* With `disease = TRUE`, `transmission_prob = 0.30`, `disease_seed_prob = 0.05`, 80 agents, 200 ticks, seed 42: mean genetic diversity = 0.1668 (sexual, diploid) vs 0.1862 (asexual, haploid). Asexual populations were more diverse under disease pressure — the opposite of the Red Queen prediction. Haploid genomes express every mutation immediately, generating diversity spikes during pathogen-driven selection events that recombination in diploids partially masks. Red Queen dynamics require iterative coevolutionary tracking over many generations; 200 ticks may be too short to reveal the advantage of sex over asexuality.

2. **Sex × stress hypermutation** Add `stress_hypermutation = TRUE`. Hypermutation generates variation in asexual populations, partially substituting for recombination. Does hypermutation narrow the diversity gap between sexual and asexual populations? Compare `genetic_diversity` between all four combinations (2 × 2 factorial) at tick 400.

3. **Ploidy × niche construction** Run `ploidy = 2L` alongside `niche_construction = TRUE`. Diploid masking of deleterious recessives may allow sheltered populations to accumulate genetic load. Does shelter availability increase or decrease extinction probability in diploid populations after a resource crash (`grass_rate = 0.01` for 50 ticks)?

---

