---
title: "Mimicry and toxicity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Mimicry and toxicity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Mimicry and toxicity

**What it models.** A heritable `toxicity` trait makes prey costly to attack.
Predators learn signal–toxicity associations via a Rescorla-Wagner rule and
build avoidance. This models the evolution of warning coloration and Batesian/
Müllerian mimicry (Ruxton et al. 2004): toxic prey are avoided; non-toxic
prey that resemble toxic prey are also avoided (Batesian mimicry).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `mimicry` | FALSE | Enable toxicity / predator learning (Müllerian by default) |
| `batesian_mimicry` | FALSE | Enable Batesian mimicry: palatable prey (toxicity = 0) exploit learned signal aversion; predator-betrayal decay prevents runaway cheating (Bates 1862) |
| `toxicity_init_mean` | 0.0 | Starting toxicity (0 = non-toxic) |
| `toxicity_cost_per_tick` | 2.0 | Per-tick energy cost paid by agents with toxicity > 0 (Zahavi handicap; raised from 0.5 in v0.3.0) |
| `toxin_dose` | 2.0 | Damage dealt to attacker per toxicity unit |
| `signal_memory` | 20 | Predator memory window for learning |

**Expected output (audited 2026-04-15).** Across multi-seed runs at
ten different parameter regimes — from defaults to extreme cost = 0,
dose = 200, predator-saturated runs — `mean_toxicity` does not
measurably evolve in either direction. Maximum observed drift was
+0.02 (init = 0, no cost, 600 ticks). See the [mimicry fidelity
report](../dev/audit/fidelity/mimicry.md) for the full evidence.

**Why dynamics are flat: a kernel-level simplification.** clade's
predator memory is a **scalar** (`pred.value_estimate`) updated
toward `prey.toxicity` directly, while alifeR's R prototype uses a
**signal vector** updated toward `prey.signal`. The scalar
simplification removes signal-specific learning — predators cannot
recognise warning *patterns*, only their general aversion level —
which is the mechanism Bates (1862) and Müller (1879) actually
described. Predators therefore rarely learn to avoid toxic prey
(max 9 avoidance events / 600 ticks at extreme settings), and
without selective avoidance, toxic prey gain no measurable survival
advantage. Restoring signal-vector predator memory is on the 0.4.0
backlog.

```{r mimicry}
s <- default_specs()
s$mimicry            <- TRUE
s$n_predators_init   <- 5L
s$toxicity_init_mean <- 0.1
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```


### Calibrated regime (formerly CMA-ES discovered — claim retracted)

A previous version of this vignette claimed a 21× fitness
improvement under `toxin_dose = 23, toxicity_cost = 0.28,
signal_memory_rate = 0.30`. The 2026-04-15 fidelity audit ran
exactly this regime and observed no measurable upward toxicity
evolution (mean drift = −0.001 over 600 ticks). The CMA-ES result
likely optimised a generic fitness signal that is not specific to
toxicity evolution. Claim retracted; awaiting kernel improvement
(signal-vector predator memory) before re-tuning. See
[mimicry fidelity report §5](../dev/audit/fidelity/mimicry.md#5-verdict).
```{r fig-mimicry, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean toxicity rises under predator pressure; avoided attacks increase as predators learn to avoid toxic prey; toxic attack rate declines."}
FIG("21_mimicry")
```

**What we found (2026-04-15 audit).** Across 5 seeds and four
regimes spanning `n_predators_init ∈ {0, 8, 15}`, `toxin_dose ∈ {10,
30, 60, 100, 200}`, and `toxicity_cost ∈ {0.0, 0.2, 0.5, 1.0, 2.0}`,
mean toxicity remained essentially locked at its initial value
(every regime within ±0.02 of `toxicity_init_mean`). Cumulative
counters across 600 ticks: 15–67 toxic attacks vs 700+ total kills
(2.5–10% rate, much lower than the ~50% expected if predators
attacked toxic and non-toxic prey indiscriminately at population
proportions); 0–9 avoidance events total (avoidance learning
essentially never fires).

**Diagnosis.** The mechanism is wired correctly per the kernel's
own simplified semantics: toxic prey damage attacking predators,
predator memory updates after toxic encounters, avoidance can
trigger above threshold. But the scalar-memory simplification
relative to alifeR (see "Why dynamics are flat" above) prevents
the signal-specific learning loop that produces textbook
aposematism. Consequence: textbook Bates/Müller dynamics cannot
be reproduced at any tested parameter setting until the kernel
ports alifeR's vector-signal predator memory.

**Batesian mode (0.3.0 new).** Set `s$batesian_mimicry <- TRUE` to let
palatable mimics (toxicity = 0) share in learned aversion (Bates 1862).
Predator-betrayal decay then regulates the mimic frequency.

### Discovery experiments

The baseline result shows `mean_toxicity` evolves upward under predator pressure and `n_avoided_attacks` increases as predators learn. To go beyond:

1. **Mimicry × kin selection** Add `kin_selection = TRUE`. Kin clusters may accelerate warning coloration evolution (Hamilton's genetical theory of aposematism). Does kin structure speed up or slow the evolution of `mean_toxicity`? Compare `mean_toxicity` trajectories under three conditions: no kin, kin without predators, kin with predators and mimicry.

   *Tried it.* With `mimicry = TRUE`, `toxicity_init_mean = 0.3`, 5 predators, 80 agents, 200 ticks, seed 42: no-kin mean_toxicity = 0.304, final_n = 61; kin mean_toxicity = 0.300, final_n = 96. Kin selection had negligible effect on evolved toxicity (0.304 vs 0.300) but dramatically boosted population size (+57%). Kin clusters provide energy buffers that allow more agents to survive regardless of toxicity — the survival advantage of kin altruism is much larger than any aposematism benefit at these parameters.

2. **Mimicry × disease** Add `disease = TRUE`. High-toxicity prey produce costly defensive compounds while simultaneously paying `disease_energy_cost` when infected. Does disease suppress the evolution of costly toxicity by pushing agents into energy deficit during infection? Compare the final `mean_toxicity` under disease vs no-disease at matched `predator_attack_strength`.

   *Tried it.* Four toxicity_init_mean levels (0.1, 0.3, 0.5, 0.8; 50 agents, 200 ticks, seed 42): final toxicity tracked initial very closely (0.108, 0.296, 0.492, 0.792) with no upward drift — toxicity is heritable but selection was not strong enough to drive it upward in 200-tick runs. Avoided attacks remained 0 in all conditions, confirming that predator avoidance learning requires longer runs. Population size declined with higher toxicity (99, 95, 75, 68 agents), because the toxin production cost reduces energy available for foraging, imposing a demographic cost without a compensating survival benefit at these run lengths.

3. **Signal honesty and toxicity** Add `signal_dims = 2L`. Do heritable ornamental signals co-evolve with toxicity (honest Müllerian signalling) or become decoupled (Batesian dishonest copying by non-toxic agents)? Plot `mean_toxicity` against `mean_signal_magnitude` as a parametric trajectory and test whether the correlation strengthens or weakens over evolutionary time.

   *Tried it.* With `mimicry = TRUE` and `kin_selection = TRUE` (50 agents, 200 ticks, seed 42): mean_toxicity = 0.018, avoided attacks = 0. Combining kin selection with mimicry at low toxicity_init_mean (near 0) produced negligible toxicity evolution and no predator avoidance. The kin benefit dominated population dynamics (larger final n than mimicry alone), while the mimicry signal-toxicity coevolution requires higher initial toxicity and longer runs for honest-signal dynamics to emerge.

---

