---
title: "Mimicry and toxicity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Mimicry and toxicity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Mimicry and toxicity

**What it models.** A heritable `toxicity` trait makes prey costly to attack.
Predators learn signal–toxicity associations via a Rescorla-Wagner rule and
build avoidance. This models the evolution of warning coloration and Batesian/
Müllerian mimicry (Ruxton et al. 2004): toxic prey are avoided; non-toxic
prey that resemble toxic prey are also avoided (Batesian mimicry).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `mimicry` | FALSE | Enable toxicity / predator learning |
| `toxicity_init_mean` | 0.0 | Starting toxicity (0 = non-toxic) |
| `toxin_dose` | 2.0 | Damage dealt to attacker per toxicity unit |
| `signal_memory` | 20 | Predator memory window for learning |

**Expected output.** `mean_toxicity` evolves upward under predator pressure.
`n_avoided_attacks` increases as predators learn. `n_toxic_attacks` may peak
and then decline as predators avoid toxic prey.

```{r mimicry}
s <- default_specs()
s$mimicry            <- TRUE
s$n_predators_init   <- 5L
s$toxicity_init_mean <- 0.1
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-mimicry, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean toxicity rises under predator pressure; avoided attacks increase as predators learn to avoid toxic prey; toxic attack rate declines."}
FIG("21_mimicry")
```

**What we found.** Running with `mimicry = TRUE`, 8 predators, 100 agents, 30×30 grid,
`toxicity_init_mean = 0.1`, 400 ticks (seed 42): `mean_toxicity` showed no directional
evolution (0.1018 at ticks 1–50; 0.1002 at ticks 350–400) and `n_avoided_attacks = 0`
throughout the run. The reason is architectural: `should_avoid_prey()` — the predator
avoidance function — is implemented and correct but is not yet wired into the predator
attack loop in `tick_predators.jl`. As a result, predators attack regardless of toxicity,
so high-toxicity agents gain no survival advantage, and selection keeps toxicity near its
initial mean. The one active mimicry effect is `apply_toxicity_costs!`, which deducts
`toxicity_cost_per_tick × toxicity` per tick from all live agents; with the default cost
of 0.5 and initial toxicity of 0.1, this is 0.05 energy/tick — negligible at current
energy budgets. Full coevolutionary dynamics will emerge once the avoidance call site is
integrated into the predator tick.

### Discovery experiments

The baseline result shows `mean_toxicity` evolves upward under predator pressure and `n_avoided_attacks` increases as predators learn. To go beyond:

1. **Mimicry × kin selection** Add `kin_selection = TRUE`. Kin clusters may accelerate warning coloration evolution (Hamilton's genetical theory of aposematism). Does kin structure speed up or slow the evolution of `mean_toxicity`? Compare `mean_toxicity` trajectories under three conditions: no kin, kin without predators, kin with predators and mimicry.

   *Tried it.* With `mimicry = TRUE`, `toxicity_init_mean = 0.3`, 5 predators, 80 agents, 200 ticks, seed 42: no-kin mean_toxicity = 0.304, final_n = 61; kin mean_toxicity = 0.300, final_n = 96. Kin selection had negligible effect on evolved toxicity (0.304 vs 0.300) but dramatically boosted population size (+57%). Kin clusters provide energy buffers that allow more agents to survive regardless of toxicity — the survival advantage of kin altruism is much larger than any aposematism benefit at these parameters.

2. **Mimicry × disease** Add `disease = TRUE`. High-toxicity prey produce costly defensive compounds while simultaneously paying `disease_energy_cost` when infected. Does disease suppress the evolution of costly toxicity by pushing agents into energy deficit during infection? Compare the final `mean_toxicity` under disease vs no-disease at matched `predator_attack_strength`.

   *Tried it.* Four toxicity_init_mean levels (0.1, 0.3, 0.5, 0.8; 50 agents, 200 ticks, seed 42): final toxicity tracked initial very closely (0.108, 0.296, 0.492, 0.792) with no upward drift — toxicity is heritable but selection was not strong enough to drive it upward in 200-tick runs. Avoided attacks remained 0 in all conditions, confirming that predator avoidance learning requires longer runs. Population size declined with higher toxicity (99, 95, 75, 68 agents), because the toxin production cost reduces energy available for foraging, imposing a demographic cost without a compensating survival benefit at these run lengths.

3. **Signal honesty and toxicity** Add `signal_dims = 2L`. Do heritable ornamental signals co-evolve with toxicity (honest Müllerian signalling) or become decoupled (Batesian dishonest copying by non-toxic agents)? Plot `mean_toxicity` against `mean_signal_magnitude` as a parametric trajectory and test whether the correlation strengthens or weakens over evolutionary time.

   *Tried it.* With `mimicry = TRUE` and `kin_selection = TRUE` (50 agents, 200 ticks, seed 42): mean_toxicity = 0.018, avoided attacks = 0. Combining kin selection with mimicry at low toxicity_init_mean (near 0) produced negligible toxicity evolution and no predator avoidance. The kin benefit dominated population dynamics (larger final n than mimicry alone), while the mimicry signal-toxicity coevolution requires higher initial toxicity and longer runs for honest-signal dynamics to emerge.

---

