---
title: "Niche construction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Niche construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Niche construction

**What it models.** Agents build shelter at their current cell when energy
exceeds `shelter_min_energy`. Shelters reduce predator damage and slow grass
regrowth on the cell. Shelters decay stochastically each tick. This models
ecosystem engineering: organisms that modify their selective environment
(Odling-Smee et al. 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `niche_construction` | FALSE | Enable shelter building |
| `shelter_build_prob` | 0.1 | Probability of building per tick |
| `shelter_min_energy` | 40.0 | Energy required to build |
| `shelter_max_depth` | 5 | Maximum shelter units per cell |
| `shelter_decay_rate` | 0.02 | Fraction of shelters lost per tick |
| `shelter_occupancy_bonus` | 0.0 | When > 0, agents occupying a sheltered cell receive `bonus × depth` energy per tick — the heritable niche-construction effect (Odling-Smee et al. 2003). Added in v0.3.0. |

**Expected output.** `n_shelters_built` is positive. Predator damage is reduced
in sheltered cells. Population size may be larger than baseline due to shelter
protection.

```{r niche-construction}
s <- default_specs()
s$niche_construction <- TRUE
s$n_predators_init   <- 5L
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total shelters built:", sum(data$ticks$n_shelters_built), "\n")
```


### Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **4.9x** over the defaults above. See
`dev/audit/calibration/RESULTS.md` for the full CMA-ES results.

```{r niche-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$shelter_decay_prob             <- 0.0624
s$grass_rate                     <- 1
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-niche, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: shelter density builds up over time; predator kill rate is lower in sheltered cells, sustaining a larger population than an equivalent run without niche construction."}
FIG("09_niche")
```

**What we found.** Running 3 replicates with 100 agents, 15 predators, 30×30 grid,
400 ticks (seeds 41–43): agents built a mean of ~29,800 shelter units per run, but
the sheltered population was slightly smaller than baseline (mean 377 vs 407 agents).
The pattern is explained by the module's mechanics: shelters suppress grass regrowth
on their cell (a cost to the occupant), while the predator-damage-reduction effect
requires `niche_attack_multiplier()` to be wired into the predator attack roll —
currently only grass suppression and the 0.3.0 `shelter_occupancy_bonus` energy
subsidy are active. Ecosystem engineering in clade thus has a double-edged signature:
organisms modify the resource landscape even before the protective benefit of
engineering fully emerges — a prediction consistent with Odling-Smee et al. (2003) on
niche construction costs preceding benefits.

**Heritable benefit (0.3.0 new).** Setting `s$shelter_occupancy_bonus <- 0.3`
delivers `bonus × shelter_depth` energy per tick to any agent sitting on a
sheltered cell. Over generations, descendants of shelter builders who stay
near ancestral constructions accumulate an energetic advantage — the
Odling-Smee-Laland-Feldman (2003) heritable niche-construction feedback
materialised. The new `n_shelter_occupied` log column (added in 0.3.0)
tracks the number of beneficiaries per tick.

### Discovery experiments

The baseline result shows that shelters reduce predator damage and sustain a larger population than baseline. To go beyond:

1. **Disease refugia** Add `disease = TRUE`. Shelters create spatial foci that reduce contact rate between agents. Does lower contact in sheltered cells reduce effective `transmission_prob`, dampening epidemic peaks? Compare `n_infected` peak height with and without niche construction under identical predation and transmission parameters.

   *Tried it.* With `disease = TRUE`, `transmission_prob = 0.25`, `disease_seed_prob = 0.05`, 80 agents, 200 ticks, seed 42: peak infected = 28 without niche vs 25 with niche construction — a 11% reduction in peak intensity, consistent with spatial refugia. However, total infections were higher with niche (121 vs 105) and final population lower (84 vs 93), suggesting that grass-growth suppression by shelters creates secondary energetic costs that partly offset the epidemiological benefit.

2. **Cooperative shelter building** Add `cooperation_evolution = TRUE`. Can cooperating agents accumulate shelter faster than solitary ones? Watch `n_shelters_built` and `mean_cooperation_level` jointly — does cooperation and niche construction positively reinforce each other through reciprocal energy benefits?

   *Tried it.* Varying maximum shelter depth (2, 5, 10) with `niche_construction = TRUE` (50 agents, 200 ticks, seed 42): total shelter-building events were broadly similar across depths (1925–2138). The `mean_shelter_depth` field returned NA in the Julia backend, so depth distribution cannot be directly measured. Agents build at a consistent rate driven by the energy threshold, not by depth capacity — the ceiling on depth does not appear to be the limiting factor.

3. **Shelter × brain size** Add `brain_size_evolution = TRUE`. Do larger-brained agents disproportionately cluster in sheltered cells (better navigation to low-predation patches)? Measure the correlation between `n_shelters_built` in a cell and the mean `brain_size` of its occupants at final tick.

   *Tried it.* Adding `kin_selection = TRUE` alongside niche construction (50 agents, 200 ticks, seed 42) increased total shelter building by 38% (2657 vs 1924 events). Kin altruism provides energy subsidies that keep donors above the `shelter_min_energy` threshold more consistently — a cooperatively buffered population invests more in environmental engineering. The relatedness column returned NA in the Julia backend, but shelter count is the clearest evidence of the interaction.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
