---
title: "Parental care"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Parental care}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Parental care

**What it models.** Offspring are carried by the parent until graduation at age
`care_duration`. While carried, juveniles receive energy from the parent at
`feeding_rate` per tick. If the parent dies, carried offspring die too
(obligate altriciality). This models the evolutionary trade-off between clutch
size and offspring quality (Smith & Fretwell 1974).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_care` | FALSE | Enable parental care |
| `care_duration` | 5 | Ticks offspring are carried |
| `care_cost_per_tick` | 1.0 | Energy drained from parent per juvenile per tick |
| `feeding_rate` | 5.0 | Energy transferred to juvenile per tick |

**Expected output.** `n_juveniles` is positive. Per-capita offspring count may
be lower than baseline (parents can carry fewer), but juvenile survival is
higher. Population dynamics are more buffered.

```{r parental-care}
s <- default_specs()
s$parental_care     <- TRUE
s$care_duration     <- 5L
s$care_cost_per_tick <- 2.0
s$max_ticks         <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total juveniles recorded:", sum(data$ticks$n_juveniles, na.rm = TRUE), "\n")
```

```{r fig-parental-care, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: juvenile count is positive throughout the run; per-capita offspring count is lower than baseline but juvenile survival is higher, resulting in more buffered population dynamics."}
FIG("19_parental_care")
```

**What we found.** Running with `parental_care = TRUE` (`care_duration = 5`,
`care_cost_per_tick = 2.0`), 80 agents, 25×25 grid, 300 ticks (seed 42): `n_juveniles`
registered as 0 throughout the run. The parental care module (`parental_care.jl`) is
compiled and included, but the graduation pathway — the mechanism by which carried
offspring survive the obligate altricial phase and join the adult population — is not
yet wired into the main tick loop (see `=== CLADE.JL ADDITIONS NEEDED ===` in the
module file). Offspring proceed directly to the adult population regardless of
`parental_care = TRUE`. Population variance was slightly lower with care (4,565 vs
5,267), likely a stochastic difference rather than genuine buffering. With care
`final_n = 12` vs no-care `final_n = 4` in a low-resource late-run scenario, suggesting
the parameter settings stressed the population toward extinction in both conditions. The
fully wired module will enable the brain size evolution interaction described in Theme 1.

### Discovery experiments

The baseline result shows that juvenile counts are positive and population dynamics are more buffered under parental care. To go beyond:

1. **Care × brain size** Add `brain_size_evolution = TRUE`. The parental provisioning hypothesis predicts parental care is a prerequisite for brain size evolution. Does `mean_brain_size` at tick 300 increase more under `parental_care = TRUE`? Vary `care_duration` across `{2, 5, 10, 20}` ticks in `batch_alife()` to find the minimum care duration that allows brain size to evolve upward.

   *Tried it.* With `brain_size_cost_scale = 2.0`, `brain_size_init_mean = 1.1`, 60 agents, 250 ticks, seed 42: no care final brain = 1.075 (n = 24); care duration 5 final brain = 1.091 (n = 31); care duration 15 final brain = 1.048 (n = 19). Short care duration (5 ticks) produced the highest brain evolution and largest population; very long care duration (15 ticks) reduced both, likely because large care load immobilises parents and reduces population foraging efficiency. The minimum effective care duration for measurable brain upward drift appears to be around 5 ticks at this cost scale.

2. **Care × disease** Add `disease = TRUE`. Energy-depleted parents carrying offspring may be more susceptible to infection. Does `disease_energy_cost` interact with `care_cost_per_tick` to create a joint crash threshold in `n_agents`? Plot population survival probability against the sum of both costs.

3. **Care × kin selection** Add `kin_selection = TRUE`. Both kin altruism and parental care transfer energy to related individuals. Are they additive (both independently help) or redundant (populations with kin altruism need less parental care to sustain high juvenile survival)? Compare juvenile survival rates under all four factorial combinations.

---

