---
title: "Pace-of-life syndromes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Pace-of-life syndromes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Pace-of-life syndromes

**What it models.** Pace-of-life syndromes (Réale et al. 2010) describe the
co-variation of life-history traits, behaviour, and physiology along a slow–fast
continuum. Fast-living organisms have high metabolic rates, reproduce early and
often, and die young; slow-living organisms invest in maintenance, defer
reproduction, and achieve longer lifespans. Stearns (1992) formalised this as a
consequence of the age-specific mortality schedule: high extrinsic mortality
selects for early reproduction and fast metabolism. This scenario fixes metabolic
rate across three values to isolate the direct effect of metabolic pace on the
age–reproduction–energy trade-off.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `metabolic_rate_init_mean` | 1.0 | Initial (and fixed) metabolic rate |
| `metabolic_rate_evolution` | FALSE | Keep metabolic rate fixed for this scenario |
| `metabolic_rate_min` | 0.1 | Lower clamp (relevant when evolution is enabled) |
| `metabolic_rate_max` | 5.0 | Upper clamp |

**Expected output.** Fast-pace agents (`metabolic_rate_init_mean = 2.0`)
exhibit higher `n_births` per tick, lower `mean_age`, and higher energy
throughput, but their populations are more volatile. Slow-pace agents
(`metabolic_rate_init_mean = 0.5`) show lower birth rate, higher `mean_age`,
and more stable `mean_energy`. Fast agents dominate in productive environments;
slow agents persist more reliably when resources are lean.

```{r pace-of-life}
library(clade)
library(ggplot2)

make_run <- function(rate, seed = 99L) {
  s <- default_specs()
  s$metabolic_rate_init_mean <- rate
  s$metabolic_rate_evolution <- FALSE
  s$max_ticks                <- 400L
  s$random_seed              <- seed
  get_run_data(run_alife(s))$ticks
}

d_slow     <- make_run(0.5)
d_baseline <- make_run(1.0)
d_fast     <- make_run(2.0)

df <- rbind(
  cbind(d_slow[,     c("t", "mean_age", "n_births")], pace = "Slow (0.5)"),
  cbind(d_baseline[, c("t", "mean_age", "n_births")], pace = "Baseline (1.0)"),
  cbind(d_fast[,     c("t", "mean_age", "n_births")], pace = "Fast (2.0)")
)

ggplot(df, aes(t, mean_age, colour = pace)) +
  geom_line() +
  scale_colour_manual(
    values = c("Slow (0.5)" = "#377eb8", "Baseline (1.0)" = "#4daf4a",
               "Fast (2.0)" = "#e41a1c"),
    name = NULL) +
  labs(title = "Pace-of-life: mean age across metabolic rates",
       x = "Tick", y = "Mean agent age") +
  theme_minimal()
```

```{r fig-pace-of-life, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: slow-pace agents (blue) maintain the highest mean age; fast-pace agents (red) have the lowest mean age but the highest per-tick birth rate. The three trajectories illustrate the life-history trade-off along the slow-fast continuum."}
FIG("pace_of_life")
```

**What we found.** Running with `senescence_rate = 0.05` (fast pace, `repro_senescence
= 0.03`) vs `senescence_rate = 0.001` (slow pace, no reproductive senescence), 80 agents,
25×25 grid, `grass_rate = 0.15`, 400 ticks (3 replicates): the extreme fast pace drove
near-extinction (mean population 5 vs 188 for slow pace). Fast-pace agents had mean age
2.2 ticks (vs 85 ticks slow), mean energy 22 (vs 131 slow), and total births of 88 (vs
433 slow). The Gompertz senescence rate of 0.05/tick creates a ~5% per-tick hazard rate
that is independent of current age, causing rapid population collapse because offspring
cannot reach reproductive age before dying. No age-related deaths were logged in the fast
condition (agents died too young to trigger the age-death pathway), whereas 245 age-deaths
were logged in the slow condition. The contrast illustrates that `senescence_rate` is not
a mild parameter: values above ~0.01 impose near-lethal selection pressure. Calibrate by
starting with `senescence_rate = 0.005` and increasing in steps; watch `mean_age` to
confirm viable demography before comparing conditions.

### Discovery experiments

The baseline result shows the pace-of-life trade-off: fast-pace agents have higher birth rates and lower mean age; slow-pace agents have lower birth rates and higher mean age and energy stability. To go beyond:

1. **Pace × disease** Add `disease = TRUE`. High metabolic rate means fast-paced organisms have less energy margin; `disease_energy_cost` should push fast-paced agents into energy deficit more readily. Does disease selectively eliminate fast-paced phenotypes when `metabolic_rate_evolution = TRUE`, driving evolved `mean_metabolic_rate` downward?

   *Tried it.* With `metabolic_rate_evolution = TRUE`, `transmission_prob = 0.25`, `disease_seed_prob = 0.05`, 60 agents, 200 ticks, seed 42: no-disease final metabolic rate = 1.004; with disease = 0.988. Disease reduced evolved metabolic rate by ~1.6%, consistent with directional selection against high-energy-burn phenotypes under infection. Population size was modestly lower (102 vs 106). The effect is small but in the correct direction: disease imposes a pace-of-life cost that selects toward slower, more energy-conservative strategies.

2. **Pace × brain size** Enable both `metabolic_rate_evolution = TRUE` and `brain_size_evolution = TRUE`. Do fast-paced, high-energy agents evolve larger brains (more metabolic resources available for expensive brain tissue), or does metabolic competition prevent it? Test whether `mean_brain_size` and `mean_metabolic_rate` are positively or negatively correlated at final tick.

3. **Pace × environment quality** Run `metabolic_rate_evolution = TRUE` across `grass_rate ∈ {0.02, 0.1, 0.4}` in `batch_alife()`. Theory predicts rich environments select for fast pace. Does evolved `mean_metabolic_rate` increase monotonically with `grass_rate`, or is there a non-monotone relationship (very high resources remove mortality pressure on fast-paced energy burn)?

---


## Theme 5 — Species interactions and arms races {#theme-5}

*Predator–prey coevolution, infectious disease, warning coloration, and group defence. These scenarios operationalise the major classes of antagonistic species interactions and ask how each reshapes phenotypic and genetic diversity.*

