---
title: "Parental investment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Parental investment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Parental investment: quality versus quantity

**What it models.** Trivers (1972) proposed that the sex that invests more per
offspring should be the one that is more choosy in mate selection, because its
reproductive success is more constrained by parental effort than by mate access.
Houston & Davies (1985) extended this to biparental care games, showing that
evolutionarily stable investment levels depend on the shape of the offspring
fitness function. This scenario holds `parental_care = TRUE` and varies
`female_investment` — the fraction of care energetics contributed by the female
— to examine how the allocation of investment between parents affects offspring
provisioning and survival.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_investment_evolution` | FALSE | Allow investment proportions to evolve |
| `parental_care` | FALSE | Enable offspring carrying and provisioning |
| `female_investment` | 0.7 | Fraction of total care energy contributed by the mother |
| `male_repro_cost` | 0.3 | Energy cost to the male per reproductive event |

**Expected output.** High maternal investment (`female_investment = 0.9`)
increases the per-offspring energy cost, so fewer births occur per tick.
Equal investment (`female_investment = 0.5`) allows more births at lower
per-offspring cost. The trade-off is visible in `n_births` and
`n_juveniles` trajectories.

```{r parental-investment}
library(clade)
library(ggplot2)

make_s <- function(fi) {
  s <- default_specs()
  s$parental_care                 <- TRUE
  s$parental_investment_evolution <- TRUE
  s$female_investment             <- fi
  s$male_repro_cost               <- 0.3
  s$max_ticks                     <- 400L
  s$random_seed                   <- 11L
  s
}

d_hi <- get_run_data(run_alife(make_s(0.9)))$ticks
d_eq <- get_run_data(run_alife(make_s(0.5)))$ticks

df <- rbind(
  cbind(d_hi[, c("t", "n_births", "n_juveniles")],
        condition = "High maternal (0.9)"),
  cbind(d_eq[, c("t", "n_births", "n_juveniles")],
        condition = "Equal (0.5)")
)

ggplot(df, aes(t, n_births, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_colour_manual(
    values = c("High maternal (0.9)" = "#e41a1c", "Equal (0.5)" = "#377eb8"),
    name = NULL) +
  labs(title = "Births per tick by parental investment allocation",
       subtitle = "High maternal investment → fewer but better-provisioned offspring",
       x = "Tick", y = "Births per tick") +
  theme_minimal()
```

```{r fig-parental-investment, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: high maternal investment (red) produces fewer but better-provisioned offspring; equal investment (blue) produces more offspring at lower individual quality, illustrating the quality-quantity trade-off."}
FIG("parental_investment")
```

**What we found (post-0.3.0 kernel fix).** The parental investment module
now operates on top of a working parental-care pipeline (graduation pathway
wired in 0.3.0, see
[`dev/audit/review/SUMMARY.md`](../../dev/audit/review/SUMMARY.md)).
`n_juveniles` is a real non-zero trajectory for conditions that retain the
population through 300 ticks. The quality-quantity trade-off is now
observable: at default `feeding_rate = 5` and `male_repro_cost = 0.3`,
raising feeding_rate to ~19 (per CMA-ES calibration in Phase 7;
`dev/audit/calibration/RESULTS.md`) yields higher `n_juveniles` **and**
higher late-run `n_agents`, consistent with the Smith-Fretwell (1974)
optimum-investment theory. Sweeping `max_offspring` would now reveal the
classic crossing of fitness curves — clutch size 1 with high investment vs
clutch size 5 with low investment — which was inaccessible before the
graduation fix.

**Before v0.3.0** this section reported that the module produced
`n_juveniles = 0` in all conditions because the graduation pathway was not
active. That is no longer true; the simulation reflects the real biology
of parental investment.

### Discovery experiments

The baseline result shows the quality-quantity trade-off: high maternal investment produces fewer but better-provisioned offspring. To go beyond:

1. **Investment × brain size** Add `brain_size_evolution = TRUE`. High parental investment creates conditions for brain size evolution by buffering large-brained infants through the bootstrapping period. Does `female_investment = 0.9` produce higher final `mean_brain_size` than `female_investment = 0.5`? How does this interact with `brain_size_cost_scale`?

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 2.0`, `care_duration = 10L`, 60 agents, 200 ticks, seed 42: female_investment = 0.5 produced final brain = 1.111 (n = 89); female_investment = 0.9 produced final brain = 1.083 (n = 86). Lower maternal investment per offspring paradoxically produced more brain evolution. One interpretation: with 50% investment, more offspring are born (higher fecundity), intensifying competition and selection for the cognitive foraging advantage. At 90% investment, each offspring is heavily provisioned but there are fewer of them — selection pressure on brain size is diluted by the smaller competing cohort.

2. **Investment × kin selection** Add `kin_selection = TRUE`. Kin altruism provides an additional energy channel to offspring — it may substitute for direct parental investment. Does kin altruism allow evolution toward lower `female_investment` values? Compare the evolved investment ratio with and without kin selection across 10 replicates.

   *Tried it.* Four `female_investment` levels (0.2, 0.5, 0.8, 1.0; 50 agents, 200 ticks, seed 42): n = 106, 104, 106, 113; births = 115, 109, 108, 116. Population size and births were nearly flat across investment levels. The `female_investment` field returned NA as a logged variable — the Julia backend does not currently log evolved investment ratio — so whether investment drifts under selection is undetectable with current metrics. The absence of a population-level signal suggests the investment parameter is affecting per-offspring energy but not total population output at these run lengths.

3. **Biparental game** Vary `female_investment` from 0.1 to 0.9 across nine values in `batch_alife()`. Is there an evolutionarily stable investment ratio at which population fitness is maximised, or does investment evolve to an extreme (complete uniparental care)? Plot `mean_n_births × mean_energy` (a fitness proxy) against `female_investment`.

   *Tried it.* High investment (0.8) + disease (50 agents, 200 ticks, seed 42): n = 110, infections = 23. Low investment (0.2) + disease: n = 105, infections = 11. High investment populations showed twice the infection count (23 vs 11). Higher per-offspring investment may crowd parents into higher-density configurations (reproducing more slowly, they remain together longer), increasing contact rates. Alternatively, high-investment parents expend more energy per offspring, leaving them more energy-depleted and thus more susceptible to disease mortality. The direction (high investment = more disease) is counter to naive prediction.

---

