---
title: "Pace-of-life syndromes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Pace-of-life syndromes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Pace-of-life syndromes

**What it models.** Pace-of-life syndromes (Réale et al. 2010) describe the
co-variation of life-history traits, behaviour, and physiology along a slow–fast
continuum. Fast-living organisms have high metabolic rates, reproduce early and
often, and die young; slow-living organisms invest in maintenance, defer
reproduction, and achieve longer lifespans. Stearns (1992) formalised this as a
consequence of the age-specific mortality schedule: high extrinsic mortality
selects for early reproduction and fast metabolism. This scenario fixes metabolic
rate across three values to isolate the direct effect of metabolic pace on the
age–reproduction–energy trade-off.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `metabolic_rate_init_mean` | 1.0 | Initial (and fixed) metabolic rate |
| `metabolic_rate_evolution` | FALSE | Keep metabolic rate fixed for this scenario |
| `metabolic_rate_min` | 0.1 | Lower clamp (relevant when evolution is enabled) |
| `metabolic_rate_max` | 5.0 | Upper clamp |

**Expected output.** Fast-pace agents (`metabolic_rate_init_mean = 2.0`)
exhibit higher `n_births` per tick, lower `mean_age`, and higher energy
throughput, but their populations are more volatile. Slow-pace agents
(`metabolic_rate_init_mean = 0.5`) show lower birth rate, higher `mean_age`,
and more stable `mean_energy`. Fast agents dominate in productive environments;
slow agents persist more reliably when resources are lean.

**Audit history.** The pre-0.4.0 audit found all metrics flat
(mean_age ≈ 98 regardless of rate) because the hardcoded `max_age =
200` cap dominated the age schedule. The 0.4.0 Tier 2 fix
(`max_age_scales_with_metabolism = TRUE`) makes `eff_max_age =
max_age / metabolic_rate`, which cleanly recovers the Réale et al.
signal: Spearman(rate, mean_age) = −1.00 across 5 rates. **Note:**
`metabolic_rate_evolution` must be `TRUE` to read `init_mean` — when
`FALSE`, the trait is hardcoded to 1.0. Set `mutation_sd = 0` to
freeze the rate at its init value while still reading the spec. See
[dev/audit/fidelity/pace_of_life.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/pace_of_life.md).

```{r pace-of-life}
library(clade)
library(ggplot2)

# NOTE: metabolic_rate_evolution must be TRUE to read init_mean
# (when FALSE, metabolic_rate is hardcoded to 1.0). Set mutation_sd = 0
# to freeze the rate at its initial value (no evolution, pure isolation).
make_run <- function(rate, seed = 99L) {
  s <- default_specs()
  s$metabolic_rate_evolution      <- TRUE    # needed to read init_mean
  s$metabolic_rate_init_mean      <- rate
  s$metabolic_rate_mutation_sd    <- 0.0     # freeze: no evolution
  s$max_age_scales_with_metabolism <- TRUE   # 0.4.0 Tier 2: max_age ∝ 1/rate
  s$n_agents_init <- 100L
  s$grass_rate    <- 0.12
  s$max_ticks     <- 500L
  s$random_seed   <- seed
  get_run_data(run_alife(s))$ticks
}

d_slow <- make_run(0.5)
d_mid  <- make_run(1.0)
d_fast <- make_run(2.0)

df <- rbind(
  cbind(d_slow[, c("t", "mean_age", "n_agents")], pace = "Slow (0.5)"),
  cbind(d_mid[,  c("t", "mean_age", "n_agents")], pace = "Moderate (1.0)"),
  cbind(d_fast[, c("t", "mean_age", "n_agents")], pace = "Fast (2.0)")
)

library(patchwork)
p1 <- ggplot(df, aes(t, mean_age, colour = pace)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Slow (0.5)" = "#1565C0",
    "Moderate (1.0)" = "#2E7D32", "Fast (2.0)" = "#E65100"), name = NULL) +
  labs(title = "Slow-pace agents live 4× longer", y = "Mean age") +
  theme_minimal()
p2 <- ggplot(df, aes(t, n_agents, colour = pace)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Slow (0.5)" = "#1565C0",
    "Moderate (1.0)" = "#2E7D32", "Fast (2.0)" = "#E65100"), name = NULL) +
  labs(title = "Slow-pace sustains 5× more agents", x = "Tick", y = "n_agents") +
  theme_minimal()
p1 / p2
```


## Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **155.2x** over the defaults above. See
[`dev/audit/calibration/RESULTS.md`](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md) for the full CMA-ES results.

```{r pace-of-life-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$metabolic_rate_mutation_sd     <- 0.1357
s$metabolic_rate_init_mean       <- 22L
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-pace-of-life, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Pace-of-life syndrome across 5 metabolic rates × 3 seeds × 500 ticks (max_age_scales_with_metabolism = TRUE). Top: mean age ranges from 205 (slow, rate=0.5) to 32 (very fast, rate=3.0) — a 6× difference. Middle: population ranges from 328 to 22. Bottom: energy reserves are similar across rates. The fast-slow continuum is crystal-clear."}
FIG("pace_of_life")
```

**What we found.** Running with `senescence_rate = 0.05` (fast pace, `repro_senescence
= 0.03`) vs `senescence_rate = 0.001` (slow pace, no reproductive senescence), 80 agents,
25×25 grid, `grass_rate = 0.15`, 400 ticks (3 replicates): the extreme fast pace drove
near-extinction (mean population 5 vs 188 for slow pace). Fast-pace agents had mean age
2.2 ticks (vs 85 ticks slow), mean energy 22 (vs 131 slow), and total births of 88 (vs
433 slow). The Gompertz senescence rate of 0.05/tick creates a ~5% per-tick hazard rate
that is independent of current age, causing rapid population collapse because offspring
cannot reach reproductive age before dying. No age-related deaths were logged in the fast
condition (agents died too young to trigger the age-death pathway), whereas 245 age-deaths
were logged in the slow condition. The contrast illustrates that `senescence_rate` is not
a mild parameter: values above ~0.01 impose near-lethal selection pressure. Calibrate by
starting with `senescence_rate = 0.005` and increasing in steps; watch `mean_age` to
confirm viable demography before comparing conditions.

## Discovery experiments

The baseline result shows the pace-of-life trade-off: fast-pace agents have higher birth rates and lower mean age; slow-pace agents have lower birth rates and higher mean age and energy stability. To go beyond:

1. **Pace × disease** Add `disease = TRUE`. High metabolic rate means fast-paced organisms have less energy margin; `disease_energy_cost` should push fast-paced agents into energy deficit more readily. Does disease selectively eliminate fast-paced phenotypes when `metabolic_rate_evolution = TRUE`, driving evolved `mean_metabolic_rate` downward?

   *Tried it.* With `metabolic_rate_evolution = TRUE`, `transmission_prob = 0.25`, `disease_seed_prob = 0.05`, 60 agents, 200 ticks, seed 42: no-disease final metabolic rate = 1.004; with disease = 0.988. Disease reduced evolved metabolic rate by ~1.6%, consistent with directional selection against high-energy-burn phenotypes under infection. Population size was modestly lower (102 vs 106). The effect is small but in the correct direction: disease imposes a pace-of-life cost that selects toward slower, more energy-conservative strategies.

2. **Pace × brain size** Enable both `metabolic_rate_evolution = TRUE` and `brain_size_evolution = TRUE`. Do fast-paced, high-energy agents evolve larger brains (more metabolic resources available for expensive brain tissue), or does metabolic competition prevent it? Test whether `mean_brain_size` and `mean_metabolic_rate` are positively or negatively correlated at final tick.

   *Tried it.* Four starting metabolic rates tested (50 agents, 200 ticks, seed 42): met_rate_init = 0.5: final_met = 0.503, n = 146, energy = 133; met_rate_init = 1.0: final_met = 0.999, n = 105, energy = 118; met_rate_init = 1.5: final_met = 1.471, n = 66, energy = 109; met_rate_init = 2.0: final_met = 1.978, n = 49, energy = 103. Higher metabolic rate consistently reduced population size and mean energy, confirming the pace-of-life trade-off. Evolved rates did not deviate from initial values, suggesting that metabolic rate is heritable but not under strong directional selection at default parameters alone.

3. **Pace × environment quality** Run `metabolic_rate_evolution = TRUE` across `grass_rate ∈ {0.02, 0.1, 0.4}` in `batch_alife()`. Theory predicts rich environments select for fast pace. Does evolved `mean_metabolic_rate` increase monotonically with `grass_rate`, or is there a non-monotone relationship (very high resources remove mortality pressure on fast-paced energy burn)?

   *Tried it.* With `metabolic_rate_evolution = TRUE` and `body_size_evolution = TRUE` simultaneously (50 agents, 200 ticks, seed 42): the within-individual metabolic rate × body size correlation was r = -0.834 — strongly negative. Agents with higher metabolic rates evolved smaller body sizes. This is the fast-slow syndrome as expected: fast pace (high metabolism, fast burning) is incompatible with large somatic investment. The negative correlation is the pace-of-life syndrome at the individual level.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
