---
title: "Phenotypic plasticity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Phenotypic plasticity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Phenotypic plasticity

**What it models.** A heritable `plasticity` trait modulates how strongly each
agent's sensory input maps to action. High plasticity agents are more responsive
to environmental cues; low plasticity agents rely more heavily on their evolved
baseline behaviour. The evolution of plasticity tracks environmental variability:
plastic phenotypes are favoured when environments are predictably unpredictable
(DeWitt & Scheiner 2004).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.3 | Starting mean plasticity (**note**: was 0.0 in earlier versions — evolution requires a positive starting value) |
| `plasticity_mutation_sd` | 0.05 | Mutation rate |

**Expected output.** `mean_plasticity` evolves upward in seasonal
environments and stays lower (or drifts) in stable environments —
but only if **season length ≤ agent lifetime**. The DeWitt-
Scheiner prediction is about *within-lifetime* environmental
variability: if an agent lives through only one seasonal state,
genetic adaptation to that state is just as good as plasticity,
and there is no selection gradient. See "What we found" below for
the 2026-04-17 season-length sweep that surfaced this constraint.

```{r plasticity}
s <- fast_specs()                     # max_age = 30, ~66 generations in 2000 ticks
s$phenotypic_plasticity    <- TRUE
s$plasticity_init_mean     <- 0.3
s$plasticity_mutation_sd   <- 0.05
s$bnn_sigma_source         <- "trait"
s$brain_energy_sigma_scale <- 0.05
s$seasonal_amplitude       <- 0.35
s$season_length            <- 10L     # < max_age so agents see multiple seasons per life

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final plasticity:", tail(data$ticks$mean_plasticity, 1L), "\n")
```


### Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **38.1x** over the defaults above. See
`dev/audit/calibration/RESULTS.md` for the full CMA-ES results.

```{r plasticity-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$plasticity_init_mean           <- 0.6972
s$plasticity_mutation_sd         <- 0.1405
s$grass_rate                     <- 0.0264
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-plasticity, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="fast_specs() + season_length = 10 demo (5 seeds × 2000 ticks = 66 generations). Seasons are shorter than a lifetime, so each agent experiences multiple seasonal states. Seasonal environments maintain higher plasticity than stable ones — the DeWitt-Scheiner (2004) prediction. Δdelta = +0.014 across 5 seeds (direction PASS, below the 0.02 promotion threshold but 7× larger than at season_length = 100 where the signal reverses). See \"What we found (2026-04-17)\" for the full season-length sweep."}
FIG("22_plasticity")
```

**Latest (0.5.18 ✅ promotion via `seasonal_spatial_bias` regime).**
Pre-0.5.18 seasonality was a *uniform* stressor: `seasonal_amplitude`
modulated grass_rate uniformly across the grid, and
`winter_death_prob` applied equal mortality to every agent. Both
phenotype-agnostic, so neither created the fluctuating selection
DeWitt & Scheiner 2004 requires. The new `seasonal_spatial_bias`
spec makes grass growth *phenotype-relevant*: the optimal foraging
direction flips between seasons (north in summer, south in winter).
Plastic agents track the flip; canalized ones cannot.

3 conditions × 16 seeds × 2000 ticks at `default_specs` with RL on:

| condition | `seasonal_amplitude` | `seasonal_spatial_bias` | mean_prior_sigma |
|---|---|---|---|
| stable | 0 | 0 | 0.376 |
| amp_only (uniform) | 0.5 | 0 | 0.380 |
| **flipping** | **0.5** | **0.9** | **0.399** |

Pairwise differentials:

| comparison | Δσ | t | verdict |
|---|---|---|---|
| amp_only − stable | +0.004 | +0.91 | uniform stressor ≈ null |
| **flipping − stable** | **+0.0227 ± 0.0049** | **+4.58 PASS** | |
| **flipping − amp_only** | **+0.0186 ± 0.0040** | **+4.60 PASS** | |

Fluctuating selection raises prior σ by ~6% relative to stable —
decisive direction + magnitude for the DeWitt-Scheiner prediction.
Equilibrium population is ~28% lower in flipping (exploration has
a cost), consistent with DeWitt-Scheiner's theoretical caveat. Full
protocol:
[dev/audit/fidelity/plasticity_baldwin_promotion.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/plasticity_baldwin_promotion.md).

**Earlier (pre-0.5.18, 🟠 season-length sweep).**
The critical parameter turned out to be `season_length / max_age`:
plasticity evolves in the DeWitt-Scheiner direction (seasonal >
stable) only when each agent experiences multiple seasonal states
during its lifetime. See
[dev/audit/fidelity/plasticity_within_lifetime_sweep.R](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/plasticity_within_lifetime_sweep.R)
for the full sweep; results at fast_specs (`max_age = 30`, 40×40
grid, 180 agents init, 5 seeds × 2000 ticks):

| season_length | stable Δ | seasonal Δ | Δdelta | P1 |
|---|---|---|---|---|
| 10  | +0.004 | +0.018 | **+0.014** | PASS |
| 20  | +0.006 | +0.012 | +0.006 | PASS |
| 30  | +0.014 | +0.023 | +0.009 | PASS |
| 60  | +0.024 | +0.021 | −0.003 | FAIL |
| 100 | +0.020 | +0.010 | −0.010 | FAIL |

The direction flips at `season_length > max_age`. When seasons
are longer than a lifetime, each agent experiences only one
seasonal state, so genetic adaptation to the current state is
just as good as plasticity and the selection gradient on
plasticity disappears. When seasons are shorter than a lifetime,
plasticity pays off because it lets one agent respond to multiple
states. This is the correct DeWitt-Scheiner reading — "plasticity
tracks environmental variability at the timescale of the organism"
— and it explains why the 2026-04-17 first-pass multi-seed audit
(at `season_length = 100`) reported direction flips and a −0.084
Δdelta: that audit was testing a scenario where plasticity
*shouldn't* evolve.

The demo chunk in this vignette now uses `season_length = 10`
to demonstrate the regime where the prediction holds. Status
stays 🟠.

An **8-seed re-audit** (`dev/audit/fidelity/plasticity_8seed.R`)
of the sl=10 regime showed the 5-seed +0.014 claim was optimistic:

- 5-seed Δdelta = +0.014 (reported above)
- 8-seed Δdelta = +0.005 ± SE 0.007, *t* ≈ 0.74

Direction PASS is robust (seasonal > stable across both audits),
but the magnitude shrinks as more seeds are added — consistent
with the "signal on the edge of drift" interpretation. The
scenario is kernel-limited after all: clade's plasticity-to-sigma
coupling produces the correct sign but the effect is too small
relative to drift variance at 5–8 seeds to push past the 0.02
threshold. Genuine promotion likely needs a kernel lift (decouple
plasticity cost from BNN sigma, or a dedicated "plasticity_cost"
trait with a differential-across-individuals structure).

**Earlier 0.4.2 audit state (superseded by the above).** Full
protocol: [dev/audit/fidelity/plasticity.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/plasticity.md).

At pre-0.4.0 defaults the trajectories were flat — plasticity
barely moved in either environment (Δ ≈ −0.001/−0.002) because
sigma was tied to heterozygosity and the plasticity trait had no
fitness channel. Two kernel fixes across 0.4.0/0.4.1 addressed
that:

1. **0.4.0 Tier 5A** — `bnn_sigma_source = "trait"` couples BNN
   prior width directly to the `TRAIT_PLASTICITY` gene.
2. **0.4.1 Tier 5C** — `brain_energy_sigma_scale > 0` adds a
   log-scaled metabolic cost to sigma, creating the selection
   gradient DeWitt-Scheiner predict.

Post-fix (4 seeds × 1500 ticks under the 0.4.2 audit with
`brain_energy_sigma_scale = 0.05`):

| Condition | init → final | Δ |
|---|---|---|
| Stable | 0.301 → 0.304 | +0.003 |
| Seasonal (amp=0.7) | 0.301 → 0.306 | +0.005 |

**Direction correct (P1 PASS)**: seasonal env maintains higher
plasticity than stable — the DeWitt-Scheiner contrast. Magnitude
is modest (Δdelta = +0.002) at current parameters. The full
Hinton-Nowlan prediction (stable *canalises* downward) is not
reproduced — both envs drift slightly upward — because clade's
sigma also mediates behavioural variance, so selection against
noisy actions competes with selection against learning width
regardless of environmental stability. Stays 🟠 with direction
correct, magnitude small. See `dev/audit/fidelity/plasticity.md`
and the s-baldwin vignette for the shared kernel-limitation
caveat.

### Discovery experiments

The baseline result shows plasticity evolves downward in stable environments and is maintained at intermediate values under seasonal variation. To go beyond:

1. **Plasticity × stress hypermutation** Add `stress_hypermutation = TRUE`. Both mechanisms generate phenotypic variance under stress. Do they substitute for each other (plastic populations evolve lower hypermutation rates because plasticity already buffers stress) or accumulate (stressed populations use both mechanisms simultaneously)?

   *Tried it.* With `phenotypic_plasticity = TRUE`, `grass_rate = 0.05`, 60 agents, 200 ticks, seed 42: without hypermutation — plasticity = 0.010, n = 107. With `stress_hypermutation = TRUE` — plasticity = 0.012, n = 109. The two mechanisms are weakly additive: adding hypermutation slightly increased both plasticity and population size. They do not substitute for each other at these parameters — the mechanisms act on different timescales (within-lifetime vs genetic) and appear to complement rather than compete.

2. **Plasticity × social learning** Add `social_learning = TRUE`. Social copying propagates successful strategies; individual plasticity discovers them. Does social learning substitute for individual plasticity, causing `mean_plasticity` to decline more rapidly when social learning is available?

   *Tried it.* Four `plasticity_mutation_sd` values (0.01, 0.05, 0.10, 0.20; 50 agents, 200 ticks, seed 42): mean_plasticity = 0 in all conditions, gd = 0.182–0.188, n = 99–114. The `mean_plasticity` field is 0 throughout — the plasticity module may require explicit initialisation of the plasticity trait above 0. Without initial heritable plasticity variation, selection cannot drive the predicted canalization pattern. Run with `plasticity_init_mean = 0.5` to seed initial plasticity and observe the decline in stable environments.

3. **Plasticity × seasonality** Add `seasonal_amplitude = 0.7`. DeWitt & Scheiner (2004) predict that predictably variable environments maintain intermediate plasticity. Does `mean_plasticity` stabilise at an intermediate value under seasonality, contrasting with the monotone decline in the stable-environment baseline? Test across three amplitude values.

   *Tried it.* Plasticity + seasonality combined (50 agents, 200 ticks, seed 42): mean_plasticity = 0, gd = 0.190. Without non-zero initial plasticity, seasonal amplitude cannot maintain or evolve plasticity. The zero result is a calibration limitation: `plasticity_init_mean` must be set to a positive value before the canalization-vs-flexibility trade-off can be tested. The genetic diversity increase (0.190 vs 0.185 baseline) reflects seasonal amplitude independently of plasticity.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
