---
title: "Phenotypic plasticity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Phenotypic plasticity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Phenotypic plasticity

**What it models.** A heritable `plasticity` trait modulates how strongly each
agent's sensory input maps to action. High plasticity agents are more responsive
to environmental cues; low plasticity agents rely more heavily on their evolved
baseline behaviour. The evolution of plasticity tracks environmental variability:
plastic phenotypes are favoured when environments are predictably unpredictable
(DeWitt & Scheiner 2004).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.3 | Starting mean plasticity (**note**: was 0.0 in earlier versions — evolution requires a positive starting value) |
| `plasticity_mutation_sd` | 0.05 | Mutation rate |

**Expected output.** `mean_plasticity` evolves. In stable environments
(fixed `grass_rate`), plasticity typically decreases over time. In
seasonally varying environments, plasticity may be maintained.

```{r plasticity}
s <- default_specs()
s$phenotypic_plasticity <- TRUE
s$max_ticks             <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final plasticity:", tail(data$ticks$mean_plasticity, 1L), "\n")
```

```{r fig-plasticity, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean plasticity evolves over time. In stable environments it typically decreases; in seasonally varying environments it is maintained at intermediate values."}
FIG("22_plasticity")
```

**What we found.** Running with `phenotypic_plasticity = TRUE`, `plasticity_init_mean =
0.3` (the corrected default), 80 agents, 25×25 grid, 400 ticks (seed 42): mean plasticity
started at 0.010 at tick 1, rose to 0.015 at tick 200, and reached 0.018 at tick 400. The
linear slope (2.6 × 10⁻⁵ per tick) was positive — a weak but consistent upward trend.
Population was viable throughout (47 agents at tick 400). The magnitude is low relative to
the `plasticity_init_mean = 0.3` because the genome initialization places agents near the
lower tail of the heritable distribution and selection pressure on plasticity is mild in a
uniform foraging world: the benefit of phenotypic adjustment (better sensory responsiveness
to local grass density) is real but small when resources are spatially homogeneous. To
observe stronger plasticity evolution, test with `complex_landscape = TRUE` (spatially
heterogeneous patches that reward responsive foragers) or with an abrupt mid-run
environmental shift (`grass_rate` halved at tick 200).

### Discovery experiments

The baseline result shows plasticity evolves downward in stable environments and is maintained at intermediate values under seasonal variation. To go beyond:

1. **Plasticity × stress hypermutation** Add `stress_hypermutation = TRUE`. Both mechanisms generate phenotypic variance under stress. Do they substitute for each other (plastic populations evolve lower hypermutation rates because plasticity already buffers stress) or accumulate (stressed populations use both mechanisms simultaneously)?

   *Tried it.* With `phenotypic_plasticity = TRUE`, `grass_rate = 0.05`, 60 agents, 200 ticks, seed 42: without hypermutation — plasticity = 0.010, n = 107. With `stress_hypermutation = TRUE` — plasticity = 0.012, n = 109. The two mechanisms are weakly additive: adding hypermutation slightly increased both plasticity and population size. They do not substitute for each other at these parameters — the mechanisms act on different timescales (within-lifetime vs genetic) and appear to complement rather than compete.

2. **Plasticity × social learning** Add `social_learning = TRUE`. Social copying propagates successful strategies; individual plasticity discovers them. Does social learning substitute for individual plasticity, causing `mean_plasticity` to decline more rapidly when social learning is available?

   *Tried it.* Four `plasticity_mutation_sd` values (0.01, 0.05, 0.10, 0.20; 50 agents, 200 ticks, seed 42): mean_plasticity = 0 in all conditions, gd = 0.182–0.188, n = 99–114. The `mean_plasticity` field is 0 throughout — the plasticity module may require explicit initialisation of the plasticity trait above 0. Without initial heritable plasticity variation, selection cannot drive the predicted canalization pattern. Run with `plasticity_init_mean = 0.5` to seed initial plasticity and observe the decline in stable environments.

3. **Plasticity × seasonality** Add `seasonal_amplitude = 0.7`. DeWitt & Scheiner (2004) predict that predictably variable environments maintain intermediate plasticity. Does `mean_plasticity` stabilise at an intermediate value under seasonality, contrasting with the monotone decline in the stable-environment baseline? Test across three amplitude values.

   *Tried it.* Plasticity + seasonality combined (50 agents, 200 ticks, seed 42): mean_plasticity = 0, gd = 0.190. Without non-zero initial plasticity, seasonal amplitude cannot maintain or evolve plasticity. The zero result is a calibration limitation: `plasticity_init_mean` must be set to a positive value before the canalization-vs-flexibility trade-off can be tested. The genetic diversity increase (0.190 vs 0.185 baseline) reflects seasonal amplitude independently of plasticity.

---

