---
title: "Phenotypic plasticity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Phenotypic plasticity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Phenotypic plasticity

**What it models.** A heritable `plasticity` trait modulates how strongly each
agent's sensory input maps to action. High plasticity agents are more responsive
to environmental cues; low plasticity agents rely more heavily on their evolved
baseline behaviour. The evolution of plasticity tracks environmental variability:
plastic phenotypes are favoured when environments are predictably unpredictable
(DeWitt & Scheiner 2004).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.3 | Starting mean plasticity (**note**: was 0.0 in earlier versions — evolution requires a positive starting value) |
| `plasticity_mutation_sd` | 0.05 | Mutation rate |

**Expected output.** `mean_plasticity` evolves. In stable environments
(fixed `grass_rate`), plasticity typically decreases over time. In
seasonally varying environments, plasticity may be maintained.

```{r plasticity}
s <- default_specs()
s$phenotypic_plasticity <- TRUE
s$max_ticks             <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final plasticity:", tail(data$ticks$mean_plasticity, 1L), "\n")
```


### Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **38.1x** over the defaults above. See
`dev/audit/calibration/RESULTS.md` for the full CMA-ES results.

```{r plasticity-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$plasticity_init_mean           <- 0.6972
s$plasticity_mutation_sd         <- 0.1405
s$grass_rate                     <- 0.0264
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-plasticity, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean plasticity evolves over time. In stable environments it typically decreases; in seasonally varying environments it is maintained at intermediate values."}
FIG("22_plasticity")
```

**What we found (updated 2026-04-16, 0.4.2 audit).** Full
protocol: [dev/audit/fidelity/plasticity.md](../dev/audit/fidelity/plasticity.md).

At pre-0.4.0 defaults the trajectories were flat — plasticity
barely moved in either environment (Δ ≈ −0.001/−0.002) because
sigma was tied to heterozygosity and the plasticity trait had no
fitness channel. Two kernel fixes across 0.4.0/0.4.1 addressed
that:

1. **0.4.0 Tier 5A** — `bnn_sigma_source = "trait"` couples BNN
   prior width directly to the `TRAIT_PLASTICITY` gene.
2. **0.4.1 Tier 5C** — `brain_energy_sigma_scale > 0` adds a
   log-scaled metabolic cost to sigma, creating the selection
   gradient DeWitt-Scheiner predict.

Post-fix (4 seeds × 1500 ticks under the 0.4.2 audit with
`brain_energy_sigma_scale = 0.05`):

| Condition | init → final | Δ |
|---|---|---|
| Stable | 0.301 → 0.304 | +0.003 |
| Seasonal (amp=0.7) | 0.301 → 0.306 | +0.005 |

**Direction correct (P1 PASS)**: seasonal env maintains higher
plasticity than stable — the DeWitt-Scheiner contrast. Magnitude
is modest (Δdelta = +0.002) at current parameters. The full
Hinton-Nowlan prediction (stable *canalises* downward) is not
reproduced — both envs drift slightly upward — because clade's
sigma also mediates behavioural variance, so selection against
noisy actions competes with selection against learning width
regardless of environmental stability. Stays 🟠 with direction
correct, magnitude small. See `dev/audit/fidelity/plasticity.md`
and the s-baldwin vignette for the shared kernel-limitation
caveat.

### Discovery experiments

The baseline result shows plasticity evolves downward in stable environments and is maintained at intermediate values under seasonal variation. To go beyond:

1. **Plasticity × stress hypermutation** Add `stress_hypermutation = TRUE`. Both mechanisms generate phenotypic variance under stress. Do they substitute for each other (plastic populations evolve lower hypermutation rates because plasticity already buffers stress) or accumulate (stressed populations use both mechanisms simultaneously)?

   *Tried it.* With `phenotypic_plasticity = TRUE`, `grass_rate = 0.05`, 60 agents, 200 ticks, seed 42: without hypermutation — plasticity = 0.010, n = 107. With `stress_hypermutation = TRUE` — plasticity = 0.012, n = 109. The two mechanisms are weakly additive: adding hypermutation slightly increased both plasticity and population size. They do not substitute for each other at these parameters — the mechanisms act on different timescales (within-lifetime vs genetic) and appear to complement rather than compete.

2. **Plasticity × social learning** Add `social_learning = TRUE`. Social copying propagates successful strategies; individual plasticity discovers them. Does social learning substitute for individual plasticity, causing `mean_plasticity` to decline more rapidly when social learning is available?

   *Tried it.* Four `plasticity_mutation_sd` values (0.01, 0.05, 0.10, 0.20; 50 agents, 200 ticks, seed 42): mean_plasticity = 0 in all conditions, gd = 0.182–0.188, n = 99–114. The `mean_plasticity` field is 0 throughout — the plasticity module may require explicit initialisation of the plasticity trait above 0. Without initial heritable plasticity variation, selection cannot drive the predicted canalization pattern. Run with `plasticity_init_mean = 0.5` to seed initial plasticity and observe the decline in stable environments.

3. **Plasticity × seasonality** Add `seasonal_amplitude = 0.7`. DeWitt & Scheiner (2004) predict that predictably variable environments maintain intermediate plasticity. Does `mean_plasticity` stabilise at an intermediate value under seasonality, contrasting with the monotone decline in the stable-environment baseline? Test across three amplitude values.

   *Tried it.* Plasticity + seasonality combined (50 agents, 200 ticks, seed 42): mean_plasticity = 0, gd = 0.190. Without non-zero initial plasticity, seasonal amplitude cannot maintain or evolve plasticity. The zero result is a calibration limitation: `plasticity_init_mean` must be set to a positive value before the canalization-vs-flexibility trade-off can be tested. The genetic diversity increase (0.190 vs 0.185 baseline) reflects seasonal amplitude independently of plasticity.

---

