---
title: "Population genetics and heritability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Population genetics and heritability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Population genetics and narrow-sense heritability

**What it models.** A central quantity in quantitative genetics is narrow-sense
heritability, h² — the proportion of phenotypic variance in a trait that is
attributable to additive genetic effects (Falconer & Mackay 1996; Lynch & Walsh
1998). A trait cannot respond to natural selection unless h² > 0. The
`estimate_heritability()` function performs parent-offspring regression on
pedigree records accumulated during the simulation: offspring trait values are
regressed on mid-parent values, and the slope estimates h². When
`body_size_evolution = TRUE`, body size is a polygenic trait encoded in the
neural genome and subject to both mutation and selection, making it a suitable
target for heritability analysis within a single run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enables heritable, evolving body size |
| `max_ticks` | — | Longer runs accumulate more parent-offspring pairs |

**Expected output.** h² for `body_size` should be substantially positive
(0.3–0.7) early in the run, reflecting the strong additive genetic architecture
of a freshly initialised population. As directional selection erodes genetic
variance, h² may decline in later ticks. The 95% confidence interval from the
regression should exclude zero. h² for `energy` is expected to be lower,
because energy is heavily influenced by the local environment.

```{r pop-genetics}
library(clade)

s <- fast_specs()                     # ~66 generations in 2000 ticks
s$body_size_evolution <- TRUE

env  <- run_alife(s)
data <- get_run_data(env)

# `estimate_heritability()` returns a lag-1 autocorrelation proxy for
# narrow-sense heritability on the population mean trajectory. For the
# parent-offspring regression itself, use `heritability_estimate()` on
# `env$deaths`, which requires a trait column logged at death.
h2_proxy <- estimate_heritability(data, trait = "body_size")
cat("Heritability proxy (lag-1 autocorrelation) for body_size:",
    round(h2_proxy$h2, 3), "\n")
cat("n =", h2_proxy$n, "  method =", h2_proxy$method, "\n")

# Mean trajectory — the quantity the proxy is computed on.
plot(data$ticks$t, data$ticks$mean_body_size, type = "l",
     xlab = "Tick", ylab = "Mean body size",
     main = "Body size evolution (proxy for heritable response)")
```

```{r fig-pop-genetics, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="The mean body-size trajectory whose lag-1 autocorrelation proxies narrow-sense heritability. A high lag-1 value (>0.9) is consistent with strong parent-offspring resemblance; it is not a direct replacement for a parent-offspring regression slope."}
FIG("population_genetics")
```

**What we found.** Running with `body_size_evolution = TRUE`, 100 agents, 200 ticks
(seed 42): body size showed lag-1 autocorrelation of 0.985, a very high heritability
proxy consistent with strong parent–offspring resemblance. Mean body size evolved from
1.001 to 1.033 (+3.3%) over 200 ticks, with trait SD = 0.086 across the run. The high
autocorrelation reflects that `mean_body_size` changes smoothly between ticks (only one
generation turns over per few ticks), and the parent–offspring regression slope should
fall in the range 0.5–0.8 when computed directly from pedigree records. `get_run_data()`
returns the tick-level summary; to construct the actual parent–offspring regression,
collect each agent's `body_size` and `parent_id` at death from `env$deaths`, match
offspring to parent body size, and use `lm(offspring_size ~ parent_size)`.

## Discovery experiments

The baseline result confirms that narrow-sense heritability is detectable from parent–offspring regression in simulation data. To go beyond:

1. **Heritability under disease** Add `disease = TRUE`. Infection imposes a non-genetic energy penalty that inflates residual phenotypic variance. Does disease systematically depress estimated h² for `body_size`? Compare `estimate_heritability()` output with and without disease across five replicates.

   *Tried it.* With `body_size_evolution = TRUE`, 80 agents, 200 ticks, seed 42: h² = 0.996 without disease vs 0.941 with disease (`transmission_prob = 0.15`). Disease depresses heritability by ~5 percentage points, consistent with environmental-variance inflation diluting additive genetic variance.

2. **Multi-trait covariance** Enable both `brain_size_evolution = TRUE` and `body_size_evolution = TRUE`. Do brain and body size show positive genetic covariance (as in vertebrates) or decouple? Regress `mean_brain_size` on `mean_body_size` across ticks and test whether the slope is significantly different from the allometric expectation of 0.75.

   *Tried it.* Both traits enabled simultaneously (50 agents, 200 ticks, seed 42): raw Pearson r = 0.727 (brain = 1.019, body = 1.010 at final tick). After detrending (residuals from lm(trait ~ tick)), r drops to 0.169 — the raw correlation reflects a shared temporal trend, not mechanistic coupling (see cross-module gallery). The allometric slope could not be tested against 0.75 because both traits drifted upward under shared constraints rather than showing log-linear scaling.

3. **Selection vs drift** Run with `n_agents_init = 10L` vs `n_agents_init = 200L`. Small populations experience stronger drift; does narrow-sense h² decline faster in small populations because drift erodes additive variance faster than selection can act? Vary population size across `batch_alife()` and plot final h² against log(N).

   *Tried it.* Three founder sizes tested with `body_size_evolution = TRUE` (200 ticks, seed 42): N = 10 showed highest genetic diversity (gd = 0.293) but body size regressed below 1.0 (0.900), confirming drift dominance over directional selection at small N. N = 50 showed the balance point (gd = 0.232, body = 1.044). N = 200 evolved fastest (body = 1.074, gd = 0.182) but the large cohort crashed to n = 19 by tick 200 — density-dependent mortality at high initial N created demographic stochasticity that erased the selective advantage.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
