---
title: "Predator-prey dynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predator-prey dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained sinusoidal oscillations in which
predator abundance tracks prey abundance with a quarter-cycle lag (Lotka
1925; Volterra 1926). Crucially, that mean-field prediction assumes
**fixed, mass-action predators**. In `clade`, predators carry their own
evolving neural-network genome — they learn, mutate, and adapt across
generations. This single change is enough to break the textbook LV signature,
in a specific and well-documented way (see "What we found" below).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output (evolutionary ABM, not mean-field LV).** Watch `n_agents`
(prey) and `n_predators` over time at the calibrated parameters below. Two
distinct phases emerge:

1. **Bootstrap (t ≈ 0–30).** Initial predators have random neural genomes
   and hunt poorly. Prey climb rapidly toward their grass-imposed carrying
   capacity. Predators that survive long enough to reproduce pass on
   slightly-better-than-random hunting brains.

2. **Arms-race equilibrium (t ≈ 30+).** Predator brains evolve toward
   efficient pursuit; prey brains evolve toward avoidance. Predators
   saturate at their cap (`predator_max_agents`) and **stay there** even
   when prey are sparse — efficient hunters extract enough food from any
   non-zero prey density to maintain the cap-limited population. Prey
   continue to oscillate (driven by grass renewal × predation pressure),
   but the classic quarter-cycle predator lag of textbook LV is **muted**:
   the predator decline phase that LV requires is suppressed by ongoing
   pursuit-trait evolution.

This is the documented behaviour of evolutionary predator-prey ABMs (see
the [predator-prey fidelity report](../dev/audit/fidelity/predator_prey.md)
for cross-reference to the alifeR prototype, which contains the same
explanation in §12 of its showcase vignette). To recover textbook
sustained sinusoidal LV, you would need a fixed-policy predator (not yet
exposed as a spec) or non-toroidal patchy spatial structure (Huffaker
1958; future scenario).

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init          <- 10L
s$n_agents_init             <- 100L
s$grid_rows                 <- 30L
s$grid_cols                 <- 30L
s$predator_energy_gain      <- 30
s$predator_min_repro_energy <- 50
s$predator_max_agents       <- 100L
s$grass_rate                <- 0.20
s$max_ticks                 <- 1000L
s$random_seed               <- 42L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Evolutionary predator-prey: prey oscillate, predators arms-race to cap"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Calibrated predator-prey regime (seed 42, 1000 ticks). Predators bootstrap from 10 to their cap of 100 by tick ~30 and stay there for the remainder — the arms-race equilibrium. Prey climb to a peak of ~460 by tick 200, crash to ~250 as predator pressure builds, then continue oscillating at amplitude ±30 around a mean of ~300. The classic LV quarter-cycle predator lag is muted because evolved predators track prey efficiently regardless of density."}
FIG("14_predators")
```

**What we found.** At the calibrated parameters above, seed 42, 1000 ticks:
predators bootstrap from 10 to the cap of 100 by tick ~30 and remain
saturated for the rest of the run — this is the arms-race equilibrium,
not a kernel limit being hit (the multi-seed audit confirmed that runs
with `predator_max_agents = 800` show identical saturation behaviour at
~6–12 individuals when energy is granted only on kills). Prey climb from
150 to 466 by tick 200, crash to 245 over ticks 200–250 as cap-limited
predators consume them faster than grass renewal can replace them, then
continue oscillating around ~300 with amplitude ±30. Across 10 seeds the
prey oscillation score (magnitude of the most negative autocorrelation in
the lag-20-to-lag-100 range after tick 100) is 0.39 ± 0.14 — a
reproducible cycle. Predator oscillation score is ≈ 0 by design: once
evolved, they track prey efficiently and never enter the LV decline phase.

This is the **expected** signature of an evolutionary predator-prey ABM,
and matches the alifeR R prototype's documented behaviour (see
[`alifeR/vignettes/showcase.Rmd`](https://github.com/itchyshin/alifeR)
§12 — the package author's own explanation of why evolving predators
suppress textbook LV oscillations). It is **not** Lotka 1925's mean-field
sustained sinusoidal result, and we no longer claim to be. Full
cross-reference table and audit protocol:
[dev/audit/fidelity/predator_prey.md](dev/audit/fidelity/predator_prey.md).

### Discovery experiments

The calibrated regime reproduces the documented evolutionary-ABM
predator-prey signature: prey oscillate, predators saturate at cap.
Three open directions for going beyond:

1. **Can textbook LV be recovered with fixed-policy predators?** Lotka
   1925's sinusoidal cycles require non-evolving predators. A future
   spec (`predator_evolution = FALSE` or a fixed `policy_fn`) would let
   us test whether the pure mass-action mechanism reproduces the
   quarter-cycle lag. Currently the kernel always evolves predator
   brains, so this scenario can't yet be tested directly.

2. **Can spatial-refugia LV be recovered?** Huffaker (1958) achieved
   sustained oscillations only by breaking the grid into patches with
   dispersal barriers. With `complex_landscape = TRUE` plus
   `toroidal = FALSE` and a much larger grid, the same mechanism may
   produce sustained cycles even with evolving predators. Worth a
   dedicated `s-spatial-refugia` scenario.

3. **Prey trait evolution under predation.** Turn on
   `body_size_evolution = TRUE`. Under predation, are larger prey
   selected (harder to kill) or smaller (faster maturation)? Run 10
   seeds each at 0, 5, and 15 predators for 1000+ ticks and compare
   final `mean_body_size`.

---

