---
title: "Predator-prey dynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predator-prey dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained oscillations in which predator
abundance tracks prey abundance with a characteristic lag (Lotka 1925;
Volterra 1928). In `clade`, predators are autonomous agents with their own
energy budgets, movement rules, and reproduction thresholds, so oscillations
emerge from individual behaviour rather than being imposed by differential
equations.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output.** Watch `n_agents` (prey) and `n_predators` over time. At
the **default** parameters predators barely reproduce, so prey simply rise to
their grass-imposed carrying capacity and stay there — no interesting dynamics.
At the **calibrated** parameters in the chunk below (discovered via a grid
search over `predator_energy_gain × predator_min_repro_energy × n_predators_init
× grass_rate` in `dev/audit/fidelity/predator_prey.R`), the system produces one
clear boom-bust cycle — prey climb to ~460 while predators grow to their cap,
then crash at tick ~200 as predation pressure exceeds prey renewal — followed
by damped coexistence near a prey equilibrium of ~300. This is the classic
signature of spatially-structured Lotka-Volterra dynamics: the first cycle is
visible, but sustained oscillations are damped by local refugia (Huffaker 1958;
Comins & Hassell 1996; Murdoch et al. 1992).

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init          <- 10L
s$n_agents_init             <- 100L
s$grid_rows                 <- 30L
s$grid_cols                 <- 30L
s$predator_energy_gain      <- 30
s$predator_min_repro_energy <- 50
s$predator_max_agents       <- 100L
s$grass_rate                <- 0.20
s$max_ticks                 <- 1000L
s$random_seed               <- 42L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Predator-prey: boom-bust cycle at calibrated parameters"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Calibrated predator-prey regime (seed 42, 1000 ticks). Predators saturate at their cap by tick 30; prey climb to ~460 by tick 200, crash to ~250 as predator pressure builds, then settle into damped oscillations near 300. One clear boom-bust cycle — sustained LV oscillations are damped by spatial refugia in the agent-based model."}
FIG("14_predators")
```

**What we found.** At the calibrated parameters shown above, seed 42, 1000
ticks: predators grow from 10 to the hard cap of 100 by tick ~30 as prey
provide abundant energy; prey climb from 150 to a peak of 466 by tick 200;
prey crash to 245 over ticks 200–250 as predator pressure exceeds prey
renewal; the population then settles into damped oscillations around a mean
of ~300 prey with amplitude ±30 for the rest of the run. Across 10 seeds
(`dev/audit/fidelity/predator_prey.R`), the oscillation score — the magnitude
of the most negative autocorrelation in the lag-20-to-lag-100 range after
tick 100 — is 0.39 ± 0.14, confirming a reproducible (but damped) cycle.
Sustained LV oscillations comparable to Lotka's 1925 mean-field derivation
are **not** reproduced here: the spatial ABM dampens them via local prey
refugia, consistent with Huffaker (1958) and Comins & Hassell (1996). The
prior vignette claimed a classical 30-tick LV lag, which is not supported
by this audit; the first crash lag is ~50 ticks and subsequent cycles are
not strongly periodic. Companion fidelity report:
[dev/audit/fidelity/predator_prey.md](dev/audit/fidelity/predator_prey.md).

### Discovery experiments

The calibrated regime reproduces one boom-bust cycle + damped coexistence.
Three open questions for going beyond:

1. **Can sustained LV oscillations be recovered?** Huffaker (1958) achieved
   sustained oscillations in a spatial mite system only by breaking the grid
   into patches with dispersal barriers. Try lowering
   `predator_max_agents` from 100 back to 50 or 25, reducing
   `predator_sense_radius` to sharpen the patch scale, and raising
   `grid_rows`/`grid_cols` so local refugia persist. Measure the
   oscillation score across a grid-size sweep: does score rise with area?

2. **Prey trait evolution under predation.** Turn on `body_size_evolution =
   TRUE`. Under predation, are larger prey selected for (harder to kill)
   or smaller prey (faster maturation)? Run 10 seeds each at 0, 5, and 15
   predators for 1000+ ticks and compare final `mean_body_size`.

3. **Predator-side calibration: what makes `predator_attack_strength`
   matter?** The vignette's original "CMA-ES calibrated regime" with
   `predator_attack_strength = 1` and `grass_rate = 4.22e-06` was a
   numerical-optimisation artefact, not biology. A proper sweep of
   `predator_attack_strength` in the [10, 120] range at the calibrated
   baseline above would map the hunting-efficiency dose-response curve.

---

