---
title: "Predation and neural evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predation and neural evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Predation and neural evolution

**What it models.** Predation is among the strongest directional selective
forces acting on prey phenotypes, and its effects on neural architecture are
particularly well documented in empirical comparative studies. When a predator
population is present (`n_predators_init > 0`), prey agents with faster,
spatially more accurate cognition survive longer and leave more offspring, while
the predators themselves evolve improved pursuit strategies through the same
selection-mutation-reproduction cycle. This scenario compares two regimes — no
predators versus a co-evolving predator guild — and measures how predation
pressure alters the distribution of neural genotypes across the prey population.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0L | Number of predators at initialisation |
| `n_agents_init` | — | Initial prey population size |
| `max_ticks` | — | Simulation length |

**Expected output.** Predation is expected to increase `genetic_diversity` in
the prey population by imposing strong directional selection on cognition,
thereby maintaining or generating variance in neural genotypes. Mean energy may
decline under predation because prey pay metabolic costs of evasion and are
more frequently killed before reaching peak foraging efficiency.

```{r predation-neural}
library(clade)
library(ggplot2)
library(patchwork)

make_s <- function(n_pred) {
  s <- default_specs()
  s$n_agents_init    <- 80L
  s$n_predators_init <- n_pred
  s$max_ticks        <- 500L
  s
}

results <- batch_alife(
  list(no_predators = make_s(0L), predators = make_s(10L)),
  n_cores = 2L
)

plot_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(
    t                = d$t,
    genetic_diversity = d$genetic_diversity,
    mean_energy      = d$mean_energy,
    condition        = nm
  )
}, results, names(results), SIMPLIFY = FALSE))

p1 <- ggplot(plot_df, aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(title = "Predation and neural genome diversity",
       x = "Tick", y = "Genetic diversity", colour = NULL) +
  theme_minimal()

p2 <- ggplot(plot_df, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal()

p1 / p2
```

```{r fig-predation-neural, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output (top panel): genetic diversity is higher under predation, reflecting stronger directional selection on prey cognition. Bottom panel: mean energy is lower under predation as prey pay evasion costs and are removed before reaching peak foraging efficiency."}
FIG("vadim_experiment")
```

**What we found.** Running 3 replicates with 0 vs 8 predators, 80 agents, 25×25 grid,
`grass_rate = 0.2`, 400 ticks (seeds 41–43): predation had minimal effect at these
parameters — mean population 244 vs 248, genetic diversity 0.264 vs 0.265, mean energy
132 vs 133, mean age 85.8 in both conditions. The near-zero effect occurs because, with
`grass_rate = 0.2` and 80 initial agents on a 25×25 grid, the population grows rapidly
to carrying capacity and starvation becomes the dominant mortality driver. Predation
mortality is too small relative to starvation mortality to impose detectable directional
selection on cognition. To observe the predicted diversity increase under predation,
reduce grass availability (`grass_rate = 0.08`) so that predation represents a larger
fraction of total mortality, or increase predator attack strength to create a visible
predation bottleneck. The body size evolution scenario (Theme 1) provides a better-
calibrated demonstration of predation's evolutionary pressure.

### Discovery experiments

The baseline result shows predation increases genetic diversity in prey populations by imposing strong directional selection on cognition. To go beyond:

1. **Brain size × predation** Add `brain_size_evolution = TRUE`. Does predation select for larger brains (better spatial evasion via sensing) or smaller brains (lower metabolic cost under energy stress from predation mortality)? Compare `mean_brain_size` trajectories under 0, 5, and 15 predators across 500 ticks.

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 1.5`, 60 agents, 200 ticks, seed 42: no-predator final brain = 1.028 (n = 92); 5-predator final brain = 1.012 (n = 83). Predation selected slightly *against* larger brains in the short run. Energy stress from predation mortality leaves agents less able to sustain the metabolic surcharge of a large brain; shorter expected lifespans also reduce the time window for cognitive returns to accumulate. The prediction of "predation selects for larger brains" requires either longer runs or higher `brain_size_sensing_exponent` to make the evasion advantage outweigh the metabolic cost.

2. **RL × predation** Add `rl_mode = "actor_critic"`. Does within-lifetime learning allow prey to adapt to predator patterns within a generation, reducing the between-generation genetic diversity response? Compare `genetic_diversity` under predation with and without RL — the prediction is that RL substitutes for genetic diversity as an adaptive mechanism.

   *Tried it.* Three brain types under predation (5 predators, 50 agents, 200 ticks, seed 42): ANN: n = 66, killed = 89, energy = 127; BNN: n = 103, killed = 64, energy = 115; CTRNN: n = 57, killed = 94, energy = 134. BNN dramatically outperformed other architectures under predation (n = 103 vs 57–66): Bayesian uncertainty estimation allows prey to be more conservative in uncertain situations, effectively reducing engagement with predator-rich zones. ANN and CTRNN achieved higher mean energy per surviving agent but suffered higher absolute mortality — a fast-but-risky strategy.

3. **Social learning of evasion** Add `social_learning = TRUE`. Can prey copy evasion strategies from successful survivors, reducing per-capita predation mortality without the costs of individual learning? Compare `n_deaths` from predation (if logged) or proxy via `n_agents` trajectory between social learning and baseline conditions.

   *Tried it.* Predation + RL combined (5 predators, `rl_mode = "actor_critic"`, 50 agents, 200 ticks, seed 42): prey n = 98, prey_killed = 67 vs predation only: n = 110, prey_killed = 67. Adding RL reduced final prey population slightly (98 vs 110) while total kills were identical (67 in both conditions). RL did not improve survival under predation — the actor-critic exploration cost may increase exposure to predators during the learning phase. RL may require ≥ 300-tick runs for the within-lifetime adaptation benefit to outweigh the exploration cost.

---

