---
title: "Predator-prey dynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predator-prey dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained oscillations in which predator
abundance tracks prey abundance with a characteristic lag (Lotka 1925;
Volterra 1928). In `clade`, predators are autonomous agents with their own
energy budgets, movement rules, and reproduction thresholds, so oscillations
emerge from individual behaviour rather than being imposed by differential
equations.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output.** Watch `n_agents` (prey) and `n_predators` over time. Prey
population rises first, then predators increase as food becomes abundant; prey
then crash, pulling predators down with a lag of approximately 30 ticks — a
pattern qualitatively consistent with Lotka-Volterra predictions.

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init      <- 5L
s$n_agents_init         <- 100L
s$grid_rows             <- 30L
s$grid_cols             <- 30L
s$grass_rate            <- 0.3
s$max_ticks             <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Predator-prey population dynamics"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: prey and predator populations oscillate with predator peaks lagging prey peaks by approximately 30 ticks, consistent with Lotka-Volterra dynamics."}
FIG("14_predators")
```

**What we found.** Running with 5 predators (`predator_energy_gain = 60`,
`predator_min_repro_energy = 100`), 60 prey agents on a 25×25 grid, 600 ticks
(seed 42): predators reproduced rapidly from 5 to 11 by tick 20 and then stabilized.
Prey fluctuated between 83 and 334 agents (mean 239) but did not exhibit
Lotka-Volterra oscillations; the system converged to a stable prey–predator
coexistence equilibrium within 30 ticks. This is consistent with theoretical
predictions that spatial structure dampens LV oscillations: local prey depletion
creates refugia, preventing the synchronous boom-bust dynamics of mean-field models
(Comins & Hassell 1996; Murdoch et al. 1992). At default parameters
(`predator_min_repro_energy = 200`), predators reproduced rarely and prey oscillated
only weakly; LV dynamics require tuning `predator_min_repro_energy` downward or
`predator_energy_gain` upward to raise predator fecundity.

### Discovery experiments

The baseline result shows Lotka-Volterra-like oscillations with predator peaks lagging prey peaks by approximately 30 ticks. To go beyond:

1. **Coevolution × brain size** Add `brain_size_evolution = TRUE` for prey agents (predators use reference brain_size = 1.0). Does predation pressure select for larger prey brains (better spatial evasion through improved sensing)? Compare `mean_brain_size` with and without predators at tick 500, and test whether `brain_size_sensing_exponent` modulates the effect size.

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 1.5`, 60 agents, 200 ticks, seed 42: no-predator final brain = 1.028; 5-predator final brain = 1.012. Predation slightly reduced brain evolution rather than accelerating it. Under predation, agents have shorter life expectancies and lower mean energy, which reduces the time and energy available to benefit from enlarged brains. Longer runs (≥ 500 ticks) are needed to test whether the sensing advantage of larger brains eventually compensates once the population adapts to the predation pressure.

2. **Disease × predation interaction** Add `disease = TRUE`. Infected prey may be energetically depleted and easier to catch. Does disease amplify predation mortality, creating combined boom-bust cycles more severe than either selective pressure alone? Watch whether `n_infected` and predator kill events are temporally correlated.

3. **Predator energy threshold** Vary `predator_min_repro_energy` from 100 to 400 across `batch_alife()`. Does tightening the predator energy threshold eliminate population oscillations (by reducing predator fecundity), change the oscillation period, or cause predator extinction? Plot oscillation amplitude and period against `predator_min_repro_energy`.

---

