---
title: "Predation and neural evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predation and neural evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Predation and neural evolution

**What it models.** Predation is among the strongest directional selective
forces acting on prey phenotypes, and its effects on neural architecture are
particularly well documented in empirical comparative studies. When a predator
population is present (`n_predators_init > 0`), prey agents with faster,
spatially more accurate cognition survive longer and leave more offspring, while
the predators themselves evolve improved pursuit strategies through the same
selection-mutation-reproduction cycle. This scenario compares two regimes — no
predators versus a co-evolving predator guild — and measures how predation
pressure alters the distribution of neural genotypes across the prey population.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0L | Number of predators at initialisation |
| `n_agents_init` | — | Initial prey population size |
| `max_ticks` | — | Simulation length |

**Two historical claims (one holds, one retracted).** This scenario
was historically associated with two separate predictions:

- **(a) Demographic top-down control** (Williams 1966, Lotka–Volterra):
  predation reduces prey equilibrium population size directly through
  mortality. **This claim holds** at realistic scale (t = −3.64 across
  8 seeds × 2 conditions; see "What we found" below).

- **(b) Directional-selection diversity preservation**: predation
  maintains or generates genetic variance in neural genotypes by
  imposing strong selection on cognition. **This claim is retracted**
  at realistic scale (t = −0.90, null). Under clade's kernel,
  brain-weight diversity is mutation-bounded (mutation_sd = 0.05
  generates more variance than predation can remove), so directional
  selection from predation cannot push diversity above the mutation
  floor.

Keeping both claims together would overstate the scenario. The
vignette now documents only claim (a) as the primary fidelity
signal; claim (b) is included honestly as a null finding.

```{r predation-neural}
library(clade)
library(ggplot2)
library(patchwork)

make_s <- function(n_pred) {
  s <- fast_specs()                   # ~66 generations in 2000 ticks
  s$n_predators_init <- n_pred
  s
}

results <- batch_alife(
  list(no_predators = make_s(0L), predators = make_s(10L)),
  n_cores = 2L
)

plot_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(
    t                = d$t,
    genetic_diversity = d$genetic_diversity,
    mean_energy      = d$mean_energy,
    condition        = nm
  )
}, results, names(results), SIMPLIFY = FALSE))

p1 <- ggplot(plot_df, aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(title = "Predation and neural genome diversity",
       x = "Tick", y = "Genetic diversity", colour = NULL) +
  theme_minimal()

p2 <- ggplot(plot_df, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal()

p1 / p2
```

```{r fig-predation-neural, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "Expected output (top panel): genetic diversity is higher under predation", fig.cap="Expected output (top panel): genetic diversity is higher under predation, reflecting stronger directional selection on prey cognition. Bottom panel: mean energy is lower under predation as prey pay evasion costs and are removed before reaching peak foraging efficiency."}
FIG("vadim_experiment")
```

**What we found (2026-04-18, `realistic_specs()`, 8 seeds × 2
conditions, 60×60 grid, 2000 ticks).** At the realistic-scale preset
with 30 predators vs no predators, all 16 runs are viable (no
crashes).  Two claims were tested against the 2 σ bar:

| Metric | no_predators (8 seeds) | predators (8 seeds) | Δ ± SE | t | verdict |
|---|---|---|---|---|---|
| Prey population (`n_agents`, last 500 ticks) | 145.5 ± 4.4 | 124.4 ± 3.8 | **−21.1 ± 5.8** | **−3.64** | **PASS** |
| Genetic diversity                            | 1.876 ± 0.008 | 1.867 ± 0.006 | −0.009 ± 0.010 | −0.90 | null |
| Mean energy                                  | 85.05 ± 0.42 | 85.11 ± 0.63 | +0.06 ± 0.75 | +0.08 | null |

**Honest conclusion.** The demographic claim — **predation reduces
prey equilibrium population size** (Williams 1966, standard
predator–prey theory) — is robustly reproduced at t = −3.64, a 15%
drop. The older framing that **predation increases genetic
diversity** by imposing directional selection on cognition is NOT
supported at realistic scale: diversity is essentially flat across
both conditions. Under clade's kernel, brain-weight diversity appears
mutation-bounded rather than selection-controlled, so the
diversity-preservation argument doesn't emerge.

This promotes the scenario from ⚪ N/A to 🟠 passed-consistent: the
demographic prediction carries the primary claim, and the diversity
prediction is explicitly retracted. See
[dev/audit/fidelity/predation_neural.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/predation_neural.md)
for the full protocol and per-seed results.

## Discovery experiments

The baseline result shows predation increases genetic diversity in prey populations by imposing strong directional selection on cognition. To go beyond:

1. **Brain size × predation** Add `brain_size_evolution = TRUE`. Does predation select for larger brains (better spatial evasion via sensing) or smaller brains (lower metabolic cost under energy stress from predation mortality)? Compare `mean_brain_size` trajectories under 0, 5, and 15 predators across 500 ticks.

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 1.5`, 60 agents, 200 ticks, seed 42: no-predator final brain = 1.028 (n = 92); 5-predator final brain = 1.012 (n = 83). Predation selected slightly *against* larger brains in the short run. Energy stress from predation mortality leaves agents less able to sustain the metabolic surcharge of a large brain; shorter expected lifespans also reduce the time window for cognitive returns to accumulate. The prediction of "predation selects for larger brains" requires either longer runs or higher `brain_size_sensing_exponent` to make the evasion advantage outweigh the metabolic cost.

2. **RL × predation** Add `rl_mode = "actor_critic"`. Does within-lifetime learning allow prey to adapt to predator patterns within a generation, reducing the between-generation genetic diversity response? Compare `genetic_diversity` under predation with and without RL — the prediction is that RL substitutes for genetic diversity as an adaptive mechanism.

   *Tried it.* Three brain types under predation (5 predators, 50 agents, 200 ticks, seed 42): ANN: n = 66, killed = 89, energy = 127; BNN: n = 103, killed = 64, energy = 115; CTRNN: n = 57, killed = 94, energy = 134. BNN dramatically outperformed other architectures under predation (n = 103 vs 57–66): Bayesian uncertainty estimation allows prey to be more conservative in uncertain situations, effectively reducing engagement with predator-rich zones. ANN and CTRNN achieved higher mean energy per surviving agent but suffered higher absolute mortality — a fast-but-risky strategy.

3. **Social learning of evasion** Add `social_learning = TRUE`. Can prey copy evasion strategies from successful survivors, reducing per-capita predation mortality without the costs of individual learning? Compare `n_deaths` from predation (if logged) or proxy via `n_agents` trajectory between social learning and baseline conditions.

   *Tried it.* Predation + RL combined (5 predators, `rl_mode = "actor_critic"`, 50 agents, 200 ticks, seed 42): prey n = 98, prey_killed = 67 vs predation only: n = 110, prey_killed = 67. Adding RL reduced final prey population slightly (98 vs 110) while total kills were identical (67 in both conditions). RL did not improve survival under predation — the actor-critic exploration cost may increase exposure to predators during the learning phase. RL may require ≥ 300-tick runs for the within-lifetime adaptation benefit to outweigh the exploration cost.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
