---
title: "Predator-prey dynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predator-prey dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained sinusoidal oscillations in which
predator abundance tracks prey abundance with a quarter-cycle lag (Lotka
1925; Volterra 1926). Crucially, that mean-field prediction assumes
**fixed, mass-action predators**. In `clade`, predators carry their own
evolving neural-network genome — they learn, mutate, and adapt across
generations. This single change is enough to break the textbook LV signature,
in a specific and well-documented way (see "What we found" below).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output (evolutionary ABM, not mean-field LV).** Watch `n_agents`
(prey) and `n_predators` over time at the calibrated parameters below. Two
distinct phases emerge:

1. **Bootstrap (t ≈ 0–30).** Initial predators have random neural genomes
   and hunt poorly. Prey climb rapidly toward their grass-imposed carrying
   capacity. Predators that survive long enough to reproduce pass on
   slightly-better-than-random hunting brains.

2. **Arms-race equilibrium (t ≈ 30+).** Predator brains evolve toward
   efficient pursuit; prey brains evolve toward avoidance. Predators
   saturate at their cap (`predator_max_agents`) and **stay there** even
   when prey are sparse — efficient hunters extract enough food from any
   non-zero prey density to maintain the cap-limited population. Prey
   continue to oscillate (driven by grass renewal × predation pressure),
   but the classic quarter-cycle predator lag of textbook LV is **muted**:
   the predator decline phase that LV requires is suppressed by ongoing
   pursuit-trait evolution.

This is the documented behaviour of evolutionary predator-prey ABMs (see
the [predator-prey fidelity report](../dev/audit/fidelity/predator_prey.md)
for cross-reference to the alifeR prototype, which contains the same
explanation in §12 of its showcase vignette). To recover textbook
sustained sinusoidal LV, you would need a fixed-policy predator (not yet
exposed as a spec) or non-toroidal patchy spatial structure (Huffaker
1958; future scenario).

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init          <- 10L
s$n_agents_init             <- 100L
s$grid_rows                 <- 30L
s$grid_cols                 <- 30L
s$predator_energy_gain      <- 30
s$predator_min_repro_energy <- 50
s$predator_max_agents       <- 100L
s$grass_rate                <- 0.20
s$max_ticks                 <- 1000L
s$random_seed               <- 42L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Evolutionary predator-prey: prey oscillate, predators arms-race to cap"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Calibrated predator-prey regime (seed 42, 1000 ticks). Predators bootstrap from 10 to their cap of 100 by tick ~30 and stay there for the remainder — the arms-race equilibrium. Prey climb to a peak of ~460 by tick 200, crash to ~250 as predator pressure builds, then continue oscillating at amplitude ±30 around a mean of ~300. The classic LV quarter-cycle predator lag is muted because evolved predators track prey efficiently regardless of density."}
FIG("14_predators")
```

**What we found.** At the calibrated parameters above, seed 42, 1000 ticks:
predators bootstrap from 10 to the cap of 100 by tick ~30 and remain
saturated for the rest of the run — this is the arms-race equilibrium,
not a kernel limit being hit (the multi-seed audit confirmed that runs
with `predator_max_agents = 800` show identical saturation behaviour at
~6–12 individuals when energy is granted only on kills). Prey climb from
150 to 466 by tick 200, crash to 245 over ticks 200–250 as cap-limited
predators consume them faster than grass renewal can replace them, then
continue oscillating around ~300 with amplitude ±30. Across 10 seeds the
prey oscillation score (magnitude of the most negative autocorrelation in
the lag-20-to-lag-100 range after tick 100) is 0.39 ± 0.14 — a
reproducible cycle. Predator oscillation score is ≈ 0 by design: once
evolved, they track prey efficiently and never enter the LV decline phase.

This is the **expected** signature of an evolutionary predator-prey ABM,
and matches the alifeR R prototype's documented behaviour (see
[`alifeR/vignettes/showcase.Rmd`](https://github.com/itchyshin/alifeR)
§12 — the package author's own explanation of why evolving predators
suppress textbook LV oscillations). It is **not** Lotka 1925's mean-field
sustained sinusoidal result, and we no longer claim to be. Full
cross-reference table and audit protocol:
[dev/audit/fidelity/predator_prey.md](../dev/audit/fidelity/predator_prey.md).

### Discovery experiments

The calibrated regime reproduces the documented evolutionary-ABM
predator-prey signature: prey oscillate, predators saturate at cap.
Three open directions for going beyond:

1. **Can textbook LV be recovered?** Lotka 1925's sinusoidal cycles
   require fixed, non-adaptive predators with linear predation rates.

   *Tried it (indirectly).* Tested `predator_mutation_sd ∈ {0, 0.1,
   1.0, 5.0}` × 5 seeds × 1000 ticks, plus `predator_max_agents ∈
   {20, 50, 200, 500}` × 2 mutation rates × 5 seeds × 1500 ticks.
   Also checked cross-correlation for the LV quarter-cycle lag.

   Prey oscillation scores range 0.25–0.34 across all regimes (moderate,
   no textbook sinusoidal). Predator populations have **zero variance**
   after the bootstrap phase (sd = 0 in every condition): once evolved
   past random, predators efficiently hunt at any non-zero prey density
   and saturate at cap. Cross-correlation between prey and predator
   timeseries cannot be computed (predator series is constant).

   **Why.** Evolved neural-network predators are adaptively nonlinear —
   they track prey efficiently regardless of density. This removes the
   quarter-cycle phase lag LV requires (textbook model assumes density-
   proportional predation). The predator-decline phase that creates the
   second half of the LV cycle never occurs because evolved predators
   are too smart to fail at hunting. This is biologically realistic
   (real predators with behavioural plasticity also suppress LV) but
   means textbook LV needs a `predator_evolution = FALSE` kernel spec
   that freezes predator brains. Deferred to 0.6+.

2. **Oscillation strength vs predator density.** Does increasing
   predator density produce stronger LV-like oscillations? Measured
   oscillation score (magnitude of most negative autocorrelation at
   lags 20–100 after tick 50) across 4 predator densities, 10 seeds
   each, 500 ticks.

   *Tried it.* `n_predators_init ∈ {5, 10, 20, 40}`, 100 prey, 30×30
   grid, `grass_rate = 0.15`: oscillation scores = 0.217, 0.248,
   0.221, 0.257. Moderate oscillation (~0.22–0.26) across all
   densities with no strong trend. The oscillation amplitude is
   insensitive to predator count because evolved predators
   efficiently track prey regardless of density — they saturate
   behaviourally rather than dynamically. This is why clade produces
   damped LV rather than sustained: the predator arms race removes
   the phase lag that sustains textbook sinusoidal cycles.

3. **Prey body-size evolution under predation.** Does predation
   select for larger or smaller prey? Run 5 seeds each at 0, 5, and
   15 predators for 600 ticks with `body_size_evolution = TRUE`.

   *Tried it.* `n_pred=0`: Δbody_size = +0.080 ± 0.014;
   `n_pred=5`: Δ = +0.082 ± 0.021; `n_pred=15`: Δ = +0.115 ± 0.007.
   All three conditions drift upward (Cope's rule). At 5 seeds the
   predation-accelerates trend (n_pred=15 > n_pred=0) is suggestive,
   but the 0.5.2 body-size P2 16-seed audit showed this direction
   is NOT statistically robust — both the "predation accelerates"
   and "predation slows" framings are inside the 5-seed noise band.
   The honest finding: predation does not modulate body-size drift
   at default parameters. See [s-body-size](s-body-size.html) for
   the full 16-seed resolution.

4. **Spatial refugia — Huffaker (1958).** Huffaker (1958) achieved
   sustained oscillations by breaking the grid into patches with
   dispersal barriers. clade actually exposes `toroidal` as a
   first-class spec (`TRUE` = wraparound, `FALSE` = clamped
   boundaries), so we can factorially cross it with
   `complex_landscape` to test the Huffaker prediction cleanly.

   *Tried it.* 2×2 factorial of
   `toroidal ∈ {TRUE, FALSE}` × `complex_landscape ∈ {FALSE, TRUE}`,
   50×50 grid, 250 prey init, 25 predators init,
   `max_agents = 1000` (lifted so cycling isn't clipped by the cap),
   5 seeds × 800 ticks each. Prey oscillation scores:

   | toroidal | complex_landscape | prey_osc (mean ± sd) | prey_mean |
   |---|---|---|---|
   | TRUE  | FALSE (flat)   | 0.299 ± 0.032 | 643 |
   | TRUE  | TRUE (patchy)  | **0.637 ± 0.009** | 841 |
   | FALSE | FALSE (flat)   | 0.000 ± 0.000 | 276 |
   | FALSE | TRUE (patchy)  | 0.000 ± 0.000 | 348 |

   Two surprises, neither of which is the textbook Huffaker result.

   **(a) Complex landscape with toroidal wrap DOUBLES the oscillation
   score.** At 0.64 ± 0.009 this is the strongest, cleanest LV-like
   cycling any regime of clade has produced. The habitat patches
   (shrubs + canopy) create local heterogeneity in predator-encounter
   rate while the toroidal wrap keeps populations well-mixed globally.
   Interestingly, prey mean is also *higher* under patchy habitat
   (841 vs 643) — the refugia effect raises carrying capacity, which
   is the opposite of what a "refugia let predators cycle off and
   lose prey" reading would suggest.

   **(b) Non-toroidal kills all cycling** (osc = 0.00 in both
   landscape conditions). This is *not* a clean Huffaker refutation
   because the kernel's non-toroidal implementation clamps at
   boundaries (`wrap_or_clamp` → `clamp(x, 1, n)` in
   `inst/julia/src/Clade.jl:48`) — agents that try to step across
   the edge end up pinned at the edge rather than having their move
   rejected. The result is pile-up at the perimeter, which halves
   the effective prey population (276 vs 643) and damps cycling
   monotonically. A proper Huffaker test needs reflective boundaries
   (reject-if-blocked), not clamp. Filed as a kernel-polish item.

   Net finding: **patchy habitat within a connected (toroidal) world
   amplifies LV cycling**, which is the first mechanism in clade we
   have found that partially defeats the arms-race absorption seen
   in Exps 1–3 and 6. This is the cleanest positive LV-enhancing
   manipulation this scenario has surfaced and deserves a dedicated
   `s-spatial-heterogeneity` follow-up.

5. **Grass density × predator density — does LV damping depend on
   resources?** LV damping should be strongest when prey are
   resource-limited (small amplitude) and weakest when prey are
   resource-rich (runaway growth, crashes). Tested
   `grass_rate ∈ {0.05, 0.15, 0.30}` × `n_predators_init ∈ {5, 20}`
   × 5 seeds × 800 ticks.

   *Tried it.* Prey oscillation scores (mean ± sd across 5 seeds):

   | grass_rate | 5 preds     | 20 preds    |
   |---|---|---|
   | 0.05 (scarce) | 0.19 ± 0.07 | 0.24 ± 0.02 |
   | 0.15 (moderate) | **0.32 ± 0.04** | **0.32 ± 0.01** |
   | 0.30 (rich)   | 0.28 ± 0.03 | 0.27 ± 0.05 |

   The result is **non-monotone in grass density**: oscillation
   peaks at moderate grass (~0.15) and damps at both extremes. With
   scarce grass (0.05), prey stay near ~40 individuals — too few
   for coherent cycling. With rich grass (0.30), prey saturate near
   ~315 — near the ceiling, with little room to oscillate. The
   moderate regime (grass = 0.15, prey ~190) sits in the sweet spot
   where carrying-capacity × predation pressure produces the
   cleanest cycle. Predator initial count has negligible effect
   within each grass level (the arms-race saturation of Exp 2 holds:
   evolved predators fill the cap regardless of starting number).

6. **Group defense × LV — Hamilton (1971) dilution.** Hamilton's
   selfish-herd and dilution arguments predict that group defense
   reduces per-prey attack risk, which should dampen LV oscillations
   by reducing predator efficiency. Tested `group_defense ∈ {FALSE,
   TRUE}` (radius 2, strength 0.3) × 5 seeds × 800 ticks at 20
   predators.

   *Tried it.* Prey oscillation score: 0.307 ± 0.013 (off) vs
   0.300 ± 0.020 (on). Effectively identical (difference = 0.007,
   inside the 0.02 standard deviation of either condition). Prey
   mean size: 241 vs 245 — also effectively unchanged. Why?
   Evolved predators appear to compensate behaviourally for the
   per-attack risk reduction: the hunt-trait evolution already
   tracks prey availability density by density, so a uniform
   reduction in attack success is absorbed by the same arms-race
   saturation mechanism that flattens the predator-count effect in
   Exp 2. A Hamiltonian dilution signal would likely require either
   (a) frozen predators (no arms-race absorption) or (b) variance
   in defense *within* the prey population so that the
   differential benefit of clustering can be detected genetically.
   Neither is currently exposed.

---

