---
title: "Predator-prey dynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Predator-prey dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained sinusoidal oscillations in which
predator abundance tracks prey abundance with a quarter-cycle lag (Lotka
1925; Volterra 1926). Crucially, that mean-field prediction assumes
**fixed, mass-action predators**. In `clade`, predators carry their own
evolving neural-network genome — they learn, mutate, and adapt across
generations. This single change is enough to break the textbook LV signature,
in a specific and well-documented way (see "What we found" below).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output (evolutionary ABM, not mean-field LV).** Watch `n_agents`
(prey) and `n_predators` over time at the calibrated parameters below. Two
distinct phases emerge:

1. **Bootstrap (t ≈ 0–30).** Initial predators have random neural genomes
   and hunt poorly. Prey climb rapidly toward their grass-imposed carrying
   capacity. Predators that survive long enough to reproduce pass on
   slightly-better-than-random hunting brains.

2. **Arms-race equilibrium (t ≈ 30+).** Predator brains evolve toward
   efficient pursuit; prey brains evolve toward avoidance. Predators
   saturate at their cap (`predator_max_agents`) and **stay there** even
   when prey are sparse — efficient hunters extract enough food from any
   non-zero prey density to maintain the cap-limited population. Prey
   continue to oscillate (driven by grass renewal × predation pressure),
   but the classic quarter-cycle predator lag of textbook LV is **muted**:
   the predator decline phase that LV requires is suppressed by ongoing
   pursuit-trait evolution.

This is the documented behaviour of evolutionary predator-prey ABMs (see
the [predator-prey fidelity report](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/predator_prey.md)
for cross-reference to the alifeR prototype, which contains the same
explanation in §12 of its showcase vignette). To recover textbook
sustained sinusoidal LV, you would need a fixed-policy predator (not yet
exposed as a spec) or non-toroidal patchy spatial structure (Huffaker
1958; future scenario).

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init          <- 10L
s$n_agents_init             <- 100L
s$grid_rows                 <- 30L
s$grid_cols                 <- 30L
s$predator_energy_gain      <- 30
s$predator_min_repro_energy <- 50
s$predator_max_agents       <- 100L
s$grass_rate                <- 0.20
s$max_ticks                 <- 1000L
s$random_seed               <- 42L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Evolutionary predator-prey: prey oscillate, predators arms-race to cap"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Calibrated predator-prey regime (seed 42, 1000 ticks). Predators bootstrap from 10 to their cap of 100 by tick ~30 and stay there for the remainder — the arms-race equilibrium. Prey climb to a peak of ~460 by tick 200, crash to ~250 as predator pressure builds, then continue oscillating at amplitude ±30 around a mean of ~300. The classic LV quarter-cycle predator lag is muted because evolved predators track prey efficiently regardless of density."}
FIG("14_predators")
```

**What we found.** At the calibrated parameters above, seed 42, 1000 ticks:
predators bootstrap from 10 to the cap of 100 by tick ~30 and remain
saturated for the rest of the run — this is the arms-race equilibrium,
not a kernel limit being hit (the multi-seed audit confirmed that runs
with `predator_max_agents = 800` show identical saturation behaviour at
~6–12 individuals when energy is granted only on kills). Prey climb from
150 to 466 by tick 200, crash to 245 over ticks 200–250 as cap-limited
predators consume them faster than grass renewal can replace them, then
continue oscillating around ~300 with amplitude ±30. Across 10 seeds the
prey oscillation score (magnitude of the most negative autocorrelation in
the lag-20-to-lag-100 range after tick 100) is 0.39 ± 0.14 — a
reproducible cycle. Predator oscillation score is ≈ 0 by design: once
evolved, they track prey efficiently and never enter the LV decline phase.

This is the **expected** signature of an evolutionary predator-prey ABM,
and matches the alifeR R prototype's documented behaviour (see
[`alifeR/vignettes/showcase.Rmd`](https://github.com/itchyshin/alifeR)
§12 — the package author's own explanation of why evolving predators
suppress textbook LV oscillations). It is **not** Lotka 1925's mean-field
sustained sinusoidal result, and we no longer claim to be. Full
cross-reference table and audit protocol:
[dev/audit/fidelity/predator_prey.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/predator_prey.md).

### Discovery experiments

The calibrated regime reproduces the documented evolutionary-ABM
predator-prey signature: prey oscillate, predators saturate at cap.
Three open directions for going beyond:

1. **Can textbook LV be recovered?** Lotka 1925's sinusoidal cycles
   require per-capita attack rates that depend on density, not on
   accumulated hunting skill. Natural LV systems (lynx–hare, Gause's
   *Didinium*–*Paramecium*) work because predator hunting efficiency
   evolves *slowly* relative to the cycle period.

   *Tried it (4 conditions × 5 seeds, 2000 ticks, realistic_specs).*

   - **evolving** (default): `predator_mutation_sd = 0.1`,
     `predator_max_age = 60` — predators stabilise at ~30 within 30
     ticks; `pred_var = 0` from t=500 onwards. Arms-race-to-equilibrium.
   - **frozen**: `predator_mutation_sd = 0` — predators stabilise
     at ~30; `pred_var = 0`. Freezing mutation doesn't help because
     selection on initial brain variance still runs to completion.
   - **longlived** (slow pace-of-life): `predator_max_age = 300`
     (10× prey lifespan; lynx > hare). Also `pred_var = 0`. Slower
     evolution still produces saturation.
   - **fast_turnover**: short-lived + easy reproduction
     (`predator_max_age = 20`, `predator_min_repro_energy = 60`,
     `predator_mutation_sd = 0`). Predators now cycle — but the cycle
     is a boom-bust runaway, not LV: population explodes from 40 to
     10,000 (cap) by t ≈ 900 while prey crash from 150 to ~5. The
     phase plot shows a one-way trajectory, not a closed LV loop.

   **Why textbook LV doesn't emerge in any of these.** LV requires
   `dP/dt ∝ prey × predator` — **predator births proportional to prey
   density**. clade's predator reproduction is energy-threshold-gated:
   a predator reproduces when its own energy exceeds
   `predator_min_repro_energy`, regardless of current prey density.
   That produces either **saturation** (if the threshold is hard,
   easy predators can always find enough to top up) or **runaway**
   (if the threshold is easy, boom until prey extinction), but never
   density-matched cycling. To get LV, you would need either:

   - `predator_repro_rate = kP * mean(prey_density_in_territory)`
     — explicit density-dependent reproduction rate, OR
   - A chemostat-style steady-state where predator mortality is
     forced to track prey scarcity (e.g. a harsh per-tick metabolic
     cost that exceeds what marginal kills can cover).

   Neither is in the current kernel. The lynx-hare cycles exist in
   nature because lynx reproduction IS density-dependent through
   gestation/provisioning constraints; clade's predator module is
   more like a bacterial predator (constant metabolic cost, eat-
   when-encountered) than a mammalian carnivore. Deferred as a
   potential 0.6+ kernel extension.

   Companion figure: `vignettes/figures/showcase_14b_predators_lv_comparison.png`
   (4-panel time series + phase plot of the fast_turnover boom-bust).
   Audit script: [`predator_prey_lv.R`](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/predator_prey_lv.R).

2. **Oscillation strength vs predator density.** Does increasing
   predator density produce stronger LV-like oscillations? Measured
   oscillation score (magnitude of most negative autocorrelation at
   lags 20–100 after tick 50) across 4 predator densities, 10 seeds
   each, 500 ticks.

   *Tried it.* `n_predators_init ∈ {5, 10, 20, 40}`, 100 prey, 30×30
   grid, `grass_rate = 0.15`: oscillation scores = 0.217, 0.248,
   0.221, 0.257. Moderate oscillation (~0.22–0.26) across all
   densities with no strong trend. The oscillation amplitude is
   insensitive to predator count because evolved predators
   efficiently track prey regardless of density — they saturate
   behaviourally rather than dynamically. This is why clade produces
   damped LV rather than sustained: the predator arms race removes
   the phase lag that sustains textbook sinusoidal cycles.

3. **Prey body-size evolution under predation.** Does predation
   select for larger or smaller prey? Run 5 seeds each at 0, 5, and
   15 predators for 600 ticks with `body_size_evolution = TRUE`.

   *Tried it.* `n_pred=0`: Δbody_size = +0.080 ± 0.014;
   `n_pred=5`: Δ = +0.082 ± 0.021; `n_pred=15`: Δ = +0.115 ± 0.007.
   All three conditions drift upward (Cope's rule). At 5 seeds the
   predation-accelerates trend (n_pred=15 > n_pred=0) is suggestive,
   but the 0.5.2 body-size P2 16-seed audit showed this direction
   is NOT statistically robust — both the "predation accelerates"
   and "predation slows" framings are inside the 5-seed noise band.
   The honest finding: predation does not modulate body-size drift
   at default parameters. See [s-body-size](s-body-size.html) for
   the full 16-seed resolution.

4. **Spatial heterogeneity and resource layering.** Two separate
   theoretical predictions about spatial structure: Huffaker (1958)
   said dispersal barriers between patches sustain predator-prey
   cycles, and Rosenzweig (1971) — "paradox of enrichment" — said
   richer prey carrying capacity *amplifies* oscillations even in
   a well-mixed world. clade exposes `toroidal` as a first-class
   spec (`TRUE` = wraparound, `FALSE` = clamped boundaries) and a
   `complex_landscape` module that adds shrub + canopy resource
   layers on top of ground grass. The 2×2 of these tests both
   predictions at once.

   *Tried it.* 2×2 factorial of
   `toroidal ∈ {TRUE, FALSE}` × `complex_landscape ∈ {FALSE, TRUE}`,
   50×50 grid, 250 prey init, 25 predators init,
   `max_agents = 1000` (lifted so cycling isn't clipped by the cap),
   5 seeds × 800 ticks each. Prey oscillation scores:

   | toroidal | complex_landscape | prey_osc (mean ± sd) | prey_mean |
   |---|---|---|---|
   | TRUE  | FALSE (flat)   | 0.299 ± 0.032 | 643 |
   | TRUE  | TRUE (patchy)  | **0.637 ± 0.009** | 841 |
   | FALSE | FALSE (flat)   | 0.000 ± 0.000 | 276 |
   | FALSE | TRUE (patchy)  | 0.000 ± 0.000 | 348 |

   Two surprises, and the mechanism is neither a clean Huffaker
   nor a clean Rosenzweig.

   **(a) `complex_landscape` within a toroidal world DOUBLES the
   oscillation score at 50×50** (0.338 → 0.650) and raises the prey
   equilibrium (663 → 841). But a follow-up 30×30 replicate shows
   the effect is **grid-scale dependent**:

   | grid  | landscape | prey_osc (mean ± sd) | prey_mean |
   |-------|-----------|----------------------|-----------|
   | 30×30 | flat      | 0.307 ± 0.048 | 240 |
   | 30×30 | patchy    | 0.310 ± 0.024 | 308 |
   | 50×50 | flat      | 0.338 ± 0.022 | 663 |
   | 50×50 | patchy    | **0.650 ± 0.013** | 841 |

   At 30×30, patchy vs flat gives essentially no change in cycling
   (0.307 vs 0.310). The amplification appears only at 50×50. This
   rules out pure Rosenzweig-1971 enrichment (which would amplify at
   any scale once K is raised) and instead points to a spatial-
   decoupling mechanism: `complex_landscape` adds resource layers
   (shrubs + canopy — see `inst/julia/src/modules/complex_landscape.jl`)
   that create local resource-density pockets, and those pockets
   only matter when the grid is large enough that predator and prey
   trajectories can desynchronise across them. At 30×30 the board
   is too small for predators and prey to "lose" each other between
   patches; at 50×50 the patches are far enough apart that local
   predator depletion and prey recovery run out of phase.

   This is clade's first mechanism that partially defeats the
   predator arms-race absorption seen in Exps 1–3 and 6. It is also
   the strongest, cleanest LV-like cycling any clade regime has
   produced — and it requires *both* resource layering *and* enough
   grid area for the layering to create effective spatial decoupling.

   **(b) Non-toroidal zeros all sustained cycling** (osc = 0.00 in
   both landscape conditions). This is *not* a clean Huffaker
   refutation because the kernel's non-toroidal implementation
   clamps at boundaries (`wrap_or_clamp` → `clamp(x, 1, n)` in
   `inst/julia/src/Clade.jl:48`) — agents that try to step across
   the edge end up pinned at the edge rather than having their move
   rejected. The result is pile-up at the perimeter, which halves
   the effective prey population (276 vs 643) and damps the ACF
   monotonically (see the figure below: the bounded panels have
   visible early transients but a declining trend dominates so the
   autocorrelation never dips below zero). A proper Huffaker test
   needs reflective boundaries (reject-if-blocked), not clamp.
   Filed as a kernel-polish item in `dev/docs/kernel-as-biology/
   tick.md §6 Variants`.

   Net finding: **resource layering within a connected (toroidal)
   world amplifies LV cycling** — clade's first clean Rosenzweig
   1971 enrichment signature, and the first manipulation that
   partially defeats the predator arms-race absorption seen in
   every other experiment here. Deserves a dedicated
   `s-spatial-heterogeneity` scenario. The Huffaker 1958 refugia
   prediction remains untested until the reject-if-blocked
   boundary is implemented.

   ```{r fig-refugia-trajectories, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Prey trajectories across the 2×2 Huffaker audit (seed 7, 800 ticks, max_agents = 1000). Top-right (toroidal + patchy) is the only condition with sustained cycling — three clear peak-trough excursions. Top-left (toroidal + flat) shows one overshoot-and-damp. The bottom row (bounded) both show one transient excursion followed by monotonic decline; the ACF-based oscillation score reads 0.00 because the declining trend dominates the autocorrelation even though visible mini-cycles are present early on."}
   FIG("14_predators_refugia")
   ```

5. **Grass density × predator density — does LV damping depend on
   resources?** LV damping should be strongest when prey are
   resource-limited (small amplitude) and weakest when prey are
   resource-rich (runaway growth, crashes). Tested
   `grass_rate ∈ {0.05, 0.15, 0.30}` × `n_predators_init ∈ {5, 20}`
   × 5 seeds × 800 ticks.

   *Tried it.* Prey oscillation scores (mean ± sd across 5 seeds):

   | grass_rate | 5 preds     | 20 preds    |
   |---|---|---|
   | 0.05 (scarce) | 0.19 ± 0.07 | 0.24 ± 0.02 |
   | 0.15 (moderate) | **0.32 ± 0.04** | **0.32 ± 0.01** |
   | 0.30 (rich)   | 0.28 ± 0.03 | 0.27 ± 0.05 |

   The result is **non-monotone in grass density**: oscillation
   peaks at moderate grass (~0.15) and damps at both extremes. With
   scarce grass (0.05), prey stay near ~40 individuals — too few
   for coherent cycling. With rich grass (0.30), prey saturate near
   ~315 — near the ceiling, with little room to oscillate. The
   moderate regime (grass = 0.15, prey ~190) sits in the sweet spot
   where carrying-capacity × predation pressure produces the
   cleanest cycle. Predator initial count has negligible effect
   within each grass level (the arms-race saturation of Exp 2 holds:
   evolved predators fill the cap regardless of starting number).

6. **Group defense × LV — Hamilton (1971) dilution.** Hamilton's
   selfish-herd and dilution arguments predict that group defense
   reduces per-prey attack risk, which should dampen LV oscillations
   by reducing predator efficiency. Tested `group_defense ∈ {FALSE,
   TRUE}` (radius 2, strength 0.3) × 5 seeds × 800 ticks at 20
   predators.

   *Tried it.* Prey oscillation score: 0.307 ± 0.013 (off) vs
   0.300 ± 0.020 (on). Effectively identical (difference = 0.007,
   inside the 0.02 standard deviation of either condition). Prey
   mean size: 241 vs 245 — also effectively unchanged. Why?
   Evolved predators appear to compensate behaviourally for the
   per-attack risk reduction: the hunt-trait evolution already
   tracks prey availability density by density, so a uniform
   reduction in attack success is absorbed by the same arms-race
   saturation mechanism that flattens the predator-count effect in
   Exp 2. A Hamiltonian dilution signal would likely require either
   (a) frozen predators (no arms-race absorption) or (b) variance
   in defense *within* the prey population so that the
   differential benefit of clustering can be detected genetically.
   Neither is currently exposed.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
