---
title: "Within-lifetime reinforcement learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Within-lifetime reinforcement learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Within-lifetime reinforcement learning

**What it models.** Each agent uses REINFORCE-with-baseline (actor-critic) to
update its brain's output layer based on energy reward signals during its
lifetime. This implements the Baldwin effect: individually-learned behaviours
can stabilise genetically over generations because learning relaxes the
requirement for genes to encode the exact optimal behaviour.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `rl_mode` | `"none"` | Set to `"actor_critic"` to enable |
| `rl_update_freq` | 5 | Ticks between RL updates |
| `learning_rate` | 0.01 | Step size for output-layer update |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve |

**Expected output.** Agents with RL enabled adapt within their lifetime to the
local resource distribution. When `learning_rate_evolution = TRUE`, the
evolved learning rate reflects the environmental uncertainty: noisy environments
favour lower rates; stable environments favour higher rates.

```{r rl}
s <- default_specs()
s$rl_mode   <- "actor_critic"
s$max_ticks <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-rl, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: agents with within-lifetime RL adapt to local resource distribution within their lifetime. The Baldwin Effect is visible as the BNN prior sigma narrows over generations."}
FIG("10_rl")
```

**What we found.** Running 3 replicates with `rl_mode = "actor_critic"` vs `"none"`,
80 agents, 25×25 grid, `grass_rate = 0.15`, 400 ticks (seeds 41–43): mean population
was 208 (RL) vs 207 (no RL); mean energy 124.8 vs 124.85. BNN prior sigma rose from
~0.30 early to ~0.50 late in *both* conditions — the opposite of the expected Baldwin
Effect canalization. At `learning_rate = 0.01` and `rl_update_freq = 5`, each REINFORCE
gradient step is small relative to the BNN's per-tick weight resampling from a broad
sigma prior. The copied delta is diluted before the next update, preventing sustained
policy improvement. The Baldwin Effect (sigma narrowing through genetic assimilation)
is predicted to require: (i) runs longer than 400 ticks to allow genetic fixation; or
(ii) a narrower initial sigma that is itself heritable and under selection. A dedicated
BNN canalization scenario (see below) uses a 500-tick run that is more likely to reveal
the sigma trajectory.

### Discovery experiments

The baseline result shows agents with RL adapt within their lifetime to local resource distribution, and the Baldwin Effect is visible as BNN prior sigma narrows. To go beyond:

1. **RL × brain size** Add `brain_size_evolution = TRUE`. Does within-lifetime RL reduce selection pressure on brain size (because learning compensates for cognitive limitations) or amplify it (because larger-brained agents learn more efficiently via the sensing advantage)? Compare `mean_brain_size` at tick 500 with and without `rl_mode = "actor_critic"`.

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 1.5`, 60 agents, 200 ticks, seed 42: no-RL final brain = 1.034 (n = 88); with RL final brain = 1.012 (n = 97). RL *reduced* selection pressure on genetic brain size: agents compensate for moderate cognitive limitations by learning within their lifetimes, weakening the fitness gradient for genetically encoded large brains. Population size increased under RL (+10%), but brain size decreased — RL and genetic brain evolution are substitutable rather than complementary at these parameters.

2. **RL × social learning** Add `social_learning = TRUE`. Do agents that can both learn individually (RL) and copy socially converge faster than either mechanism alone? Watch `mean_energy` and `genetic_diversity`: does the combination flatten genetic diversity by reducing between-individual variance in foraging success more than either mechanism separately?

3. **RL under seasonality** Add `seasonal_amplitude = 0.8`. Does within-lifetime RL improve survival during resource troughs by allowing agents to adapt their foraging strategy to winter conditions within a single season? Compare mean agent survival rates across the seasonal cycle with and without RL.

---

