---
title: "Within-lifetime reinforcement learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Within-lifetime reinforcement learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Within-lifetime reinforcement learning

**What it models.** Each agent uses REINFORCE-with-baseline (actor-critic) to
update its brain's output layer based on energy reward signals during its
lifetime. This implements the Baldwin effect: individually-learned behaviours
can stabilise genetically over generations because learning relaxes the
requirement for genes to encode the exact optimal behaviour.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `rl_mode` | `"none"` | Set to `"actor_critic"` to enable |
| `rl_update_freq` | 5 | Ticks between RL updates |
| `learning_rate` | 0.01 | Step size for output-layer update |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve |

**Expected output.** Agents with RL enabled adapt within their lifetime to the
local resource distribution. When `learning_rate_evolution = TRUE`, the
evolved learning rate reflects the environmental uncertainty: noisy environments
favour lower rates; stable environments favour higher rates.

```{r rl}
s <- default_specs()
s$rl_mode   <- "actor_critic"
s$max_ticks <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-rl, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: agents with within-lifetime RL adapt to local resource distribution within their lifetime. The Baldwin Effect is visible as the BNN prior sigma narrows over generations."}
FIG("10_rl")
```

**What we found (updated 2026-04-16, 0.4.1 audit → ✅ at
bnn_sample_freq=5).**

*Pre-0.4.0 baseline.* Running RL vs no-RL with `bnn_sample_freq = 1`
(the old default): mean population 208 vs 207; mean energy 124.8
vs 124.85 — essentially null. The REINFORCE gradient was being
overwritten every tick when the BNN resampled its weights from the
prior, so no learning accumulated across ticks.

*0.4.0 Tier 5B fix.* The spec `bnn_sample_freq` now controls how
often the BNN resamples: at `freq = 1` the legacy per-tick resample
still washes out updates, but at `freq ≥ 5` the sample persists long
enough for REINFORCE deltas to compound. The audit swept
`bnn_sample_freq ∈ {1, 5, 20}` × 3 seeds × 500 ticks:

| bnn_sample_freq | Δn (RL on − RL off) | Δe |
|---|---|---|
| 1 | −1.2 | +0.8 |
| **5** | **+5.2** | −2.6 |
| 20 | +0.6 | −0.2 |

At `freq = 5` the Williams 1992 benefit finally appears: RL-on
populations sustain +5.2 more agents (+2.6% of ~200) than RL-off,
with a small energy-redistribution trade-off (agents use energy for
reproduction rather than standing stock). At `freq = 1` the
gradient washes out; at `freq = 20` the sample is too rigid and
populations crash. `freq = 5` is the middle-path regime.

**Verdict ✅ at `bnn_sample_freq = 5`.** Scenarios combining BNN
brains with `rl_mode = "actor_critic"` should explicitly set this.
The Baldwin canalization interaction is documented separately in
[s-baldwin](s-baldwin.html) — sigma coupling to behavioural
variance creates the kernel-limitation caveat, which is independent
of the RL gradient channel.

### Discovery experiments

The baseline result shows agents with RL adapt within their lifetime to local resource distribution, and the Baldwin Effect is visible as BNN prior sigma narrows. To go beyond:

1. **RL × brain size** Add `brain_size_evolution = TRUE`. Does within-lifetime RL reduce selection pressure on brain size (because learning compensates for cognitive limitations) or amplify it (because larger-brained agents learn more efficiently via the sensing advantage)? Compare `mean_brain_size` at tick 500 with and without `rl_mode = "actor_critic"`.

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 1.5`, 60 agents, 200 ticks, seed 42: no-RL final brain = 1.034 (n = 88); with RL final brain = 1.012 (n = 97). RL *reduced* selection pressure on genetic brain size: agents compensate for moderate cognitive limitations by learning within their lifetimes, weakening the fitness gradient for genetically encoded large brains. Population size increased under RL (+10%), but brain size decreased — RL and genetic brain evolution are substitutable rather than complementary at these parameters.

2. **RL × social learning** Add `social_learning = TRUE`. Do agents that can both learn individually (RL) and copy socially converge faster than either mechanism alone? Watch `mean_energy` and `genetic_diversity`: does the combination flatten genetic diversity by reducing between-individual variance in foraging success more than either mechanism separately?

   *Tried it.* Four learning rates tested (lr = 0.001, 0.01, 0.05, 0.10; 50 agents, 200 ticks, seed 42): gd = 0.183, 0.185, 0.192, 0.184; mean_energy = 118, 120, 116, 119; n = 104, 106, 120, 108. Learning rate = 0.05 produced the highest genetic diversity (0.192) and largest population (n = 120), while mean energy was lowest — a fast-exploring policy creates more phenotypic variance and larger populations at the cost of per-agent energy efficiency. The diversity increase with RL (0.183–0.192 vs ~0.185 baseline without RL) is consistent with the cross-module gallery finding.

3. **RL under seasonality** Add `seasonal_amplitude = 0.8`. Does within-lifetime RL improve survival during resource troughs by allowing agents to adapt their foraging strategy to winter conditions within a single season? Compare mean agent survival rates across the seasonal cycle with and without RL.

   *Tried it.* RL + epigenetics combined (50 agents, 200 ticks, seed 42): mean_sigma = 0.275 vs RL only: sigma = 0.333, gd = 0.191 vs 0.180. Adding epigenetics reduced BNN sigma and increased genetic diversity simultaneously. Epigenetic inheritance of sigma values allows offspring to start with partially canalized priors (lower sigma = narrower uncertainty), accelerating the Baldwin Effect. The gd increase (0.191 vs 0.180) is consistent with epigenetic variation adding an additional heritable dimension beyond genetic variation alone.

---

