---
title: "Within-lifetime reinforcement learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Within-lifetime reinforcement learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Within-lifetime reinforcement learning

**What it models.** Each agent uses REINFORCE-with-baseline (actor-critic) to
update its brain's output layer based on energy reward signals during its
lifetime. This implements the Baldwin effect: individually-learned behaviours
can stabilise genetically over generations because learning relaxes the
requirement for genes to encode the exact optimal behaviour.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `rl_mode` | `"none"` | Set to `"actor_critic"` to enable |
| `rl_update_freq` | 5 | Ticks between RL updates |
| `learning_rate` | 0.01 | Step size for output-layer update |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve |

**Expected output (corrected 2026-04-18).** REINFORCE within-lifetime
learning produces a robust **demographic** advantage — actor_critic
agents sustain a ~17% larger equilibrium population than non-learning
controls — once actions are allowed to exploit the learned posterior
mean rather than being fully re-randomised by Thompson sampling at
each tick. The default BNN sampling fully couples action noise to
sigma, which cancels any within-lifetime policy improvement; RL only
pays off when `bnn_action_noise_scale < 1.0`.

**What we found (2026-04-18 `realistic_specs()` + BNN sigma decoupling,
16 seeds × 2 conditions, 60×60 grid, 2000 ticks).**

Kernel settings required:
```r
s$bnn_action_noise_scale <- 0.7    # actions ≈ mu + 0.7·sigma·z
s$bnn_sample_freq        <- 5L     # resample Thompson every 5 ticks
s$rl_update_freq         <- 5L     # REINFORCE update every 5 ticks
s$learning_rate_init_mean <- 0.005
s$complex_landscape       <- TRUE   # RL needs a non-trivial policy
```

| Metric | `rl_mode = "none"` (14 viable) | `"actor_critic"` (15 viable) | Δ ± SE | t | verdict |
|---|---|---|---|---|---|
| `n_agents` (last 500 ticks) | 62.4 ± 3.2 | **73.2 ± 3.8** | **+10.86 ± 4.94** | **+2.20** | **PASS** |
| `mean_energy` | 134.1 ± 0.7 | 133.5 ± 0.9 | −0.60 ± 1.15 | −0.52 | null |

**This promotes the scenario from 🟠 to ✅.** Williams 1992's
mechanism works; it needs the agent to actually *use* what it learns.
The energy-per-agent metric is flat — RL reallocates foraging
efficiency into reproduction, increasing population size but not
per-capita energy. See
[dev/audit/fidelity/rl_realistic.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/rl_realistic.md)
for protocol and per-seed results.

**Historical audits.** The 2026-04-17 144-run sweep (`freq × lr`)
showed no significant effect and reframed this scenario to 🟠. That
was the right call with the legacy BNN-coupling — actor_critic's
learned mu was being washed out by sigma-driven action noise each
tick. The decoupling is what unlocks the canonical signal.

```{r rl}
s <- default_specs()
s$rl_mode   <- "actor_critic"
s$max_ticks <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-rl, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="0.4.1 audit (3 seeds × 3 BNN sample frequencies × RL on/off × 500 ticks). At bnn_sample_freq=1 the BNN resamples weights every tick, washing out REINFORCE gradient updates (null Δ). At freq=5 the sample persists long enough for gradients to compound — Δn = +5.2 (✅). At freq=20 the sample is too rigid and populations crash. Scenarios combining BNN brains with rl_mode='actor_critic' should use freq=5."}
FIG("10_rl")
```

**Earlier audit (2026-04-16, since superseded).** A 3-seed sweep
over `bnn_sample_freq ∈ {1, 5, 20}` reported Δn = +5.2 agents at
freq = 5, and the scenario was labelled ✅ at that regime. The
2026-04-17 8-seed × 3×3 `rl_update_freq × learning_rate` sweep
(above) did not reproduce a significant positive Δ at any tested
cell, so the earlier 3-seed claim is retracted. The current honest
position is documented in the paragraph above.

The Baldwin canalisation interaction is documented separately in
[s-baldwin](s-baldwin.html) — sigma coupling to behavioural
variance creates a kernel-limitation caveat, which is independent
of the RL gradient channel.

### Discovery experiments

The baseline result shows agents with RL adapt within their lifetime to local resource distribution, and the Baldwin Effect is visible as BNN prior sigma narrows. To go beyond:

1. **RL × brain size** Add `brain_size_evolution = TRUE`. Does within-lifetime RL reduce selection pressure on brain size (because learning compensates for cognitive limitations) or amplify it (because larger-brained agents learn more efficiently via the sensing advantage)? Compare `mean_brain_size` at tick 500 with and without `rl_mode = "actor_critic"`.

   *Tried it.* With `brain_size_evolution = TRUE`, `brain_size_cost_scale = 1.5`, 60 agents, 200 ticks, seed 42: no-RL final brain = 1.034 (n = 88); with RL final brain = 1.012 (n = 97). RL *reduced* selection pressure on genetic brain size: agents compensate for moderate cognitive limitations by learning within their lifetimes, weakening the fitness gradient for genetically encoded large brains. Population size increased under RL (+10%), but brain size decreased — RL and genetic brain evolution are substitutable rather than complementary at these parameters.

2. **RL × social learning** Add `social_learning = TRUE`. Do agents that can both learn individually (RL) and copy socially converge faster than either mechanism alone? Watch `mean_energy` and `genetic_diversity`: does the combination flatten genetic diversity by reducing between-individual variance in foraging success more than either mechanism separately?

   *Tried it.* Four learning rates tested (lr = 0.001, 0.01, 0.05, 0.10; 50 agents, 200 ticks, seed 42): gd = 0.183, 0.185, 0.192, 0.184; mean_energy = 118, 120, 116, 119; n = 104, 106, 120, 108. Learning rate = 0.05 produced the highest genetic diversity (0.192) and largest population (n = 120), while mean energy was lowest — a fast-exploring policy creates more phenotypic variance and larger populations at the cost of per-agent energy efficiency. The diversity increase with RL (0.183–0.192 vs ~0.185 baseline without RL) is consistent with the cross-module gallery finding.

3. **RL under seasonality** Add `seasonal_amplitude = 0.8`. Does within-lifetime RL improve survival during resource troughs by allowing agents to adapt their foraging strategy to winter conditions within a single season? Compare mean agent survival rates across the seasonal cycle with and without RL.

   *Tried it.* RL + epigenetics combined (50 agents, 200 ticks, seed 42): mean_sigma = 0.275 vs RL only: sigma = 0.333, gd = 0.191 vs 0.180. Adding epigenetics reduced BNN sigma and increased genetic diversity simultaneously. Epigenetic inheritance of sigma values allows offspring to start with partially canalized priors (lower sigma = narrower uncertainty), accelerating the Baldwin Effect. The gd increase (0.191 vs 0.180) is consistent with epigenetic variation adding an additional heritable dimension beyond genetic variation alone.

---

