---
title: "Scavenging and carrion dynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Scavenging and carrion dynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Scavenging and carrion dynamics

**What it models.** Scavenging — the consumption of carrion left by dead
conspecifics — is a widespread foraging strategy that provides a resource
buffer when primary productivity is low (DeVault et al. 2003). In `clade`,
when `scavenging = TRUE`, agent deaths deposit a fraction of their remaining
energy as carrion on the grid cell where they died. Carrion decays each tick
at rate `carrion_decay_rate` and can be consumed by any agent that moves onto
the cell, yielding `carrion_eat_gain` energy units.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `scavenging` | FALSE | Enables carrion deposition and consumption |
| `carrion_fraction` | 0.5 | Fraction of a dead agent's energy deposited as carrion |
| `carrion_decay_rate` | 0.1 | Proportional decay of carrion per tick |
| `carrion_eat_gain` | 3.0 | Energy gained by a scavenger per unit of carrion consumed |
| `carrion_transmission_prob` | 0.0 | Probability that eating from an infected carcass transmits disease |
| `grass_rate` | 0.05 | Set low to create resource scarcity that accentuates the scavenging advantage |

**Expected output (latest: 0.5.x ✅ via realistic_specs + predator guild).**
Earlier default-scale sweeps found no DeVault 2003 energy benefit
because the carrion supply was too thin (small grid, short runs,
sparse predators). At `realistic_specs` (60×60, 2000 ticks, 30 init
predators, `predator_max_agents = 120`, `predator_energy_gain = 20`)
the predator guild produces enough carrion for the scavenging
channel to express as a measurable energy boost:

| Metric | scavenging OFF | scavenging ON | Δ ± SE | t |
|---|---|---|---|---|
| `mean_energy` (last 500 ticks) | — | — | **+3.42 ± 0.71** | **+4.83 PASS** |
| `n_agents` | — | — | **+14.9 ± 6.1** | **+2.46 PASS** |

DeVault 2003 holds when the predator guild supplies adequate
carcasses. Full protocol: [dev/audit/fidelity/scavenging_realistic.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/scavenging_realistic.md).

**Earlier (pre-realistic-specs, 🟠).** A 2×3×2 parameter sweep
(192 runs, `scavenging_strength_sweep.R`) over `grass_rate` ∈
{0.05, 0.10} × `carrion_eat_gain` ∈ {3, 8, 15} × `carrion_fraction`
∈ {0.5, 1.0} found no cell giving Δenergy > 0 at *t* ≥ 2 in the
canonical direction. Diagnosis: carrion supply was too thin at
`default_specs` scale (30×30 grid, 500-tick runs) — scavenging is a
population-density × grid-size × run-length phenomenon and needs
the realistic regime to express.

```{r scavenging}
library(clade)
library(ggplot2)

base_specs <- function() {
  s <- default_specs()
  s$n_agents_init <- 100L
  s$grid_rows     <- 30L
  s$grid_cols     <- 30L
  s$grass_rate    <- 0.15
  s$max_ticks     <- 400L
  s
}

s_no <- base_specs(); s_no$scavenging <- FALSE
s_sc <- base_specs()
s_sc$scavenging        <- TRUE
s_sc$carrion_fraction  <- 0.5
s_sc$carrion_decay_rate <- 0.1
s_sc$carrion_eat_gain  <- 3.0

d_no <- get_run_data(run_alife(s_no))$ticks
d_sc <- get_run_data(run_alife(s_sc))$ticks

d_no$condition <- "No scavenging"
d_sc$condition <- "Scavenging"
dat <- rbind(d_no, d_sc)

ggplot(dat, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("No scavenging" = "#E53935", "Scavenging" = "#FB8C00"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Mean agent energy",
    title = "Scavenging sustains energy budgets under resource scarcity"
  ) +
  theme_classic(base_size = 12)
```


## Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **12.0x** over the defaults above. See
[`dev/audit/calibration/RESULTS.md`](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md) for the full CMA-ES results.

```{r scavenging-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$grass_rate                     <- 1
s$idle_cost                      <- 1.528
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-scavenging, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: scavengers (orange) maintain higher mean energy than non-scavengers (red) under identical low-grass conditions, with carrion providing a density-dependent buffer against starvation."}
FIG("scavenging")
```

**What we found (2026-04-18 `realistic_specs()` with predator
guild, 8 seeds × 2 conds, 60×60 grid, 2000 ticks).** DeVault 2003's
carrion-as-energy-channel prediction is robustly reproduced at
realistic scale:

| Metric | scav_off (8 seeds) | scav_on (8 seeds) | Δ (on − off) | t | verdict |
|---|---|---|---|---|---|
| `mean_energy` (last 500 ticks) | 84.53 ± 0.46 | **87.94 ± 0.54** | **+3.42 ± 0.71** | **+4.83** | **PASS** |
| `n_agents` (last 500 ticks)    | 132.8 ± 3.4  | **147.6 ± 5.0**  | **+14.88 ± 6.06** | **+2.46** | **PASS** |

Both metrics cross 2 σ. Scavenging gives ~4% more energy and ~11%
more population size. **This promotes the scenario from 🟠 to ✅**
— the earlier 2026-04-17 null result (192-run sweep × 12 cells at
default `30×30` / 500-tick scale) was scale-limited: the predator
guild at default scale is too sparse to generate a detectable
carrion channel. At `realistic_specs()` scale (60×60 grid, 30
predators with `predator_max_age = 60`, 2000 ticks) carrion
accumulates into a measurable foraging channel. See
[dev/audit/fidelity/scavenging_realistic.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/scavenging_realistic.md)
for protocol and per-seed results.

Earlier small-scale finding (3 replicates, 25×25 grid, `grass_rate
= 0.06`, 400 ticks): scavenging increased final population by 21%
when food was scarce and the run was short enough that a
starvation-bottleneck dominated. The small-scale effect was real
but conditional; the realistic-scale result is the robust claim.

## Discovery experiments

The baseline result shows that scavenging sustains higher mean energy than baseline under identical low-grass conditions. To go beyond:

1. **Carrion as pathogen reservoir** When `disease = TRUE` and `carrion_transmission_prob > 0`, agents that die while infected deposit a flagged carcass; any agent that eats from it becomes infected with the specified probability. This models fomite transmission observed in vultures, hyenas, and other obligate scavengers (Ogada et al. 2012). The question: does carrion-mediated transmission reverse the previously observed anti-epidemic effect of scavenging?

```{r carrion-pathogen}
library(clade)
library(ggplot2)

make_specs <- function(carrion_tp = 0.0) {
  s <- default_specs()
  s$n_agents_init           <- 200L
  s$grid_rows               <- 25L
  s$grid_cols               <- 25L
  s$grass_rate              <- 0.08
  s$max_ticks               <- 300L
  s$scavenging              <- TRUE
  s$carrion_fraction        <- 0.5
  s$carrion_decay_rate      <- 0.1
  s$disease                 <- TRUE
  s$transmission_prob       <- 0.15
  s$disease_seed_prob       <- 0.03
  s$carrion_transmission_prob <- carrion_tp
  s
}

d_safe <- get_run_data(run_alife(make_specs(0.0)))$ticks
d_risk <- get_run_data(run_alife(make_specs(0.3)))$ticks

d_safe$condition <- "Carrion safe (prob = 0)"
d_risk$condition <- "Carrion risky (prob = 0.3)"
dat <- rbind(d_safe, d_risk)

ggplot(dat, aes(x = t, y = n_new_infections, colour = condition)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  scale_colour_manual(
    values = c("Carrion safe (prob = 0)" = "#43A047",
               "Carrion risky (prob = 0.3)" = "#E53935"),
    name = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "New infections per tick",
    title = "Carrion as pathogen reservoir amplifies epidemics"
  ) +
  theme_classic(base_size = 12)
```

**Expected output.** When `carrion_transmission_prob = 0.3`, `n_new_infections` should be higher than the safe-carrion baseline, particularly during population troughs when scavenging rates are highest. The carrion_decay_rate acts as a natural quarantine: fast decay reduces fomite persistence and dampens transmission. Watch whether the epidemic peak shifts earlier (higher early transmission from dense infection) or later (carcasses accumulate as the epidemic matures).

2. **Kin scavenging** Add `kin_selection = TRUE`. Both kin altruism and scavenging buffer energy deficits — are they additive (both help independently) or redundant (populations with kin altruism scavenge less because energy transfers substitute for carrion)? Watch `n_altruistic_acts` and `mean_energy` jointly under combined and single-module conditions.

   *Tried it.* With scavenging + disease enabled (50 agents, 200 ticks, seed 42): despite scavenge_events = 0, disease transmission was 37% lower in the scavenging condition (22 vs 35 infections). The scavenging module requires larger populations (≥ 200 agents) or longer runs (≥ 500 ticks) for consistent dead-agent turnover that generates detectable carrion. The epidemiological signal likely reflects energy-state differences between conditions rather than literal carrion contact.

3. **Scavenging × learning** Add `rl_mode = "actor_critic"`. Can within-lifetime RL discover scavenging opportunities faster than pure genetic evolution can encode the behaviour? Compare `mean_energy` trajectories between RL and no-RL conditions in a scavenging-enabled world at tick 100 (early learning phase) vs tick 400 (late genetic phase).

   *Tried it.* Combining scavenging and kin selection (50 agents, 200 ticks, seed 42): kin altruism operated normally (1278 acts, n = 132), but scavenge_events = 0. Both modules function independently at small population scales — kin altruism operates via live energy transfers while scavenging requires dead-agent corpse density. The two energy-buffering mechanisms are additive in principle but non-overlapping in the conditions that activate them.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
