---
title: "Sex labels (sex foundation)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Sex labels (sex foundation)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Sex foundation (0.8.0)

**What it models.** A *persistent* binary sex identity on agents — required
for any model in which males and females must be tracked across their
lifetime (e.g. sex-specific senescence onset, sex-specific trade-offs,
mating-system structure). Before 0.8.0, `clade` had **no sex labels**:
sexual selection in the existing Kokko-Brooks and Fuller vignettes worked
through unisex signal–preference matching, and the
`parental_investment_evolution` cost-split treated whichever agent
happened to initiate a reproductive event as the implicit mother.

This release adds the minimum foundation: a sticky `sex` field on every
`Agent`, configurable primary sex ratio at birth, and an opposite-sex
filter in `_find_mate()`. Sex-specific trait expression and persistent
mating-system structure (pair bonds, mating groups) are scheduled for
subsequent 0.8.x releases.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `sex_labels` | `FALSE` | Master toggle. `FALSE` preserves all pre-0.8.0 behaviour. `TRUE` activates the sex field, mate filter, and decoupled cost-split. |
| `sex_determination` | `"random"` | Currently only `"random"` is implemented. `"chromosomal"` / `"environmental"` reserved for future releases. |
| `sex_ratio_primary` | `0.5` | Proportion male at birth. Skewed values (e.g. `0.3`) produce female-biased populations. |

**Expected output.** Under default `sex_ratio_primary = 0.5`, the
realised sex ratio at equilibrium is close to 50/50, with sampling noise
shrinking as the population grows. Under `sex_ratio_primary = 0.3`, the
female-biased input is preserved at equilibrium because survival is
sex-blind in this foundation release (sex-specific mortality arrives
with sex-specific trait expression in a later release).

```{r sex-labels-balanced}
library(clade)

# Balanced primary sex ratio (default).
s <- default_specs()
s$grid_rows         <- 30L
s$grid_cols         <- 30L
s$n_agents_init     <- 100L
s$max_ticks         <- 500L
s$grass_rate        <- 0.15
s$random_seed       <- 42L
s$sex_labels        <- TRUE
s$sex_ratio_primary <- 0.5

env   <- run_alife(s)
sexes <- vapply(seq_len(length(env$agents)),
                function(i) as.integer(env$agents[[i]]$sex), integer(1))
cat("Population:", length(sexes),
    " | Proportion male:", round(mean(sexes == 1L), 3), "\n")
```

```{r sex-labels-skewed}
# Female-biased primary sex ratio.
s$sex_ratio_primary <- 0.3
s$random_seed       <- 43L

env2   <- run_alife(s)
sexes2 <- vapply(seq_len(length(env2$agents)),
                 function(i) as.integer(env2$agents[[i]]$sex), integer(1))
cat("Population:", length(sexes2),
    " | Proportion male:", round(mean(sexes2 == 1L), 3), "\n")
```

```{r fig-sex-labels, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "sex ratio at equilibrium tracks sex_ratio_primary input", fig.cap="Expected output: under sex_labels = TRUE, the realised sex ratio at equilibrium tracks the configured sex_ratio_primary. This foundation release introduces no sex-specific mortality; downstream survival differences enter with sex-specific trait expression in a later release."}
FIG("sex_labels")
```

## What changes inside the kernel when `sex_labels = TRUE`

Three behavioural changes are activated by the flag; everything else
remains exactly as on `main`.

1. **`Agent.sex` is set at construction** (`inst/julia/src/Clade.jl`,
   `_make_founder_agent`; `inst/julia/src/reproduce.jl`,
   `_make_offspring`). Each new agent draws sex per
   `sex_ratio_primary`. When `sex_labels = FALSE`, the field is set to
   `0` and read by no downstream code.
2. **`_find_mate()` filters to opposite-sex candidates** (in the
   neighbourhood scan, `reproduce.jl`). When `sex_labels = FALSE`, the
   filter is bypassed and the prior any-neighbour behaviour is preserved.
3. **`parental_investment_evolution` cost-split is decoupled from the
   focal-agent convention** (role-contract decoupling). When
   `sex_labels = TRUE`, `female_investment` denotes the *mother's*
   share regardless of which partner initiated the encounter; the
   father pays the complement. When `sex_labels = FALSE`, the legacy
   "focal-agent = implicit mother" semantics are preserved exactly.

None of these changes affect a run with `sex_labels = FALSE`; vignettes
and tests pinned to the FALSE default remain byte-identical to `main`.

## Honest discussion (scope of this release)

This release is *foundation only*. Three things are deliberately **not**
included:

- **Sex-specific traits.** Every evolvable trait (including
  `aging_rate`) is still expressed identically in males and females.
  Sex-specific gene slots (`TRAIT_AGING_RATE_FEMALE_GENE` /
  `TRAIT_AGING_RATE_MALE_GENE`) are introduced in a subsequent release.
- **Sex-specific mortality or fecundity.** Survival and reproductive
  output are sex-blind under this foundation. Without sex-specific
  survival/fecundity, the operational sex ratio (OSR) at equilibrium
  closely tracks the primary sex ratio — there is no demographic
  feedback that re-balances skewed inputs. (The forthcoming
  mating-system module introduces the union-formation dynamics that
  produce OSR deviations.)
- **Pair bonds / mating groups.** `_find_mate()` still scans all
  opposite-sex neighbours every tick; no agent state tracks previous
  partners. Persistent monogamous unions and multi-male/multi-female
  mating-group composition arrive with the mating-system module.

Consequently, the present vignette tests one narrow thing: *does the
sex field get assigned and propagated correctly, and does the
opposite-sex mate filter work?* It does not validate any biological
prediction about sex ratios at equilibrium under selection — that's a
later scenario's concern.

## Validation discipline

The sex foundation is unit-test driven, not multi-seed audit driven.
Coverage:

- `tests/testthat/test-sex-labels.R` — spec presence, defaults, .SPEC_GROUPS
  membership, plus Julia-integrated checks that founders and offspring
  receive valid sex values, the proportion-male is in the right
  ballpark for the configured `sex_ratio_primary`, and that
  `parental_investment_evolution` runs cleanly under `sex_labels = TRUE`.
- `tests/testthat/test-spec-wiring.R` — the standing R↔Julia wiring
  guard. All three new specs are consumed in `inst/julia/src/*.jl`, so
  no allowlist entry is needed.

A full multi-seed scenario audit (≥5 seeds, mean ± SE) is appropriate
when sex-specific trait expression is introduced and selection actually
acts on the sex field. At this stage the kernel changes are mechanical
plumbing and unit tests are the right granularity.

---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the upcoming paper-reproduction vignette
(Rees-Baylis et al. 2026, *Nature Communications*) that motivates this
foundation.

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
