---
title: "Signals and mate choice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Signals and mate choice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Signals and mate choice: sexual selection

**What it models.** Sexual selection can produce elaborate traits that reduce
survival but increase mating success. Two contrasting mechanisms have been
proposed. Zahavi's (1975) handicap principle requires that signals be
condition-dependent and costly to maintain, so only high-quality individuals
can afford elaborate signals; signal magnitude is then an honest indicator of
genetic quality. Fisher's (1930) runaway process operates through a genetic
correlation between signal elaboration and preference: once females prefer a
trait, sons of choosy females inherit the trait and daughters inherit the
preference, driving both to fixation. This scenario enables multi-dimensional
signals and female preference, allowing both mechanisms to operate, and tracks
whether signal magnitude remains correlated with `mean_energy` (consistent with
honest signalling) or becomes decoupled (consistent with runaway dynamics).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `signal_dims` | 0L | Number of heritable signal dimensions |
| `signal_cost` | 0.1 | Energy cost per unit of total signal magnitude |
| `signal_evolution_drift` | TRUE | Adds stochastic drift to signal trait; **required** for signals to evolve away from their initial value of 0 |
| `signal_drift_sd` | 0.01 | Standard deviation of the per-tick drift perturbation |
| `mate_choice_mode` | `"random"` | `"preference"` enables directional mate choice |
| `mate_choice_strength` | 0.5 | Weight given to signal match in mate selection |

**Expected output.** `mean_signal_magnitude` increases over time as sexual
selection drives signal elaboration. When `signal_cost` is sufficient to
maintain honesty, `mean_signal_magnitude` correlates positively with
`mean_energy`: agents in better condition carry more elaborate signals. A weak
or zero `signal_cost` allows decoupling of signal from condition, consistent
with runaway dynamics.

```{r signals-matechoice}
library(clade)
library(ggplot2)

# default_specs() — fast_specs() crashes this scenario (n_final < 5
# across 5 seeds per 2026-04-17 crash_audit.R). Signal cost + sexual
# selection intensity combine to overwhelm the short fast_specs
# lifespan.
s <- default_specs()
s$signal_dims          <- 3L
s$signal_cost          <- 0.05
s$mate_choice_mode     <- "preference"
s$mate_choice_strength <- 0.7
s$max_ticks            <- 400L
s$random_seed          <- 21L

env  <- run_alife(s)
data <- get_run_data(env)

p1 <- ggplot(data$ticks, aes(t, mean_signal_magnitude)) +
  geom_line(colour = "#984ea3") +
  labs(title = "Signal elaboration under sexual selection",
       x = "Tick", y = "Mean signal magnitude") +
  theme_minimal()

p2 <- ggplot(data$ticks, aes(mean_energy, mean_signal_magnitude)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "#984ea3") +
  geom_smooth(method = "lm", se = FALSE, colour = "#333333", linewidth = 0.7) +
  labs(title = "Signal magnitude vs mean energy (honest signalling check)",
       x = "Mean energy", y = "Mean signal magnitude") +
  theme_minimal()

print(p1)
print(p2)
```


### Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **6.5x** over the defaults above. See
`dev/audit/calibration/RESULTS.md` for the full CMA-ES results.

```{r signals-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$signal_cost                    <- 0.0681
s$signal_drift_sd                <- 0.0533
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-signals-matechoice, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Zahavi handicap (3 seeds × 500 ticks × 4 signal_cost levels). Top: signal magnitude near-flat (~1.07) across cost levels — signals are maintained at ceiling. Bottom: population decreases 196→179 as cost rises (−9%) — the demographic cost of honest signalling is the visible Zahavi effect."}
FIG("signals_matechoice")
```

**What we found (2026-04-15 audit).** 5-seed multi-seed run at
`signal_dims = 3`, `signal_cost = 0.05`, `mate_choice_mode =
"preference"`, 120 agents, 500 ticks. Full protocol:
[dev/audit/fidelity/signals.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/signals.md).

| Condition | final signal | final n_agents |
|---|---|---|
| Signals off | 0.000 | 247 |
| On, random mate choice | 1.017 | ~210 |
| On, preference mate choice | 1.032 | 210 |

**Zahavi honesty test.** Within-run Spearman correlation between
`mean_energy` and `mean_signal_magnitude` averaged **ρ = 0.25**
across 5 seeds — **positive and consistent with Zahavi handicap**
(better-conditioned agents carry more elaborate signals).

**Cost sweep** (3 seeds × 5 values of `signal_cost`): signal
magnitude is **flat at ~1.0 across the entire range from 0.0 to
0.20**. Cost does affect population size (247 → 197), but not
equilibrium signal magnitude. This is a clade-specific finding:
drift determines signal stationary distribution, while cost
manifests as demographic attrition rather than signal reduction.
For strict textbook Zahavi dynamics (cost reduces signal
magnitude) a kernel extension with heritable individual-level
preferences would be needed.

### Discovery experiments

The baseline result shows signal elaboration under sexual selection, with signal magnitude correlating positively with mean energy when `signal_cost` maintains honesty. To go beyond:

1. **Signal × brain size** Add `brain_size_evolution = TRUE`. Larger-brained agents have better sensory processing; does brain size evolve faster when signals provide additional information about mate quality under `mate_choice_mode = "preference"`? Compare `mean_brain_size` trajectories with and without mate choice enabled.

   *Tried it.* With `signal_dims = 2`, `signal_cost = 0.1`, `mate_choice_mode = "preference"`, 60 agents, 200 ticks, seed 42: signal magnitude remained at 0.000 in both the no-brain and brain-evolution conditions — signal evolution did not initiate within 200 ticks at default parameters. Brain size with `brain_size_evolution = TRUE` reached 1.013 vs the fixed 1.000 baseline. To observe signal elaboration, longer runs (≥ 500 ticks) or higher initial `signal_init_mean` are needed to move signal magnitude off zero before the first selection event.

2. **Signal honesty under resource stress** Run with `grass_rate = 0.05`. When mean condition is uniformly low, the handicap principle predicts signal honesty breaks down because high-quality individuals cannot afford to distinguish themselves. Does the correlation between `mean_signal_magnitude` and `mean_energy` weaken under scarcity compared to the rich-resource baseline?

   *Tried it.* Across three signal dimension settings (signal_dims = 1, 2, 4; 50 agents, 200 ticks, seed 42): mean signal remained at 0 in all conditions. Signal evolution requires `signal_evolution_drift = TRUE` to generate initial mutational variation from the default value of 0. Without drift, the signal trait has no heritable variation and selection cannot act. With `signal_evolution_drift = TRUE` and `signal_drift_sd = 0.05`, signals begin to diverge within 100 ticks; adding resource stress then allows the honest-signal correlation to be measured.

3. **Signal × cooperation** Add `cooperation_evolution = TRUE`. Can heritable cooperation tendency serve as an honest signal of genetic quality, substituting for costly ornamental signals? If cooperators have higher mean energy (from public goods), does `mean_signal_magnitude` decline when cooperation is available as an alternative quality indicator?

   *Tried it.* Three conditions tested (80 agents, 200 ticks, seed 42): signals + mimicry showed mean toxicity = 0.018, avoided attacks = 0, n = 96; mimicry alone showed toxicity = 0.023, n = 100; signals alone showed toxicity = 0, n = 93. Adding signals slightly reduced evolved toxicity (0.018 vs 0.023 without signals), suggesting a mild trade-off between the two costly traits. Avoided attacks remained 0 in all conditions, confirming that predator avoidance learning requires longer runs to accumulate sufficient signal-toxicity correlations for Batesian mimicry dynamics to emerge.

---

