---
title: "Speciation and genetic divergence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Speciation and genetic divergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

## Speciation and genetic divergence

**What it models.** Speciation by ecological character displacement occurs when
disruptive selection on a resource-use trait, combined with assortative mating
among similar phenotypes, drives a single population to split into
reproductively isolated lineages (Dieckmann & Doebeli 1999). Once genomes
diverge beyond a threshold, gene flow between incipient species is reduced,
accelerating further divergence. Coyne & Orr (2004) reviewed extensive
empirical evidence that reproductive isolation accumulates approximately
linearly with genetic distance. This scenario enables the speciation module,
tracks the number of distinct genetic clusters (`n_species`) detected at
regular intervals, and illustrates how within-species diversity compares to
between-species divergence over the course of the run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `speciation` | FALSE | Enable genetic-cluster detection and isolation |
| `isolation_threshold` | 0.5 | Minimum genetic distance required for reproductive isolation |
| `speciation_cluster_interval` | 10L | Ticks between cluster re-detection passes |

**Expected output.** `n_species` starts at 1 and rises to 2–4 as accumulated
genome divergence exceeds `isolation_threshold` in geographically or
ecologically separated subpopulations. Within-cluster genetic diversity remains
lower than between-cluster diversity once isolation is established. The timing
of the first split depends on population size, mutation rate, and the
stringency of `isolation_threshold`.

```{r speciation}
library(clade)
library(ggplot2)

# fast_specs() gives ~66 generations over 2000 ticks, which comfortably
# resolves the speciation burst (~100 clusters in the first few hundred
# ticks) through consolidation to ~10 persistent clusters.
s <- fast_specs()
s$speciation                  <- TRUE
s$isolation_threshold         <- 0.15   # lower = easier speciation
s$mutation_sd                 <- 0.15   # higher = faster divergence
s$speciation_cluster_interval <- 10L
s$random_seed                 <- 55L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(t, n_species)) +
  geom_step(colour = "#ff7f00", linewidth = 0.9) +
  scale_y_continuous(breaks = 1:8) +
  labs(title = "Speciation: number of distinct genetic clusters over time",
       x = "Tick", y = "Number of species (clusters)") +
  theme_minimal()
```


## Calibrated regime (CMA-ES discovered)

Running Phase 7 auto-calibration (`dev/audit/calibration/`) over the
scenario's parameter subspace discovered the following regime, which
produces a fitness improvement of **189.1x** over the defaults above. See
[`dev/audit/calibration/RESULTS.md`](https://github.com/itchyshin/clade/blob/main/dev/audit/calibration/RESULTS.md) for the full CMA-ES results.

```{r speciation-calibrated, eval=FALSE}
# Parameter overrides discovered by CMA-ES (see dev/audit/calibration/):
s <- default_specs()
s$isolation_threshold            <- 0.5127
s$mutation_sd                    <- 1.31
# env <- run_alife(s)   # uncomment to run the calibrated regime
```
```{r fig-18-speciation, eval=TRUE, echo=FALSE, out.width="100%", fig.alt = "n_species rises from 1 to 2-4 as genome divergence accumulates beyond the isolation threshold", fig.cap="Expected output: n_species rises from 1 to 2-4 as genome divergence accumulates beyond the isolation threshold. Each step in the staircase represents a speciation event."}
FIG("18_speciation")
```

**What we found (2026-04-15 audit).** 3 seeds × 1000 ticks, 100
agents init, 30×30 grid, `isolation_threshold = 0.15`. Full
protocol: [dev/audit/fidelity/speciation.md](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/speciation.md).

| Regime | max n_species |
|---|---|
| Default `mutation_sd = 0.1` | 96 |
| Aggressive `mutation_sd = 0.3` | 200 |

**Isolation threshold sweep** (5 levels × 3 seeds, mut_sd = 0.2):
Spearman ρ = −0.97 between `isolation_threshold` and max n_species
— tighter threshold, fewer clusters, exactly as theory predicts.

**Caveat on n_species interpretation.** Counts in the hundreds
(from 500-agent populations) indicate that the pairwise-distance
clustering algorithm at high mutation rates flags every distinct
genome as a separate lineage. The sensitivity signs (to
`mutation_sd` and `isolation_threshold`) are correct, but the
absolute count should be read as "number of genetic clusters
detected at this resolution," not "Mayr-style biological species."
For textbook-style 2–4 distinct species, use lower mutation
(0.05–0.08) and longer runs (2000+ ticks).

## Discovery experiments

The baseline result shows `n_species` rising from 1 to 2–4 as genetic distance accumulates past `isolation_threshold`. To go beyond:

1. **Speciation × disease** Add `disease = TRUE`. Genetically distinct species share fewer contact opportunities (reproductive isolation requires spatial separation), which may create inter-species barriers to pathogen transmission. Does speciation reduce epidemic peak `n_infected` by fragmenting the contact network into partially isolated sub-populations?

   *Tried it.* With `speciation = TRUE`, `isolation_threshold = 0.5`, `transmission_prob = 0.25`, 80 agents, 200 ticks, seed 42: speciation did not produce multiple species in 200 ticks (n_species = 1 in both conditions); `isolation_threshold = 0.5` requires more genetic divergence than 200 ticks allows. The disease added a 14% population reduction (79 vs 91 final). To test the fragmentation hypothesis, run for ≥ 500 ticks with `isolation_threshold = 0.3` and seed diversity via `mutation_sd ≥ 0.2`.

2. **Speciation × brain size** Add `brain_size_evolution = TRUE`. Do incipient species diverge in brain size as a by-product of ecological niche partitioning, or does brain size converge across species via parallel evolution toward the same cognitive optimum? Compare `mean_brain_size` across detected genetic clusters at final tick.

   *Tried it.* With `n_species = 1L`, `isolation_threshold = 0.3`, `mutation_sd = 0.1`, 80 agents, 25×25 grid, 500 ticks (seed 42): n_species_max = 0 (no speciation detected) in both the brain-evolution and no-brain conditions, with very small final populations (n = 31–32). The populations struggled demographically at these parameters — 25×25 grid with 80 agents and high mutation may deplete energy faster than recovery. Speciation and brain-size divergence cannot be compared when speciation itself fails. This experiment requires parameter tuning (larger populations, lower mutation_sd, longer runs) before the brain-divergence question can be tested.

3. **Isolation threshold sensitivity** Vary `isolation_threshold` from 0.2 to 0.8 across seven values in `batch_alife()`. Does speciation rate (slope of `n_species` vs time) respond monotonically to threshold, or is there a critical threshold below which sympatric speciation cannot initiate even given sufficient time?

   *Tried it.* Seven isolation thresholds tested (0.2 to 0.8; `mutation_sd = 0.1`, 80 agents, 300 ticks, seed 42): no speciation detected at any threshold (n_species_max = 0 in all conditions). Final populations were very small (11–34 agents), suggesting demographic collapse is the primary constraint at 300 ticks, not the isolation threshold. The scenarios "What we found" notes that speciation requires >1,000 ticks at default mutation rates. Threshold sensitivity is best tested at ≥ 1,000 ticks with a minimum of 100 agents and `mutation_sd ≥ 0.15`.

---


---

## Citation

If you use this scenario in published work, please cite both the
`clade` package and the primary literature the scenario references.
The theory-to-scenario mapping is catalogued in the
[fidelity audit dashboard](https://github.com/itchyshin/clade/blob/main/dev/audit/fidelity/DASHBOARD.md).

```bibtex
@misc{clade2026,
  author  = {Nakagawa, Shinichi},
  title   = {clade: evolve behaviour, minds, and brains in R},
  year    = {2026},
  note    = {R package},
  url     = {https://github.com/itchyshin/clade}
}
```
