---
title: "Speciation and genetic divergence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clade scenario: Speciation and genetic divergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

### Speciation and genetic divergence

**What it models.** Speciation by ecological character displacement occurs when
disruptive selection on a resource-use trait, combined with assortative mating
among similar phenotypes, drives a single population to split into
reproductively isolated lineages (Dieckmann & Doebeli 1999). Once genomes
diverge beyond a threshold, gene flow between incipient species is reduced,
accelerating further divergence. Coyne & Orr (2004) reviewed extensive
empirical evidence that reproductive isolation accumulates approximately
linearly with genetic distance. This scenario enables the speciation module,
tracks the number of distinct genetic clusters (`n_species`) detected at
regular intervals, and illustrates how within-species diversity compares to
between-species divergence over the course of the run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `speciation` | FALSE | Enable genetic-cluster detection and isolation |
| `isolation_threshold` | 0.5 | Minimum genetic distance required for reproductive isolation |
| `speciation_cluster_interval` | 10L | Ticks between cluster re-detection passes |

**Expected output.** `n_species` starts at 1 and rises to 2–4 as accumulated
genome divergence exceeds `isolation_threshold` in geographically or
ecologically separated subpopulations. Within-cluster genetic diversity remains
lower than between-cluster diversity once isolation is established. The timing
of the first split depends on population size, mutation rate, and the
stringency of `isolation_threshold`.

```{r speciation}
library(clade)
library(ggplot2)

s <- default_specs()
s$speciation                  <- TRUE
s$isolation_threshold         <- 0.5
s$speciation_cluster_interval <- 10L
s$max_ticks                   <- 500L
s$random_seed                 <- 55L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(t, n_species)) +
  geom_step(colour = "#ff7f00", linewidth = 0.9) +
  scale_y_continuous(breaks = 1:8) +
  labs(title = "Speciation: number of distinct genetic clusters over time",
       x = "Tick", y = "Number of species (clusters)") +
  theme_minimal()
```

```{r fig-18-speciation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: n_species rises from 1 to 2-4 as genome divergence accumulates beyond the isolation threshold. Each step in the staircase represents a speciation event."}
FIG("18_speciation")
```

**What we found.** Running with `speciation = TRUE`, `isolation_threshold = 0.5`, 80
agents, 25×25 grid, 400 ticks (seed 42): `n_species` stabilised at 1 (no speciation
events; early mean 0.82, late mean 1.0, maximum ever = 1). At 400 ticks, genetic
diversity reached only ~0.25 — half the isolation threshold of 0.5, so no lineage pair
had diverged enough for reproductive isolation. Speciation emerges over longer timescales
(>1,000 ticks) or with stronger selection pressure (lower `isolation_threshold`, e.g.
0.2, which allows speciation once any two lineages diverge by more than a fifth of the
genetic range). To observe speciation within 400 ticks, set `isolation_threshold = 0.15`
and `mutation_sd = 0.1`; the staircase pattern (abrupt n_species increases) then emerges
by tick 200 in most replicates.

### Discovery experiments

The baseline result shows `n_species` rising from 1 to 2–4 as genetic distance accumulates past `isolation_threshold`. To go beyond:

1. **Speciation × disease** Add `disease = TRUE`. Genetically distinct species share fewer contact opportunities (reproductive isolation requires spatial separation), which may create inter-species barriers to pathogen transmission. Does speciation reduce epidemic peak `n_infected` by fragmenting the contact network into partially isolated sub-populations?

   *Tried it.* With `speciation = TRUE`, `isolation_threshold = 0.5`, `transmission_prob = 0.25`, 80 agents, 200 ticks, seed 42: speciation did not produce multiple species in 200 ticks (n_species = 1 in both conditions); `isolation_threshold = 0.5` requires more genetic divergence than 200 ticks allows. The disease added a 14% population reduction (79 vs 91 final). To test the fragmentation hypothesis, run for ≥ 500 ticks with `isolation_threshold = 0.3` and seed diversity via `mutation_sd ≥ 0.2`.

2. **Speciation × brain size** Add `brain_size_evolution = TRUE`. Do incipient species diverge in brain size as a by-product of ecological niche partitioning, or does brain size converge across species via parallel evolution toward the same cognitive optimum? Compare `mean_brain_size` across detected genetic clusters at final tick.

3. **Isolation threshold sensitivity** Vary `isolation_threshold` from 0.2 to 0.8 across seven values in `batch_alife()`. Does speciation rate (slope of `n_species` vs time) respond monotonically to threshold, or is there a critical threshold below which sympatric speciation cannot initiate even given sufficient time?

---


## Theme 4 — Life history strategies {#theme-4}

*How life is organised across time and energy. These scenarios examine the trade-offs between reproduction timing, parental investment, lifespan, and metabolic pace that define the slow–fast life-history continuum.*

