---
title: "Biological Scenarios and Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biological Scenarios and Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Each module in `clade` implements a biological mechanism that can be enabled
independently or combined. This article documents every module: the biological
motivation, the key parameters, what to look for in the output, and a minimal
working example.

All examples require Julia. Wrap any call in `if (julia_is_ready()) { ... }`
for scripts that may run on machines without Julia.

---

## Baseline world

**What it models.** Agents forage on a renewable grass resource and reproduce
when energy exceeds a threshold. The neural-network genome (brain weights)
evolves under natural selection with no additional modules. This is the
simplest demonstration that evolution can occur.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `grass_rate` | 0.10 | Probability that empty cell regrows per tick |
| `mutation_sd` | 0.05 | Gaussian noise on brain weights per offspring |
| `brain_type` | `"bnn"` | Neural architecture |
| `ploidy` | 1 | Haploid (1) or diploid (2) |

**Expected output.** Population stabilises well below `max_agents`. Genetic
diversity is maintained: neither fixation nor unbounded growth.

```{r baseline}
library(clade)

s <- default_specs()
s$grid_rows     <- 20L
s$grid_cols     <- 20L
s$n_agents_init <- 40L
s$max_ticks     <- 300L
s$random_seed   <- 1L

env  <- run_alife(s)
data <- get_run_data(env)
plot_run(data)
```

---

## Complex landscape (forest world)

**What it models.** A three-layer habitat: ground (grass), shrubs (mid-level),
and canopy (top). Shrubs and canopy have higher energy density but slower
regrowth. Canopy access requires a minimum `wing_size` (a heritable trait).
This creates a selection pressure for wing-size evolution and niche
partitioning. The scenario is structurally analogous to the
*cephalopod paradox* (Liedtke & Fromhage 2019): a rich resource is accessible
only to agents that have evolved the morphology to exploit it.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `complex_landscape` | FALSE | Enable this module |
| `shrub_density` | 0.35 | Fraction of cells with shrubs |
| `shrub_energy` | 25.0 | Energy gained from shrubs |
| `shrub_growth_rate` | 0.04 | Shrub regrowth rate |
| `canopy_density` | 0.15 | Fraction of cells with canopy |
| `canopy_energy` | 55.0 | Energy gained from canopy |
| `canopy_threshold` | 0.6 | Minimum wing_size for canopy access |
| `wing_size_init_mean` | 0.1 | Starting wing size |
| `wing_size_mutation_sd` | 0.05 | Wing size mutation rate |

**Expected output.** `mean_wing_size` rises over time. `n_canopy_agents`
increases as wing-size evolution crosses the `canopy_threshold`.
Niche entropy (ground/shrub/canopy mix) first rises then stabilises.

```{r complex-landscape}
s <- default_specs()
s$complex_landscape   <- TRUE
s$shrub_density       <- 0.35
s$canopy_density      <- 0.15
s$canopy_energy       <- 55.0
s$canopy_threshold    <- 0.6
s$wing_size_init_mean <- 0.1
s$max_ticks           <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

# Wing size evolution
plot(data$ticks$t, data$ticks$mean_wing_size, type = "l",
     xlab = "Tick", ylab = "Mean wing size",
     main = "Wing size evolution in the forest world")
```

Use `tune_complex_landscape()` to find parameter combinations that
produce strong wing-size evolution with CMA-ES:

```{r tune-landscape}
tuned <- tune_complex_landscape(default_specs(), n_iterations = 80L)
tuned$specs   # optimal landscape parameters
```

---

## Spatial sorting

**What it models.** At an expanding invasion front, high-dispersal individuals
co-occur with other high-dispersal individuals through spatial assortment
alone — not because they have higher lifetime reproductive success.
Dispersal-enhancing alleles "surf the wave" (Shine et al. 2011). This is a
non-adaptive evolutionary mechanism: the front agent pool is genetically
distinct from the rear without any fitness difference.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal tendency |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation rate for dispersal |
| `spatial_sorting` | FALSE | Enable invasion-front mating assortment |
| `sorting_front_threshold` | 0.75 | Fraction from front to define "front zone" |
| `sorting_mating_boost` | 3.0 | Fold-increase in mate encounters at front |

**Expected output.** `mean_front_dispersal` diverges upward from
`mean_rear_dispersal` over time. The magnitude of the gap reflects the
strength of spatial sorting. Rear dispersal may drift downward if there is any
cost to dispersal.

```{r spatial-sorting}
s <- default_specs()
s$dispersal_evolution   <- TRUE
s$dispersal_init_mean   <- 0.3
s$dispersal_mutation_sd <- 0.04
s$spatial_sorting       <- TRUE
s$sorting_mating_boost  <- 3.0
s$max_ticks             <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

library(ggplot2)
ggplot(data$ticks, aes(t)) +
  geom_line(aes(y = mean_front_dispersal, colour = "Front")) +
  geom_line(aes(y = mean_rear_dispersal,  colour = "Rear"), linetype = "dashed") +
  scale_colour_manual(values = c(Front = "#e41a1c", Rear = "#377eb8"),
                      name = NULL) +
  labs(title = "Spatial sorting: dispersal diverges at the invasion front",
       x = "Tick", y = "Mean dispersal tendency") +
  theme_minimal()
```

---

## IFfolk inclusive fitness

**What it models.** Agents can transfer energy to nearby relatives. The
transfer amount and eligibility radius are controlled by `iffolk_*` parameters.
Cooperative behaviour evolves because it improves inclusive fitness: the
donor's fitness cost is recovered through the reproductive gains of related
recipients (Hamilton 1964). `parliament_suppression` enforces cooperation by
penalising defectors (low `helper_tendency`) surrounded by cooperators,
modelling the *parliament of genes* (Fromhage & Jennions 2019).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `iffolk_selection` | FALSE | Enable IFfolk transfers |
| `iffolk_r_min` | 0.125 | Minimum relatedness (cousins and closer) |
| `iffolk_radius` | 5 | Neighbourhood radius for kin search |
| `iffolk_transfer` | 3.0 | Energy transferred per act |
| `iffolk_min_energy` | 60.0 | Donor must exceed this energy |
| `parliament_suppression` | FALSE | Penalise defectors among cooperators |
| `parliament_cost` | 0.5 | Energy penalty for defectors |
| `cooperative_breeding` | FALSE | Enable `helper_tendency` trait |
| `helper_tendency_init_mean` | 0.2 | Starting helper tendency |

**Expected output.** `mean_helper_tendency` increases over time.
`n_iffolk_transfers` is positive and grows with population size.
With `parliament_suppression = TRUE`, the increase is faster and converges
near 1.0 for high-relatedness environments.

```{r iffolk}
s <- default_specs()
s$iffolk_selection        <- TRUE
s$iffolk_r_min            <- 0.125
s$iffolk_transfer         <- 3.0
s$parliament_suppression  <- TRUE
s$parliament_cost         <- 0.5
s$cooperative_breeding    <- TRUE
s$helper_tendency_init_mean <- 0.2
s$max_ticks               <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

plot(data$ticks$t, data$ticks$mean_helper_tendency, type = "l",
     xlab = "Tick", ylab = "Mean helper tendency",
     main = "Helper tendency evolves under IFfolk + parliament suppression")
```

---

## Kin selection

**What it models.** Agents with energy above a threshold donate a fixed cost to
the most closely related Moore-neighbourhood agent with relatedness
r >= `kin_altruism_r_min`. Relatedness is computed from the pedigree (parent
chains). This is Hamilton's rule operationalised at each tick: the altruistic
act evolves when Hamilton's rule holds: r &times; benefit &gt; cost.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `kin_selection` | FALSE | Enable kin altruism |
| `kin_altruism_r_min` | 0.25 | Minimum relatedness to donate (siblings) |
| `kin_altruism_cost` | 2.0 | Energy transferred per act |
| `kin_altruism_min_donor_energy` | 50.0 | Donor energy floor |

**Expected output.** `n_altruistic_acts` is positive. Population size is
generally higher than baseline (altruism increases mean fitness). Genetic
diversity may be slightly lower if relatedness clustering creates local
inbreeding.

```{r kin-selection}
s <- default_specs()
s$kin_selection               <- TRUE
s$kin_altruism_r_min          <- 0.25
s$kin_altruism_cost           <- 2.0
s$max_ticks                   <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total altruistic acts:", sum(data$ticks$n_altruistic_acts), "\n")
```

---

## SIR disease

**What it models.** A Susceptible-Infected-Recovered epidemic model. Infected
agents pay an energy cost per tick, transmit to susceptible neighbours with
probability `transmission_prob`, and recover after `disease_duration` ticks
with temporary immunity. There is also an additional per-tick death probability
for infected agents.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `disease` | FALSE | Enable SIR dynamics |
| `transmission_prob` | 0.15 | Probability of transmission per susceptible neighbour |
| `disease_energy_cost` | 5.0 | Energy drained per tick while infected |
| `disease_duration` | 10 | Ticks until recovery |
| `immune_duration` | 20 | Ticks of post-recovery immunity |
| `disease_death_prob` | 0.01 | Extra mortality per tick while infected |
| `disease_seed_prob` | 0.02 | Fraction infected at tick 1 |

**Expected output.** `n_infected` shows epidemic peaks. Population size dips
sharply during outbreaks and recovers as immunity builds. Multiple epidemic
cycles are possible if immunity wanes before the next susceptible cohort matures.

```{r disease}
s <- default_specs()
s$disease          <- TRUE
s$transmission_prob <- 0.20
s$max_ticks        <- 300L

env  <- run_alife(s)
data <- get_run_data(env)

plot(data$ticks$t, data$ticks$n_infected, type = "l",
     xlab = "Tick", ylab = "Infected agents",
     main = "SIR epidemic dynamics")
```

---

## Niche construction

**What it models.** Agents build shelter at their current cell when energy
exceeds `shelter_min_energy`. Shelters reduce predator damage and slow grass
regrowth on the cell. Shelters decay stochastically each tick. This models
ecosystem engineering: organisms that modify their selective environment
(Odling-Smee et al. 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `niche_construction` | FALSE | Enable shelter building |
| `shelter_build_prob` | 0.1 | Probability of building per tick |
| `shelter_min_energy` | 40.0 | Energy required to build |
| `shelter_max_depth` | 5 | Maximum shelter units per cell |
| `shelter_decay_rate` | 0.02 | Fraction of shelters lost per tick |

**Expected output.** `n_shelters_built` is positive. Predator damage is reduced
in sheltered cells. Population size may be larger than baseline due to shelter
protection.

```{r niche-construction}
s <- default_specs()
s$niche_construction <- TRUE
s$n_predators_init   <- 5L
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total shelters built:", sum(data$ticks$n_shelters_built), "\n")
```

---

## Body size evolution

**What it models.** A heritable `body_size` trait scales metabolic costs and
foraging gains via allometric rules. Larger body size increases move cost but
also foraging efficiency. Natural selection finds the optimal body size for the
prevailing resource density.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enable body size as a heritable trait |
| `body_size_init_mean` | 1.0 | Starting body size (relative units) |
| `body_size_mutation_sd` | 0.05 | Mutation rate |
| `body_size_min` | 0.1 | Lower clamp |
| `body_size_max` | 5.0 | Upper clamp |

**Expected output.** `mean_body_size` evolves to a stable value. High
`grass_rate` (rich environment) typically selects for larger body size;
sparse environments select for smaller.

```{r body-size}
s <- default_specs()
s$body_size_evolution  <- TRUE
s$body_size_init_mean  <- 1.0
s$max_ticks            <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final mean body size:", tail(data$ticks$mean_body_size, 1L), "\n")
```

---

## Dispersal evolution

**What it models.** A heritable `dispersal_tendency` trait governs how far
agents move from their birthplace. This allows evolution to optimise dispersal
in response to local resource density and competition. Can be combined with
`spatial_sorting` to study invasion dynamics.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation rate |
| `dispersal_min / dispersal_max` | 0.0 / 1.0 | Clamp range |

**Expected output.** `mean_dispersal` evolves in response to crowding.
Dense, resource-rich environments typically select for lower dispersal
(local knowledge is valuable); sparse, patchy environments select for higher.

```{r dispersal}
s <- default_specs()
s$dispersal_evolution <- TRUE
s$max_ticks           <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
# Dispersal summary is in data$ticks$n_dispersal_events
```

---

## Social learning

**What it models.** Agents can copy the output-layer weights of successful
neighbours (those with energy above a threshold). This is a model of vertical-
and horizontal-cultural transmission: information about which actions are
rewarding propagates socially, accelerating adaptation relative to genetic
evolution alone (Henrich & McElreath 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `social_learning` | FALSE | Enable social learning |
| `social_learning_freq` | 10 | Ticks between learning events |
| `social_learning_radius` | 3 | Search radius for models |

**Expected output.** Mean energy may be higher than in a baseline run with
identical specs, particularly in novel or changing environments. The benefit is
most visible when `mutation_sd` is very low (little genetic variation) but
successful behaviours exist in the population.

```{r social-learning}
s <- default_specs()
s$social_learning      <- TRUE
s$social_learning_freq <- 10L
s$max_ticks            <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

---

## Parental care

**What it models.** Offspring are carried by the parent until graduation at age
`care_duration`. While carried, juveniles receive energy from the parent at
`juvenile_energy_gain` per tick. If the parent dies, carried offspring die too
(obligate altriciality). This models the evolutionary trade-off between clutch
size and offspring quality (Smith & Fretwell 1974).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_care` | FALSE | Enable parental care |
| `care_duration` | 5 | Ticks offspring are carried |
| `care_energy_cost` | 2.0 | Energy drained from parent per juvenile per tick |
| `juvenile_energy_gain` | 3.0 | Energy transferred to juvenile per tick |

**Expected output.** `n_juveniles` is positive. Per-capita offspring count may
be lower than baseline (parents can carry fewer), but juvenile survival is
higher. Population dynamics are more buffered.

```{r parental-care}
s <- default_specs()
s$parental_care     <- TRUE
s$care_duration     <- 5L
s$care_energy_cost  <- 2.0
s$max_ticks         <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total juveniles recorded:", sum(data$ticks$n_juveniles, na.rm = TRUE), "\n")
```

---

## Mimicry and toxicity

**What it models.** A heritable `toxicity` trait makes prey costly to attack.
Predators learn signal–toxicity associations via a Rescorla-Wagner rule and
build avoidance. This models the evolution of warning coloration and Batesian/
Mullerian mimicry (Ruxton et al. 2004): toxic prey are avoided; non-toxic
prey that resemble toxic prey are also avoided (Batesian mimicry).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `mimicry` | FALSE | Enable toxicity / predator learning |
| `toxicity_init_mean` | 0.0 | Starting toxicity (0 = non-toxic) |
| `toxin_dose` | 2.0 | Damage dealt to attacker per toxicity unit |
| `signal_memory` | 20 | Predator memory window for learning |

**Expected output.** `mean_toxicity` evolves upward under predator pressure.
`n_avoided_attacks` increases as predators learn. `n_toxic_attacks` may peak
and then decline as predators avoid toxic prey.

```{r mimicry}
s <- default_specs()
s$mimicry           <- TRUE
s$n_predators_init  <- 5L
s$toxicity_init_mean <- 0.1
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

---

## Within-lifetime reinforcement learning

**What it models.** Each agent uses REINFORCE-with-baseline (actor-critic) to
update its brain's output layer based on energy reward signals during its
lifetime. This implements the Baldwin effect: individually-learned behaviours
can stabilise genetically over generations because learning relaxes the
requirement for genes to encode the exact optimal behaviour.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `rl_mode` | `"none"` | Set to `"actor_critic"` to enable |
| `rl_update_freq` | 5 | Ticks between RL updates |
| `learning_rate` | 0.01 | Step size for output-layer update |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve |

**Expected output.** Agents with RL enabled adapt within their lifetime to the
local resource distribution. When `learning_rate_evolution = TRUE`, the
evolved learning rate reflects the environmental uncertainty: noisy environments
favour lower rates; stable environments favour higher rates.

```{r rl}
s <- default_specs()
s$rl_mode    <- "actor_critic"
s$max_ticks  <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

---

## Phenotypic plasticity

**What it models.** A heritable `plasticity` trait modulates how strongly each
agent's sensory input maps to action. High plasticity agents are more
responsive to environmental cues; low plasticity agents rely more heavily on
their evolved baseline behaviour. The evolution of plasticity tracks
environmental variability: plastic phenotypes are favoured when environments
are predictably unpredictable (DeWitt & Scheiner 2004).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.5 | Starting plasticity |
| `plasticity_mutation_sd` | 0.05 | Mutation rate |

**Expected output.** `mean_plasticity` evolves. In stable environments
(fixed `grass_rate`), plasticity typically decreases over time. In
seasonally varying environments, plasticity may be maintained.

```{r plasticity}
s <- default_specs()
s$phenotypic_plasticity <- TRUE
s$max_ticks             <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final plasticity:", tail(data$ticks$mean_plasticity, 1L), "\n")
```

---

## Pure-R scenario: Evolution of bad science

**What it models.** A pure-R simulation (no Julia required) of Smaldino &
McElreath (2016). Labs compete for citations. Publication probability depends
on finding a positive result. Low statistical power (low `effort`) increases
false-positive rate (FPR). Under publication pressure, low-effort strategies
spread because they produce more positive results per unit cost. Replication
slows but does not stop the deterioration.

```{r bad-science}
# No Julia required — runs entirely in R
df_norep <- run_bad_science(n_ticks = 500L, replication_rate = 0.0, seed = 1L)
df_rep   <- run_bad_science(n_ticks = 500L, replication_rate = 0.5, seed = 1L)

library(ggplot2)
df <- rbind(
  cbind(df_norep, rate = "No replication"),
  cbind(df_rep,   rate = "50% replication")
)
ggplot(df, aes(t, mean_fpr, colour = rate)) +
  geom_line() +
  labs(title = "False-positive rate under publication pressure",
       x = "Tick", y = "Mean FPR", colour = NULL) +
  theme_minimal()
```

---

## References

DeWitt, T.J. & Scheiner, S.M. (2004) *Phenotypic Plasticity: Functional and
Conceptual Approaches*. Oxford University Press.

Fromhage, L. & Jennions, M.D. (2019) The evolution of mutualism from
reciprocal altruism by a selection regime shift. *Nature Communications*
10:1294.

Hamilton, W.D. (1964) The genetical evolution of social behaviour I & II.
*Journal of Theoretical Biology* 7(1):1-52.

Henrich, J. & McElreath, R. (2003) The evolution of cultural evolution.
*Evolutionary Anthropology* 12(3):123-135.

Liedtke, J. & Fromhage, L. (2019) The evolution of learning in the
cephalopod paradox. *Proceedings of the Royal Society B* 286:20190918.

Odling-Smee, F.J., Laland, K.N. & Feldman, M.W. (2003) *Niche Construction:
The Neglected Process in Evolution*. Princeton University Press.

Ruxton, G.D., Sherratt, T.N. & Speed, M.P. (2004) *Avoiding Attack:
The Evolutionary Ecology of Crypsis, Warning Signals and Mimicry*. Oxford
University Press.

Shine, R., Brown, G.P. & Phillips, B.L. (2011) An evolutionary process that
assembles phenotypes through space rather than through time. *Proceedings of
the National Academy of Sciences* 108(14):5708-5711.

Smaldino, P.E. & McElreath, R. (2016) The natural selection of bad science.
*Royal Society Open Science* 3:160384.

Smith, C.C. & Fretwell, S.D. (1974) The optimal balance between size and
number of offspring. *American Naturalist* 108(962):499-506.
