---
title: "Biological Scenarios and Modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biological Scenarios and Modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

Each module in `clade` implements a biological mechanism that can be enabled
independently or combined. This article documents every module: the biological
motivation, the key parameters, what to look for in the output, and a minimal
working example.

All examples require Julia. Wrap any call in `if (julia_is_ready()) { ... }`
for scripts that may run on machines without Julia.

---

## Part I — Core ecology and evolution

### Baseline world

**What it models.** Agents forage on a renewable grass resource and reproduce
when energy exceeds a threshold. The neural-network genome (brain weights)
evolves under natural selection with no additional modules. This is the
simplest demonstration that evolution can occur.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `grass_rate` | 0.10 | Probability that empty cell regrows per tick |
| `mutation_sd` | 0.05 | Gaussian noise on brain weights per offspring |
| `brain_type` | `"bnn"` | Neural architecture |
| `ploidy` | 2 | Haploid (1) or diploid (2) |

**Expected output.** Population stabilises well below `max_agents`. Genetic
diversity is maintained: neither fixation nor unbounded growth.

```{r baseline}
library(clade)

s <- default_specs()
s$grid_rows     <- 20L
s$grid_cols     <- 20L
s$n_agents_init <- 40L
s$max_ticks     <- 300L
s$random_seed   <- 1L

env  <- run_alife(s)
data <- get_run_data(env)
plot_run(data)
```

```{r fig-baseline, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: six-panel run summary showing population dynamics, mean energy, genetic diversity, births/deaths, grass coverage, and BNN sigma."}
FIG("01_run_dashboard")
```

---

### Complex landscape (forest world)

**What it models.** A three-layer habitat: ground (grass), shrubs (mid-level),
and canopy (top). Shrubs and canopy have higher energy density but slower
regrowth. Canopy access requires a minimum `wing_size` (a heritable trait).
This creates a selection pressure for wing-size evolution and niche
partitioning. The scenario is structurally analogous to the
*cephalopod paradox* (Liedtke & Fromhage 2019): a rich resource is accessible
only to agents that have evolved the morphology to exploit it.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `complex_landscape` | FALSE | Enable this module |
| `shrub_density` | 0.35 | Fraction of cells with shrubs |
| `shrub_energy` | 25.0 | Energy gained from shrubs |
| `shrub_growth_rate` | 0.04 | Shrub regrowth rate |
| `canopy_density` | 0.15 | Fraction of cells with canopy |
| `canopy_energy` | 55.0 | Energy gained from canopy |
| `canopy_threshold` | 0.6 | Minimum wing_size for canopy access |
| `wing_size_init_mean` | 0.1 | Starting wing size |
| `wing_size_mutation_sd` | 0.05 | Wing size mutation rate |

**Expected output.** `mean_wing_size` rises over time. `n_canopy_agents`
increases as wing-size evolution crosses the `canopy_threshold`.
Niche entropy (ground/shrub/canopy mix) first rises then stabilises.

```{r complex-landscape}
s <- default_specs()
s$complex_landscape   <- TRUE
s$shrub_density       <- 0.35
s$canopy_density      <- 0.15
s$canopy_energy       <- 55.0
s$canopy_threshold    <- 0.6
s$wing_size_init_mean <- 0.1
s$max_ticks           <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

# Wing size evolution
plot(data$ticks$t, data$ticks$mean_wing_size, type = "l",
     xlab = "Tick", ylab = "Mean wing size",
     main = "Wing size evolution in the forest world")
```

Use `tune_complex_landscape()` to find parameter combinations that
produce strong wing-size evolution with CMA-ES:

```{r tune-landscape}
tuned <- tune_complex_landscape(default_specs(), n_iterations = 80L)
tuned$specs   # optimal landscape parameters
```

```{r fig-complex-landscape, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: agents stratify across ground, shrub, and canopy layers as wing size evolves past the canopy threshold."}
FIG("16_habitat_preference")
```

---

### Spatial sorting

**What it models.** At an expanding invasion front, high-dispersal individuals
co-occur with other high-dispersal individuals through spatial assortment
alone — not because they have higher lifetime reproductive success.
Dispersal-enhancing alleles "surf the wave" (Shine et al. 2011). This is a
non-adaptive evolutionary mechanism: the front agent pool is genetically
distinct from the rear without any fitness difference.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal tendency |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation rate for dispersal |
| `spatial_sorting` | FALSE | Enable invasion-front mating assortment |
| `sorting_front_threshold` | 0.75 | Fraction from front to define "front zone" |
| `sorting_mating_boost` | 3.0 | Fold-increase in mate encounters at front |

**Expected output.** `mean_front_dispersal` diverges upward from
`mean_rear_dispersal` over time. The magnitude of the gap reflects the
strength of spatial sorting. Rear dispersal may drift downward if there is any
cost to dispersal.

```{r spatial-sorting}
s <- default_specs()
s$dispersal_evolution   <- TRUE
s$dispersal_init_mean   <- 0.3
s$dispersal_mutation_sd <- 0.04
s$spatial_sorting       <- TRUE
s$sorting_mating_boost  <- 3.0
s$max_ticks             <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

library(ggplot2)
ggplot(data$ticks, aes(t)) +
  geom_line(aes(y = mean_front_dispersal, colour = "Front")) +
  geom_line(aes(y = mean_rear_dispersal,  colour = "Rear"), linetype = "dashed") +
  scale_colour_manual(values = c(Front = "#e41a1c", Rear = "#377eb8"),
                      name = NULL) +
  labs(title = "Spatial sorting: dispersal diverges at the invasion front",
       x = "Tick", y = "Mean dispersal tendency") +
  theme_minimal()
```

```{r fig-spatial-sorting, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: front-zone dispersal tendency (red) diverges upward from rear-zone dispersal (blue dashed) as high-dispersal genotypes accumulate at the invasion front."}
FIG("05_dispersal")
```

---

### IFfolk inclusive fitness

**What it models.** Agents can transfer energy to nearby relatives. The
transfer amount and eligibility radius are controlled by `iffolk_*` parameters.
Cooperative behaviour evolves because it improves inclusive fitness: the
donor's fitness cost is recovered through the reproductive gains of related
recipients (Hamilton 1964). `parliament_suppression` enforces cooperation by
penalising defectors (low `helper_tendency`) surrounded by cooperators,
modelling the *parliament of genes* (Fromhage & Jennions 2019).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `iffolk_selection` | FALSE | Enable IFfolk transfers |
| `iffolk_r_min` | 0.125 | Minimum relatedness (cousins and closer) |
| `iffolk_radius` | 5 | Neighbourhood radius for kin search |
| `iffolk_transfer` | 3.0 | Energy transferred per act |
| `iffolk_min_energy` | 60.0 | Donor must exceed this energy |
| `parliament_suppression` | FALSE | Penalise defectors among cooperators |
| `parliament_cost` | 0.5 | Energy penalty for defectors |
| `cooperative_breeding` | FALSE | Enable `helper_tendency` trait |
| `helper_tendency_init_mean` | 0.2 | Starting helper tendency |

**Expected output.** `mean_helper_tendency` increases over time.
`n_iffolk_transfers` is positive and grows with population size.
With `parliament_suppression = TRUE`, the increase is faster and converges
near 1.0 for high-relatedness environments.

```{r iffolk}
s <- default_specs()
s$iffolk_selection        <- TRUE
s$iffolk_r_min            <- 0.125
s$iffolk_transfer         <- 3.0
s$parliament_suppression  <- TRUE
s$parliament_cost         <- 0.5
s$cooperative_breeding    <- TRUE
s$helper_tendency_init_mean <- 0.2
s$max_ticks               <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

plot(data$ticks$t, data$ticks$mean_helper_tendency, type = "l",
     xlab = "Tick", ylab = "Mean helper tendency",
     main = "Helper tendency evolves under IFfolk + parliament suppression")
```

```{r fig-iffolk, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean helper tendency rises over time as cooperative genotypes increase in frequency. Parliament suppression accelerates convergence."}
FIG("20_cooperative_breeding")
```

---

### Kin selection

**What it models.** Agents with energy above a threshold donate a fixed cost to
the most closely related Moore-neighbourhood agent with relatedness
r >= `kin_altruism_r_min`. Relatedness is computed from the pedigree (parent
chains). This is Hamilton's rule operationalised at each tick: the altruistic
act evolves when Hamilton's rule holds: r &times; benefit &gt; cost.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `kin_selection` | FALSE | Enable kin altruism |
| `kin_altruism_r_min` | 0.25 | Minimum relatedness to donate (siblings) |
| `kin_altruism_cost` | 2.0 | Energy transferred per act |
| `kin_altruism_min_donor_energy` | 50.0 | Donor energy floor |

**Expected output.** `n_altruistic_acts` is positive. Population size is
generally higher than baseline (altruism increases mean fitness). Genetic
diversity may be slightly lower if relatedness clustering creates local
inbreeding.

```{r kin-selection}
s <- default_specs()
s$kin_selection               <- TRUE
s$kin_altruism_r_min          <- 0.25
s$kin_altruism_cost           <- 2.0
s$max_ticks                   <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total altruistic acts:", sum(data$ticks$n_altruistic_acts), "\n")
```

```{r fig-kin-selection, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: population is sustained at higher density than baseline, supported by energy transfers between relatives. Altruistic acts accumulate over the run."}
FIG("07_kin_selection")
```

---

## Part II — Ecology modules

### SIR disease

**What it models.** A Susceptible-Infected-Recovered epidemic model. Infected
agents pay an energy cost per tick, transmit to susceptible neighbours with
probability `transmission_prob`, and recover after `disease_duration` ticks
with temporary immunity. There is also an additional per-tick death probability
for infected agents.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `disease` | FALSE | Enable SIR dynamics |
| `transmission_prob` | 0.15 | Probability of transmission per susceptible neighbour |
| `disease_energy_cost` | 5.0 | Energy drained per tick while infected |
| `disease_duration` | 10 | Ticks until recovery |
| `immune_duration` | 20 | Ticks of post-recovery immunity |
| `disease_death_prob` | 0.01 | Extra mortality per tick while infected |
| `disease_seed_prob` | 0.02 | Fraction infected at tick 1 |

**Expected output.** `n_infected` shows epidemic peaks. Population size dips
sharply during outbreaks and recovers as immunity builds. Multiple epidemic
cycles are possible if immunity wanes before the next susceptible cohort matures.

```{r disease}
s <- default_specs()
s$disease          <- TRUE
s$transmission_prob <- 0.20
s$max_ticks        <- 300L

env  <- run_alife(s)
data <- get_run_data(env)

plot(data$ticks$t, data$ticks$n_infected, type = "l",
     xlab = "Tick", ylab = "Infected agents",
     main = "SIR epidemic dynamics")
```

```{r fig-disease, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: epidemic waves visible as peaks in the infected count. Population dips sharply during outbreaks and recovers as immunity builds."}
FIG("06_disease")
```

---

### Niche construction

**What it models.** Agents build shelter at their current cell when energy
exceeds `shelter_min_energy`. Shelters reduce predator damage and slow grass
regrowth on the cell. Shelters decay stochastically each tick. This models
ecosystem engineering: organisms that modify their selective environment
(Odling-Smee et al. 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `niche_construction` | FALSE | Enable shelter building |
| `shelter_build_prob` | 0.1 | Probability of building per tick |
| `shelter_min_energy` | 40.0 | Energy required to build |
| `shelter_max_depth` | 5 | Maximum shelter units per cell |
| `shelter_decay_rate` | 0.02 | Fraction of shelters lost per tick |

**Expected output.** `n_shelters_built` is positive. Predator damage is reduced
in sheltered cells. Population size may be larger than baseline due to shelter
protection.

```{r niche-construction}
s <- default_specs()
s$niche_construction <- TRUE
s$n_predators_init   <- 5L
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total shelters built:", sum(data$ticks$n_shelters_built), "\n")
```

```{r fig-niche, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: shelter density builds up over time; predator kill rate is lower in sheltered cells, sustaining a larger population than an equivalent run without niche construction."}
FIG("09_niche")
```

---

### Body size evolution

**What it models.** A heritable `body_size` trait scales metabolic costs and
foraging gains via allometric rules. Larger body size increases move cost but
also foraging efficiency. Natural selection finds the optimal body size for the
prevailing resource density.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enable body size as a heritable trait |
| `body_size_init_mean` | 1.0 | Starting body size (relative units) |
| `body_size_mutation_sd` | 0.05 | Mutation rate |
| `body_size_min` | 0.1 | Lower clamp |
| `body_size_max` | 5.0 | Upper clamp |

**Expected output.** `mean_body_size` evolves to a stable value. High
`grass_rate` (rich environment) typically selects for larger body size;
sparse environments select for smaller.

```{r body-size}
s <- default_specs()
s$body_size_evolution  <- TRUE
s$body_size_init_mean  <- 1.0
s$max_ticks            <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final mean body size:", tail(data$ticks$mean_body_size, 1L), "\n")
```

```{r fig-body-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean body size (±SD ribbon) evolves to a stable intermediate value. The dashed line marks the baseline size of 1.0."}
FIG("04_body_size")
```

---

### Dispersal evolution

**What it models.** A heritable `dispersal_tendency` trait governs how far
agents move from their birthplace. This allows evolution to optimise dispersal
in response to local resource density and competition. Can be combined with
`spatial_sorting` to study invasion dynamics.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation rate |
| `dispersal_min / dispersal_max` | 0.0 / 1.0 | Clamp range |

**Expected output.** `mean_dispersal` evolves in response to crowding.
Dense, resource-rich environments typically select for lower dispersal
(local knowledge is valuable); sparse, patchy environments select for higher.

```{r dispersal}
s <- default_specs()
s$dispersal_evolution <- TRUE
s$max_ticks           <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
# Dispersal summary is in data$ticks$n_dispersal_events
```

```{r fig-dispersal, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean dispersal tendency evolves in response to local competition. Rich, dense environments typically select for lower dispersal."}
FIG("05_dispersal")
```

---

### Social learning

**What it models.** Agents can copy the output-layer weights of successful
neighbours (those with energy above a threshold). This is a model of vertical-
and horizontal-cultural transmission: information about which actions are
rewarding propagates socially, accelerating adaptation relative to genetic
evolution alone (Henrich & McElreath 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `social_learning` | FALSE | Enable social learning |
| `social_learning_freq` | 10 | Ticks between learning events |
| `social_learning_radius` | 3 | Search radius for models |

**Expected output.** Mean energy may be higher than in a baseline run with
identical specs, particularly in novel or changing environments. The benefit is
most visible when `mutation_sd` is very low (little genetic variation) but
successful behaviours exist in the population.

```{r social-learning}
s <- default_specs()
s$social_learning      <- TRUE
s$social_learning_freq <- 10L
s$max_ticks            <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-social-learning, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean energy and genetic diversity are maintained at higher levels than a baseline run, as successful foraging behaviours propagate through the social network."}
FIG("11_social_learning")
```

---

### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained oscillations in which predator
abundance tracks prey abundance with a characteristic lag (Lotka 1925;
Volterra 1928). In `clade`, predators are autonomous agents with their own
energy budgets, movement rules, and reproduction thresholds, so oscillations
emerge from individual behaviour rather than being imposed by differential
equations.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output.** Watch `n_agents` (prey) and `n_predators` over time. Prey
population rises first, then predators increase as food becomes abundant; prey
then crash, pulling predators down with a lag of approximately 30 ticks — a
pattern qualitatively consistent with Lotka-Volterra predictions.

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init      <- 5L
s$n_agents_init         <- 100L
s$grid_rows             <- 30L
s$grid_cols             <- 30L
s$grass_rate            <- 0.3
s$max_ticks             <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Predator-prey population dynamics"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: prey and predator populations oscillate with predator peaks lagging prey peaks by approximately 30 ticks, consistent with Lotka-Volterra dynamics."}
FIG("14_predators")
```

---

### Group defense (dilution of risk)

**What it models.** Hamilton's (1971) selfish herd hypothesis proposes that
aggregation reduces individual predation risk by diluting attacks across a
larger group. When any single predator can only attack one agent per encounter,
each additional group member decreases the per-capita probability of being the
target. In `clade`, group defense is implemented as a collective
damage-reduction mechanism: agents within `group_defense_radius` cells of an
attacked individual jointly absorb a fraction of the attack, reducing effective
predation pressure on each member.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `group_defense` | FALSE | Enables collective damage reduction in groups |
| `group_defense_radius` | 2 | Neighbourhood radius (cells) over which defense is pooled |
| `group_defense_strength` | 0.3 | Fraction by which attack damage is reduced per additional neighbour |
| `n_predators_init` | 0 | Set > 0 to activate predation pressure |
| `predator_attack_strength` | 40.0 | Baseline damage before group reduction is applied |

**Expected output.** In the `group_defense = TRUE` condition, mean population
size should be substantially higher than the baseline, and survival curves
should diverge after the first predator boom. The mechanism is dilution:
per-capita attack rate falls as group size grows, creating a positive feedback
between aggregation and survival.

```{r group-defense}
library(clade)
library(ggplot2)

base_specs <- function() {
  s <- default_specs()
  s$n_predators_init <- 5L
  s$n_agents_init    <- 100L
  s$grid_rows        <- 30L
  s$grid_cols        <- 30L
  s$max_ticks        <- 400L
  s
}

s_no <- base_specs()
s_no$group_defense <- FALSE

s_gd <- base_specs()
s_gd$group_defense          <- TRUE
s_gd$group_defense_radius   <- 2L
s_gd$group_defense_strength <- 0.3

d_no <- get_run_data(run_alife(s_no))$ticks
d_gd <- get_run_data(run_alife(s_gd))$ticks

d_no$condition <- "No group defense"
d_gd$condition <- "Group defense"
dat <- rbind(d_no, d_gd)

ggplot(dat, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("No group defense" = "#E53935", "Group defense" = "#43A047"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Group defense reduces predation-driven extinction risk"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-group-defense, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: group defense (green) sustains a larger and more stable population than the baseline (red) under identical predation pressure, illustrating the dilution-of-risk effect."}
FIG("15_group_defense")
```

---

### Seasonal dynamics

**What it models.** Seasonality imposes periodic variation in resource
availability and mortality risk that drives boom-bust dynamics across a wide
range of taxa (Boyce 1979). In `clade`, grass productivity oscillates
sinusoidally with period `season_length` and amplitude `seasonal_amplitude`.
During winter phases, grass regrowth slows and a per-tick death probability
(`winter_death_prob`) adds additional mortality pressure. The result is a
structured environment in which agents must survive periods of scarcity to
reproduce in periods of abundance — a minimal model of temperate seasonality.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `seasonal_amplitude` | 0.0 | Amplitude of the grass productivity oscillation (0 = no seasons) |
| `season_length` | 100 | Length of one full season cycle in ticks |
| `winter_death_prob` | 0.0 | Additional per-agent death probability during the winter phase |
| `grass_rate` | 0.05 | Baseline regrowth rate, modulated by the seasonal signal |

**Expected output.** Population size (`n_agents`) should track grass coverage
with a short lag, producing recurrent crashes in winter and recoveries in
spring. Multiple complete cycles should be visible over 500 ticks given
`season_length = 100`.

```{r seasonal}
library(clade)
library(ggplot2)
library(tidyr)

s <- default_specs()
s$seasonal_amplitude <- 0.8
s$season_length      <- 100L
s$winter_death_prob  <- 0.05
s$n_agents_init      <- 100L
s$grid_rows          <- 30L
s$grid_cols          <- 30L
s$max_ticks          <- 500L

env  <- run_alife(s)
tks  <- get_run_data(env)$ticks

tks$grass_scaled <- tks$grass_coverage /
  max(tks$grass_coverage, na.rm = TRUE) *
  max(tks$n_agents,       na.rm = TRUE)

dat_long <- pivot_longer(
  tks[, c("t", "n_agents", "grass_scaled")],
  cols      = c("n_agents", "grass_scaled"),
  names_to  = "variable",
  values_to = "value"
)
dat_long$variable <- ifelse(dat_long$variable == "n_agents",
                            "Population size", "Grass coverage (scaled)")

ggplot(dat_long, aes(x = t, y = value, colour = variable)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("Population size" = "#1565C0",
               "Grass coverage (scaled)" = "#2E7D32"),
    name = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Value (agents / scaled grass)",
    title = "Seasonal boom-bust dynamics"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-seasonal, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: population size (blue) tracks scaled grass productivity (green) with a short lag, producing recurrent winter crashes and spring recoveries over approximately five complete cycles."}
FIG("17_seasons")
```

---

### Habitat preference and ideal free distribution

**What it models.** The ideal free distribution (IFD) predicts that, under
perfect information and free movement, individuals should distribute themselves
across habitat patches proportionally to resource availability, equalising
per-capita intake rates (Fretwell & Lucas 1970). In `clade`, habitat preference
is a heritable, continuously varying trait. Positive values bias movement toward
high-grass cells; negative values toward low-grass cells. When
`habitat_preference_evolution = TRUE`, selection acts on this trait because
agents that track resource-rich patches acquire more energy, survive longer, and
leave more offspring.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `habitat_preference_evolution` | FALSE | Enables evolution of the habitat preference trait |
| `habitat_preference_init_mean` | 0.0 | Starting mean preference (0 = no bias) |
| `habitat_preference_mutation_sd` | 0.03 | Mutational step size per generation |
| `habitat_preference_strength` | 0.5 | Strength of bias applied to movement decisions |

**Expected output.** Watch `mean_habitat_preference` over ticks. Starting near
zero, the trait mean should drift toward positive values as agents that prefer
grass-rich cells out-compete those that do not. Spatial plots at the end of the
run should show agents concentrated near dense grass patches, consistent with
the IFD prediction.

```{r habitat-pref}
library(clade)
library(ggplot2)

s <- default_specs()
s$habitat_preference_evolution   <- TRUE
s$habitat_preference_init_mean   <- 0.0
s$habitat_preference_mutation_sd <- 0.03
s$habitat_preference_strength    <- 0.5
s$n_agents_init                  <- 100L
s$grid_rows                      <- 30L
s$grid_cols                      <- 30L
s$max_ticks                      <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_habitat_preference)) +
  geom_line(colour = "#6A1B9A", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    x     = "Tick",
    y     = "Mean habitat preference",
    title = "Evolution of habitat preference toward grass-rich patches"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-habitat-pref, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean habitat preference rises from zero toward positive values as selection favours individuals that concentrate near high-grass cells, consistent with ideal free distribution theory."}
FIG("16_habitat_preference")
```

---

### Scavenging and carrion dynamics

**What it models.** Scavenging — the consumption of carrion left by dead
conspecifics — is a widespread foraging strategy that provides a resource
buffer when primary productivity is low (DeVault et al. 2003). In `clade`,
when `scavenging = TRUE`, agent deaths deposit a fraction of their remaining
energy as carrion on the grid cell where they died. Carrion decays each tick
at rate `carrion_decay_rate` and can be consumed by any agent that moves onto
the cell, yielding `carrion_eat_gain` energy units.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `scavenging` | FALSE | Enables carrion deposition and consumption |
| `carrion_fraction` | 0.5 | Fraction of a dead agent's energy deposited as carrion |
| `carrion_decay_rate` | 0.1 | Proportional decay of carrion per tick |
| `carrion_eat_gain` | 3.0 | Energy gained by a scavenger per unit of carrion consumed |
| `grass_rate` | 0.05 | Set low to create resource scarcity that accentuates the scavenging advantage |

**Expected output.** Under scarce grass conditions, mean agent energy should be
higher in the scavenging condition than the baseline, and the total number of
deaths should be lower. Carrion provides a density-dependent buffer: more
deaths in a tick produce more carrion, partially compensating for the energy
deficit that caused those deaths.

```{r scavenging}
library(clade)
library(ggplot2)

base_specs <- function() {
  s <- default_specs()
  s$n_agents_init <- 100L
  s$grid_rows     <- 30L
  s$grid_cols     <- 30L
  s$grass_rate    <- 0.15
  s$max_ticks     <- 400L
  s
}

s_no <- base_specs(); s_no$scavenging <- FALSE
s_sc <- base_specs()
s_sc$scavenging        <- TRUE
s_sc$carrion_fraction  <- 0.5
s_sc$carrion_decay_rate <- 0.1
s_sc$carrion_eat_gain  <- 3.0

d_no <- get_run_data(run_alife(s_no))$ticks
d_sc <- get_run_data(run_alife(s_sc))$ticks

d_no$condition <- "No scavenging"
d_sc$condition <- "Scavenging"
dat <- rbind(d_no, d_sc)

ggplot(dat, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("No scavenging" = "#E53935", "Scavenging" = "#FB8C00"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Mean agent energy",
    title = "Scavenging sustains energy budgets under resource scarcity"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-scavenging, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: scavengers (orange) maintain higher mean energy than non-scavengers (red) under identical low-grass conditions, with carrion providing a density-dependent buffer against starvation."}
FIG("scavenging")
```

---

## Part III — Reproduction and life history

### Parental care

**What it models.** Offspring are carried by the parent until graduation at age
`care_duration`. While carried, juveniles receive energy from the parent at
`feeding_rate` per tick. If the parent dies, carried offspring die too
(obligate altriciality). This models the evolutionary trade-off between clutch
size and offspring quality (Smith & Fretwell 1974).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_care` | FALSE | Enable parental care |
| `care_duration` | 5 | Ticks offspring are carried |
| `care_cost_per_tick` | 1.0 | Energy drained from parent per juvenile per tick |
| `feeding_rate` | 5.0 | Energy transferred to juvenile per tick |

**Expected output.** `n_juveniles` is positive. Per-capita offspring count may
be lower than baseline (parents can carry fewer), but juvenile survival is
higher. Population dynamics are more buffered.

```{r parental-care}
s <- default_specs()
s$parental_care     <- TRUE
s$care_duration     <- 5L
s$care_cost_per_tick <- 2.0
s$max_ticks         <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total juveniles recorded:", sum(data$ticks$n_juveniles, na.rm = TRUE), "\n")
```

```{r fig-parental-care, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: juvenile count is positive throughout the run; per-capita offspring count is lower than baseline but juvenile survival is higher, resulting in more buffered population dynamics."}
FIG("19_parental_care")
```

---

### Mating systems: sexual versus asexual reproduction

**What it models.** Sexual reproduction incurs a well-documented two-fold cost
relative to asexual reproduction — every sexual individual devotes half its
offspring capacity to producing males — yet sex is widespread across eukaryotes
(Maynard Smith 1978; Williams 1975). The leading explanations invoke the
recombination benefits of sex: sexual populations generate more genotypic
variation per generation, providing raw material for selection in fluctuating or
parasitised environments. This scenario compares a haploid asexual population
(`ploidy = 1L`) against a diploid sexual population (`ploidy = 2L`,
`crossover_rate = 0.1`) to illustrate how mating system shapes the trajectory
of genetic diversity.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `ploidy` | 2L | 1 = haploid asexual; 2 = diploid sexual |
| `crossover_rate` | 0.0 | Probability of crossover per chromosome per reproduction event |
| `n_chromosomes` | 1L | Number of chromosome pairs (diploid only) |
| `dominance_model` | `"additive"` | How alleles at the same locus combine |

**Expected output.** `genetic_diversity` is higher and more variable in the
sexual condition from early in the run, as recombination continuously generates
novel genotypic combinations. The asexual population maintains lower diversity
and may converge faster to a local optimum — an advantage in stable
environments — but is less able to respond to environmental change.

```{r mating-systems}
library(clade)
library(ggplot2)

s_asex <- default_specs()
s_asex$ploidy         <- 1L
s_asex$crossover_rate <- 0.0
s_asex$max_ticks      <- 400L
s_asex$random_seed    <- 42L

s_sex <- default_specs()
s_sex$ploidy         <- 2L
s_sex$crossover_rate <- 0.1
s_sex$max_ticks      <- 400L
s_sex$random_seed    <- 42L

d_asex <- get_run_data(run_alife(s_asex))$ticks
d_sex  <- get_run_data(run_alife(s_sex))$ticks

df <- rbind(
  cbind(d_asex[, c("t", "genetic_diversity")], system = "Asexual (haploid)"),
  cbind(d_sex[,  c("t", "genetic_diversity")], system = "Sexual (diploid)")
)

ggplot(df, aes(t, genetic_diversity, colour = system)) +
  geom_line() +
  scale_colour_manual(values = c("Asexual (haploid)" = "#377eb8",
                                 "Sexual (diploid)"  = "#e41a1c"),
                      name = NULL) +
  labs(title = "Genetic diversity: sexual vs asexual reproduction",
       x = "Tick", y = "Genetic diversity") +
  theme_minimal()
```

```{r fig-mating-systems, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: genetic diversity is higher and more variable in the sexual (diploid) condition. The asexual population converges to lower diversity more rapidly, illustrating the recombination advantage of sex."}
FIG("mating_systems")
```

---

### Life history: semelparous versus iteroparous strategies

**What it models.** Cole's paradox (Cole 1954) established that the fitness
advantage of iteroparity over semelparity is surprisingly small — one additional
surviving offspring is sufficient to equalise the two strategies under simple
demographic conditions. Real divergence arises from differences in juvenile
versus adult survival, resource availability, and the reliability of the
reproductive season (Roff 1992). Semelparous organisms invest all reproductive
resources in a single event and die; iteroparous organisms spread reproduction
across multiple seasons. This scenario contrasts the two life histories, holding
other parameters constant, to reveal how each shapes population age structure
and birth-rate dynamics.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `life_history` | `"iteroparous"` | `"semelparous"` triggers post-reproductive death |
| `max_age` | 200L | Maximum attainable age (iteroparous bound) |
| `senescence_rate` | 0.0 | Gompertz senescence coefficient |
| `repro_senescence` | 0.0 | Reproductive decline rate with age |

**Expected output.** The semelparous population shows a shorter `mean_age` and
episodic bursts in `n_births` coinciding with cohort turnover. The iteroparous
population maintains a smoother, more continuous birth rate and a higher
`mean_age`, with population size fluctuating less dramatically between
generations.

```{r life-history}
library(clade)
library(ggplot2)

s_sem <- default_specs()
s_sem$life_history <- "semelparous"
s_sem$max_ticks    <- 400L
s_sem$random_seed  <- 7L

s_ite <- default_specs()
s_ite$life_history <- "iteroparous"
s_ite$max_ticks    <- 400L
s_ite$random_seed  <- 7L

d_sem <- get_run_data(run_alife(s_sem))$ticks
d_ite <- get_run_data(run_alife(s_ite))$ticks

df <- rbind(
  cbind(d_sem[, c("t", "mean_age", "n_births")], strategy = "Semelparous"),
  cbind(d_ite[, c("t", "mean_age", "n_births")], strategy = "Iteroparous")
)

ggplot(df, aes(t, mean_age, colour = strategy)) +
  geom_line() +
  scale_colour_manual(values = c(Semelparous = "#e41a1c", Iteroparous = "#4daf4a"),
                      name = NULL) +
  labs(title = "Mean age: semelparous vs iteroparous life history",
       x = "Tick", y = "Mean agent age") +
  theme_minimal()
```

```{r fig-life-history, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: the semelparous population (red) shows lower mean age and episodic birth bursts; the iteroparous population (green) sustains a higher mean age and a smoother birth-rate trajectory."}
FIG("life_history")
```

---

### Clutch size evolution and r/K strategy

**What it models.** The Lack clutch-size hypothesis (Lack 1947) proposes that
parents should produce the number of offspring that maximises the number
successfully raised, given the resources available per offspring. Smith &
Fretwell (1974) formalised the quantity–quality trade-off: when resources are
scarce, producing fewer, better-provisioned offspring (K-strategy) yields higher
per-offspring survival than producing many poorly provisioned ones (r-strategy).
This scenario allows clutch size to evolve freely between `clutch_size_min` and
`clutch_size_max` under two contrasting resource environments.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `clutch_size_evolution` | FALSE | Enable heritable clutch size |
| `clutch_size_min` | 1L | Minimum evolvable clutch size |
| `clutch_size_max` | 5L | Maximum evolvable clutch size |
| `clutch_size_mutation_sd` | 0.3 | Gaussian noise on clutch size per offspring |
| `grass_rate` | 0.10 | Resource renewal rate (varied between conditions) |

**Expected output.** Under rich resources (`grass_rate = 0.4`),
per-tick births are higher as large clutches are sustainable.
Under scarce resources (`grass_rate = 0.05`), per-tick births are lower
and the population is constrained by energy — consistent with the K end
of the r/K continuum. The two conditions diverge substantially after
approximately 150 ticks.

```{r clutch-size}
library(clade)
library(ggplot2)

make_s <- function(gr) {
  s <- default_specs()
  s$clutch_size_evolution   <- TRUE
  s$clutch_size_min         <- 1L
  s$clutch_size_max         <- 5L
  s$clutch_size_mutation_sd <- 0.3
  s$grass_rate              <- gr
  s$max_ticks               <- 400L
  s$random_seed             <- 3L
  s
}

d_rich <- get_run_data(run_alife(make_s(0.4)))$ticks
d_scar <- get_run_data(run_alife(make_s(0.05)))$ticks

df <- rbind(
  cbind(d_rich[, c("t", "n_births")], environment = "Rich (0.4)"),
  cbind(d_scar[, c("t", "n_births")], environment = "Scarce (0.05)")
)

ggplot(df, aes(t, n_births, colour = environment)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_colour_manual(
    values = c("Rich (0.4)" = "#e41a1c", "Scarce (0.05)" = "#377eb8"),
    name = NULL) +
  labs(title = "Births per tick under contrasting resource availability",
       subtitle = "Clutch size evolution enabled",
       x = "Tick", y = "Births per tick") +
  theme_minimal()
```

```{r fig-clutch-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean clutch size diverges between resource environments. Rich conditions (red) select for larger clutches; scarce conditions (blue) select for smaller clutches, illustrating the r/K continuum."}
FIG("clutch_size")
```

---

### Parental investment: quality versus quantity

**What it models.** Trivers (1972) proposed that the sex that invests more per
offspring should be the one that is more choosy in mate selection, because its
reproductive success is more constrained by parental effort than by mate access.
Houston & Davies (1985) extended this to biparental care games, showing that
evolutionarily stable investment levels depend on the shape of the offspring
fitness function. This scenario holds `parental_care = TRUE` and varies
`female_investment` — the fraction of care energetics contributed by the female
— to examine how the allocation of investment between parents affects offspring
provisioning and survival.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_investment_evolution` | FALSE | Allow investment proportions to evolve |
| `parental_care` | FALSE | Enable offspring carrying and provisioning |
| `female_investment` | 0.7 | Fraction of total care energy contributed by the mother |
| `male_repro_cost` | 0.3 | Energy cost to the male per reproductive event |

**Expected output.** High maternal investment (`female_investment = 0.9`)
increases the per-offspring energy cost, so fewer births occur per tick.
Equal investment (`female_investment = 0.5`) allows more births at lower
per-offspring cost. The trade-off is visible in `n_births` and
`n_juveniles` trajectories.

```{r parental-investment}
library(clade)
library(ggplot2)

make_s <- function(fi) {
  s <- default_specs()
  s$parental_care                 <- TRUE
  s$parental_investment_evolution <- TRUE
  s$female_investment             <- fi
  s$male_repro_cost               <- 0.3
  s$max_ticks                     <- 400L
  s$random_seed                   <- 11L
  s
}

d_hi <- get_run_data(run_alife(make_s(0.9)))$ticks
d_eq <- get_run_data(run_alife(make_s(0.5)))$ticks

df <- rbind(
  cbind(d_hi[, c("t", "n_births", "n_juveniles")],
        condition = "High maternal (0.9)"),
  cbind(d_eq[, c("t", "n_births", "n_juveniles")],
        condition = "Equal (0.5)")
)

ggplot(df, aes(t, n_births, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_colour_manual(
    values = c("High maternal (0.9)" = "#e41a1c", "Equal (0.5)" = "#377eb8"),
    name = NULL) +
  labs(title = "Births per tick by parental investment allocation",
       subtitle = "High maternal investment → fewer but better-provisioned offspring",
       x = "Tick", y = "Births per tick") +
  theme_minimal()
```

```{r fig-parental-investment, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: high maternal investment (red) produces fewer but better-provisioned offspring; equal investment (blue) produces more offspring at lower individual quality, illustrating the quality-quantity trade-off."}
FIG("parental_investment")
```

---

### Pace-of-life syndromes

**What it models.** Pace-of-life syndromes (Réale et al. 2010) describe the
co-variation of life-history traits, behaviour, and physiology along a slow–fast
continuum. Fast-living organisms have high metabolic rates, reproduce early and
often, and die young; slow-living organisms invest in maintenance, defer
reproduction, and achieve longer lifespans. Stearns (1992) formalised this as a
consequence of the age-specific mortality schedule: high extrinsic mortality
selects for early reproduction and fast metabolism. This scenario fixes metabolic
rate across three values to isolate the direct effect of metabolic pace on the
age–reproduction–energy trade-off.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `metabolic_rate_init_mean` | 1.0 | Initial (and fixed) metabolic rate |
| `metabolic_rate_evolution` | FALSE | Keep metabolic rate fixed for this scenario |
| `metabolic_rate_min` | 0.1 | Lower clamp (relevant when evolution is enabled) |
| `metabolic_rate_max` | 5.0 | Upper clamp |

**Expected output.** Fast-pace agents (`metabolic_rate_init_mean = 2.0`)
exhibit higher `n_births` per tick, lower `mean_age`, and higher energy
throughput, but their populations are more volatile. Slow-pace agents
(`metabolic_rate_init_mean = 0.5`) show lower birth rate, higher `mean_age`,
and more stable `mean_energy`. Fast agents dominate in productive environments;
slow agents persist more reliably when resources are lean.

```{r pace-of-life}
library(clade)
library(ggplot2)

make_run <- function(rate, seed = 99L) {
  s <- default_specs()
  s$metabolic_rate_init_mean <- rate
  s$metabolic_rate_evolution <- FALSE
  s$max_ticks                <- 400L
  s$random_seed              <- seed
  get_run_data(run_alife(s))$ticks
}

d_slow     <- make_run(0.5)
d_baseline <- make_run(1.0)
d_fast     <- make_run(2.0)

df <- rbind(
  cbind(d_slow[,     c("t", "mean_age", "n_births")], pace = "Slow (0.5)"),
  cbind(d_baseline[, c("t", "mean_age", "n_births")], pace = "Baseline (1.0)"),
  cbind(d_fast[,     c("t", "mean_age", "n_births")], pace = "Fast (2.0)")
)

ggplot(df, aes(t, mean_age, colour = pace)) +
  geom_line() +
  scale_colour_manual(
    values = c("Slow (0.5)" = "#377eb8", "Baseline (1.0)" = "#4daf4a",
               "Fast (2.0)" = "#e41a1c"),
    name = NULL) +
  labs(title = "Pace-of-life: mean age across metabolic rates",
       x = "Tick", y = "Mean agent age") +
  theme_minimal()
```

```{r fig-pace-of-life, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: slow-pace agents (blue) maintain the highest mean age; fast-pace agents (red) have the lowest mean age but the highest per-tick birth rate. The three trajectories illustrate the life-history trade-off along the slow-fast continuum."}
FIG("pace_of_life")
```

---

## Part IV — Social and coevolutionary dynamics

### Mimicry and toxicity

**What it models.** A heritable `toxicity` trait makes prey costly to attack.
Predators learn signal–toxicity associations via a Rescorla-Wagner rule and
build avoidance. This models the evolution of warning coloration and Batesian/
Müllerian mimicry (Ruxton et al. 2004): toxic prey are avoided; non-toxic
prey that resemble toxic prey are also avoided (Batesian mimicry).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `mimicry` | FALSE | Enable toxicity / predator learning |
| `toxicity_init_mean` | 0.0 | Starting toxicity (0 = non-toxic) |
| `toxin_dose` | 2.0 | Damage dealt to attacker per toxicity unit |
| `signal_memory` | 20 | Predator memory window for learning |

**Expected output.** `mean_toxicity` evolves upward under predator pressure.
`n_avoided_attacks` increases as predators learn. `n_toxic_attacks` may peak
and then decline as predators avoid toxic prey.

```{r mimicry}
s <- default_specs()
s$mimicry            <- TRUE
s$n_predators_init   <- 5L
s$toxicity_init_mean <- 0.1
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-mimicry, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean toxicity rises under predator pressure; avoided attacks increase as predators learn to avoid toxic prey; toxic attack rate declines."}
FIG("21_mimicry")
```

---

### Cooperative breeding and public goods

**What it models.** When `cooperation_evolution = TRUE`, each agent carries a
heritable cooperation tendency trait that determines its contribution to a local
public goods pool. The pool benefit is shared among all agents in the
neighbourhood, scaled by `cooperation_multiplier`, while contributors pay a
fixed `cooperation_cost`. This implements a spatially explicit Prisoner's
Dilemma: defectors (low tendency) free-ride on cooperators but, because
interactions are local, spatial clustering of cooperators can shield them from
invasion by defectors — a result established analytically by Nowak & May (1992)
and extended to continuous strategies by Hauert et al. (2002).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `cooperation_evolution` | FALSE | Enables heritable cooperation tendency |
| `cooperation_multiplier` | 2.0 | Return multiplier on pooled contributions |
| `cooperation_cost` | 1.0 | Per-tick energy cost paid by the contributor |
| `cooperation_init_mean` | 0.5 | Initial mean cooperation tendency |
| `cooperation_mutation_sd` | 0.05 | Mutation standard deviation on the trait |

**Expected output.** Track `mean_cooperation_level` over ticks. When
`cooperation_multiplier` exceeds the cost-benefit threshold (here
2.5 > 1.0 / 0.5 = 2.0), cooperation is favoured and the population-mean
tendency should rise toward 1.0; spatial structure sustains cooperation even at
intermediate global frequencies.

```{r cooperation}
library(clade)
library(ggplot2)

s <- default_specs()
s$cooperation_evolution  <- TRUE
s$cooperation_multiplier <- 2.5
s$cooperation_cost       <- 1.0
s$max_ticks              <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_cooperation_level)) +
  geom_line(colour = "#1b7837", linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, colour = "#762a83",
              fill = "#c2a5cf", alpha = 0.3) +
  labs(
    title    = "Evolution of cooperation tendency",
    subtitle = "cooperation_multiplier = 2.5, cost = 1.0",
    x = "Tick", y = "Mean cooperation tendency"
  ) +
  theme_minimal()
```

```{r fig-cooperation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean cooperation tendency rises over evolutionary time when the multiplier exceeds the cost-benefit threshold. Spatial clustering accelerates the transition by shielding cooperator clusters from defector invasion."}
FIG("08_cooperation")
```

---

### Signals and mate choice: sexual selection

**What it models.** Sexual selection can produce elaborate traits that reduce
survival but increase mating success. Two contrasting mechanisms have been
proposed. Zahavi's (1975) handicap principle requires that signals be
condition-dependent and costly to maintain, so only high-quality individuals
can afford elaborate signals; signal magnitude is then an honest indicator of
genetic quality. Fisher's (1930) runaway process operates through a genetic
correlation between signal elaboration and preference: once females prefer a
trait, sons of choosy females inherit the trait and daughters inherit the
preference, driving both to fixation. This scenario enables multi-dimensional
signals and female preference, allowing both mechanisms to operate, and tracks
whether signal magnitude remains correlated with `mean_energy` (consistent with
honest signalling) or becomes decoupled (consistent with runaway dynamics).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `signal_dims` | 0L | Number of heritable signal dimensions |
| `signal_cost` | 0.1 | Energy cost per unit of total signal magnitude |
| `mate_choice_mode` | `"random"` | `"preference"` enables directional mate choice |
| `mate_choice_strength` | 0.5 | Weight given to signal match in mate selection |

**Expected output.** `mean_signal_magnitude` increases over time as sexual
selection drives signal elaboration. When `signal_cost` is sufficient to
maintain honesty, `mean_signal_magnitude` correlates positively with
`mean_energy`: agents in better condition carry more elaborate signals. A weak
or zero `signal_cost` allows decoupling of signal from condition, consistent
with runaway dynamics.

```{r signals-matechoice}
library(clade)
library(ggplot2)

s <- default_specs()
s$signal_dims          <- 3L
s$signal_cost          <- 0.05
s$mate_choice_mode     <- "preference"
s$mate_choice_strength <- 0.7
s$max_ticks            <- 400L
s$random_seed          <- 21L

env  <- run_alife(s)
data <- get_run_data(env)

p1 <- ggplot(data$ticks, aes(t, mean_signal_magnitude)) +
  geom_line(colour = "#984ea3") +
  labs(title = "Signal elaboration under sexual selection",
       x = "Tick", y = "Mean signal magnitude") +
  theme_minimal()

p2 <- ggplot(data$ticks, aes(mean_energy, mean_signal_magnitude)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "#984ea3") +
  geom_smooth(method = "lm", se = FALSE, colour = "#333333", linewidth = 0.7) +
  labs(title = "Signal magnitude vs mean energy (honest signalling check)",
       x = "Mean energy", y = "Mean signal magnitude") +
  theme_minimal()

print(p1)
print(p2)
```

```{r fig-signals-matechoice, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean signal magnitude rises over time as sexual selection drives elaboration. A positive correlation between signal magnitude and mean energy indicates that the handicap mechanism is maintaining signal honesty."}
FIG("signals_matechoice")
```

---

### Speciation and genetic divergence

**What it models.** Speciation by ecological character displacement occurs when
disruptive selection on a resource-use trait, combined with assortative mating
among similar phenotypes, drives a single population to split into
reproductively isolated lineages (Dieckmann & Doebeli 1999). Once genomes
diverge beyond a threshold, gene flow between incipient species is reduced,
accelerating further divergence. Coyne & Orr (2004) reviewed extensive
empirical evidence that reproductive isolation accumulates approximately
linearly with genetic distance. This scenario enables the speciation module,
tracks the number of distinct genetic clusters (`n_species`) detected at
regular intervals, and illustrates how within-species diversity compares to
between-species divergence over the course of the run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `speciation` | FALSE | Enable genetic-cluster detection and isolation |
| `isolation_threshold` | 0.5 | Minimum genetic distance required for reproductive isolation |
| `speciation_cluster_interval` | 10L | Ticks between cluster re-detection passes |

**Expected output.** `n_species` starts at 1 and rises to 2–4 as accumulated
genome divergence exceeds `isolation_threshold` in geographically or
ecologically separated subpopulations. Within-cluster genetic diversity remains
lower than between-cluster diversity once isolation is established. The timing
of the first split depends on population size, mutation rate, and the
stringency of `isolation_threshold`.

```{r speciation}
library(clade)
library(ggplot2)

s <- default_specs()
s$speciation                  <- TRUE
s$isolation_threshold         <- 0.5
s$speciation_cluster_interval <- 10L
s$max_ticks                   <- 500L
s$random_seed                 <- 55L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(t, n_species)) +
  geom_step(colour = "#ff7f00", linewidth = 0.9) +
  scale_y_continuous(breaks = 1:8) +
  labs(title = "Speciation: number of distinct genetic clusters over time",
       x = "Tick", y = "Number of species (clusters)") +
  theme_minimal()
```

```{r fig-18-speciation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: n_species rises from 1 to 2-4 as genome divergence accumulates beyond the isolation threshold. Each step in the staircase represents a speciation event."}
FIG("18_speciation")
```

---

## Part V — Learning and cognition

### Within-lifetime reinforcement learning

**What it models.** Each agent uses REINFORCE-with-baseline (actor-critic) to
update its brain's output layer based on energy reward signals during its
lifetime. This implements the Baldwin effect: individually-learned behaviours
can stabilise genetically over generations because learning relaxes the
requirement for genes to encode the exact optimal behaviour.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `rl_mode` | `"none"` | Set to `"actor_critic"` to enable |
| `rl_update_freq` | 5 | Ticks between RL updates |
| `learning_rate` | 0.01 | Step size for output-layer update |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve |

**Expected output.** Agents with RL enabled adapt within their lifetime to the
local resource distribution. When `learning_rate_evolution = TRUE`, the
evolved learning rate reflects the environmental uncertainty: noisy environments
favour lower rates; stable environments favour higher rates.

```{r rl}
s <- default_specs()
s$rl_mode   <- "actor_critic"
s$max_ticks <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-rl, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: agents with within-lifetime RL adapt to local resource distribution within their lifetime. The Baldwin Effect is visible as the BNN prior sigma narrows over generations."}
FIG("10_rl")
```

---

### Phenotypic plasticity

**What it models.** A heritable `plasticity` trait modulates how strongly each
agent's sensory input maps to action. High plasticity agents are more responsive
to environmental cues; low plasticity agents rely more heavily on their evolved
baseline behaviour. The evolution of plasticity tracks environmental variability:
plastic phenotypes are favoured when environments are predictably unpredictable
(DeWitt & Scheiner 2004).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.5 | Starting plasticity |
| `plasticity_mutation_sd` | 0.05 | Mutation rate |

**Expected output.** `mean_plasticity` evolves. In stable environments
(fixed `grass_rate`), plasticity typically decreases over time. In
seasonally varying environments, plasticity may be maintained.

```{r plasticity}
s <- default_specs()
s$phenotypic_plasticity <- TRUE
s$max_ticks             <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final plasticity:", tail(data$ticks$mean_plasticity, 1L), "\n")
```

```{r fig-plasticity, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean plasticity evolves over time. In stable environments it typically decreases; in seasonally varying environments it is maintained at intermediate values."}
FIG("22_plasticity")
```

---

### BNN uncertainty canalization — the Baldwin Effect

**What it models.** The Bayesian neural network (BNN) brain type
(`brain_type = "bnn"`) places a prior distribution over each synaptic weight,
parameterised by a standard deviation σ. A large σ encodes behavioural
flexibility: the agent samples widely from its weight posterior and can learn
within its lifetime. A small σ encodes genetic assimilation: the agent's
behaviour is largely determined by the prior mean — the population has, over
generations of selection, canalized a formerly learned behaviour into the
genome. This is the computational signature of the Baldwin Effect (Baldwin
1896), formalised computationally by Hinton & Nowlan (1987) and analysed
analytically by Mayley (1996): learning guides evolution toward fixed adaptive
solutions, after which the learning machinery becomes redundant.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `brain_type` | `"bnn"` | Must be `"bnn"` to enable Bayesian weight distributions |
| `learning_rate_init_mean` | 0.01 | Initial within-lifetime learning rate |
| `max_ticks` | — | Longer runs show fuller canalization |

**Expected output.** Track `mean_prior_sigma` in `data$ticks`. The column
should decline from approximately 0.3 (broad, flexible priors) toward 0.05
(narrow, assimilated priors) as selection repeatedly rewards the same weight
configurations. The decline rate should accelerate when genetic diversity is
high, because a large pool of variants provides a stronger selection gradient.

```{r bnn-uncertainty}
library(clade)
library(ggplot2)

s <- default_specs()
s$brain_type <- "bnn"
s$max_ticks  <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_prior_sigma)) +
  geom_line(colour = "#2166ac", linewidth = 0.8, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "#d6604d", fill = "#f4a582", alpha = 0.4) +
  labs(
    title    = "Genetic assimilation of learned behaviour (Baldwin Effect)",
    subtitle = "BNN prior sigma declines as behaviour canalizes into the genome",
    x = "Tick", y = expression("Mean prior " * sigma)
  ) +
  theme_minimal()
```

```{r fig-bnn-uncertainty, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean prior sigma declines monotonically from broad (flexible, ~0.3) to narrow (assimilated, ~0.05), with the steepest drop in periods of high genetic diversity. A fitted linear trend confirms the directional pattern."}
FIG("bnn_uncertainty")
```

---

### The cephalopod paradox — short lifespan selects for fast learning

**What it models.** Classical life-history theory predicts that long-lived
organisms should invest more heavily in learning, because only a long lifespan
provides enough time to recoup the energetic costs of a large brain (the payback
period argument). The cephalopod paradox inverts this prediction: octopuses and
squid have among the shortest lifespans of any large-brained animal, yet they
express sophisticated within-lifetime learning. Liedtke & Fromhage (2019)
formalised the resolution: when lifespan is too brief for genetic evolution to
converge on food-finding solutions, within-lifetime learning becomes the only
viable mechanism for adaptive behaviour. Here, `complex_landscape = TRUE`
creates a heterogeneous resource topology that cannot be solved by a fixed
behavioural rule, and `learning_rate_evolution = TRUE` allows the genome to
optimise how quickly agents update their neural weights from experience.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `max_age` | — | Maximum agent lifespan in ticks |
| `complex_landscape` | FALSE | Enables spatially heterogeneous, shifting resource patches |
| `learning_rate_evolution` | FALSE | Makes the within-lifetime learning rate a heritable trait |
| `learning_rate_init_mean` | 0.01 | Initial mean evolved learning rate |

**Expected output.** The evolved mean learning rate (recorded at the final tick
of each run) should be highest at short `max_age` values and decline as
lifespan increases, because long-lived agents can afford to rely on genetic
evolution to encode solutions directly. The relationship need not be monotone
at very short lifespans, where agents have too little experience to benefit
from any learning.

```{r cephalopod}
library(clade)
library(ggplot2)

lifespans  <- c(20L, 50L, 100L, 200L)
specs_list <- lapply(lifespans, function(ma) {
  s <- default_specs()
  s$max_age               <- ma
  s$complex_landscape     <- TRUE
  s$learning_rate_evolution <- TRUE
  s$max_ticks             <- 400L
  s
})
names(specs_list) <- paste0("max_age_", lifespans)

results <- batch_alife(specs_list, n_cores = 4L)

lr_df <- do.call(rbind, mapply(function(env, ma) {
  d <- get_run_data(env)$ticks
  data.frame(
    max_age       = ma,
    mean_lr_final = tail(d$mean_learning_rate, 1L)
  )
}, results, lifespans, SIMPLIFY = FALSE))

ggplot(lr_df, aes(x = max_age, y = mean_lr_final)) +
  geom_line(colour = "#e08214", linewidth = 1) +
  geom_point(size = 3, colour = "#8073ac") +
  labs(
    title    = "The cephalopod paradox: short lifespan selects for fast learning",
    subtitle = "Evolved mean learning rate at end of run, by maximum lifespan",
    x = "Maximum lifespan (ticks)", y = "Evolved mean learning rate"
  ) +
  theme_minimal()
```

```{r fig-cephalopod, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: evolved learning rate is highest under the shortest lifespan (max_age = 20) and declines as lifespan increases, consistent with the cephalopod paradox. Genetic evolution substitutes for within-lifetime learning when the payback period is long enough."}
FIG("cephalopod_paradox")
```

---

### Stress hypermutation — the SOS response

**What it models.** In bacteria, severe DNA damage or prolonged resource
depletion triggers the SOS response: a suite of error-prone polymerases that
dramatically increase the genomic mutation rate (McKenzie & Rosenberg 2001).
This is a bet-hedging strategy — most mutations are deleterious, but the
expanded variance in offspring phenotype increases the probability that at least
one lineage will escape the current fitness trough. When
`stress_hypermutation = TRUE`, agents whose energy falls below `stress_threshold`
apply a mutation rate multiplied by `stress_mutation_multiplier` to their neural
genome. Pairing this with a scarce-resource environment (`grass_rate = 0.05`)
ensures that stress is frequent enough to detect the effect within a tractable
run length.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `stress_hypermutation` | FALSE | Enables condition-dependent mutation rate elevation |
| `stress_threshold` | 20.0 | Energy level below which hypermutation activates |
| `stress_mutation_multiplier` | 3.0 | Factor by which mutation rate is scaled under stress |
| `grass_rate` | — | Resource regrowth rate; reduce to impose chronic scarcity |

**Expected output.** Compare `genetic_diversity` trajectories across the two
conditions. Under stress hypermutation, diversity should spike transiently
during resource crashes (low `mean_energy` epochs), followed by faster adaptive
recovery as beneficial mutants sweep. The baseline condition should show slower,
smoother recovery after equivalent crashes.

```{r stress-hypermutation}
library(clade)
library(ggplot2)

make_s <- function(hypermut) {
  s <- default_specs()
  s$stress_hypermutation       <- hypermut
  s$stress_threshold           <- 20.0
  s$stress_mutation_multiplier <- 5.0
  s$grass_rate                 <- 0.05
  s$max_ticks                  <- 500L
  s
}

results <- batch_alife(
  list(baseline = make_s(FALSE), hypermutation = make_s(TRUE)),
  n_cores = 2L
)

plot_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(t = d$t, genetic_diversity = d$genetic_diversity, condition = nm)
}, results, names(results), SIMPLIFY = FALSE))

ggplot(plot_df, aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  scale_colour_manual(values = c(baseline = "#878787",
                                 hypermutation = "#d6604d")) +
  labs(
    title    = "Stress hypermutation increases genetic diversity during resource crashes",
    subtitle = "grass_rate = 0.05; stress_mutation_multiplier = 5",
    x = "Tick", y = "Genetic diversity", colour = "Condition"
  ) +
  theme_minimal()
```

```{r fig-stress-hypermutation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: genetic diversity is transiently elevated in the stress-hypermutation condition during resource-crash epochs. Adaptive recovery is faster in the hypermutation condition as beneficial variants spread more rapidly through the population."}
FIG("stress_hypermutation")
```

---

## Part VI — Population genetics

### Population genetics and narrow-sense heritability

**What it models.** A central quantity in quantitative genetics is narrow-sense
heritability, h² — the proportion of phenotypic variance in a trait that is
attributable to additive genetic effects (Falconer & Mackay 1996; Lynch & Walsh
1998). A trait cannot respond to natural selection unless h² > 0. The
`estimate_heritability()` function performs parent-offspring regression on
pedigree records accumulated during the simulation: offspring trait values are
regressed on mid-parent values, and the slope estimates h². When
`body_size_evolution = TRUE`, body size is a polygenic trait encoded in the
neural genome and subject to both mutation and selection, making it a suitable
target for heritability analysis within a single run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enables heritable, evolving body size |
| `max_ticks` | — | Longer runs accumulate more parent-offspring pairs |

**Expected output.** h² for `body_size` should be substantially positive
(0.3–0.7) early in the run, reflecting the strong additive genetic architecture
of a freshly initialised population. As directional selection erodes genetic
variance, h² may decline in later ticks. The 95% confidence interval from the
regression should exclude zero. h² for `energy` is expected to be lower,
because energy is heavily influenced by the local environment.

```{r pop-genetics}
library(clade)

s <- default_specs()
s$body_size_evolution <- TRUE
s$max_ticks           <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

h2 <- estimate_heritability(data, trait = "body_size")

cat("Heritability h\u00b2:", round(h2$h2, 3), "\n")
cat("95% CI: [", round(h2$ci[1], 3), ",", round(h2$ci[2], 3), "]\n")

plot(h2)  # parent-offspring regression scatter with fitted line
```

```{r fig-pop-genetics, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: parent-offspring regression for body size with a positive slope (h² = 0.3-0.7). Each point is one offspring plotted against its mid-parent value. The shaded band is the 95% confidence interval on the regression line."}
FIG("population_genetics")
```

---

## Part VII — Comparative experiments

### Predation and neural evolution

**What it models.** Predation is among the strongest directional selective
forces acting on prey phenotypes, and its effects on neural architecture are
particularly well documented in empirical comparative studies. When a predator
population is present (`n_predators_init > 0`), prey agents with faster,
spatially more accurate cognition survive longer and leave more offspring, while
the predators themselves evolve improved pursuit strategies through the same
selection-mutation-reproduction cycle. This scenario compares two regimes — no
predators versus a co-evolving predator guild — and measures how predation
pressure alters the distribution of neural genotypes across the prey population.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0L | Number of predators at initialisation |
| `n_agents_init` | — | Initial prey population size |
| `max_ticks` | — | Simulation length |

**Expected output.** Predation is expected to increase `genetic_diversity` in
the prey population by imposing strong directional selection on cognition,
thereby maintaining or generating variance in neural genotypes. Mean energy may
decline under predation because prey pay metabolic costs of evasion and are
more frequently killed before reaching peak foraging efficiency.

```{r predation-neural}
library(clade)
library(ggplot2)
library(patchwork)

make_s <- function(n_pred) {
  s <- default_specs()
  s$n_agents_init    <- 80L
  s$n_predators_init <- n_pred
  s$max_ticks        <- 500L
  s
}

results <- batch_alife(
  list(no_predators = make_s(0L), predators = make_s(10L)),
  n_cores = 2L
)

plot_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(
    t                = d$t,
    genetic_diversity = d$genetic_diversity,
    mean_energy      = d$mean_energy,
    condition        = nm
  )
}, results, names(results), SIMPLIFY = FALSE))

p1 <- ggplot(plot_df, aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(title = "Predation and neural genome diversity",
       x = "Tick", y = "Genetic diversity", colour = NULL) +
  theme_minimal()

p2 <- ggplot(plot_df, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal()

p1 / p2
```

```{r fig-predation-neural, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output (top panel): genetic diversity is higher under predation, reflecting stronger directional selection on prey cognition. Bottom panel: mean energy is lower under predation as prey pay evasion costs and are removed before reaching peak foraging efficiency."}
FIG("vadim_experiment")
```

---

### Module comparison — 14-condition experiment

**What it models.** Biological systems are shaped by multiple interacting
selective pressures, but their individual contributions are difficult to
disentangle in nature. This scenario applies a one-module-at-a-time factorial
design: each condition activates exactly one module above the shared baseline,
and three summary statistics — mean genetic diversity, mean population size, and
mean energy — are recorded across the full run. Comparing modules in isolation
before combining them allows each module's unique contribution to be identified;
the all-on condition then reveals emergent interactions that cannot be predicted
from the single-module results alone.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0L | Adds co-evolving predator pressure |
| `disease` | FALSE | Activates SIR epidemic dynamics |
| `kin_selection` | FALSE | Enables pedigree-based altruistic transfers |
| `niche_construction` | FALSE | Agents build and inherit shelters |
| `rl_mode` | `"none"` | Set to `"actor_critic"` for within-lifetime RL |
| `social_learning` | FALSE | Copies output-layer weights from successful neighbours |
| `body_size_evolution` | FALSE | Heritable metabolic scaling |
| `dispersal_evolution` | FALSE | Heritable dispersal tendency |
| `cooperation_evolution` | FALSE | Heritable cooperation trait with public goods |
| `complex_landscape` | FALSE | Spatially heterogeneous, shifting resource patches |

**Expected output.** Modules imposing strong directional selection — predators
and disease — tend to increase genetic diversity by maintaining a selective
gradient; modules that smooth energy acquisition — niche construction, social
learning — may reduce within-population variance. The all-on condition typically
reveals non-additive interactions: cooperation and kin selection can partially
compensate for disease mortality, and niche construction can buffer the
energetic cost of predation evasion.

```{r module-comparison}
library(clade)
library(ggplot2)

make_specs <- function(...) {
  s <- default_specs()
  s$max_ticks <- 300L
  args <- list(...)
  for (nm in names(args)) s[[nm]] <- args[[nm]]
  s
}

specs_list <- list(
  baseline    = make_specs(),
  predators   = make_specs(n_predators_init = 5L),
  disease     = make_specs(disease = TRUE),
  kin         = make_specs(kin_selection = TRUE),
  niche       = make_specs(niche_construction = TRUE),
  rl          = make_specs(rl_mode = "actor_critic"),
  social      = make_specs(social_learning = TRUE),
  body_size   = make_specs(body_size_evolution = TRUE),
  dispersal   = make_specs(dispersal_evolution = TRUE),
  care        = make_specs(parental_care = TRUE),
  mimicry     = make_specs(mimicry = TRUE, n_predators_init = 5L),
  landscape   = make_specs(complex_landscape = TRUE),
  cooperation = make_specs(cooperation_evolution = TRUE),
  all_on      = make_specs(
    n_predators_init      = 5L,
    disease               = TRUE,
    kin_selection         = TRUE,
    niche_construction    = TRUE,
    rl_mode               = "actor_critic",
    social_learning       = TRUE,
    body_size_evolution   = TRUE,
    dispersal_evolution   = TRUE,
    cooperation_evolution = TRUE
  )
)

results <- batch_alife(specs_list, n_cores = 4L)

summary_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(
    condition      = nm,
    mean_diversity = mean(d$genetic_diversity, na.rm = TRUE),
    mean_pop       = mean(d$n_agents,          na.rm = TRUE),
    mean_energy    = mean(d$mean_energy,       na.rm = TRUE)
  )
}, results, names(results), SIMPLIFY = FALSE))

ggplot(summary_df,
       aes(x = reorder(condition, mean_diversity), y = mean_diversity)) +
  geom_col(fill = "#4dac26") +
  coord_flip() +
  labs(
    title = "Mean genetic diversity by module condition",
    x     = NULL,
    y     = "Mean genetic diversity"
  ) +
  theme_minimal()
```

```{r fig-module-comparison, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: horizontal bar chart of mean genetic diversity across all 14 conditions. Conditions imposing strong directional selection (predators, disease, mimicry) appear near the top; the all-on condition reveals emergent multi-module interactions."}
FIG("12_kitchen_dashboard")
```

---

### Combining modules — a kitchen-sink run

**What it models.** The preceding sections isolate individual mechanisms to
establish clear cause-and-effect relationships. Real ecosystems combine many
selective pressures simultaneously, and their interactions are often
non-additive. A multi-module run demonstrates that `clade`'s design is
genuinely compositional: any combination of modules can be enabled together
without special-casing, and the emergent dynamics are richer than the sum of
their parts.

**Key parameters.**

| Parameter | Setting | Module |
|---|---|---|
| `complex_landscape` | TRUE | Heterogeneous habitat structure |
| `n_predators_init` | 5 | Predation pressure |
| `disease` | TRUE | SIR infectious disease dynamics |
| `kin_selection` | TRUE | Kin-based energy transfer |
| `social_learning` | TRUE | Copy successful neighbours' policies |
| `rl_mode` | `"actor_critic"` | Within-lifetime policy gradient learning |
| `niche_construction` | TRUE | Agents build shelters that modify the landscape |
| `max_ticks` | 500 | Long enough for multiple waves of each process |

**Expected output.** No single output metric tells the full story. Look for
disease waves in `n_infected`, predator oscillations in `n_predators`, and
shelter accumulation in `n_shelters`. The interaction between modules is
emergent: shelters reduce predation, which alters the population structure on
which kin selection and disease act.

```{r kitchen-sink}
library(clade)
library(ggplot2)
library(tidyr)

s <- default_specs()
s$complex_landscape      <- TRUE
s$grid_rows              <- 40L
s$grid_cols              <- 40L
s$n_agents_init          <- 120L
s$n_predators_init       <- 5L
s$disease                <- TRUE
s$kin_selection          <- TRUE
s$social_learning        <- TRUE
s$rl_mode                <- "actor_critic"
s$niche_construction     <- TRUE
s$max_ticks              <- 500L

env  <- run_alife(s)
tks  <- get_run_data(env)$ticks

dat_long <- pivot_longer(
  tks[, c("t", "n_agents", "n_predators", "n_infected")],
  cols      = c("n_agents", "n_predators", "n_infected"),
  names_to  = "variable",
  values_to = "value"
)
dat_long$variable <- factor(
  dat_long$variable,
  levels = c("n_agents", "n_predators", "n_infected"),
  labels = c("Prey agents", "Predators", "Infected agents")
)

ggplot(dat_long, aes(x = t, y = value, colour = variable)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_colour_manual(
    values = c("Prey agents"     = "#1565C0",
               "Predators"       = "#B71C1C",
               "Infected agents" = "#F57F17"),
    name = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Count",
    title = "Kitchen-sink run: predation, disease, kin selection, RL, and niche construction"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-kitchen-sink, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: prey (blue), predator (red), and infected-agent (orange) time series show interacting waves. Disease suppresses prey density, which in turn reduces predator abundance, while niche construction and social learning modulate recovery rates."}
FIG("12_kitchen_sink")
```

---

## Part VIII — alifeR experiment reference guide

`clade` is the direct successor to the `alifeR` R package, which used an Rcpp
backend for simulation. All biological modules, parameter names, and
experimental designs are preserved: the sole change required to run any alifeR
experiment in `clade` is replacing `alifeR::run_alife(specs)` with
`clade::run_alife(specs)`. The Julia backend provides approximately 25× faster
throughput for populations above 200 agents, making parameter sweeps and
replicated batch experiments tractable.

**Key translated experiments.**

| alifeR script | clade equivalent | Key parameters |
|---|---|---|
| `grand_experiment.R` | Module comparison (above) | 14 conditions × `batch_alife()` |
| `kin_selection_experiment.R` | Kin selection section | `kin_selection`, `kin_altruism_r_min` |
| `niche_experiment.R` | Niche construction section | `niche_construction`, `shelter_build_prob` |
| `body_size_experiment.R` | Body size + predation | `body_size_evolution`, `n_predators_init` |
| `mimicry_experiment.R` | Mimicry and toxicity section | `mimicry`, `toxicity_init_mean`, `n_predators_init` |
| `social_learning_experiment.R` | Social learning section | `social_learning`, `social_learning_freq` |
| `rl_experiment.R` | Within-lifetime RL section | `rl_mode`, `learning_rate`, `rl_update_freq` |

**Migration checklist.** Replace the package namespace in all calls
(`alifeR::` → `clade::`). Confirm that `default_specs()` returns the same
parameter names as the alifeR version being migrated; any alifeR parameters
introduced after the clade fork will need manual addition via the
`default_specs()` extension pattern documented in `?default_specs`. No changes
to downstream analysis functions (`estimate_heritability()`, `get_run_data()`,
`plot_run()`, `compare_conditions()`) are required.

---

## Part IX — Pure-R scenario: evolution of bad science

**What it models.** A pure-R simulation (no Julia required) of Smaldino &
McElreath (2016). Labs compete for citations. Publication probability depends
on finding a positive result. Low statistical power (low `effort`) increases
false-positive rate (FPR). Under publication pressure, low-effort strategies
spread because they produce more positive results per unit cost. Replication
slows but does not stop the deterioration.

```{r bad-science}
# No Julia required — runs entirely in R
df_norep <- run_bad_science(n_ticks = 500L, replication_rate = 0.0, seed = 1L)
df_rep   <- run_bad_science(n_ticks = 500L, replication_rate = 0.5, seed = 1L)

library(ggplot2)
df <- rbind(
  cbind(df_norep, rate = "No replication"),
  cbind(df_rep,   rate = "50% replication")
)
ggplot(df, aes(t, mean_fpr, colour = rate)) +
  geom_line() +
  labs(title = "False-positive rate under publication pressure",
       x = "Tick", y = "Mean FPR", colour = NULL) +
  theme_minimal()
```

```{r fig-bad-science, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: false-positive rate rises under publication pressure without replication (red). Replication (blue) slows but does not prevent deterioration."}
FIG("bad_science")
```

---

## References

Baldwin, J.M. (1896) A new factor in evolution. *American Naturalist*
30(354): 441–451.

Boyce, M.S. (1979) Seasonality and patterns of natural selection for life
histories. *American Naturalist* 114(4): 569–583.

Cole, L.C. (1954) The population consequences of life history phenomena.
*Quarterly Review of Biology* 29(2): 103–137.

Coyne, J.A. & Orr, H.A. (2004) *Speciation*. Sinauer Associates, Sunderland,
MA.

DeVault, T.L., Rhodes, O.E. & Shivik, J.A. (2003) Scavenging by vertebrates:
behavioral, ecological, and evolutionary perspectives on an important energy
transfer pathway in terrestrial ecosystems. *Oikos* 102(2): 225–234.

DeWitt, T.J. & Scheiner, S.M. (2004) *Phenotypic Plasticity: Functional and
Conceptual Approaches*. Oxford University Press.

Dieckmann, U. & Doebeli, M. (1999) On the origin of species by sympatric
speciation. *Nature* 400(6742): 354–357.

Falconer, D.S. & Mackay, T.F.C. (1996) *Introduction to Quantitative
Genetics*, 4th edn. Longman, Harlow.

Fisher, R.A. (1930) *The Genetical Theory of Natural Selection*. Oxford
University Press.

Fretwell, S.D. & Lucas, H.L. (1970) On territorial behaviour and other
factors influencing habitat distribution in birds. *Acta Biotheoretica*
19(1): 16–36.

Fromhage, L. & Jennions, M.D. (2019) The evolution of mutualism from
reciprocal altruism by a selection regime shift. *Nature Communications*
10: 1294.

Hamilton, W.D. (1964) The genetical evolution of social behaviour I & II.
*Journal of Theoretical Biology* 7(1): 1–52.

Hamilton, W.D. (1971) Geometry for the selfish herd. *Journal of Theoretical
Biology* 31(2): 295–311.

Hauert, C., De Monte, S., Hofbauer, J. & Sigmund, K. (2002) Volunteering as
Red Queen mechanism for cooperation in public goods games. *Science*
296(5570): 1129–1132.

Henrich, J. & McElreath, R. (2003) The evolution of cultural evolution.
*Evolutionary Anthropology* 12(3): 123–135.

Hinton, G.E. & Nowlan, S.J. (1987) How learning can guide evolution. *Complex
Systems* 1(3): 495–502.

Houston, A.I. & Davies, N.B. (1985) The evolution of cooperation and life
history in the dunnock, *Prunella modularis*. In: Sibley, R.M. & Smith, R.H.
(eds.) *Behavioural Ecology*. Blackwell, Oxford, pp. 471–487.

Lack, D. (1947) The significance of clutch-size. *Ibis* 89(2): 302–352.

Liedtke, J. & Fromhage, L. (2019) The evolution of learning in the cephalopod
paradox. *Proceedings of the Royal Society B* 286: 20190918.

Lotka, A.J. (1925) *Elements of Physical Biology*. Williams & Wilkins,
Baltimore.

Lynch, M. & Walsh, B. (1998) *Genetics and Analysis of Quantitative Traits*.
Sinauer Associates, Sunderland, MA.

Mayley, G. (1996) Landscapes, learning costs and genetic assimilation.
*Evolutionary Computation* 4(3): 213–234.

Maynard Smith, J. (1978) *The Evolution of Sex*. Cambridge University Press.

McKenzie, G.J. & Rosenberg, S.M. (2001) Adaptive mutations, mutator DNA
polymerases and genetic change: strategies of survival for the long-term
starving cell. *Current Opinion in Microbiology* 4(5): 586–594.

Nowak, M.A. & May, R.M. (1992) Evolutionary games and spatial chaos. *Nature*
359(6398): 826–829.

Odling-Smee, F.J., Laland, K.N. & Feldman, M.W. (2003) *Niche Construction:
The Neglected Process in Evolution*. Princeton University Press.

Réale, D., Garant, D., Humphries, M.M., Bergeron, P., Careau, V. & Montiglio,
P.-O. (2010) Personality and the emergence of the pace-of-life syndrome
concept at the population level. *Philosophical Transactions of the Royal
Society B* 365(1560): 4051–4063.

Roff, D.A. (1992) *The Evolution of Life Histories*. Chapman & Hall, New York.

Ruxton, G.D., Sherratt, T.N. & Speed, M.P. (2004) *Avoiding Attack: The
Evolutionary Ecology of Crypsis, Warning Signals and Mimicry*. Oxford
University Press.

Shine, R., Brown, G.P. & Phillips, B.L. (2011) An evolutionary process that
assembles phenotypes through space rather than through time. *Proceedings of
the National Academy of Sciences* 108(14): 5708–5711.

Smaldino, P.E. & McElreath, R. (2016) The natural selection of bad science.
*Royal Society Open Science* 3: 160384.

Smith, C.C. & Fretwell, S.D. (1974) The optimal balance between size and
number of offspring. *American Naturalist* 108(962): 499–506.

Stearns, S.C. (1992) *The Evolution of Life Histories*. Oxford University
Press.

Trivers, R.L. (1972) Parental investment and sexual selection. In: Campbell, B.
(ed.) *Sexual Selection and the Descent of Man 1871–1971*. Aldine, Chicago,
pp. 136–179.

Volterra, V. (1928) Variations and fluctuations of the number of individuals
in animal species living together. *Journal of the Council International pour
l'Exploration de la Mer* 3(1): 3–51.

Williams, G.C. (1975) *Sex and Evolution*. Princeton University Press.

Zahavi, A. (1975) Mate selection — a selection for a handicap. *Journal of
Theoretical Biology* 53(1): 205–214.
