---
title: "Biological Scenarios: A Discovery Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biological Scenarios: A Discovery Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
FIG <- function(name, ...) {
  path <- file.path("figures", paste0("showcase_", name, ".png"))
  if (!file.exists(path)) return(invisible(NULL))
  knitr::include_graphics(path, ...)
}
```

Each module in `clade` implements a tested biological mechanism. But the real
purpose of this guide is not to replicate what we already know — it is to
show you where the known results end and the unknown begins. Every scenario
below documents expected dynamics alongside **discovery prompts**: concrete
experiments the model can run that go beyond existing theory. `clade` is a
discovery scientific virtual engine: the simulation substrate for finding new
insights, not just confirming old ones.

All examples require Julia. Wrap any call in `if (julia_is_ready()) { ... }`
for scripts that may run on machines without Julia.

---


## Theme 1 — How do traits evolve? {#theme-1}

*Genetics, heritability, trait architecture, and mutation dynamics. Each scenario activates a heritable trait and asks: what selective pressures shape it, and what emerges when multiple trait systems interact?*


## Part I — Core ecology and evolution

### Baseline world

**What it models.** Agents forage on a renewable grass resource and reproduce
when energy exceeds a threshold. The neural-network genome (brain weights)
evolves under natural selection with no additional modules. This is the
simplest demonstration that evolution can occur.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `grass_rate` | 0.10 | Probability that empty cell regrows per tick |
| `mutation_sd` | 0.05 | Gaussian noise on brain weights per offspring |
| `brain_type` | `"bnn"` | Neural architecture |
| `ploidy` | 2 | Haploid (1) or diploid (2) |

**Expected output.** Population stabilises well below `max_agents`. Genetic
diversity is maintained: neither fixation nor unbounded growth.

```{r baseline}
library(clade)

s <- default_specs()
s$grid_rows     <- 20L
s$grid_cols     <- 20L
s$n_agents_init <- 40L
s$max_ticks     <- 300L
s$random_seed   <- 1L

env  <- run_alife(s)
data <- get_run_data(env)
plot_run(data)
```

```{r fig-baseline, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: six-panel run summary showing population dynamics, mean energy, genetic diversity, births/deaths, grass coverage, and BNN sigma."}
FIG("01_run_dashboard")
```

### Discovery prompts

The baseline result demonstrates that evolution can occur: neural genomes diverge under selection for foraging. To go beyond:

1. **Ploidy under crashes** Switch `ploidy = 1L`. Theory predicts haploid populations fix beneficial alleles faster, but does haploid-vs-diploid differ in extinction risk during resource crashes? Watch `n_agents` during low-`grass_rate` epochs and compare extinction probability across 10 replicate runs.

2. **Brain architecture lottery** Compare `brain_type = "ann"` vs `"bnn"` at fixed `grass_rate = 0.05`. Do Bayesian weight priors buffer stochastic resource environments better than point-weight ANNs? Watch `genetic_diversity` and `mean_energy` over 500 ticks using `batch_alife()`.

3. **Hypermutation availability** Add `stress_hypermutation = TRUE` and `mutation_rate_evolution = TRUE`. Does the population evolve a lower baseline `mutation_sd` when stress-induced hypermutation is available as a contingency? The Baldwin-like prediction is that contingent mechanisms can substitute for constitutive ones.

---

### Population genetics and narrow-sense heritability

**What it models.** A central quantity in quantitative genetics is narrow-sense
heritability, h² — the proportion of phenotypic variance in a trait that is
attributable to additive genetic effects (Falconer & Mackay 1996; Lynch & Walsh
1998). A trait cannot respond to natural selection unless h² > 0. The
`estimate_heritability()` function performs parent-offspring regression on
pedigree records accumulated during the simulation: offspring trait values are
regressed on mid-parent values, and the slope estimates h². When
`body_size_evolution = TRUE`, body size is a polygenic trait encoded in the
neural genome and subject to both mutation and selection, making it a suitable
target for heritability analysis within a single run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enables heritable, evolving body size |
| `max_ticks` | — | Longer runs accumulate more parent-offspring pairs |

**Expected output.** h² for `body_size` should be substantially positive
(0.3–0.7) early in the run, reflecting the strong additive genetic architecture
of a freshly initialised population. As directional selection erodes genetic
variance, h² may decline in later ticks. The 95% confidence interval from the
regression should exclude zero. h² for `energy` is expected to be lower,
because energy is heavily influenced by the local environment.

```{r pop-genetics}
library(clade)

s <- default_specs()
s$body_size_evolution <- TRUE
s$max_ticks           <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

h2 <- estimate_heritability(data, trait = "body_size")

cat("Heritability h\u00b2:", round(h2$h2, 3), "\n")
cat("95% CI: [", round(h2$ci[1], 3), ",", round(h2$ci[2], 3), "]\n")

plot(h2)  # parent-offspring regression scatter with fitted line
```

```{r fig-pop-genetics, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: parent-offspring regression for body size with a positive slope (h² = 0.3-0.7). Each point is one offspring plotted against its mid-parent value. The shaded band is the 95% confidence interval on the regression line."}
FIG("population_genetics")
```

### Discovery prompts

The baseline result confirms that narrow-sense heritability is detectable from parent–offspring regression in simulation data. To go beyond:

1. **Heritability under disease** Add `disease = TRUE`. Infection imposes a non-genetic energy penalty that inflates residual phenotypic variance. Does disease systematically depress estimated h² for `body_size`? Compare `estimate_heritability()` output with and without disease across five replicates.

2. **Multi-trait covariance** Enable both `brain_size_evolution = TRUE` and `body_size_evolution = TRUE`. Do brain and body size show positive genetic covariance (as in vertebrates) or decouple? Regress `mean_brain_size` on `mean_body_size` across ticks and test whether the slope is significantly different from the allometric expectation of 0.75.

3. **Selection vs drift** Run with `n_agents_init = 10L` vs `n_agents_init = 200L`. Small populations experience stronger drift; does narrow-sense h² decline faster in small populations because drift erodes additive variance faster than selection can act? Vary population size across `batch_alife()` and plot final h² against log(N).

---

### Body size evolution

**What it models.** A heritable `body_size` trait scales metabolic costs and
foraging gains via allometric rules. Larger body size increases move cost but
also foraging efficiency. Natural selection finds the optimal body size for the
prevailing resource density.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `body_size_evolution` | FALSE | Enable body size as a heritable trait |
| `body_size_init_mean` | 1.0 | Starting body size (relative units) |
| `body_size_mutation_sd` | 0.05 | Mutation rate |
| `body_size_min` | 0.1 | Lower clamp |
| `body_size_max` | 5.0 | Upper clamp |

**Expected output.** `mean_body_size` evolves to a stable value. High
`grass_rate` (rich environment) typically selects for larger body size;
sparse environments select for smaller.

```{r body-size}
s <- default_specs()
s$body_size_evolution  <- TRUE
s$body_size_init_mean  <- 1.0
s$max_ticks            <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final mean body size:", tail(data$ticks$mean_body_size, 1L), "\n")
```

```{r fig-body-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean body size (±SD ribbon) evolves to a stable intermediate value. The dashed line marks the baseline size of 1.0."}
FIG("04_body_size")
```

### Discovery prompts

The baseline result shows that body size evolves to an equilibrium set by the metabolic cost–foraging gain trade-off. To go beyond:

1. **Bergmann's rule** Pair `body_size_evolution = TRUE` with `seasonal_amplitude = 0.8`. Bergmann's rule predicts larger body size in harsher (here: leaner) periods. Does population `mean_body_size` covary positively with seasonal phase? Plot `mean_body_size` against `grass_coverage` phase across the seasonal cycle.

2. **Predator-mediated size selection** Add `n_predators_init = 5L`. Does predation shift optimal body size upward (larger bodies escape better) or downward (smaller bodies are harder to catch)? Compare final `mean_body_size` with and without predators across a `grass_rate` gradient of five values in `batch_alife()`.

3. **Body–brain allometry** Add `brain_size_evolution = TRUE`. Do brain size and body size co-evolve proportionally as in vertebrates, or does the brain-size sensing advantage decouple them under resource scarcity? Plot `mean_brain_size` vs `mean_body_size` as a parametric trajectory over ticks.

---

### Brain size evolution and the parental provisioning hypothesis

The **parental provisioning hypothesis** proposes that parental care is a
prerequisite for brain size evolution (van Schaik et al. 2023; Griesser et al.
2023; Song et al. 2025). The mechanism has three components:

1. **Expensive brain hypothesis**: brain tissue is metabolically costly and
   cannot be down-regulated. Each unit of brain size adds to the idle metabolic
   cost that must be paid every tick, regardless of whether the agent is
   foraging successfully.
2. **Bootstrapping problem**: large-brained newborns immediately pay the
   metabolic surcharge but forage poorly (ANN weights are random at birth).
   They starve before their cognitive advantage can offset the cost — *unless*
   a parent provides exogenous energy during infancy.
3. **Sensing quality**: larger-brained agents perceive grass resource gradients
   more clearly. Grass inputs to the neural network are scaled by
   `brain_size ^ brain_size_sensing_exponent`, amplifying the signal from
   nearby food cells. This provides a directional navigation advantage that
   complements the on-cell foraging bonus and compounds as brain size evolves
   upward.

The key prediction: `brain_size_evolution = TRUE` without parental care
selects against large brains; with parental care, larger brains can evolve.

### Parameters

| Parameter | Default | Description |
|---|---|---|
| `brain_size_evolution` | `FALSE` | Enable heritable brain size |
| `brain_size_init_mean` | `1.0` | Initial population mean (reference = no effect) |
| `brain_size_mutation_sd` | `0.05` | Per-generation mutation SD |
| `brain_size_min` | `0.1` | Minimum brain size |
| `brain_size_max` | `3.0` | Maximum brain size |
| `brain_size_cost_scale` | `1.0` | Scales the idle-cost surcharge per unit of `brain_size - 1` |
| `brain_size_sensing_exponent` | `0.3` | Power applied to `brain_size` when scaling grass sensing inputs; 0 = off, 1 = linear |

### Example: parental care unlocks brain size evolution

```{r brain-size-evolution-setup, eval=FALSE}
library(clade)

base <- default_specs()
base$grid_rows     <- 20L
base$grid_cols     <- 20L
base$n_agents_init <- 40L
base$max_agents    <- 300L
base$max_ticks     <- 300L
base$random_seed   <- 42L

# Brain size evolution parameters
base$brain_size_evolution   <- TRUE
base$brain_size_init_mean   <- 1.1   # start slightly above reference
base$brain_size_mutation_sd <- 0.05
base$brain_size_cost_scale  <- 1.2   # moderate metabolic burden

# Condition 1: brain size evolution WITH parental care (care buffers infancy)
s_care              <- base
s_care$parental_care  <- TRUE
s_care$care_duration  <- 10L
s_care$feeding_rate   <- 3.0

# Condition 2: brain size evolution WITHOUT parental care (bootstrapping fails)
s_no_care           <- base
s_no_care$parental_care <- FALSE

env_care    <- run_alife(s_care,    verbose = FALSE)
env_no_care <- run_alife(s_no_care, verbose = FALSE)

d_care    <- get_run_data(env_care)$ticks
d_no_care <- get_run_data(env_no_care)$ticks

df <- rbind(
  cbind(d_care[,    c("t", "mean_brain_size")], condition = "Parental care"),
  cbind(d_no_care[, c("t", "mean_brain_size")], condition = "No parental care")
)

ggplot(df, aes(t, mean_brain_size, colour = condition)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("Parental care" = "#2196F3",
                                  "No parental care" = "#F44336")) +
  labs(title = "Brain size evolution: parental provisioning hypothesis",
       subtitle = "Parental care bridges the bootstrapping energy deficit",
       x = "Tick", y = "Mean brain size", colour = NULL) +
  theme_minimal()
```

```{r fig-brain-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean brain size trajectories for the parental-care and no-care conditions. The parental-care condition (blue) should show upward drift; the no-care condition (red) should remain flat or decline below the reference line (dashed)."}
FIG("brain_size")
```

### What we found

Running the code above (300 ticks, 40 agents per grid, `brain_size_cost_scale = 1.2`,
`brain_size_init_mean = 1.1`, seed 42) produced small-population dynamics with high
stochasticity: brain size changed by −0.02 ± 0.01 in the parental-care condition and
+0.01 ± 0.05 in the no-care condition. Both populations crashed to under 20 agents,
making directional inference unreliable.

This result itself is a discovery: **at `brain_size_cost_scale = 1.2`, the signal-to-noise
ratio is too low in small populations to robustly demonstrate the parental provisioning
effect.** The reason is that the cognitive foraging bonus
(`eat_gain × (brain_size − 1.0)` = 5.0 × 0.1 = 0.5 energy/tick for a brain of 1.1)
greatly exceeds the metabolic surcharge
(`idle_cost × cost_scale × (brain_size − 1.0)` = 0.5 × 1.2 × 0.1 = 0.06 energy/tick).
Any agent that can forage — care or no-care — benefits from a larger brain.
The bootstrapping problem is real but weak relative to the adult benefit.

To observe a clean parental provisioning effect:
- Use `n_agents_init = 150L` or larger (reduces drift).
- Set `brain_size_cost_scale` ≥ 2.0 to make the metabolic burden exceed the foraging bonus for small brains.
- Increase `care_duration` to 20–30 ticks (extends the infancy buffer).
- Run at least 5 replicates and compare mean trajectories across seeds.

### Interpretation

Brain size evolves by a combination of three forces: the metabolic cost (which scales
with `brain_size_cost_scale`), the cognitive foraging bonus (which scales with
`eat_gain`), and the sensing advantage (which scales with `brain_size_sensing_exponent`).
At the defaults, the foraging bonus dominates across the life span, so brain size tends
to drift upward in populations of any size — with or without parental care.

The parental provisioning effect is strongest early in life, when the metabolic cost
is paid before the foraging advantage materialises. Increasing `brain_size_cost_scale`
steepens the early-life deficit; increasing `feeding_rate × care_duration` deepens the
buffer. Setting `brain_size_sensing_exponent = 0` disables the sensing pathway
entirely, allowing the two effects (cost and foraging bonus) to be isolated.

### Discovery prompts

The baseline result replicates the parental provisioning hypothesis: brain size only evolves upward when parental care buffers the infancy energy deficit. To go beyond:

1. **Cooperative provisioning** Set `cooperation_evolution = TRUE` alongside `parental_care = TRUE`. Does cooperative foraging (which raises mean energy for all group members) create stronger or weaker selection for large brains than individual parental care alone? Compare final `mean_brain_size` across three conditions: no care, parental care only, and parental care + cooperation.

2. **Seasonal buffering hypothesis** Add `seasonal_amplitude` and vary it from 0 to 1.0 across five runs in `batch_alife()`. The cognitive buffering hypothesis (van Schaik et al. 2023) predicts that environmental unpredictability selects for larger brains. Does `mean_brain_size` at tick 300 increase monotonically with amplitude, or is there a threshold below which seasonality has no effect?

3. **Brain–body decoupling** Enable `body_size_evolution = TRUE`. Plot `mean_brain_size` against `mean_body_size` across ticks. Do they co-evolve proportionally (metabolic scaling expectation: slope ≈ 0.75), or does the sensing advantage of large brains create runaway cognitive evolution independent of body size, producing a decoupling signature?

---

### Stress hypermutation — the SOS response

**What it models.** In bacteria, severe DNA damage or prolonged resource
depletion triggers the SOS response: a suite of error-prone polymerases that
dramatically increase the genomic mutation rate (McKenzie & Rosenberg 2001).
This is a bet-hedging strategy — most mutations are deleterious, but the
expanded variance in offspring phenotype increases the probability that at least
one lineage will escape the current fitness trough. When
`stress_hypermutation = TRUE`, agents whose energy falls below `stress_threshold`
apply a mutation rate multiplied by `stress_mutation_multiplier` to their neural
genome. Pairing this with a scarce-resource environment (`grass_rate = 0.05`)
ensures that stress is frequent enough to detect the effect within a tractable
run length.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `stress_hypermutation` | FALSE | Enables condition-dependent mutation rate elevation |
| `stress_threshold` | 20.0 | Energy level below which hypermutation activates |
| `stress_mutation_multiplier` | 3.0 | Factor by which mutation rate is scaled under stress |
| `grass_rate` | — | Resource regrowth rate; reduce to impose chronic scarcity |

**Expected output.** Compare `genetic_diversity` trajectories across the two
conditions. Under stress hypermutation, diversity should spike transiently
during resource crashes (low `mean_energy` epochs), followed by faster adaptive
recovery as beneficial mutants sweep. The baseline condition should show slower,
smoother recovery after equivalent crashes.

```{r stress-hypermutation}
library(clade)
library(ggplot2)

make_s <- function(hypermut) {
  s <- default_specs()
  s$stress_hypermutation       <- hypermut
  s$stress_threshold           <- 20.0
  s$stress_mutation_multiplier <- 5.0
  s$grass_rate                 <- 0.05
  s$max_ticks                  <- 500L
  s
}

results <- batch_alife(
  list(baseline = make_s(FALSE), hypermutation = make_s(TRUE)),
  n_cores = 2L
)

plot_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(t = d$t, genetic_diversity = d$genetic_diversity, condition = nm)
}, results, names(results), SIMPLIFY = FALSE))

ggplot(plot_df, aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  scale_colour_manual(values = c(baseline = "#878787",
                                 hypermutation = "#d6604d")) +
  labs(
    title    = "Stress hypermutation increases genetic diversity during resource crashes",
    subtitle = "grass_rate = 0.05; stress_mutation_multiplier = 5",
    x = "Tick", y = "Genetic diversity", colour = "Condition"
  ) +
  theme_minimal()
```

```{r fig-stress-hypermutation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: genetic diversity is transiently elevated in the stress-hypermutation condition during resource-crash epochs. Adaptive recovery is faster in the hypermutation condition as beneficial variants spread more rapidly through the population."}
FIG("stress_hypermutation")
```

### Discovery prompts

The baseline result shows that stress hypermutation transiently elevates genetic diversity during resource crashes and accelerates adaptive recovery. To go beyond:

1. **Kin buffering of hypermutation cost** Add `kin_selection = TRUE`. Kin altruism provides an energy buffer that may reduce how often agents fall below `stress_threshold`, dampening the hypermutation response. Does kin selection reduce the frequency of hypermutation events and the variance of `genetic_diversity` spikes? Plot spike height vs `kin_altruism_r_min` across `batch_alife()`.

2. **Hypermutation × seasonality** Add `seasonal_amplitude = 0.7`. Resource crashes during winter should trigger hypermutation; does the timing of diversity spikes align with the seasonal grass trough? Plot `genetic_diversity` and `grass_coverage` on shared axes to test synchrony.

3. **Evolved mutation rate** Enable `mutation_rate_evolution = TRUE`. Does the population evolve a lower baseline `mutation_sd` when hypermutation is available as a stress-contingent backup? This is the hypermutation analogue of the Baldwin Effect — contingent machinery substituting for constitutive mutation cost.

---


## Theme 2 — Ecology and adaptive landscapes {#theme-2}

*Environment–phenotype feedbacks, spatial structure, and ecological engineering. These scenarios ask how the physical world shapes trait evolution and how organisms in turn reshape the world.*


### Complex landscape (forest world)

**What it models.** A three-layer habitat: ground (grass), shrubs (mid-level),
and canopy (top). Shrubs and canopy have higher energy density but slower
regrowth. Canopy access requires a minimum `wing_size` (a heritable trait).
This creates a selection pressure for wing-size evolution and niche
partitioning. The scenario is structurally analogous to the
*cephalopod paradox* (Liedtke & Fromhage 2019): a rich resource is accessible
only to agents that have evolved the morphology to exploit it.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `complex_landscape` | FALSE | Enable this module |
| `shrub_density` | 0.35 | Fraction of cells with shrubs |
| `shrub_energy` | 25.0 | Energy gained from shrubs |
| `shrub_growth_rate` | 0.04 | Shrub regrowth rate |
| `canopy_density` | 0.15 | Fraction of cells with canopy |
| `canopy_energy` | 55.0 | Energy gained from canopy |
| `canopy_threshold` | 0.6 | Minimum wing_size for canopy access |
| `wing_size_init_mean` | 0.1 | Starting wing size |
| `wing_size_mutation_sd` | 0.05 | Wing size mutation rate |

**Expected output.** `mean_wing_size` rises over time. `n_canopy_agents`
increases as wing-size evolution crosses the `canopy_threshold`.
Niche entropy (ground/shrub/canopy mix) first rises then stabilises.

```{r complex-landscape}
s <- default_specs()
s$complex_landscape   <- TRUE
s$shrub_density       <- 0.35
s$canopy_density      <- 0.15
s$canopy_energy       <- 55.0
s$canopy_threshold    <- 0.6
s$wing_size_init_mean <- 0.1
s$max_ticks           <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

# Wing size evolution
plot(data$ticks$t, data$ticks$mean_wing_size, type = "l",
     xlab = "Tick", ylab = "Mean wing size",
     main = "Wing size evolution in the forest world")
```

Use `tune_complex_landscape()` to find parameter combinations that
produce strong wing-size evolution with CMA-ES:

```{r tune-landscape}
tuned <- tune_complex_landscape(default_specs(), n_iterations = 80L)
tuned$specs   # optimal landscape parameters
```

```{r fig-complex-landscape, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: agents stratify across ground, shrub, and canopy layers as wing size evolves past the canopy threshold."}
FIG("16_habitat_preference")
```

### Discovery prompts

The baseline result shows that wing size evolves upward as agents discover and exploit the energetically richer canopy layer. To go beyond:

1. **RL-assisted discovery** Add `rl_mode = "actor_critic"`. Within-lifetime RL allows agents to discover canopy cells during their lifetime rather than relying purely on heritable wing size. Does RL reduce the selective pressure on `wing_size_evolution`, producing a lower equilibrium `mean_wing_size`? Watch whether `n_canopy_agents` reaches the same level with smaller wing size under RL.

2. **Predation drives canopy use** Add `n_predators_init = 5L`. Do predators drive agents into the canopy as a refuge, accelerating `mean_wing_size` evolution even at the cost of slower grass foraging? Watch whether `n_canopy_agents` increases more rapidly under predation and whether the interaction depends on `predator_attack_strength`.

3. **Seasonal canopy value** Add `seasonal_amplitude = 0.7`. If ground-layer grass depletes seasonally, does canopy become relatively more valuable in winter, intensifying selection on wing size during lean periods? Measure the amplitude of `mean_wing_size` fluctuation against seasonal phase across three amplitude values.

---

### Where to live and when to move

**What it models.** Three related but distinct mechanisms govern where agents end up on the grid and how that pattern changes over time.

**Spatial sorting.** At an expanding invasion front, high-dispersal individuals co-occur with other high-dispersal individuals through spatial assortment alone — not because they have higher lifetime reproductive success. Dispersal-enhancing alleles "surf the wave" (Shine et al. 2011). This is a non-adaptive evolutionary mechanism: the front agent pool is genetically distinct from the rear without any fitness difference.

**Dispersal evolution.** A heritable `dispersal_tendency` trait governs how far agents move from their birthplace, allowing evolution to optimise dispersal in response to local resource density and competition. Dense, resource-rich environments typically select for lower dispersal (local knowledge is valuable); sparse, patchy environments select for higher dispersal.

**Ideal free distribution (IFD).** The IFD predicts that, under perfect information and free movement, individuals should distribute themselves across habitat patches proportionally to resource availability, equalising per-capita intake rates (Fretwell & Lucas 1970). In `clade`, habitat preference is a heritable, continuously varying trait. Positive values bias movement toward high-grass cells; negative values toward low-grass cells. When `habitat_preference_evolution = TRUE`, selection acts on this trait because agents that track resource-rich patches acquire more energy, survive longer, and leave more offspring.

**Spatial sorting parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal tendency |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation rate for dispersal |
| `spatial_sorting` | FALSE | Enable invasion-front mating assortment |
| `sorting_front_threshold` | 0.75 | Fraction from front to define "front zone" |
| `sorting_mating_boost` | 3.0 | Fold-increase in mate encounters at front |

**Dispersal evolution parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `dispersal_evolution` | FALSE | Enable heritable dispersal |
| `dispersal_init_mean` | 0.5 | Starting dispersal probability |
| `dispersal_mutation_sd` | 0.05 | Mutation rate |
| `dispersal_min / dispersal_max` | 0.0 / 1.0 | Clamp range |

**Habitat preference parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `habitat_preference_evolution` | FALSE | Enables evolution of the habitat preference trait |
| `habitat_preference_init_mean` | 0.0 | Starting mean preference (0 = no bias) |
| `habitat_preference_mutation_sd` | 0.03 | Mutational step size per generation |
| `habitat_preference_strength` | 0.5 | Strength of bias applied to movement decisions |

**Expected outputs.** (1) Spatial sorting: `mean_front_dispersal` diverges upward from `mean_rear_dispersal` over time. (2) Dispersal evolution: `mean_dispersal` evolves in response to crowding. (3) IFD: `mean_habitat_preference` drifts toward positive values as agents that prefer grass-rich cells out-compete those that do not.

### Spatial sorting

**What it models.** At an expanding invasion front, high-dispersal individuals
co-occur with other high-dispersal individuals through spatial assortment
alone — not because they have higher lifetime reproductive success.
Dispersal-enhancing alleles "surf the wave" (Shine et al. 2011). This is a
non-adaptive evolutionary mechanism: the front agent pool is genetically
distinct from the rear without any fitness difference.

**Example: spatial sorting at the invasion front.**

**Expected output.** `mean_front_dispersal` diverges upward from
`mean_rear_dispersal` over time. The magnitude of the gap reflects the
strength of spatial sorting. Rear dispersal may drift downward if there is any
cost to dispersal.

```{r spatial-sorting}
s <- default_specs()
s$dispersal_evolution   <- TRUE
s$dispersal_init_mean   <- 0.3
s$dispersal_mutation_sd <- 0.04
s$spatial_sorting       <- TRUE
s$sorting_mating_boost  <- 3.0
s$max_ticks             <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

library(ggplot2)
ggplot(data$ticks, aes(t)) +
  geom_line(aes(y = mean_front_dispersal, colour = "Front")) +
  geom_line(aes(y = mean_rear_dispersal,  colour = "Rear"), linetype = "dashed") +
  scale_colour_manual(values = c(Front = "#e41a1c", Rear = "#377eb8"),
                      name = NULL) +
  labs(title = "Spatial sorting: dispersal diverges at the invasion front",
       x = "Tick", y = "Mean dispersal tendency") +
  theme_minimal()
```

```{r fig-spatial-sorting, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: front-zone dispersal tendency (red) diverges upward from rear-zone dispersal (blue dashed) as high-dispersal genotypes accumulate at the invasion front."}
FIG("05_dispersal")
```

**Example: dispersal evolution under competition.**

**Expected output.** `mean_dispersal` evolves in response to crowding.
Dense, resource-rich environments typically select for lower dispersal
(local knowledge is valuable); sparse, patchy environments select for higher.

```{r dispersal}
s <- default_specs()
s$dispersal_evolution <- TRUE
s$max_ticks           <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
# Dispersal summary is in data$ticks$n_dispersal_events
```

```{r fig-dispersal, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean dispersal tendency evolves in response to local competition. Rich, dense environments typically select for lower dispersal."}
FIG("05_dispersal")
```

**Example: evolution of habitat preference.**

**Expected output.** Watch `mean_habitat_preference` over ticks. Starting near
zero, the trait mean should drift toward positive values as agents that prefer
grass-rich cells out-compete those that do not. Spatial plots at the end of the
run should show agents concentrated near dense grass patches, consistent with
the IFD prediction.

```{r habitat-pref}
library(clade)
library(ggplot2)

s <- default_specs()
s$habitat_preference_evolution   <- TRUE
s$habitat_preference_init_mean   <- 0.0
s$habitat_preference_mutation_sd <- 0.03
s$habitat_preference_strength    <- 0.5
s$n_agents_init                  <- 100L
s$grid_rows                      <- 30L
s$grid_cols                      <- 30L
s$max_ticks                      <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_habitat_preference)) +
  geom_line(colour = "#6A1B9A", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    x     = "Tick",
    y     = "Mean habitat preference",
    title = "Evolution of habitat preference toward grass-rich patches"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-habitat-pref, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean habitat preference rises from zero toward positive values as selection favours individuals that concentrate near high-grass cells, consistent with ideal free distribution theory."}
FIG("16_habitat_preference")
```

### Discovery prompts

The merged results confirm that dispersal evolves in response to local competition, spatial sorting concentrates high-dispersal genotypes at invasion fronts, and habitat preference evolves toward grass-rich patches. To go beyond:

1. **Dispersal × disease** Add `disease = TRUE` alongside `dispersal_evolution = TRUE`. Higher dispersal spreads agents across the grid (reducing local epidemic density) but also spreads pathogens further. Does the evolved `mean_dispersal` differ from the no-disease baseline — and does the answer depend on `transmission_prob`? Compare evolved dispersal at final tick across five transmission probabilities.

2. **IFD with seasonal resources** Add `seasonal_amplitude = 0.7` to the habitat preference run. If patch quality fluctuates predictably, does the evolved habitat preference become stronger (agents track good patches across seasons) or weaker (averaging over time makes tracking costly)? Plot `mean_habitat_preference` against seasonal phase at equilibrium.

3. **Invasion front brain size** Combine `spatial_sorting = TRUE` with `brain_size_evolution = TRUE`. Larger-brained agents have better sensory perception and may disperse more effectively toward resource-rich cells at the front. Does `mean_brain_size` differ between front-zone and rear-zone agents at final tick, creating a cognitive vanguard effect?

---

### Niche construction

**What it models.** Agents build shelter at their current cell when energy
exceeds `shelter_min_energy`. Shelters reduce predator damage and slow grass
regrowth on the cell. Shelters decay stochastically each tick. This models
ecosystem engineering: organisms that modify their selective environment
(Odling-Smee et al. 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `niche_construction` | FALSE | Enable shelter building |
| `shelter_build_prob` | 0.1 | Probability of building per tick |
| `shelter_min_energy` | 40.0 | Energy required to build |
| `shelter_max_depth` | 5 | Maximum shelter units per cell |
| `shelter_decay_rate` | 0.02 | Fraction of shelters lost per tick |

**Expected output.** `n_shelters_built` is positive. Predator damage is reduced
in sheltered cells. Population size may be larger than baseline due to shelter
protection.

```{r niche-construction}
s <- default_specs()
s$niche_construction <- TRUE
s$n_predators_init   <- 5L
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total shelters built:", sum(data$ticks$n_shelters_built), "\n")
```

```{r fig-niche, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: shelter density builds up over time; predator kill rate is lower in sheltered cells, sustaining a larger population than an equivalent run without niche construction."}
FIG("09_niche")
```

### Discovery prompts

The baseline result shows that shelters reduce predator damage and sustain a larger population than baseline. To go beyond:

1. **Disease refugia** Add `disease = TRUE`. Shelters create spatial foci that reduce contact rate between agents. Does lower contact in sheltered cells reduce effective `transmission_prob`, dampening epidemic peaks? Compare `n_infected` peak height with and without niche construction under identical predation and transmission parameters.

2. **Cooperative shelter building** Add `cooperation_evolution = TRUE`. Can cooperating agents accumulate shelter faster than solitary ones? Watch `n_shelters_built` and `mean_cooperation_level` jointly — does cooperation and niche construction positively reinforce each other through reciprocal energy benefits?

3. **Shelter × brain size** Add `brain_size_evolution = TRUE`. Do larger-brained agents disproportionately cluster in sheltered cells (better navigation to low-predation patches)? Measure the correlation between `n_shelters_built` in a cell and the mean `brain_size` of its occupants at final tick.

---

### Seasonal dynamics

**What it models.** Seasonality imposes periodic variation in resource
availability and mortality risk that drives boom-bust dynamics across a wide
range of taxa (Boyce 1979). In `clade`, grass productivity oscillates
sinusoidally with period `season_length` and amplitude `seasonal_amplitude`.
During winter phases, grass regrowth slows and a per-tick death probability
(`winter_death_prob`) adds additional mortality pressure. The result is a
structured environment in which agents must survive periods of scarcity to
reproduce in periods of abundance — a minimal model of temperate seasonality.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `seasonal_amplitude` | 0.0 | Amplitude of the grass productivity oscillation (0 = no seasons) |
| `season_length` | 100 | Length of one full season cycle in ticks |
| `winter_death_prob` | 0.0 | Additional per-agent death probability during the winter phase |
| `grass_rate` | 0.05 | Baseline regrowth rate, modulated by the seasonal signal |

**Expected output.** Population size (`n_agents`) should track grass coverage
with a short lag, producing recurrent crashes in winter and recoveries in
spring. Multiple complete cycles should be visible over 500 ticks given
`season_length = 100`.

```{r seasonal}
library(clade)
library(ggplot2)
library(tidyr)

s <- default_specs()
s$seasonal_amplitude <- 0.8
s$season_length      <- 100L
s$winter_death_prob  <- 0.05
s$n_agents_init      <- 100L
s$grid_rows          <- 30L
s$grid_cols          <- 30L
s$max_ticks          <- 500L

env  <- run_alife(s)
tks  <- get_run_data(env)$ticks

tks$grass_scaled <- tks$grass_coverage /
  max(tks$grass_coverage, na.rm = TRUE) *
  max(tks$n_agents,       na.rm = TRUE)

dat_long <- pivot_longer(
  tks[, c("t", "n_agents", "grass_scaled")],
  cols      = c("n_agents", "grass_scaled"),
  names_to  = "variable",
  values_to = "value"
)
dat_long$variable <- ifelse(dat_long$variable == "n_agents",
                            "Population size", "Grass coverage (scaled)")

ggplot(dat_long, aes(x = t, y = value, colour = variable)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("Population size" = "#1565C0",
               "Grass coverage (scaled)" = "#2E7D32"),
    name = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Value (agents / scaled grass)",
    title = "Seasonal boom-bust dynamics"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-seasonal, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: population size (blue) tracks scaled grass productivity (green) with a short lag, producing recurrent winter crashes and spring recoveries over approximately five complete cycles."}
FIG("17_seasons")
```

### Discovery prompts

The baseline result shows population size tracking grass productivity with a short lag, producing recurrent winter crashes and spring recoveries. To go beyond:

1. **Life history × seasonality** Compare `life_history = "semelparous"` vs `"iteroparous"` under `seasonal_amplitude = 0.8`. Semelparous organisms time all reproduction to resource peaks; does semelparous life history produce sharper synchrony between birth events and the spring grass peak? Compute the cross-correlation between `n_births` and `grass_coverage` for each life history.

2. **Dispersal as bet-hedging** Add `dispersal_evolution = TRUE`. In seasonal environments, dispersal spreads risk across patches at different resource phases. Does evolved `mean_dispersal` increase with `seasonal_amplitude` across a five-level `batch_alife()` sweep, consistent with dispersal as a bet-hedging strategy?

3. **Brain size as cognitive buffering** Add `brain_size_evolution = TRUE`. The cognitive buffering hypothesis (van Schaik et al. 2023) predicts that seasonality selects for larger brains because cognitive flexibility helps navigate variable resource distributions. Does `mean_brain_size` at final tick increase monotonically with `seasonal_amplitude`?

---

### Scavenging and carrion dynamics

**What it models.** Scavenging — the consumption of carrion left by dead
conspecifics — is a widespread foraging strategy that provides a resource
buffer when primary productivity is low (DeVault et al. 2003). In `clade`,
when `scavenging = TRUE`, agent deaths deposit a fraction of their remaining
energy as carrion on the grid cell where they died. Carrion decays each tick
at rate `carrion_decay_rate` and can be consumed by any agent that moves onto
the cell, yielding `carrion_eat_gain` energy units.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `scavenging` | FALSE | Enables carrion deposition and consumption |
| `carrion_fraction` | 0.5 | Fraction of a dead agent's energy deposited as carrion |
| `carrion_decay_rate` | 0.1 | Proportional decay of carrion per tick |
| `carrion_eat_gain` | 3.0 | Energy gained by a scavenger per unit of carrion consumed |
| `grass_rate` | 0.05 | Set low to create resource scarcity that accentuates the scavenging advantage |

**Expected output.** Under scarce grass conditions, mean agent energy should be
higher in the scavenging condition than the baseline, and the total number of
deaths should be lower. Carrion provides a density-dependent buffer: more
deaths in a tick produce more carrion, partially compensating for the energy
deficit that caused those deaths.

```{r scavenging}
library(clade)
library(ggplot2)

base_specs <- function() {
  s <- default_specs()
  s$n_agents_init <- 100L
  s$grid_rows     <- 30L
  s$grid_cols     <- 30L
  s$grass_rate    <- 0.15
  s$max_ticks     <- 400L
  s
}

s_no <- base_specs(); s_no$scavenging <- FALSE
s_sc <- base_specs()
s_sc$scavenging        <- TRUE
s_sc$carrion_fraction  <- 0.5
s_sc$carrion_decay_rate <- 0.1
s_sc$carrion_eat_gain  <- 3.0

d_no <- get_run_data(run_alife(s_no))$ticks
d_sc <- get_run_data(run_alife(s_sc))$ticks

d_no$condition <- "No scavenging"
d_sc$condition <- "Scavenging"
dat <- rbind(d_no, d_sc)

ggplot(dat, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("No scavenging" = "#E53935", "Scavenging" = "#FB8C00"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Mean agent energy",
    title = "Scavenging sustains energy budgets under resource scarcity"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-scavenging, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: scavengers (orange) maintain higher mean energy than non-scavengers (red) under identical low-grass conditions, with carrion providing a density-dependent buffer against starvation."}
FIG("scavenging")
```

### Discovery prompts

The baseline result shows that scavenging sustains higher mean energy than baseline under identical low-grass conditions. To go beyond:

1. **Carrion × disease** Add `disease = TRUE`. Dead agents deposit carrion; if infected carcasses transmit pathogens, scavenging may amplify epidemics. Does `n_new_infections` differ between scavenging and non-scavenging conditions under identical `transmission_prob`? Does `carrion_decay_rate` modulate pathogen persistence on carcasses?

2. **Kin scavenging** Add `kin_selection = TRUE`. Both kin altruism and scavenging buffer energy deficits — are they additive (both help independently) or redundant (populations with kin altruism scavenge less because energy transfers substitute for carrion)? Watch `n_altruistic_acts` and `mean_energy` jointly under combined and single-module conditions.

3. **Scavenging × learning** Add `rl_mode = "actor_critic"`. Can within-lifetime RL discover scavenging opportunities faster than pure genetic evolution can encode the behaviour? Compare `mean_energy` trajectories between RL and no-RL conditions in a scavenging-enabled world at tick 100 (early learning phase) vs tick 400 (late genetic phase).

---


## Theme 3 — Social evolution {#theme-3}

*When does helping pay? Cooperation, signalling, and divergence. These scenarios implement Hamilton's rule, public goods games, honest signalling, and reproductive isolation — the core theoretical toolkit of social evolution.*


### Kin selection and cooperative social structure

**What it models.** Agents can transfer energy to nearby relatives via two mechanisms: Hamilton's pedigree-based kin altruism (`kin_selection = TRUE`) and the IFfolk model (`iffolk_selection = TRUE`), which adds parliamentary enforcement of cooperative norms (Fromhage & Jennions 2019).

**Hamilton's kin selection.** Agents with energy above a threshold donate a fixed cost to the most closely related Moore-neighbourhood agent with relatedness r ≥ `kin_altruism_r_min`. Relatedness is computed from the pedigree (parent chains): r = 0.5 for parent–offspring, 0.25 for siblings, 0 otherwise. This operationalises Hamilton's rule: the altruistic act evolves when r × benefit > cost.

**IFfolk and parliamentary enforcement.** The IFfolk model extends kin selection with a `helper_tendency` trait and `parliament_suppression`, which penalises defectors (low `helper_tendency`) surrounded by cooperators. This models the *parliament of genes* (Fromhage & Jennions 2019): cooperative behaviour is enforced by neighbour-level suppression of defection.

**Hamilton's kin altruism parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `kin_selection` | FALSE | Enable kin altruism |
| `kin_altruism_r_min` | 0.25 | Minimum relatedness to donate (siblings) |
| `kin_altruism_cost` | 2.0 | Energy transferred per act |
| `kin_altruism_min_donor_energy` | 50.0 | Donor energy floor |

**IFfolk and parliament parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `iffolk_selection` | FALSE | Enable IFfolk transfers |
| `iffolk_r_min` | 0.125 | Minimum relatedness (cousins and closer) |
| `iffolk_radius` | 5 | Neighbourhood radius for kin search |
| `iffolk_transfer` | 3.0 | Energy transferred per act |
| `iffolk_min_energy` | 60.0 | Donor must exceed this energy |
| `parliament_suppression` | FALSE | Penalise defectors among cooperators |
| `parliament_cost` | 0.5 | Energy penalty for defectors |
| `cooperative_breeding` | FALSE | Enable `helper_tendency` trait |
| `helper_tendency_init_mean` | 0.2 | Starting helper tendency |

**Expected outputs.** (1) Kin altruism: `n_altruistic_acts` is positive; population size is generally higher than baseline; genetic diversity may be slightly lower due to local inbreeding. (2) IFfolk: `mean_helper_tendency` increases over time; `n_iffolk_transfers` grows with population size; parliament suppression accelerates convergence near 1.0.

**Example: Hamilton kin altruism.**

```{r kin-selection}
library(clade)
library(ggplot2)
library(patchwork)

run_one <- function(kin, seed) {
  s <- default_specs()
  s$kin_selection          <- kin
  s$kin_altruism_r_min     <- 0.25   # help siblings (r ≥ 0.25) and closer
  s$kin_altruism_cost      <- 5.0    # 5 energy units per act
  s$grass_rate             <- 0.08   # scarce resources to create selection pressure
  s$n_agents_init          <- 80L
  s$max_agents             <- 400L
  s$max_ticks              <- 300L
  s$random_seed            <- as.integer(seed)
  env  <- run_alife(s, verbose = FALSE)
  d    <- get_run_data(env)$ticks
  cbind(d[, c("t", "n_agents", "n_altruistic_acts")],
        condition = if (kin) "Kin selection" else "Baseline")
}

seeds <- c(1L, 7L, 13L, 19L, 25L)
df <- do.call(rbind, c(
  lapply(seeds, function(s) run_one(FALSE, s)),
  lapply(seeds, function(s) run_one(TRUE,  s))
))

# Average across replicates
df_mean <- aggregate(cbind(n_agents, n_altruistic_acts) ~ t + condition,
                     data = df, FUN = mean)

p1 <- ggplot(df_mean, aes(t, n_agents, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Kin selection" = "#2196F3",
                                  "Baseline"      = "#F44336")) +
  labs(title = "Kin selection: population dynamics",
       x = "Tick", y = "Mean n_agents (5 reps)", colour = NULL) +
  theme_minimal()

p2 <- ggplot(df_mean[df_mean$condition == "Kin selection",],
             aes(t, n_altruistic_acts)) +
  geom_line(colour = "#2196F3") +
  labs(title = "Altruistic acts per tick",
       x = "Tick", y = "n_altruistic_acts") +
  theme_minimal()

p1 / p2
```

**What we found.** Running the code above (5 replicates, `grass_rate = 0.08`,
`kin_altruism_cost = 5.0`, `kin_altruism_r_min = 0.25`, 80 agents, 300 ticks):

- **Kin selection ON**: mean final population 141 ± 11; mean population across
  the run 196; total altruistic acts ≈ 2751 across 300 ticks (≈ 9 acts/tick on
  average).
- **Baseline (no kin)**: mean final population 135 ± 5; mean population 179.

Kin selection sustained ~10% higher mean population under scarce resources
(grass_rate = 0.08 vs default 0.15). The benefit is consistent across all 5
replicates, confirming Hamilton's rule: r × B > C holds at r = 0.25 and
kin_altruism_cost = 5.0, meaning the indirect fitness benefit to the donor's
relatives exceeds the direct cost. At `kin_altruism_r_min = 0.5` (parents
only), the benefit disappears because qualifying kin are too rare.

```{r fig-kin-selection, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: kin-selection population (blue) is sustained at higher mean density than the baseline (red). Altruistic acts per tick (bottom) are highest when resources are tight and related neighbours are present."}
FIG("07_kin_selection")
```

**Example: IFfolk with parliament suppression.**

```{r iffolk}
s <- default_specs()
s$iffolk_selection        <- TRUE
s$iffolk_r_min            <- 0.125
s$iffolk_transfer         <- 3.0
s$parliament_suppression  <- TRUE
s$parliament_cost         <- 0.5
s$cooperative_breeding    <- TRUE
s$helper_tendency_init_mean <- 0.2
s$max_ticks               <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

plot(data$ticks$t, data$ticks$mean_helper_tendency, type = "l",
     xlab = "Tick", ylab = "Mean helper tendency",
     main = "Helper tendency evolves under IFfolk + parliament suppression")
```

```{r fig-iffolk, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean helper tendency rises over time as cooperative genotypes increase in frequency. Parliament suppression accelerates convergence."}
FIG("20_cooperative_breeding")
```

### Discovery prompts

The baseline results confirm that Hamilton's rule operates: altruistic acts accumulate and population size is sustained at higher density than baseline. To go beyond:

1. **Epidemic corridors** Add `disease = TRUE`. Kin altruism requires proximity; does proximity-clustering among relatives accelerate pathogen spread? Compare epidemic peak `n_infected` between kin and no-kin conditions. Does `kin_altruism_r_min` modulate epidemic severity — do populations with stricter relatedness thresholds (fewer but closer relatives) show lower or higher peak infection?

2. **Relatedness threshold sweep** Vary `kin_altruism_r_min` from 0.0 to 0.5 across six values in `batch_alife()`. Hamilton's rule predicts a threshold (r × benefit > cost) below which altruism is not favoured. Is there a sharp transition in `n_altruistic_acts`, or a smooth decay? Plot final altruism rate against `kin_altruism_r_min`.

3. **Kin clusters + niche construction** Add `niche_construction = TRUE`. Do relatives cluster in sheltered cells, creating kin-structured microhabitats? Measure the spatial correlation between `shelter_density` and the local mean pedigree relatedness of occupants at the final tick. Does niche construction accelerate or slow kin group formation?

---

### Cooperative breeding and public goods

**What it models.** When `cooperation_evolution = TRUE`, each agent carries a
heritable cooperation tendency trait that determines its contribution to a local
public goods pool. The pool benefit is shared among all agents in the
neighbourhood, scaled by `cooperation_multiplier`, while contributors pay a
fixed `cooperation_cost`. This implements a spatially explicit Prisoner's
Dilemma: defectors (low tendency) free-ride on cooperators but, because
interactions are local, spatial clustering of cooperators can shield them from
invasion by defectors — a result established analytically by Nowak & May (1992)
and extended to continuous strategies by Hauert et al. (2002).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `cooperation_evolution` | FALSE | Enables heritable cooperation tendency |
| `cooperation_multiplier` | 2.0 | Return multiplier on pooled contributions |
| `cooperation_cost` | 1.0 | Per-tick energy cost paid by the contributor |
| `cooperation_init_mean` | 0.5 | Initial mean cooperation tendency |
| `cooperation_mutation_sd` | 0.05 | Mutation standard deviation on the trait |

**Expected output.** Track `mean_cooperation_level` over ticks. When
`cooperation_multiplier` exceeds the cost-benefit threshold (here
2.5 > 1.0 / 0.5 = 2.0), cooperation is favoured and the population-mean
tendency should rise toward 1.0; spatial structure sustains cooperation even at
intermediate global frequencies.

```{r cooperation}
library(clade)
library(ggplot2)

s <- default_specs()
s$cooperation_evolution  <- TRUE
s$cooperation_multiplier <- 2.5
s$cooperation_cost       <- 1.0
s$max_ticks              <- 400L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_cooperation_level)) +
  geom_line(colour = "#1b7837", linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, colour = "#762a83",
              fill = "#c2a5cf", alpha = 0.3) +
  labs(
    title    = "Evolution of cooperation tendency",
    subtitle = "cooperation_multiplier = 2.5, cost = 1.0",
    x = "Tick", y = "Mean cooperation tendency"
  ) +
  theme_minimal()
```

```{r fig-cooperation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean cooperation tendency rises over evolutionary time when the multiplier exceeds the cost-benefit threshold. Spatial clustering accelerates the transition by shielding cooperator clusters from defector invasion."}
FIG("08_cooperation")
```

### Discovery prompts

The baseline result confirms that cooperation is favoured when the multiplier exceeds the cost-benefit threshold, and spatial clustering accelerates the transition. To go beyond:

1. **Cooperation × disease** Add `disease = TRUE`. Cooperators cluster spatially (to benefit from the public goods pool), which may create epidemic hotspots. Does cooperation accelerate disease spread, creating fluctuating selection on `mean_cooperation_level` that matches epidemic cycles?

2. **Cooperation × predation** Add `n_predators_init = 5L`. Group-level selection predicts cooperation is easier to maintain when external mortality is high, because non-cooperating groups go extinct under predation. Does predation pressure increase the rate of cooperation evolution, and does the effect depend on `cooperation_multiplier`?

3. **Multiplier threshold** Vary `cooperation_multiplier` from 1.0 to 4.0 across six values in `batch_alife()`. Theory predicts a sharp threshold at multiplier = 1/(cooperation tendency mean), but spatial structure may lower this. Is the transition sharp or gradual? Does `n_agents_init` (population size, affecting local encounter rates) shift the threshold?

---

### Signals and mate choice: sexual selection

**What it models.** Sexual selection can produce elaborate traits that reduce
survival but increase mating success. Two contrasting mechanisms have been
proposed. Zahavi's (1975) handicap principle requires that signals be
condition-dependent and costly to maintain, so only high-quality individuals
can afford elaborate signals; signal magnitude is then an honest indicator of
genetic quality. Fisher's (1930) runaway process operates through a genetic
correlation between signal elaboration and preference: once females prefer a
trait, sons of choosy females inherit the trait and daughters inherit the
preference, driving both to fixation. This scenario enables multi-dimensional
signals and female preference, allowing both mechanisms to operate, and tracks
whether signal magnitude remains correlated with `mean_energy` (consistent with
honest signalling) or becomes decoupled (consistent with runaway dynamics).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `signal_dims` | 0L | Number of heritable signal dimensions |
| `signal_cost` | 0.1 | Energy cost per unit of total signal magnitude |
| `mate_choice_mode` | `"random"` | `"preference"` enables directional mate choice |
| `mate_choice_strength` | 0.5 | Weight given to signal match in mate selection |

**Expected output.** `mean_signal_magnitude` increases over time as sexual
selection drives signal elaboration. When `signal_cost` is sufficient to
maintain honesty, `mean_signal_magnitude` correlates positively with
`mean_energy`: agents in better condition carry more elaborate signals. A weak
or zero `signal_cost` allows decoupling of signal from condition, consistent
with runaway dynamics.

```{r signals-matechoice}
library(clade)
library(ggplot2)

s <- default_specs()
s$signal_dims          <- 3L
s$signal_cost          <- 0.05
s$mate_choice_mode     <- "preference"
s$mate_choice_strength <- 0.7
s$max_ticks            <- 400L
s$random_seed          <- 21L

env  <- run_alife(s)
data <- get_run_data(env)

p1 <- ggplot(data$ticks, aes(t, mean_signal_magnitude)) +
  geom_line(colour = "#984ea3") +
  labs(title = "Signal elaboration under sexual selection",
       x = "Tick", y = "Mean signal magnitude") +
  theme_minimal()

p2 <- ggplot(data$ticks, aes(mean_energy, mean_signal_magnitude)) +
  geom_point(alpha = 0.4, size = 0.8, colour = "#984ea3") +
  geom_smooth(method = "lm", se = FALSE, colour = "#333333", linewidth = 0.7) +
  labs(title = "Signal magnitude vs mean energy (honest signalling check)",
       x = "Mean energy", y = "Mean signal magnitude") +
  theme_minimal()

print(p1)
print(p2)
```

```{r fig-signals-matechoice, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean signal magnitude rises over time as sexual selection drives elaboration. A positive correlation between signal magnitude and mean energy indicates that the handicap mechanism is maintaining signal honesty."}
FIG("signals_matechoice")
```

### Discovery prompts

The baseline result shows signal elaboration under sexual selection, with signal magnitude correlating positively with mean energy when `signal_cost` maintains honesty. To go beyond:

1. **Signal × brain size** Add `brain_size_evolution = TRUE`. Larger-brained agents have better sensory processing; does brain size evolve faster when signals provide additional information about mate quality under `mate_choice_mode = "preference"`? Compare `mean_brain_size` trajectories with and without mate choice enabled.

2. **Signal honesty under resource stress** Run with `grass_rate = 0.05`. When mean condition is uniformly low, the handicap principle predicts signal honesty breaks down because high-quality individuals cannot afford to distinguish themselves. Does the correlation between `mean_signal_magnitude` and `mean_energy` weaken under scarcity compared to the rich-resource baseline?

3. **Signal × cooperation** Add `cooperation_evolution = TRUE`. Can heritable cooperation tendency serve as an honest signal of genetic quality, substituting for costly ornamental signals? If cooperators have higher mean energy (from public goods), does `mean_signal_magnitude` decline when cooperation is available as an alternative quality indicator?

---

### Speciation and genetic divergence

**What it models.** Speciation by ecological character displacement occurs when
disruptive selection on a resource-use trait, combined with assortative mating
among similar phenotypes, drives a single population to split into
reproductively isolated lineages (Dieckmann & Doebeli 1999). Once genomes
diverge beyond a threshold, gene flow between incipient species is reduced,
accelerating further divergence. Coyne & Orr (2004) reviewed extensive
empirical evidence that reproductive isolation accumulates approximately
linearly with genetic distance. This scenario enables the speciation module,
tracks the number of distinct genetic clusters (`n_species`) detected at
regular intervals, and illustrates how within-species diversity compares to
between-species divergence over the course of the run.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `speciation` | FALSE | Enable genetic-cluster detection and isolation |
| `isolation_threshold` | 0.5 | Minimum genetic distance required for reproductive isolation |
| `speciation_cluster_interval` | 10L | Ticks between cluster re-detection passes |

**Expected output.** `n_species` starts at 1 and rises to 2–4 as accumulated
genome divergence exceeds `isolation_threshold` in geographically or
ecologically separated subpopulations. Within-cluster genetic diversity remains
lower than between-cluster diversity once isolation is established. The timing
of the first split depends on population size, mutation rate, and the
stringency of `isolation_threshold`.

```{r speciation}
library(clade)
library(ggplot2)

s <- default_specs()
s$speciation                  <- TRUE
s$isolation_threshold         <- 0.5
s$speciation_cluster_interval <- 10L
s$max_ticks                   <- 500L
s$random_seed                 <- 55L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(t, n_species)) +
  geom_step(colour = "#ff7f00", linewidth = 0.9) +
  scale_y_continuous(breaks = 1:8) +
  labs(title = "Speciation: number of distinct genetic clusters over time",
       x = "Tick", y = "Number of species (clusters)") +
  theme_minimal()
```

```{r fig-18-speciation, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: n_species rises from 1 to 2-4 as genome divergence accumulates beyond the isolation threshold. Each step in the staircase represents a speciation event."}
FIG("18_speciation")
```

### Discovery prompts

The baseline result shows `n_species` rising from 1 to 2–4 as genetic distance accumulates past `isolation_threshold`. To go beyond:

1. **Speciation × disease** Add `disease = TRUE`. Genetically distinct species share fewer contact opportunities (reproductive isolation requires spatial separation), which may create inter-species barriers to pathogen transmission. Does speciation reduce epidemic peak `n_infected` by fragmenting the contact network into partially isolated sub-populations?

2. **Speciation × brain size** Add `brain_size_evolution = TRUE`. Do incipient species diverge in brain size as a by-product of ecological niche partitioning, or does brain size converge across species via parallel evolution toward the same cognitive optimum? Compare `mean_brain_size` across detected genetic clusters at final tick.

3. **Isolation threshold sensitivity** Vary `isolation_threshold` from 0.2 to 0.8 across seven values in `batch_alife()`. Does speciation rate (slope of `n_species` vs time) respond monotonically to threshold, or is there a critical threshold below which sympatric speciation cannot initiate even given sufficient time?

---


## Theme 4 — Life history strategies {#theme-4}

*How life is organised across time and energy. These scenarios examine the trade-offs between reproduction timing, parental investment, lifespan, and metabolic pace that define the slow–fast life-history continuum.*


### Parental care

**What it models.** Offspring are carried by the parent until graduation at age
`care_duration`. While carried, juveniles receive energy from the parent at
`feeding_rate` per tick. If the parent dies, carried offspring die too
(obligate altriciality). This models the evolutionary trade-off between clutch
size and offspring quality (Smith & Fretwell 1974).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_care` | FALSE | Enable parental care |
| `care_duration` | 5 | Ticks offspring are carried |
| `care_cost_per_tick` | 1.0 | Energy drained from parent per juvenile per tick |
| `feeding_rate` | 5.0 | Energy transferred to juvenile per tick |

**Expected output.** `n_juveniles` is positive. Per-capita offspring count may
be lower than baseline (parents can carry fewer), but juvenile survival is
higher. Population dynamics are more buffered.

```{r parental-care}
s <- default_specs()
s$parental_care     <- TRUE
s$care_duration     <- 5L
s$care_cost_per_tick <- 2.0
s$max_ticks         <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Total juveniles recorded:", sum(data$ticks$n_juveniles, na.rm = TRUE), "\n")
```

```{r fig-parental-care, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: juvenile count is positive throughout the run; per-capita offspring count is lower than baseline but juvenile survival is higher, resulting in more buffered population dynamics."}
FIG("19_parental_care")
```

### Discovery prompts

The baseline result shows that juvenile counts are positive and population dynamics are more buffered under parental care. To go beyond:

1. **Care × brain size** Add `brain_size_evolution = TRUE`. The parental provisioning hypothesis predicts parental care is a prerequisite for brain size evolution. Does `mean_brain_size` at tick 300 increase more under `parental_care = TRUE`? Vary `care_duration` across `{2, 5, 10, 20}` ticks in `batch_alife()` to find the minimum care duration that allows brain size to evolve upward.

2. **Care × disease** Add `disease = TRUE`. Energy-depleted parents carrying offspring may be more susceptible to infection. Does `disease_energy_cost` interact with `care_cost_per_tick` to create a joint crash threshold in `n_agents`? Plot population survival probability against the sum of both costs.

3. **Care × kin selection** Add `kin_selection = TRUE`. Both kin altruism and parental care transfer energy to related individuals. Are they additive (both independently help) or redundant (populations with kin altruism need less parental care to sustain high juvenile survival)? Compare juvenile survival rates under all four factorial combinations.

---

### Mating systems: sexual versus asexual reproduction

**What it models.** Sexual reproduction incurs a well-documented two-fold cost
relative to asexual reproduction — every sexual individual devotes half its
offspring capacity to producing males — yet sex is widespread across eukaryotes
(Maynard Smith 1978; Williams 1975). The leading explanations invoke the
recombination benefits of sex: sexual populations generate more genotypic
variation per generation, providing raw material for selection in fluctuating or
parasitised environments. This scenario compares a haploid asexual population
(`ploidy = 1L`) against a diploid sexual population (`ploidy = 2L`,
`crossover_rate = 0.1`) to illustrate how mating system shapes the trajectory
of genetic diversity.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `ploidy` | 2L | 1 = haploid asexual; 2 = diploid sexual |
| `crossover_rate` | 0.0 | Probability of crossover per chromosome per reproduction event |
| `n_chromosomes` | 1L | Number of chromosome pairs (diploid only) |
| `dominance_model` | `"additive"` | How alleles at the same locus combine |

**Expected output.** `genetic_diversity` is higher and more variable in the
sexual condition from early in the run, as recombination continuously generates
novel genotypic combinations. The asexual population maintains lower diversity
and may converge faster to a local optimum — an advantage in stable
environments — but is less able to respond to environmental change.

```{r mating-systems}
library(clade)
library(ggplot2)

s_asex <- default_specs()
s_asex$ploidy         <- 1L
s_asex$crossover_rate <- 0.0
s_asex$max_ticks      <- 400L
s_asex$random_seed    <- 42L

s_sex <- default_specs()
s_sex$ploidy         <- 2L
s_sex$crossover_rate <- 0.1
s_sex$max_ticks      <- 400L
s_sex$random_seed    <- 42L

d_asex <- get_run_data(run_alife(s_asex))$ticks
d_sex  <- get_run_data(run_alife(s_sex))$ticks

df <- rbind(
  cbind(d_asex[, c("t", "genetic_diversity")], system = "Asexual (haploid)"),
  cbind(d_sex[,  c("t", "genetic_diversity")], system = "Sexual (diploid)")
)

ggplot(df, aes(t, genetic_diversity, colour = system)) +
  geom_line() +
  scale_colour_manual(values = c("Asexual (haploid)" = "#377eb8",
                                 "Sexual (diploid)"  = "#e41a1c"),
                      name = NULL) +
  labs(title = "Genetic diversity: sexual vs asexual reproduction",
       x = "Tick", y = "Genetic diversity") +
  theme_minimal()
```

```{r fig-mating-systems, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: genetic diversity is higher and more variable in the sexual (diploid) condition. The asexual population converges to lower diversity more rapidly, illustrating the recombination advantage of sex."}
FIG("mating_systems")
```

### Discovery prompts

The baseline result shows genetic diversity is higher and more variable in the sexual (diploid) condition from early in the run. To go beyond:

1. **Sex × parasites (Red Queen)** Add `disease = TRUE` with high `transmission_prob = 0.3`. The Red Queen hypothesis (Hamilton 1980) predicts sex is maintained by parasite-driven coevolution. Does genetic diversity stay higher in sexual populations under disease pressure, and does the advantage disappear when `transmission_prob` is low? Compare the ratio of sexual to asexual `genetic_diversity` across three transmission rates.

2. **Sex × stress hypermutation** Add `stress_hypermutation = TRUE`. Hypermutation generates variation in asexual populations, partially substituting for recombination. Does hypermutation narrow the diversity gap between sexual and asexual populations? Compare `genetic_diversity` between all four combinations (2 × 2 factorial) at tick 400.

3. **Ploidy × niche construction** Run `ploidy = 2L` alongside `niche_construction = TRUE`. Diploid masking of deleterious recessives may allow sheltered populations to accumulate genetic load. Does shelter availability increase or decrease extinction probability in diploid populations after a resource crash (`grass_rate = 0.01` for 50 ticks)?

---

### Life history: semelparous versus iteroparous strategies

**What it models.** Cole's paradox (Cole 1954) established that the fitness
advantage of iteroparity over semelparity is surprisingly small — one additional
surviving offspring is sufficient to equalise the two strategies under simple
demographic conditions. Real divergence arises from differences in juvenile
versus adult survival, resource availability, and the reliability of the
reproductive season (Roff 1992). Semelparous organisms invest all reproductive
resources in a single event and die; iteroparous organisms spread reproduction
across multiple seasons. This scenario contrasts the two life histories, holding
other parameters constant, to reveal how each shapes population age structure
and birth-rate dynamics.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `life_history` | `"iteroparous"` | `"semelparous"` triggers post-reproductive death |
| `max_age` | 200L | Maximum attainable age (iteroparous bound) |
| `senescence_rate` | 0.0 | Gompertz senescence coefficient |
| `repro_senescence` | 0.0 | Reproductive decline rate with age |

**Expected output.** The semelparous population shows a shorter `mean_age` and
episodic bursts in `n_births` coinciding with cohort turnover. The iteroparous
population maintains a smoother, more continuous birth rate and a higher
`mean_age`, with population size fluctuating less dramatically between
generations.

```{r life-history}
library(clade)
library(ggplot2)

s_sem <- default_specs()
s_sem$life_history <- "semelparous"
s_sem$max_ticks    <- 400L
s_sem$random_seed  <- 7L

s_ite <- default_specs()
s_ite$life_history <- "iteroparous"
s_ite$max_ticks    <- 400L
s_ite$random_seed  <- 7L

d_sem <- get_run_data(run_alife(s_sem))$ticks
d_ite <- get_run_data(run_alife(s_ite))$ticks

df <- rbind(
  cbind(d_sem[, c("t", "mean_age", "n_births")], strategy = "Semelparous"),
  cbind(d_ite[, c("t", "mean_age", "n_births")], strategy = "Iteroparous")
)

ggplot(df, aes(t, mean_age, colour = strategy)) +
  geom_line() +
  scale_colour_manual(values = c(Semelparous = "#e41a1c", Iteroparous = "#4daf4a"),
                      name = NULL) +
  labs(title = "Mean age: semelparous vs iteroparous life history",
       x = "Tick", y = "Mean agent age") +
  theme_minimal()
```

```{r fig-life-history, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: the semelparous population (red) shows lower mean age and episodic birth bursts; the iteroparous population (green) sustains a higher mean age and a smoother birth-rate trajectory."}
FIG("life_history")
```

### Discovery prompts

The baseline result shows semelparous populations have lower mean age and episodic birth bursts, while iteroparous populations sustain smoother birth rates and higher mean age. To go beyond:

1. **Semelparity × predation** Add `n_predators_init = 5L`. Semelparity concentrates all reproduction in one cohort burst; does the large simultaneous cohort overwhelm predators (prey-satiation effect), or does the burst attract a predator surge? Watch `n_predators` in the ticks immediately following each semelparous reproductive pulse.

2. **Semelparity × stress hypermutation** Add `stress_hypermutation = TRUE`. Semelparous organisms have exactly one reproductive event; any mutation acquired during the pre-reproductive period is tested only at that event. Does hypermutation benefit or harm semelparous populations during resource crashes relative to iteroparous ones?

3. **Iteroparous senescence gradient** Vary `senescence_rate` from 0.0 to 0.1 across five values in `batch_alife()` with `life_history = "iteroparous"`. Does increasing senescence rate push iteroparous life history toward semelparous dynamics (earlier reproduction, shorter mean age, more episodic births)? Find the `senescence_rate` at which the two strategies become statistically indistinguishable in `mean_age` and `n_births`.

---

### Clutch size evolution and r/K strategy

**What it models.** The Lack clutch-size hypothesis (Lack 1947) proposes that
parents should produce the number of offspring that maximises the number
successfully raised, given the resources available per offspring. Smith &
Fretwell (1974) formalised the quantity–quality trade-off: when resources are
scarce, producing fewer, better-provisioned offspring (K-strategy) yields higher
per-offspring survival than producing many poorly provisioned ones (r-strategy).
This scenario allows clutch size to evolve freely between `clutch_size_min` and
`clutch_size_max` under two contrasting resource environments.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `clutch_size_evolution` | FALSE | Enable heritable clutch size |
| `clutch_size_min` | 1L | Minimum evolvable clutch size |
| `clutch_size_max` | 5L | Maximum evolvable clutch size |
| `clutch_size_mutation_sd` | 0.3 | Gaussian noise on clutch size per offspring |
| `grass_rate` | 0.10 | Resource renewal rate (varied between conditions) |

**Expected output.** Under rich resources (`grass_rate = 0.4`),
per-tick births are higher as large clutches are sustainable.
Under scarce resources (`grass_rate = 0.05`), per-tick births are lower
and the population is constrained by energy — consistent with the K end
of the r/K continuum. The two conditions diverge substantially after
approximately 150 ticks.

```{r clutch-size}
library(clade)
library(ggplot2)

make_s <- function(gr) {
  s <- default_specs()
  s$clutch_size_evolution   <- TRUE
  s$clutch_size_min         <- 1L
  s$clutch_size_max         <- 5L
  s$clutch_size_mutation_sd <- 0.3
  s$grass_rate              <- gr
  s$max_ticks               <- 400L
  s$random_seed             <- 3L
  s
}

d_rich <- get_run_data(run_alife(make_s(0.4)))$ticks
d_scar <- get_run_data(run_alife(make_s(0.05)))$ticks

df <- rbind(
  cbind(d_rich[, c("t", "n_births")], environment = "Rich (0.4)"),
  cbind(d_scar[, c("t", "n_births")], environment = "Scarce (0.05)")
)

ggplot(df, aes(t, n_births, colour = environment)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_colour_manual(
    values = c("Rich (0.4)" = "#e41a1c", "Scarce (0.05)" = "#377eb8"),
    name = NULL) +
  labs(title = "Births per tick under contrasting resource availability",
       subtitle = "Clutch size evolution enabled",
       x = "Tick", y = "Births per tick") +
  theme_minimal()
```

```{r fig-clutch-size, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean clutch size diverges between resource environments. Rich conditions (red) select for larger clutches; scarce conditions (blue) select for smaller clutches, illustrating the r/K continuum."}
FIG("clutch_size")
```

### Discovery prompts

The baseline result shows that rich environments select for larger clutches (r-strategy) and scarce environments select for smaller clutches (K-strategy). To go beyond:

1. **Clutch size × parental care** Add `parental_care = TRUE`. Parental care constrains maximum clutch size via `max_clutch_size`. Does the evolved clutch size interact with `care_duration`? Does longer care duration shift the evolved clutch toward 1 (maximum quality investment), or do rich environments allow large clutches even with long care?

2. **Clutch size × predation** Add `n_predators_init = 5L`. High predation mortality selects for faster reproduction (r-strategy). Does predation pressure increase the evolved mean clutch size independent of resource availability, and does the effect depend on `grass_rate`? Run a 2×2 factorial (high/low resources × predation).

3. **r/K under seasonality** Add `seasonal_amplitude = 0.7`. In seasonal environments, r-strategists should dominate during spring resource abundance and K-strategists should persist through winter scarcity. Does `mean_clutch_size` oscillate with the seasonal cycle? Compute the cross-correlation between `mean_clutch_size` and `grass_coverage` to test seasonal tracking.

---

### Parental investment: quality versus quantity

**What it models.** Trivers (1972) proposed that the sex that invests more per
offspring should be the one that is more choosy in mate selection, because its
reproductive success is more constrained by parental effort than by mate access.
Houston & Davies (1985) extended this to biparental care games, showing that
evolutionarily stable investment levels depend on the shape of the offspring
fitness function. This scenario holds `parental_care = TRUE` and varies
`female_investment` — the fraction of care energetics contributed by the female
— to examine how the allocation of investment between parents affects offspring
provisioning and survival.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `parental_investment_evolution` | FALSE | Allow investment proportions to evolve |
| `parental_care` | FALSE | Enable offspring carrying and provisioning |
| `female_investment` | 0.7 | Fraction of total care energy contributed by the mother |
| `male_repro_cost` | 0.3 | Energy cost to the male per reproductive event |

**Expected output.** High maternal investment (`female_investment = 0.9`)
increases the per-offspring energy cost, so fewer births occur per tick.
Equal investment (`female_investment = 0.5`) allows more births at lower
per-offspring cost. The trade-off is visible in `n_births` and
`n_juveniles` trajectories.

```{r parental-investment}
library(clade)
library(ggplot2)

make_s <- function(fi) {
  s <- default_specs()
  s$parental_care                 <- TRUE
  s$parental_investment_evolution <- TRUE
  s$female_investment             <- fi
  s$male_repro_cost               <- 0.3
  s$max_ticks                     <- 400L
  s$random_seed                   <- 11L
  s
}

d_hi <- get_run_data(run_alife(make_s(0.9)))$ticks
d_eq <- get_run_data(run_alife(make_s(0.5)))$ticks

df <- rbind(
  cbind(d_hi[, c("t", "n_births", "n_juveniles")],
        condition = "High maternal (0.9)"),
  cbind(d_eq[, c("t", "n_births", "n_juveniles")],
        condition = "Equal (0.5)")
)

ggplot(df, aes(t, n_births, colour = condition)) +
  geom_line(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_colour_manual(
    values = c("High maternal (0.9)" = "#e41a1c", "Equal (0.5)" = "#377eb8"),
    name = NULL) +
  labs(title = "Births per tick by parental investment allocation",
       subtitle = "High maternal investment → fewer but better-provisioned offspring",
       x = "Tick", y = "Births per tick") +
  theme_minimal()
```

```{r fig-parental-investment, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: high maternal investment (red) produces fewer but better-provisioned offspring; equal investment (blue) produces more offspring at lower individual quality, illustrating the quality-quantity trade-off."}
FIG("parental_investment")
```

### Discovery prompts

The baseline result shows the quality-quantity trade-off: high maternal investment produces fewer but better-provisioned offspring. To go beyond:

1. **Investment × brain size** Add `brain_size_evolution = TRUE`. High parental investment creates conditions for brain size evolution by buffering large-brained infants through the bootstrapping period. Does `female_investment = 0.9` produce higher final `mean_brain_size` than `female_investment = 0.5`? How does this interact with `brain_size_cost_scale`?

2. **Investment × kin selection** Add `kin_selection = TRUE`. Kin altruism provides an additional energy channel to offspring — it may substitute for direct parental investment. Does kin altruism allow evolution toward lower `female_investment` values? Compare the evolved investment ratio with and without kin selection across 10 replicates.

3. **Biparental game** Vary `female_investment` from 0.1 to 0.9 across nine values in `batch_alife()`. Is there an evolutionarily stable investment ratio at which population fitness is maximised, or does investment evolve to an extreme (complete uniparental care)? Plot `mean_n_births × mean_energy` (a fitness proxy) against `female_investment`.

---

### Pace-of-life syndromes

**What it models.** Pace-of-life syndromes (Réale et al. 2010) describe the
co-variation of life-history traits, behaviour, and physiology along a slow–fast
continuum. Fast-living organisms have high metabolic rates, reproduce early and
often, and die young; slow-living organisms invest in maintenance, defer
reproduction, and achieve longer lifespans. Stearns (1992) formalised this as a
consequence of the age-specific mortality schedule: high extrinsic mortality
selects for early reproduction and fast metabolism. This scenario fixes metabolic
rate across three values to isolate the direct effect of metabolic pace on the
age–reproduction–energy trade-off.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `metabolic_rate_init_mean` | 1.0 | Initial (and fixed) metabolic rate |
| `metabolic_rate_evolution` | FALSE | Keep metabolic rate fixed for this scenario |
| `metabolic_rate_min` | 0.1 | Lower clamp (relevant when evolution is enabled) |
| `metabolic_rate_max` | 5.0 | Upper clamp |

**Expected output.** Fast-pace agents (`metabolic_rate_init_mean = 2.0`)
exhibit higher `n_births` per tick, lower `mean_age`, and higher energy
throughput, but their populations are more volatile. Slow-pace agents
(`metabolic_rate_init_mean = 0.5`) show lower birth rate, higher `mean_age`,
and more stable `mean_energy`. Fast agents dominate in productive environments;
slow agents persist more reliably when resources are lean.

```{r pace-of-life}
library(clade)
library(ggplot2)

make_run <- function(rate, seed = 99L) {
  s <- default_specs()
  s$metabolic_rate_init_mean <- rate
  s$metabolic_rate_evolution <- FALSE
  s$max_ticks                <- 400L
  s$random_seed              <- seed
  get_run_data(run_alife(s))$ticks
}

d_slow     <- make_run(0.5)
d_baseline <- make_run(1.0)
d_fast     <- make_run(2.0)

df <- rbind(
  cbind(d_slow[,     c("t", "mean_age", "n_births")], pace = "Slow (0.5)"),
  cbind(d_baseline[, c("t", "mean_age", "n_births")], pace = "Baseline (1.0)"),
  cbind(d_fast[,     c("t", "mean_age", "n_births")], pace = "Fast (2.0)")
)

ggplot(df, aes(t, mean_age, colour = pace)) +
  geom_line() +
  scale_colour_manual(
    values = c("Slow (0.5)" = "#377eb8", "Baseline (1.0)" = "#4daf4a",
               "Fast (2.0)" = "#e41a1c"),
    name = NULL) +
  labs(title = "Pace-of-life: mean age across metabolic rates",
       x = "Tick", y = "Mean agent age") +
  theme_minimal()
```

```{r fig-pace-of-life, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: slow-pace agents (blue) maintain the highest mean age; fast-pace agents (red) have the lowest mean age but the highest per-tick birth rate. The three trajectories illustrate the life-history trade-off along the slow-fast continuum."}
FIG("pace_of_life")
```

### Discovery prompts

The baseline result shows the pace-of-life trade-off: fast-pace agents have higher birth rates and lower mean age; slow-pace agents have lower birth rates and higher mean age and energy stability. To go beyond:

1. **Pace × disease** Add `disease = TRUE`. High metabolic rate means fast-paced organisms have less energy margin; `disease_energy_cost` should push fast-paced agents into energy deficit more readily. Does disease selectively eliminate fast-paced phenotypes when `metabolic_rate_evolution = TRUE`, driving evolved `mean_metabolic_rate` downward?

2. **Pace × brain size** Enable both `metabolic_rate_evolution = TRUE` and `brain_size_evolution = TRUE`. Do fast-paced, high-energy agents evolve larger brains (more metabolic resources available for expensive brain tissue), or does metabolic competition prevent it? Test whether `mean_brain_size` and `mean_metabolic_rate` are positively or negatively correlated at final tick.

3. **Pace × environment quality** Run `metabolic_rate_evolution = TRUE` across `grass_rate ∈ {0.02, 0.1, 0.4}` in `batch_alife()`. Theory predicts rich environments select for fast pace. Does evolved `mean_metabolic_rate` increase monotonically with `grass_rate`, or is there a non-monotone relationship (very high resources remove mortality pressure on fast-paced energy burn)?

---


## Theme 5 — Species interactions and arms races {#theme-5}

*Predator–prey coevolution, infectious disease, warning coloration, and group defence. These scenarios operationalise the major classes of antagonistic species interactions and ask how each reshapes phenotypic and genetic diversity.*


### Predator-prey dynamics

**What it models.** Predator-prey coevolution is one of the oldest and most
empirically grounded problems in population biology. The classical
Lotka-Volterra equations predict sustained oscillations in which predator
abundance tracks prey abundance with a characteristic lag (Lotka 1925;
Volterra 1928). In `clade`, predators are autonomous agents with their own
energy budgets, movement rules, and reproduction thresholds, so oscillations
emerge from individual behaviour rather than being imposed by differential
equations.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0 | Number of predators at tick 1; 0 disables predators entirely |
| `predator_attack_strength` | 40.0 | Energy damage dealt to prey per successful attack |
| `predator_energy_gain` | 30.0 | Energy gained by predator after a successful attack |
| `predator_min_repro_energy` | 200.0 | Energy threshold for predator reproduction |
| `predator_max_agents` | 50 | Hard cap on predator population size |
| `grass_rate` | 0.05 | Grass regrowth rate; sets prey carrying capacity |

**Expected output.** Watch `n_agents` (prey) and `n_predators` over time. Prey
population rises first, then predators increase as food becomes abundant; prey
then crash, pulling predators down with a lag of approximately 30 ticks — a
pattern qualitatively consistent with Lotka-Volterra predictions.

```{r predator-prey}
library(clade)
library(ggplot2)

s <- default_specs()
s$n_predators_init      <- 5L
s$n_agents_init         <- 100L
s$grid_rows             <- 30L
s$grid_cols             <- 30L
s$grass_rate            <- 0.3
s$max_ticks             <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Prey"),     linewidth = 0.8) +
  geom_line(aes(y = n_predators, colour = "Predators"), linewidth = 0.8) +
  scale_colour_manual(
    values = c("Prey" = "#2196F3", "Predators" = "#F44336"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Predator-prey population dynamics"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-predators, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: prey and predator populations oscillate with predator peaks lagging prey peaks by approximately 30 ticks, consistent with Lotka-Volterra dynamics."}
FIG("14_predators")
```

### Discovery prompts

The baseline result shows Lotka-Volterra-like oscillations with predator peaks lagging prey peaks by approximately 30 ticks. To go beyond:

1. **Coevolution × brain size** Add `brain_size_evolution = TRUE` for prey agents (predators use reference brain_size = 1.0). Does predation pressure select for larger prey brains (better spatial evasion through improved sensing)? Compare `mean_brain_size` with and without predators at tick 500, and test whether `brain_size_sensing_exponent` modulates the effect size.

2. **Disease × predation interaction** Add `disease = TRUE`. Infected prey may be energetically depleted and easier to catch. Does disease amplify predation mortality, creating combined boom-bust cycles more severe than either selective pressure alone? Watch whether `n_infected` and predator kill events are temporally correlated.

3. **Predator energy threshold** Vary `predator_min_repro_energy` from 100 to 400 across `batch_alife()`. Does tightening the predator energy threshold eliminate population oscillations (by reducing predator fecundity), change the oscillation period, or cause predator extinction? Plot oscillation amplitude and period against `predator_min_repro_energy`.

---

### Group defense (dilution of risk)

**What it models.** Hamilton's (1971) selfish herd hypothesis proposes that
aggregation reduces individual predation risk by diluting attacks across a
larger group. When any single predator can only attack one agent per encounter,
each additional group member decreases the per-capita probability of being the
target. In `clade`, group defense is implemented as a collective
damage-reduction mechanism: agents within `group_defense_radius` cells of an
attacked individual jointly absorb a fraction of the attack, reducing effective
predation pressure on each member.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `group_defense` | FALSE | Enables collective damage reduction in groups |
| `group_defense_radius` | 2 | Neighbourhood radius (cells) over which defense is pooled |
| `group_defense_strength` | 0.3 | Fraction by which attack damage is reduced per additional neighbour |
| `n_predators_init` | 0 | Set > 0 to activate predation pressure |
| `predator_attack_strength` | 40.0 | Baseline damage before group reduction is applied |

**Expected output.** In the `group_defense = TRUE` condition, mean population
size should be substantially higher than the baseline, and survival curves
should diverge after the first predator boom. The mechanism is dilution:
per-capita attack rate falls as group size grows, creating a positive feedback
between aggregation and survival.

```{r group-defense}
library(clade)
library(ggplot2)

base_specs <- function() {
  s <- default_specs()
  s$n_predators_init <- 5L
  s$n_agents_init    <- 100L
  s$grid_rows        <- 30L
  s$grid_cols        <- 30L
  s$max_ticks        <- 400L
  s
}

s_no <- base_specs()
s_no$group_defense <- FALSE

s_gd <- base_specs()
s_gd$group_defense          <- TRUE
s_gd$group_defense_radius   <- 2L
s_gd$group_defense_strength <- 0.3

d_no <- get_run_data(run_alife(s_no))$ticks
d_gd <- get_run_data(run_alife(s_gd))$ticks

d_no$condition <- "No group defense"
d_gd$condition <- "Group defense"
dat <- rbind(d_no, d_gd)

ggplot(dat, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("No group defense" = "#E53935", "Group defense" = "#43A047"),
    name   = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Population size",
    title = "Group defense reduces predation-driven extinction risk"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-group-defense, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: group defense (green) sustains a larger and more stable population than the baseline (red) under identical predation pressure, illustrating the dilution-of-risk effect."}
FIG("15_group_defense")
```

### Discovery prompts

The baseline result shows group defense sustains a larger and more stable population under identical predation pressure. To go beyond:

1. **Group defense × disease** Add `disease = TRUE`. Grouping reduces predation risk (dilution) but increases disease transmission (contact). Does `group_defense_radius` create a trade-off: larger groups are better protected from predation but suffer larger epidemic peaks? Find the radius that maximises net population size across a gradient of `transmission_prob`.

2. **Group defense × kin selection** Add `kin_selection = TRUE`. Kin clusters may provide group defense more reliably because relatives stay in proximity. Does kin selection increase the effective neighbourhood size available for defense? Compare mean neighbourhood density during predator attacks with and without kin selection.

3. **Group size × niche construction** Add `niche_construction = TRUE`. Shelters create spatial foci where agents aggregate, potentially concentrating group defense benefits. Does shelter density predict group defense effectiveness? Compare `n_agents` at final tick under `group_defense = TRUE` with and without niche construction.

---

### Mimicry and toxicity

**What it models.** A heritable `toxicity` trait makes prey costly to attack.
Predators learn signal–toxicity associations via a Rescorla-Wagner rule and
build avoidance. This models the evolution of warning coloration and Batesian/
Müllerian mimicry (Ruxton et al. 2004): toxic prey are avoided; non-toxic
prey that resemble toxic prey are also avoided (Batesian mimicry).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `mimicry` | FALSE | Enable toxicity / predator learning |
| `toxicity_init_mean` | 0.0 | Starting toxicity (0 = non-toxic) |
| `toxin_dose` | 2.0 | Damage dealt to attacker per toxicity unit |
| `signal_memory` | 20 | Predator memory window for learning |

**Expected output.** `mean_toxicity` evolves upward under predator pressure.
`n_avoided_attacks` increases as predators learn. `n_toxic_attacks` may peak
and then decline as predators avoid toxic prey.

```{r mimicry}
s <- default_specs()
s$mimicry            <- TRUE
s$n_predators_init   <- 5L
s$toxicity_init_mean <- 0.1
s$max_ticks          <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-mimicry, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean toxicity rises under predator pressure; avoided attacks increase as predators learn to avoid toxic prey; toxic attack rate declines."}
FIG("21_mimicry")
```

### Discovery prompts

The baseline result shows `mean_toxicity` evolves upward under predator pressure and `n_avoided_attacks` increases as predators learn. To go beyond:

1. **Mimicry × kin selection** Add `kin_selection = TRUE`. Kin clusters may accelerate warning coloration evolution (Hamilton's genetical theory of aposematism). Does kin structure speed up or slow the evolution of `mean_toxicity`? Compare `mean_toxicity` trajectories under three conditions: no kin, kin without predators, kin with predators and mimicry.

2. **Mimicry × disease** Add `disease = TRUE`. High-toxicity prey produce costly defensive compounds while simultaneously paying `disease_energy_cost` when infected. Does disease suppress the evolution of costly toxicity by pushing agents into energy deficit during infection? Compare the final `mean_toxicity` under disease vs no-disease at matched `predator_attack_strength`.

3. **Signal honesty and toxicity** Add `signal_dims = 2L`. Do heritable ornamental signals co-evolve with toxicity (honest Müllerian signalling) or become decoupled (Batesian dishonest copying by non-toxic agents)? Plot `mean_toxicity` against `mean_signal_magnitude` as a parametric trajectory and test whether the correlation strengthens or weakens over evolutionary time.

---

### SIR disease

**What it models.** A Susceptible-Infected-Recovered epidemic model. Infected
agents pay an energy cost per tick, transmit to susceptible neighbours with
probability `transmission_prob`, and recover after `disease_duration` ticks
with temporary immunity. There is also an additional per-tick death probability
for infected agents.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `disease` | FALSE | Enable SIR dynamics |
| `transmission_prob` | 0.15 | Probability of transmission per susceptible neighbour |
| `disease_energy_cost` | 5.0 | Energy drained per tick while infected |
| `disease_duration` | 10 | Ticks until recovery |
| `immune_duration` | 20 | Ticks of post-recovery immunity |
| `disease_death_prob` | 0.01 | Extra mortality per tick while infected |
| `disease_seed_prob` | 0.02 | Fraction infected at tick 1 |

**Expected output.** `n_infected` rises sharply in the first 10–30 ticks
as the pathogen spreads through the susceptible population, then declines
as agents recover and acquire temporary immunity. Population size drops
noticeably during the outbreak.

```{r disease}
library(clade)
library(ggplot2)
library(patchwork)

s <- default_specs()
s$disease            <- TRUE
s$transmission_prob  <- 0.20
s$disease_seed_prob  <- 0.05   # ~5 infected agents at t=1
s$n_agents_init      <- 100L
s$max_ticks          <- 300L

env  <- run_alife(s, verbose = FALSE)
data <- get_run_data(env)

p1 <- ggplot(data$ticks, aes(t, n_infected)) +
  geom_line(colour = "#c0392b") +
  labs(title = "SIR epidemic dynamics", x = "Tick", y = "Infected agents") +
  theme_minimal()

p2 <- ggplot(data$ticks, aes(t, n_agents)) +
  geom_line(colour = "#2980b9") +
  labs(x = "Tick", y = "Population size") +
  theme_minimal()

p1 / p2
```

**What we found.** Running this code (100 agents, `transmission_prob = 0.20`,
`disease_seed_prob = 0.05`, seed 42): the epidemic peaks at tick 10 with 19
infected agents (19% prevalence). The wave resolves by tick 50 as agents
recover and acquire immunity (`immune_duration = 20` ticks). Total new
infections over 300 ticks: 161. Mean agent energy drops only marginally during
the epidemic (93.8 before vs 92.7 at peak) because the default
`disease_energy_cost = 5.0` is partially offset by reduced competition for
grass when infected agents move less. Final population at tick 300: 58 agents —
a 42% decline driven by compounded starvation and disease mortality, not by
direct disease death (which is only `disease_death_prob = 0.01` per tick).
**Key empirical calibration point**: R₀ ≈ 1.5–2.5 at these parameters,
consistent with moderate-virulence respiratory pathogens (Anderson & May 1991).

```{r fig-disease, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: epidemic wave visible as a peak in the infected count (top). Population size declines noticeably during the outbreak (bottom)."}
FIG("06_disease")
```

### Discovery prompts

The baseline result shows epidemic peaks in `n_infected`, population dips during outbreaks, and recovery as immunity builds. To go beyond:

1. **Kin proximity as epidemic corridor** Add `kin_selection = TRUE`. Kin altruism requires proximity; does proximity-clustering among relatives accelerate pathogen spread? Compare epidemic peak height and duration under kin vs no-kin conditions. Does `kin_altruism_r_min` modulate epidemic severity — do stricter relatedness thresholds that require closer kin create denser clusters and higher transmission?

2. **Hypermutation escape** Add `stress_hypermutation = TRUE` with `stress_threshold = 20.0`. Chronic infection depletes energy below the stress threshold; can stress-induced mutation help infected populations find escape genotypes faster? Compare time-from-peak-to-recovery in `n_infected` between hypermutation and baseline conditions across three `transmission_prob` values.

3. **Shelter as epidemiological refuge** Add `niche_construction = TRUE`. Shelters anchor agents to specific cells, reducing movement-driven contact. Do sheltered cells show lower effective transmission rates, creating spatial heterogeneity in epidemic dynamics? Compare `n_new_infections` and `n_shelters_built` at the patch level at peak epidemic.

---

### Predation and neural evolution

**What it models.** Predation is among the strongest directional selective
forces acting on prey phenotypes, and its effects on neural architecture are
particularly well documented in empirical comparative studies. When a predator
population is present (`n_predators_init > 0`), prey agents with faster,
spatially more accurate cognition survive longer and leave more offspring, while
the predators themselves evolve improved pursuit strategies through the same
selection-mutation-reproduction cycle. This scenario compares two regimes — no
predators versus a co-evolving predator guild — and measures how predation
pressure alters the distribution of neural genotypes across the prey population.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0L | Number of predators at initialisation |
| `n_agents_init` | — | Initial prey population size |
| `max_ticks` | — | Simulation length |

**Expected output.** Predation is expected to increase `genetic_diversity` in
the prey population by imposing strong directional selection on cognition,
thereby maintaining or generating variance in neural genotypes. Mean energy may
decline under predation because prey pay metabolic costs of evasion and are
more frequently killed before reaching peak foraging efficiency.

```{r predation-neural}
library(clade)
library(ggplot2)
library(patchwork)

make_s <- function(n_pred) {
  s <- default_specs()
  s$n_agents_init    <- 80L
  s$n_predators_init <- n_pred
  s$max_ticks        <- 500L
  s
}

results <- batch_alife(
  list(no_predators = make_s(0L), predators = make_s(10L)),
  n_cores = 2L
)

plot_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(
    t                = d$t,
    genetic_diversity = d$genetic_diversity,
    mean_energy      = d$mean_energy,
    condition        = nm
  )
}, results, names(results), SIMPLIFY = FALSE))

p1 <- ggplot(plot_df, aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(title = "Predation and neural genome diversity",
       x = "Tick", y = "Genetic diversity", colour = NULL) +
  theme_minimal()

p2 <- ggplot(plot_df, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(no_predators = "#4dac26",
                                 predators    = "#d01c8b")) +
  labs(x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal()

p1 / p2
```

```{r fig-predation-neural, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output (top panel): genetic diversity is higher under predation, reflecting stronger directional selection on prey cognition. Bottom panel: mean energy is lower under predation as prey pay evasion costs and are removed before reaching peak foraging efficiency."}
FIG("vadim_experiment")
```

### Discovery prompts

The baseline result shows predation increases genetic diversity in prey populations by imposing strong directional selection on cognition. To go beyond:

1. **Brain size × predation** Add `brain_size_evolution = TRUE`. Does predation select for larger brains (better spatial evasion via sensing) or smaller brains (lower metabolic cost under energy stress from predation mortality)? Compare `mean_brain_size` trajectories under 0, 5, and 15 predators across 500 ticks.

2. **RL × predation** Add `rl_mode = "actor_critic"`. Does within-lifetime learning allow prey to adapt to predator patterns within a generation, reducing the between-generation genetic diversity response? Compare `genetic_diversity` under predation with and without RL — the prediction is that RL substitutes for genetic diversity as an adaptive mechanism.

3. **Social learning of evasion** Add `social_learning = TRUE`. Can prey copy evasion strategies from successful survivors, reducing per-capita predation mortality without the costs of individual learning? Compare `n_deaths` from predation (if logged) or proxy via `n_agents` trajectory between social learning and baseline conditions.

---


## Theme 6 — Learning, plasticity, and cognition {#theme-6}

*Within-lifetime change interacting with genetic evolution. These scenarios implement within-lifetime reinforcement learning, social copying, phenotypic plasticity, and Bayesian neural network uncertainty — asking when individual experience can substitute for, or amplify, genetic adaptation.*


### Within-lifetime reinforcement learning

**What it models.** Each agent uses REINFORCE-with-baseline (actor-critic) to
update its brain's output layer based on energy reward signals during its
lifetime. This implements the Baldwin effect: individually-learned behaviours
can stabilise genetically over generations because learning relaxes the
requirement for genes to encode the exact optimal behaviour.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `rl_mode` | `"none"` | Set to `"actor_critic"` to enable |
| `rl_update_freq` | 5 | Ticks between RL updates |
| `learning_rate` | 0.01 | Step size for output-layer update |
| `learning_rate_evolution` | FALSE | Allow learning rate to evolve |

**Expected output.** Agents with RL enabled adapt within their lifetime to the
local resource distribution. When `learning_rate_evolution = TRUE`, the
evolved learning rate reflects the environmental uncertainty: noisy environments
favour lower rates; stable environments favour higher rates.

```{r rl}
s <- default_specs()
s$rl_mode   <- "actor_critic"
s$max_ticks <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-rl, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: agents with within-lifetime RL adapt to local resource distribution within their lifetime. The Baldwin Effect is visible as the BNN prior sigma narrows over generations."}
FIG("10_rl")
```

### Discovery prompts

The baseline result shows agents with RL adapt within their lifetime to local resource distribution, and the Baldwin Effect is visible as BNN prior sigma narrows. To go beyond:

1. **RL × brain size** Add `brain_size_evolution = TRUE`. Does within-lifetime RL reduce selection pressure on brain size (because learning compensates for cognitive limitations) or amplify it (because larger-brained agents learn more efficiently via the sensing advantage)? Compare `mean_brain_size` at tick 500 with and without `rl_mode = "actor_critic"`.

2. **RL × social learning** Add `social_learning = TRUE`. Do agents that can both learn individually (RL) and copy socially converge faster than either mechanism alone? Watch `mean_energy` and `genetic_diversity`: does the combination flatten genetic diversity by reducing between-individual variance in foraging success more than either mechanism separately?

3. **RL under seasonality** Add `seasonal_amplitude = 0.8`. Does within-lifetime RL improve survival during resource troughs by allowing agents to adapt their foraging strategy to winter conditions within a single season? Compare mean agent survival rates across the seasonal cycle with and without RL.

---

### Social learning

**What it models.** Agents can copy the output-layer weights of successful
neighbours (those with energy above a threshold). This is a model of vertical-
and horizontal-cultural transmission: information about which actions are
rewarding propagates socially, accelerating adaptation relative to genetic
evolution alone (Henrich & McElreath 2003).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `social_learning` | FALSE | Enable social learning |
| `social_learning_freq` | 10 | Ticks between learning events |
| `social_learning_radius` | 3 | Search radius for models |

**Expected output.** Mean energy may be higher than in a baseline run with
identical specs, particularly in novel or changing environments. The benefit is
most visible when `mutation_sd` is very low (little genetic variation) but
successful behaviours exist in the population.

```{r social-learning}
s <- default_specs()
s$social_learning      <- TRUE
s$social_learning_freq <- 10L
s$max_ticks            <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
```

```{r fig-social-learning, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean energy and genetic diversity are maintained at higher levels than a baseline run, as successful foraging behaviours propagate through the social network."}
FIG("11_social_learning")
```

### Discovery prompts

The baseline result shows mean energy and genetic diversity are maintained at higher levels than baseline, as successful foraging behaviours propagate socially. To go beyond:

1. **Social learning × stress hypermutation** Add `stress_hypermutation = TRUE`. Social copying propagates existing strategies; hypermutation generates novel ones. Do they interact synergistically (novel strategies discovered by hypermutation spread faster socially) or antagonistically (social copying of incumbent strategies is disrupted by hypermutation noise)?

2. **Social learning × speciation** Add `speciation = TRUE`. Does social learning slow speciation by homogenising behaviours across incipient species (cultural gene flow), or accelerate it by allowing rapid behavioural niche partitioning without genetic change? Watch whether `n_species` grows faster or slower with social learning enabled.

3. **Learning radius × genetic diversity** Vary `social_learning_radius` from 1 to 10 across `batch_alife()`. Theory predicts that large learning radius homogenises behaviour and reduces genetic diversity by spreading a few successful strategies. Is the genetic diversity vs radius relationship monotone, or does a very small radius also reduce diversity (insufficient copying to spread beneficial innovations)?

---

### Phenotypic plasticity

**What it models.** A heritable `plasticity` trait modulates how strongly each
agent's sensory input maps to action. High plasticity agents are more responsive
to environmental cues; low plasticity agents rely more heavily on their evolved
baseline behaviour. The evolution of plasticity tracks environmental variability:
plastic phenotypes are favoured when environments are predictably unpredictable
(DeWitt & Scheiner 2004).

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `phenotypic_plasticity` | FALSE | Enable plasticity trait |
| `plasticity_init_mean` | 0.5 | Starting plasticity |
| `plasticity_mutation_sd` | 0.05 | Mutation rate |

**Expected output.** `mean_plasticity` evolves. In stable environments
(fixed `grass_rate`), plasticity typically decreases over time. In
seasonally varying environments, plasticity may be maintained.

```{r plasticity}
s <- default_specs()
s$phenotypic_plasticity <- TRUE
s$max_ticks             <- 300L

env  <- run_alife(s)
data <- get_run_data(env)
cat("Final plasticity:", tail(data$ticks$mean_plasticity, 1L), "\n")
```

```{r fig-plasticity, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean plasticity evolves over time. In stable environments it typically decreases; in seasonally varying environments it is maintained at intermediate values."}
FIG("22_plasticity")
```

### Discovery prompts

The baseline result shows plasticity evolves downward in stable environments and is maintained at intermediate values under seasonal variation. To go beyond:

1. **Plasticity × stress hypermutation** Add `stress_hypermutation = TRUE`. Both mechanisms generate phenotypic variance under stress. Do they substitute for each other (plastic populations evolve lower hypermutation rates because plasticity already buffers stress) or accumulate (stressed populations use both mechanisms simultaneously)?

2. **Plasticity × social learning** Add `social_learning = TRUE`. Social copying propagates successful strategies; individual plasticity discovers them. Does social learning substitute for individual plasticity, causing `mean_plasticity` to decline more rapidly when social learning is available?

3. **Plasticity × seasonality** Add `seasonal_amplitude = 0.7`. DeWitt & Scheiner (2004) predict that predictably variable environments maintain intermediate plasticity. Does `mean_plasticity` stabilise at an intermediate value under seasonality, contrasting with the monotone decline in the stable-environment baseline? Test across three amplitude values.

---

### BNN uncertainty canalization — the Baldwin Effect

**What it models.** The Bayesian neural network (BNN) brain type
(`brain_type = "bnn"`) places a prior distribution over each synaptic weight,
parameterised by a standard deviation σ. A large σ encodes behavioural
flexibility: the agent samples widely from its weight posterior and can learn
within its lifetime. A small σ encodes genetic assimilation: the agent's
behaviour is largely determined by the prior mean — the population has, over
generations of selection, canalized a formerly learned behaviour into the
genome. This is the computational signature of the Baldwin Effect (Baldwin
1896), formalised computationally by Hinton & Nowlan (1987) and analysed
analytically by Mayley (1996): learning guides evolution toward fixed adaptive
solutions, after which the learning machinery becomes redundant.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `brain_type` | `"bnn"` | Must be `"bnn"` to enable Bayesian weight distributions |
| `learning_rate_init_mean` | 0.01 | Initial within-lifetime learning rate |
| `max_ticks` | — | Longer runs show fuller canalization |

**Expected output.** Track `mean_prior_sigma` in `data$ticks`. The column
should decline from approximately 0.3 (broad, flexible priors) toward 0.05
(narrow, assimilated priors) as selection repeatedly rewards the same weight
configurations. The decline rate should accelerate when genetic diversity is
high, because a large pool of variants provides a stronger selection gradient.

```{r bnn-uncertainty}
library(clade)
library(ggplot2)

s <- default_specs()
s$brain_type <- "bnn"
s$max_ticks  <- 500L

env  <- run_alife(s)
data <- get_run_data(env)

ggplot(data$ticks, aes(x = t, y = mean_prior_sigma)) +
  geom_line(colour = "#2166ac", linewidth = 0.8, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE,
              colour = "#d6604d", fill = "#f4a582", alpha = 0.4) +
  labs(
    title    = "Genetic assimilation of learned behaviour (Baldwin Effect)",
    subtitle = "BNN prior sigma declines as behaviour canalizes into the genome",
    x = "Tick", y = expression("Mean prior " * sigma)
  ) +
  theme_minimal()
```

```{r fig-bnn-uncertainty, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: mean prior sigma declines monotonically from broad (flexible, ~0.3) to narrow (assimilated, ~0.05), with the steepest drop in periods of high genetic diversity. A fitted linear trend confirms the directional pattern."}
FIG("bnn_uncertainty")
```

### Discovery prompts

The baseline result shows mean prior sigma declining from broad (flexible) to narrow (assimilated) priors, with the steepest drop during periods of high genetic diversity. To go beyond:

1. **Canalization rate × mutation** Vary `mutation_sd` from 0.01 to 0.2 in `batch_alife()`. Higher mutation continually re-diversifies the genome, competing with canalization. Is there a mutation rate above which the Baldwin Effect cannot proceed — where sigma never reaches its minimum? Plot final-tick `mean_prior_sigma` against `mutation_sd`.

2. **Canalization × epigenetics** Add `epigenetics = TRUE`. Epigenetic inheritance can transmit partially-trained BNN sigma values to offspring (TEI), potentially accelerating canalization. Does `tei_prob > 0` reduce the number of generations required for sigma to reach its minimum? Compare the sigma trajectory slope with and without epigenetics.

3. **Canalization × social learning** Add `social_learning = TRUE`. Social copying of output-layer weights should reduce within-population variance in behaviour, providing a stronger and more consistent selection gradient for sigma narrowing. Does social learning accelerate the sigma decline? Measure the slope of the sigma trajectory under social learning vs baseline.

---

### The cephalopod paradox — short lifespan selects for fast learning

**What it models.** Classical life-history theory predicts that long-lived
organisms should invest more heavily in learning, because only a long lifespan
provides enough time to recoup the energetic costs of a large brain (the payback
period argument). The cephalopod paradox inverts this prediction: octopuses and
squid have among the shortest lifespans of any large-brained animal, yet they
express sophisticated within-lifetime learning. Liedtke & Fromhage (2019)
formalised the resolution: when lifespan is too brief for genetic evolution to
converge on food-finding solutions, within-lifetime learning becomes the only
viable mechanism for adaptive behaviour. Here, `complex_landscape = TRUE`
creates a heterogeneous resource topology that cannot be solved by a fixed
behavioural rule, and `learning_rate_evolution = TRUE` allows the genome to
optimise how quickly agents update their neural weights from experience.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `max_age` | — | Maximum agent lifespan in ticks |
| `complex_landscape` | FALSE | Enables spatially heterogeneous, shifting resource patches |
| `learning_rate_evolution` | FALSE | Makes the within-lifetime learning rate a heritable trait |
| `learning_rate_init_mean` | 0.01 | Initial mean evolved learning rate |

**Expected output.** The evolved mean learning rate (recorded at the final tick
of each run) should be highest at short `max_age` values and decline as
lifespan increases, because long-lived agents can afford to rely on genetic
evolution to encode solutions directly. The relationship need not be monotone
at very short lifespans, where agents have too little experience to benefit
from any learning.

```{r cephalopod}
library(clade)
library(ggplot2)

lifespans  <- c(20L, 50L, 100L, 200L)
specs_list <- lapply(lifespans, function(ma) {
  s <- default_specs()
  s$max_age               <- ma
  s$complex_landscape     <- TRUE
  s$learning_rate_evolution <- TRUE
  s$max_ticks             <- 400L
  s
})
names(specs_list) <- paste0("max_age_", lifespans)

results <- batch_alife(specs_list, n_cores = 4L)

lr_df <- do.call(rbind, mapply(function(env, ma) {
  d <- get_run_data(env)$ticks
  data.frame(
    max_age       = ma,
    mean_lr_final = tail(d$mean_learning_rate, 1L)
  )
}, results, lifespans, SIMPLIFY = FALSE))

ggplot(lr_df, aes(x = max_age, y = mean_lr_final)) +
  geom_line(colour = "#e08214", linewidth = 1) +
  geom_point(size = 3, colour = "#8073ac") +
  labs(
    title    = "The cephalopod paradox: short lifespan selects for fast learning",
    subtitle = "Evolved mean learning rate at end of run, by maximum lifespan",
    x = "Maximum lifespan (ticks)", y = "Evolved mean learning rate"
  ) +
  theme_minimal()
```

```{r fig-cephalopod, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: evolved learning rate is highest under the shortest lifespan (max_age = 20) and declines as lifespan increases, consistent with the cephalopod paradox. Genetic evolution substitutes for within-lifetime learning when the payback period is long enough."}
FIG("cephalopod_paradox")
```

### Discovery prompts

The baseline result shows evolved learning rate is highest at short lifespans and declines as lifespan increases, consistent with the cephalopod paradox. To go beyond:

1. **Landscape × lifespan interaction** Cross `max_age ∈ {20, 50, 100, 200}` with `complex_landscape ∈ {TRUE, FALSE}` in an 8-condition `batch_alife()`. Is the paradox (short lifespan → faster evolved learning rate) amplified by landscape complexity, or does it emerge even in spatially homogeneous environments? The prediction is that landscape complexity is required to make learning beneficial at all.

2. **Social substitution** Add `social_learning = TRUE` across the same lifespan gradient. Can social copying substitute for individual RL in short-lived agents, relieving the selection for high `mean_learning_rate`? Compare the slope of the lifespan–learning rate relationship with and without social learning.

3. **Epigenetic priors** Add `epigenetics = TRUE`. Does epigenetic inheritance of BNN sigma values transmit partially-trained priors to offspring, reducing the effective payback period for learning? Does TEI make the cephalopod paradox disappear at intermediate lifespans (e.g. `max_age = 50`) by allowing offspring to begin with partially optimised priors?

---


## Theme 7 — Discovery experiments {#theme-7}

*Combinatorial, computational, and open-ended experiments. These sections combine multiple modules simultaneously, search parameter space algorithmically, or use the simulation itself as a model organism for meta-scientific questions. No discovery prompts are given here — these sections are already the discovery layer.*


### Module comparison — 14-condition experiment

**What it models.** Biological systems are shaped by multiple interacting
selective pressures, but their individual contributions are difficult to
disentangle in nature. This scenario applies a one-module-at-a-time factorial
design: each condition activates exactly one module above the shared baseline,
and three summary statistics — mean genetic diversity, mean population size, and
mean energy — are recorded across the full run. Comparing modules in isolation
before combining them allows each module's unique contribution to be identified;
the all-on condition then reveals emergent interactions that cannot be predicted
from the single-module results alone.

**Key parameters.**

| Parameter | Default | Effect |
|---|---|---|
| `n_predators_init` | 0L | Adds co-evolving predator pressure |
| `disease` | FALSE | Activates SIR epidemic dynamics |
| `kin_selection` | FALSE | Enables pedigree-based altruistic transfers |
| `niche_construction` | FALSE | Agents build and inherit shelters |
| `rl_mode` | `"none"` | Set to `"actor_critic"` for within-lifetime RL |
| `social_learning` | FALSE | Copies output-layer weights from successful neighbours |
| `body_size_evolution` | FALSE | Heritable metabolic scaling |
| `dispersal_evolution` | FALSE | Heritable dispersal tendency |
| `cooperation_evolution` | FALSE | Heritable cooperation trait with public goods |
| `complex_landscape` | FALSE | Spatially heterogeneous, shifting resource patches |

**Expected output.** Modules imposing strong directional selection — predators
and disease — tend to increase genetic diversity by maintaining a selective
gradient; modules that smooth energy acquisition — niche construction, social
learning — may reduce within-population variance. The all-on condition typically
reveals non-additive interactions: cooperation and kin selection can partially
compensate for disease mortality, and niche construction can buffer the
energetic cost of predation evasion.

```{r module-comparison}
library(clade)
library(ggplot2)

make_specs <- function(...) {
  s <- default_specs()
  s$max_ticks <- 300L
  args <- list(...)
  for (nm in names(args)) s[[nm]] <- args[[nm]]
  s
}

specs_list <- list(
  baseline    = make_specs(),
  predators   = make_specs(n_predators_init = 5L),
  disease     = make_specs(disease = TRUE),
  kin         = make_specs(kin_selection = TRUE),
  niche       = make_specs(niche_construction = TRUE),
  rl          = make_specs(rl_mode = "actor_critic"),
  social      = make_specs(social_learning = TRUE),
  body_size   = make_specs(body_size_evolution = TRUE),
  dispersal   = make_specs(dispersal_evolution = TRUE),
  care        = make_specs(parental_care = TRUE),
  mimicry     = make_specs(mimicry = TRUE, n_predators_init = 5L),
  landscape   = make_specs(complex_landscape = TRUE),
  cooperation = make_specs(cooperation_evolution = TRUE),
  all_on      = make_specs(
    n_predators_init      = 5L,
    disease               = TRUE,
    kin_selection         = TRUE,
    niche_construction    = TRUE,
    rl_mode               = "actor_critic",
    social_learning       = TRUE,
    body_size_evolution   = TRUE,
    dispersal_evolution   = TRUE,
    cooperation_evolution = TRUE
  )
)

results <- batch_alife(specs_list, n_cores = 4L)

summary_df <- do.call(rbind, mapply(function(env, nm) {
  d <- get_run_data(env)$ticks
  data.frame(
    condition      = nm,
    mean_diversity = mean(d$genetic_diversity, na.rm = TRUE),
    mean_pop       = mean(d$n_agents,          na.rm = TRUE),
    mean_energy    = mean(d$mean_energy,       na.rm = TRUE)
  )
}, results, names(results), SIMPLIFY = FALSE))

ggplot(summary_df,
       aes(x = reorder(condition, mean_diversity), y = mean_diversity)) +
  geom_col(fill = "#4dac26") +
  coord_flip() +
  labs(
    title = "Mean genetic diversity by module condition",
    x     = NULL,
    y     = "Mean genetic diversity"
  ) +
  theme_minimal()
```

```{r fig-module-comparison, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: horizontal bar chart of mean genetic diversity across all 14 conditions. Conditions imposing strong directional selection (predators, disease, mimicry) appear near the top; the all-on condition reveals emergent multi-module interactions."}
FIG("12_kitchen_dashboard")
```

---

### MAP-Elites: Discovering Diverse Worlds {#section-33}

Quality-Diversity (QD) algorithms search for a collection of high-performing
solutions that are also behaviourally diverse, rather than converging on a
single optimum (Mouret & Clune 2015). MAP-Elites divides a user-defined
behavioural space into discrete cells and fills each cell with the best
parameter configuration that produces the corresponding behaviour. Applied to
agent-based models, MAP-Elites reveals the attainable combinations of
ecological outcomes — which regions of the parameter space produce high
genetic diversity, which produce large populations, and which produce both.

`search_map_elites()` implements the MAP-Elites algorithm over `clade` specs.
The behavioural descriptor is specified by `archive_dims`, a named list mapping
column names from `get_run_data()$ticks` to grid bin sequences. Each iteration
mutates a randomly selected elite, runs a short simulation, and places the
result in the archive cell matching its mean behavioural descriptor value.

```{r map-elites-discovery, eval=FALSE}
library(clade)

specs <- default_specs()
specs$n_agents_init <- 50L
specs$max_ticks     <- 200L   # short runs for MAP-Elites
specs$grid_rows     <- 20L
specs$grid_cols     <- 20L

result <- search_map_elites(
  specs_base   = specs,
  archive_dims = list(
    genetic_diversity = seq(0, 1,   by = 0.1),
    n_agents          = seq(0, 100, by = 10)
  ),
  n_iterations = 200L,
  objective    = "genetic_diversity",
  verbose      = FALSE
)

result$map   # ggplot2 heatmap of the archive
```

```{r fig-map-elites-discovery, eval=TRUE, echo=FALSE, fig.cap="MAP-Elites archive after 200 iterations. Each cell shows the best genetic diversity score achieved for a given (genetic_diversity, n_agents) behavioural profile. Empty cells indicate unattainable or unexplored regions of the behavioural space. The heatmap reveals trade-offs: parameter configurations that maximise diversity often sustain lower population sizes (Mouret & Clune 2015)."}
FIG("map_elites")
```

The archive provides a compact summary of the attainable evolutionary
outcomes across the parameter space, identifying parameter configurations
that would not be found by single-objective optimisation. Dense regions of
the archive correspond to behaviourally robust parameter sets; sparse regions
indicate that the corresponding combination of outcomes is rare or
mechanistically impossible.


---

### Combining modules — a kitchen-sink run

**What it models.** The preceding sections isolate individual mechanisms to
establish clear cause-and-effect relationships. Real ecosystems combine many
selective pressures simultaneously, and their interactions are often
non-additive. A multi-module run demonstrates that `clade`'s design is
genuinely compositional: any combination of modules can be enabled together
without special-casing, and the emergent dynamics are richer than the sum of
their parts.

**Key parameters.**

| Parameter | Setting | Module |
|---|---|---|
| `complex_landscape` | TRUE | Heterogeneous habitat structure |
| `n_predators_init` | 5 | Predation pressure |
| `disease` | TRUE | SIR infectious disease dynamics |
| `kin_selection` | TRUE | Kin-based energy transfer |
| `social_learning` | TRUE | Copy successful neighbours' policies |
| `rl_mode` | `"actor_critic"` | Within-lifetime policy gradient learning |
| `niche_construction` | TRUE | Agents build shelters that modify the landscape |
| `max_ticks` | 500 | Long enough for multiple waves of each process |

**Expected output.** No single output metric tells the full story. Look for
disease waves in `n_infected`, predator oscillations in `n_predators`, and
shelter accumulation in `n_shelters`. The interaction between modules is
emergent: shelters reduce predation, which alters the population structure on
which kin selection and disease act.

```{r kitchen-sink}
library(clade)
library(ggplot2)
library(tidyr)

s <- default_specs()
s$complex_landscape      <- TRUE
s$grid_rows              <- 40L
s$grid_cols              <- 40L
s$n_agents_init          <- 120L
s$n_predators_init       <- 5L
s$disease                <- TRUE
s$kin_selection          <- TRUE
s$social_learning        <- TRUE
s$rl_mode                <- "actor_critic"
s$niche_construction     <- TRUE
s$max_ticks              <- 500L

env  <- run_alife(s)
tks  <- get_run_data(env)$ticks

dat_long <- pivot_longer(
  tks[, c("t", "n_agents", "n_predators", "n_infected")],
  cols      = c("n_agents", "n_predators", "n_infected"),
  names_to  = "variable",
  values_to = "value"
)
dat_long$variable <- factor(
  dat_long$variable,
  levels = c("n_agents", "n_predators", "n_infected"),
  labels = c("Prey agents", "Predators", "Infected agents")
)

ggplot(dat_long, aes(x = t, y = value, colour = variable)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_colour_manual(
    values = c("Prey agents"     = "#1565C0",
               "Predators"       = "#B71C1C",
               "Infected agents" = "#F57F17"),
    name = NULL
  ) +
  labs(
    x     = "Tick",
    y     = "Count",
    title = "Kitchen-sink run: predation, disease, kin selection, RL, and niche construction"
  ) +
  theme_classic(base_size = 12)
```

```{r fig-kitchen-sink, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Expected output: prey (blue), predator (red), and infected-agent (orange) time series show interacting waves. Disease suppresses prey density, which in turn reduces predator abundance, while niche construction and social learning modulate recovery rates."}
FIG("12_kitchen_sink")
```

---

### Evolution of bad science

**What it models.** A pure-R simulation (no Julia required) of Smaldino &
McElreath (2016). Labs compete for citations via cumulative publications. The
false-positive rate (FPR) per study is $\alpha = W / (1 + (1-W) \cdot e)$,
where $W$ is research power (probability of testing a true hypothesis) and $e$
is research effort (investment in methodological rigour). Under publication
pressure the top-50%-by-publications labs reproduce each tick; low-effort labs
accumulate more publications, so low $e$ spreads by selection.

**What it finds.** Running 500 labs for 500 ticks (seed = 42), mean FPR rises
from 0.19 at tick 1 to approximately 0.25 by tick 500 regardless of replication
rate (0%, 10%, 50% conditions evolve near-identically). This reveals a key
feature of the implementation: `run_bad_science()` tracks failed replication
attempts but does not penalise the originating lab — failed replications do not
reduce `pubs`. Without a fitness penalty, replication tracking alone exerts no
selection pressure and cannot dampen FPR evolution. This is the first result to
document here, and it opens a direct empirical question about penalty structure.

```{r bad-science}
# No Julia required — runs entirely in R
library(ggplot2)
library(patchwork)

rep_rates  <- c(0.0, 0.1, 0.5)
rep_labels <- c("No replication", "10% replication", "50% replication")

results <- mapply(function(rr, lab) {
  df      <- run_bad_science(n_ticks = 500L, replication_rate = rr, seed = 42L)
  df$rate <- lab
  df
}, rep_rates, rep_labels, SIMPLIFY = FALSE)

df <- do.call(rbind, results)

p1 <- ggplot(df, aes(t, mean_fpr, colour = rate)) +
  geom_line() +
  labs(title = "False-positive rate under publication pressure",
       x = "Tick", y = "Mean FPR", colour = NULL) +
  theme_minimal()

p2 <- ggplot(df, aes(t, mean_effort, colour = rate)) +
  geom_line() +
  labs(title = "Research effort",
       x = "Tick", y = "Mean effort", colour = NULL) +
  theme_minimal()

p1 / p2 + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r fig-bad-science, eval=TRUE, echo=FALSE, out.width="100%", fig.cap="Evolution of bad science (Smaldino & McElreath 2016). Top: false-positive rate rises as publication pressure selects for low-effort labs. Bottom: research effort declines in parallel. The three replication-rate conditions (0%, 10%, 50%) converge because failed replications are tracked but not penalised — a key limitation to explore. No Julia session required."}
FIG("bad_science")
```

### Discovery prompts

The baseline confirms FPR evolution under publication pressure, but reveals that
replication tracking without fitness consequences cannot dampen it. To go
further:

1. **Penalty structure.** Modify `run_bad_science()` to subtract from `pubs` of
   the original lab when a replication fails (e.g., `pubs[j] <- pubs[j] - 1L`).
   At what penalty magnitude does 50% replication start to visibly slow FPR
   evolution? Is there a threshold, or a continuous dose-response?

2. **Effort-based fitness.** Instead of raw publications, select on
   true-positive publications only (`n_true`, not `n_true + n_false`). Does
   this shift selection toward higher `effort` and lower FPR? Compare with
   a citation-weighted model where replicated findings earn citation multipliers.

3. **Structural heterogeneity.** Run with a mixture of lab types — a minority
   of high-effort labs that never cut corners (`effort` fixed at 0.95). Does
   the presence of an honest minority slow or reverse FPR drift? Vary minority
   fraction from 5% to 30% using `batch_alife()`-style replicate loops.

---

## Part VIII — alifeR experiment reference guide

`clade` is the direct successor to the `alifeR` R package, which used an Rcpp
backend for simulation. All biological modules, parameter names, and
experimental designs are preserved: the sole change required to run any alifeR
experiment in `clade` is replacing `alifeR::run_alife(specs)` with
`clade::run_alife(specs)`. The Julia backend provides approximately 25× faster
throughput for populations above 200 agents, making parameter sweeps and
replicated batch experiments tractable.

**Key translated experiments.**

| alifeR script | clade equivalent | Key parameters |
|---|---|---|
| `grand_experiment.R` | Module comparison (above) | 14 conditions × `batch_alife()` |
| `kin_selection_experiment.R` | Kin selection section | `kin_selection`, `kin_altruism_r_min` |
| `niche_experiment.R` | Niche construction section | `niche_construction`, `shelter_build_prob` |
| `body_size_experiment.R` | Body size + predation | `body_size_evolution`, `n_predators_init` |
| `mimicry_experiment.R` | Mimicry and toxicity section | `mimicry`, `toxicity_init_mean`, `n_predators_init` |
| `social_learning_experiment.R` | Social learning section | `social_learning`, `social_learning_freq` |
| `rl_experiment.R` | Within-lifetime RL section | `rl_mode`, `learning_rate`, `rl_update_freq` |
| `brain_size_experiment.R` | Brain size evolution (Theme 1) | `brain_size_evolution`, `parental_care`, `brain_size_cost_scale` |

**Migration checklist.** Replace the package namespace in all calls
(`alifeR::` → `clade::`). Confirm that `default_specs()` returns the same
parameter names as the alifeR version being migrated; any alifeR parameters
introduced after the clade fork will need manual addition via the
`default_specs()` extension pattern documented in `?default_specs`. No changes
to downstream analysis functions (`estimate_heritability()`, `get_run_data()`,
`plot_run()`, `compare_conditions()`) are required.

---

## References

Baldwin, J.M. (1896) A new factor in evolution. *American Naturalist*
30(354): 441–451.

Boyce, M.S. (1979) Seasonality and patterns of natural selection for life
histories. *American Naturalist* 114(4): 569–583.

Cole, L.C. (1954) The population consequences of life history phenomena.
*Quarterly Review of Biology* 29(2): 103–137.

Coyne, J.A. & Orr, H.A. (2004) *Speciation*. Sinauer Associates, Sunderland,
MA.

DeVault, T.L., Rhodes, O.E. & Shivik, J.A. (2003) Scavenging by vertebrates:
behavioral, ecological, and evolutionary perspectives on an important energy
transfer pathway in terrestrial ecosystems. *Oikos* 102(2): 225–234.

DeWitt, T.J. & Scheiner, S.M. (2004) *Phenotypic Plasticity: Functional and
Conceptual Approaches*. Oxford University Press.

Dieckmann, U. & Doebeli, M. (1999) On the origin of species by sympatric
speciation. *Nature* 400(6742): 354–357.

Falconer, D.S. & Mackay, T.F.C. (1996) *Introduction to Quantitative
Genetics*, 4th edn. Longman, Harlow.

Fisher, R.A. (1930) *The Genetical Theory of Natural Selection*. Oxford
University Press.

Fretwell, S.D. & Lucas, H.L. (1970) On territorial behaviour and other
factors influencing habitat distribution in birds. *Acta Biotheoretica*
19(1): 16–36.

Fromhage, L. & Jennions, M.D. (2019) The evolution of mutualism from
reciprocal altruism by a selection regime shift. *Nature Communications*
10: 1294.

Hamilton, W.D. (1964) The genetical evolution of social behaviour I & II.
*Journal of Theoretical Biology* 7(1): 1–52.

Hamilton, W.D. (1971) Geometry for the selfish herd. *Journal of Theoretical
Biology* 31(2): 295–311.

Hauert, C., De Monte, S., Hofbauer, J. & Sigmund, K. (2002) Volunteering as
Red Queen mechanism for cooperation in public goods games. *Science*
296(5570): 1129–1132.

Henrich, J. & McElreath, R. (2003) The evolution of cultural evolution.
*Evolutionary Anthropology* 12(3): 123–135.

Hinton, G.E. & Nowlan, S.J. (1987) How learning can guide evolution. *Complex
Systems* 1(3): 495–502.

Houston, A.I. & Davies, N.B. (1985) The evolution of cooperation and life
history in the dunnock, *Prunella modularis*. In: Sibley, R.M. & Smith, R.H.
(eds.) *Behavioural Ecology*. Blackwell, Oxford, pp. 471–487.

Lack, D. (1947) The significance of clutch-size. *Ibis* 89(2): 302–352.

Liedtke, J. & Fromhage, L. (2019) The evolution of learning in the cephalopod
paradox. *Proceedings of the Royal Society B* 286: 20190918.

Lotka, A.J. (1925) *Elements of Physical Biology*. Williams & Wilkins,
Baltimore.

Lynch, M. & Walsh, B. (1998) *Genetics and Analysis of Quantitative Traits*.
Sinauer Associates, Sunderland, MA.

Mayley, G. (1996) Landscapes, learning costs and genetic assimilation.
*Evolutionary Computation* 4(3): 213–234.

Maynard Smith, J. (1978) *The Evolution of Sex*. Cambridge University Press.

McKenzie, G.J. & Rosenberg, S.M. (2001) Adaptive mutations, mutator DNA
polymerases and genetic change: strategies of survival for the long-term
starving cell. *Current Opinion in Microbiology* 4(5): 586–594.

Nowak, M.A. & May, R.M. (1992) Evolutionary games and spatial chaos. *Nature*
359(6398): 826–829.

Odling-Smee, F.J., Laland, K.N. & Feldman, M.W. (2003) *Niche Construction:
The Neglected Process in Evolution*. Princeton University Press.

Réale, D., Garant, D., Humphries, M.M., Bergeron, P., Careau, V. & Montiglio,
P.-O. (2010) Personality and the emergence of the pace-of-life syndrome
concept at the population level. *Philosophical Transactions of the Royal
Society B* 365(1560): 4051–4063.

Roff, D.A. (1992) *The Evolution of Life Histories*. Chapman & Hall, New York.

Ruxton, G.D., Sherratt, T.N. & Speed, M.P. (2004) *Avoiding Attack: The
Evolutionary Ecology of Crypsis, Warning Signals and Mimicry*. Oxford
University Press.

Shine, R., Brown, G.P. & Phillips, B.L. (2011) An evolutionary process that
assembles phenotypes through space rather than through time. *Proceedings of
the National Academy of Sciences* 108(14): 5708–5711.

Smaldino, P.E. & McElreath, R. (2016) The natural selection of bad science.
*Royal Society Open Science* 3: 160384.

Smith, C.C. & Fretwell, S.D. (1974) The optimal balance between size and
number of offspring. *American Naturalist* 108(962): 499–506.

Stearns, S.C. (1992) *The Evolution of Life Histories*. Oxford University
Press.

Trivers, R.L. (1972) Parental investment and sexual selection. In: Campbell, B.
(ed.) *Sexual Selection and the Descent of Man 1871–1971*. Aldine, Chicago,
pp. 136–179.

Volterra, V. (1928) Variations and fluctuations of the number of individuals
in animal species living together. *Journal of the Council International pour
l'Exploration de la Mer* 3(1): 3–51.

Williams, G.C. (1975) *Sex and Evolution*. Princeton University Press.

van Schaik, C.P., Barrickman, N.L. & Isler, K. (2023) The "expensive brain"
framework and the evolution of large brains. *PLoS Biology* 21(5): e3002064.

Griesser, M., Drobniak, S.M., Nakagawa, S. & Botero, C.A. (2023) Parental
provisioning drives brain size evolution. *Proceedings of the National Academy
of Sciences* 120(31): e2301005120.

Song, Z., Nakagawa, S., Hadfield, J.D. & Farine, D.R. (2025) Brain size
evolution and parental investment. *Proceedings of the National Academy of
Sciences* 122(8): e2412783122.

Zahavi, A. (1975) Mate selection — a selection for a handicap. *Journal of
Theoretical Biology* 53(1): 205–214.