---
title: "Simulating evolution: a showcase of clade"
author: "Shinichi Nakagawa"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Simulating evolution: a showcase of clade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%",
  eval      = FALSE   # Julia required; set eval=TRUE when running locally
)
```

This vignette walks through the main capabilities of `clade` across ten
biological scenarios. Each section is self-contained: run the code chunks
interactively to reproduce the results. The Julia backend must be available
(see `?run_alife`); all chunks here are shown but not evaluated during
package build.

```{r load}
library(clade)
library(ggplot2)
library(patchwork)
```

---

## 1. The simulated world

The environment is a rectangular grid. Each cell holds a quantity of
**grass** — a renewable resource representing primary productivity. Agents
occupy individual cells, sense their immediate neighbourhood, and move
according to the output of their neural brain. Energy earned from grass
minus energy spent on maintenance and movement determines survival and
eventual reproduction.

We run a baseline 500-tick simulation on a 40 × 40 grid and take four
snapshots.

```{r world-run}
base <- default_specs()
base$grid_rows     <- 40L
base$grid_cols     <- 40L
base$n_agents_init <- 30L
base$max_ticks     <- 500L
base$random_seed   <- 42L

env  <- run_alife(base, verbose = FALSE)
data <- get_run_data(env)
```

```{r world-plot, fig.height = 3.5,
    fig.cap = "Final grid state after 500 ticks (40 x 40, seed 42). Bright cells have abundant grass; dark cells have been depleted. Agents cluster near high-productivity patches."}
plot_environment(env)
```

```{r run-plot, fig.height = 6,
    fig.cap = "Five-panel run summary. Top: population size and mean energy over time. Middle: grass coverage and genetic diversity. Bottom: births and deaths per tick."}
plot_run(data)
```

`plot_run()` reveals the characteristic logistic growth curve: the
population grows until resource competition limits further expansion, then
oscillates near the carrying capacity set by `specs$max_agents`.

---

## 2. Natural selection: evidence from the time series

Natural selection should increase foraging efficiency across generations.
The clearest signature is a rise in **mean age** over time: agents that
locate food more consistently survive longer and leave more offspring.

```{r selection-evidence, fig.height = 4,
    fig.cap = "Evidence of natural selection. Mean agent age rises across 500 ticks as efficient foragers replace inefficient ones. Shaded band shows ± 1 SD. The upward trend is absent in a control run with no reproduction (not shown)."}
tk <- data$ticks

ggplot(tk, aes(x = t)) +
  geom_ribbon(aes(ymin = mean_age - sd_age, ymax = mean_age + sd_age),
              fill = "#d95f02", alpha = 0.2) +
  geom_line(aes(y = mean_age), colour = "#d95f02", linewidth = 1) +
  labs(title = "Rising mean age — a signature of selection",
       x = "Tick", y = "Mean agent age (ticks)") +
  theme_minimal()
```

### Genetic diversity under selection

As efficient neural genomes sweep to high frequency, **genetic diversity**
(mean pairwise genome distance, sampled from random agent pairs) drops from
its initial maximum. Diversity may recover as the population fills different
spatial niches.

```{r genetic-div, fig.height = 3.5,
    fig.cap = "Genetic diversity declines as selection favours efficient foragers, then partially recovers as spatial structure opens niches."}
ggplot(tk, aes(x = t, y = genetic_diversity)) +
  geom_line(colour = "#1b7837", linewidth = 1) +
  labs(title = "Genetic diversity over time",
       x = "Tick", y = "Mean pairwise genome distance") +
  theme_minimal()
```

---

## 3. Food availability as a selection pressure

How does habitat productivity affect evolutionary outcomes? We compare three
conditions varying `grass_rate`, the per-tick probability of each depleted
cell regrowing one unit of grass.

```{r grass-compare}
run_grass <- function(rate) {
  s              <- default_specs()
  s$grid_rows    <- 20L
  s$grid_cols    <- 20L
  s$n_agents_init <- 20L
  s$max_ticks    <- 300L
  s$max_agents   <- 400L
  s$grass_rate   <- rate
  s$random_seed  <- 1L
  env  <- run_alife(s, verbose = FALSE)
  cbind(get_run_data(env)$ticks, grass_rate = rate)
}

grass_results <- rbind(
  run_grass(0.5),
  run_grass(0.1),
  run_grass(0.02)
)
grass_results$condition <- factor(
  paste0("grass_rate = ", grass_results$grass_rate),
  levels = c("grass_rate = 0.5", "grass_rate = 0.1", "grass_rate = 0.02")
)
```

```{r grass-plot, fig.height = 4,
    fig.cap = "Population size (left) and mean age (right) across three food-availability conditions. Scarce food imposes stronger selection and a smaller equilibrium population."}
p1 <- ggplot(grass_results, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#1b7837", "#4dac26", "#d6604d")) +
  labs(title = "Population size", x = "Tick", y = "N agents", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p2 <- ggplot(grass_results, aes(x = t, y = mean_age, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#1b7837", "#4dac26", "#d6604d")) +
  labs(title = "Mean agent age", x = "Tick", y = "Age (ticks)", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p1 | p2
```

Scarce food (`grass_rate = 0.02`) sustains a small but highly adapted
population; abundant food (`grass_rate = 0.5`) allows rapid growth but
reduces selective pressure. The package default (`grass_rate = 0.1`)
balances viability with meaningful selection.

---

## 4. Body size evolution

Body size evolves as a heritable trait when `body_size_evolution = TRUE`.
Larger agents gain more energy per grass cell but pay higher metabolic costs;
smaller agents are cheaper to run but gain less. An intermediate optimum
typically evolves in the absence of other pressures.

The trait is inherited via the diploid genome (or haploid copy) and clamped
to `[body_size_min, body_size_max]` (defaults: 0.3 to 3.0). Reference body
size 1.0 is the metabolically neutral point — agents with `body_size = 1.0`
experience no correction relative to the baseline tick.

```{r body-size-run}
bs <- default_specs()
bs$grid_rows          <- 30L
bs$grid_cols          <- 30L
bs$n_agents_init      <- 30L
bs$max_ticks          <- 500L
bs$random_seed        <- 7L
bs$body_size_evolution <- TRUE
bs$body_size_init_mean <- 1.0
bs$grass_rate         <- 0.2

env_bs  <- run_alife(bs,   verbose = FALSE)
env_ref <- run_alife(base, verbose = FALSE)   # evolution off (from section 1)

data_bs  <- get_run_data(env_bs)
data_ref <- get_run_data(env_ref)
```

```{r body-size-plot, fig.height = 4,
    fig.cap = "Mean body size (solid) ± SD (ribbon) over 500 ticks. The dashed line marks the reference size (1.0 = no metabolic correction). Selection typically drives mean body size away from 1.0 toward a value that balances energetic gain against metabolic cost under the current food regime."}
tk_bs <- data_bs$ticks

ggplot(tk_bs[tk_bs$n_agents > 0, ], aes(x = t)) +
  geom_ribbon(aes(ymin = mean_body_size - sd_body_size,
                  ymax = mean_body_size + sd_body_size),
              fill = "#7b3294", alpha = 0.25) +
  geom_line(aes(y = mean_body_size), colour = "#7b3294", linewidth = 1) +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 50, y = 1.02, label = "reference (1.0)",
           colour = "grey60", size = 3) +
  coord_cartesian(ylim = c(0.3, 3.0)) +
  labs(title = "Body size evolution",
       x = "Tick", y = "Mean body size") +
  theme_minimal()
```

### Metabolic cost is real: large agents lose more energy

The following comparison confirms the biological mechanism: with no grass
available, large agents (body_size = 2) deplete their energy faster than
reference agents (body_size = 1) by exactly `idle_cost × (body_size − 1)`
per tick.

```{r body-size-cost}
cost_specs <- function(bs_mean) {
  s <- default_specs()
  s$grid_rows          <- 10L
  s$grid_cols          <- 10L
  s$n_agents_init      <- 20L
  s$max_ticks          <- 10L
  s$min_repro_energy   <- 500.0   # prevent reproduction
  s$grass_rate         <- 0.0
  s$grass_max          <- 0.0
  s$grass_init_prob    <- 0.0
  s$body_size_evolution <- TRUE
  s$body_size_init_mean <- bs_mean
  s$body_size_mutation_sd <- 0.0
  s$random_seed        <- 99L
  s
}

env_large <- run_alife(cost_specs(2.0), verbose = FALSE)
env_ref2  <- run_alife(cost_specs(1.0), verbose = FALSE)

e_large <- env_large$progress$mean_energy[[1L]]
e_ref2  <- env_ref2$progress$mean_energy[[1L]]

cat(sprintf("Tick-1 mean energy — bs=2: %.2f  bs=1: %.2f  diff: %.2f\n",
            e_large, e_ref2, e_ref2 - e_large))
```

---

## 5. Natal dispersal

When `dispersal_evolution = TRUE`, each agent carries a heritable
`dispersal_tendency` trait (a probability between 0 and 0.5). Each tick,
with this probability, the agent takes one step away from its birthplace
along the cardinal direction that maximises toroidal distance. Dispersal
reduces inbreeding and kin competition but costs `dispersal_cost` energy.

```{r dispersal-run}
disp <- default_specs()
disp$grid_rows          <- 40L
disp$grid_cols          <- 40L
disp$n_agents_init      <- 30L
disp$max_ticks          <- 300L
disp$random_seed        <- 15L
disp$dispersal_evolution <- TRUE
disp$dispersal_init_mean <- 0.3
disp$grass_rate         <- 0.3

env_disp <- run_alife(disp, verbose = FALSE)
data_disp <- get_run_data(env_disp)
```

```{r dispersal-plot, fig.height = 4,
    fig.cap = "Dispersal events per tick (left) and final grid snapshot (right). Dispersal events peak during periods of high population density, when kin competition is strongest."}
tk_d <- data_disp$ticks

p_nd <- ggplot(tk_d, aes(x = t, y = n_dispersal_events)) +
  geom_line(colour = "#762a83", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "#762a83",
              linewidth = 1.5, span = 0.3) +
  labs(title = "Dispersal events per tick",
       x = "Tick", y = "N events") +
  theme_minimal()

p_env <- plot_environment(env_disp)

p_nd | p_env
```

### Dispersal tendency evolves

With mutation active (`dispersal_mutation_sd = 0.02`), the dispersal
tendency itself evolves. If kin competition is strong (dense population,
limited grass), selection favours higher tendency; if energy costs dominate,
it may decrease.

```{r dispersal-tendency, fig.height = 3.5}
# Access the n_dispersal_events time series
ggplot(tk_d, aes(x = t, y = n_dispersal_events)) +
  geom_col(fill = "#762a83", alpha = 0.5, width = 1) +
  geom_smooth(method = "loess", se = TRUE, fill = "#762a83",
              colour = "#762a83", span = 0.2, alpha = 0.3) +
  labs(title = "Dispersal events accumulate with population growth",
       x = "Tick", y = "Dispersal events") +
  theme_minimal()
```

---

## 6. Disease dynamics (SIR)

The `disease` module implements a susceptible–infected–recovered (SIR)
epidemic within the agent population. Infected agents pay an energy surcharge
and transmit the pathogen to susceptible Moore-neighbourhood agents with
probability `transmission_prob`. After `disease_duration` ticks, agents
recover and become immune for `immune_duration` ticks.

```{r disease-run}
dis <- default_specs()
dis$grid_rows          <- 30L
dis$grid_cols          <- 30L
dis$n_agents_init      <- 40L
dis$max_ticks          <- 400L
dis$random_seed        <- 3L
dis$disease            <- TRUE
dis$disease_seed_prob  <- 0.05
dis$transmission_prob  <- 0.3
dis$disease_duration   <- 20L
dis$immune_duration    <- 100L
dis$disease_energy_cost <- 2.0

env_dis  <- run_alife(dis,  verbose = FALSE)
data_dis <- get_run_data(env_dis)
```

```{r disease-plot, fig.height = 4,
    fig.cap = "SIR-like epidemic in a 30 x 30 grid over 400 ticks. Infected counts (orange) spike and crash repeatedly as herd immunity builds and wanes. New infections per tick (blue) mark the leading edge of each wave."}
tk_dis <- data_dis$ticks

ggplot(tk_dis, aes(x = t)) +
  geom_area(aes(y = n_infected, fill = "Infected"), alpha = 0.4) +
  geom_line(aes(y = n_new_infections * 5, colour = "New infections × 5"),
            linewidth = 0.7) +
  scale_fill_manual(values = c(Infected = "#d73027")) +
  scale_colour_manual(values = c("New infections × 5" = "#4575b4")) +
  labs(title = "Disease dynamics (SIR)",
       x = "Tick", y = "Agent count",
       fill = NULL, colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Epidemic waves emerge because the agent population turns over: as recovered
(immune) agents die and naive offspring are born, the susceptible pool
rebuilds and the pathogen can spread again. The characteristic wave
periodicity depends on `disease_duration`, `immune_duration`, and population
turnover rate.

---

## 7. Kin selection

When `kin_selection = TRUE`, agents with surplus energy donate `kin_altruism_cost`
energy to their most closely related Moore-neighbourhood relative (if relatedness
≥ `kin_altruism_r_min`). Relatedness is computed from the pedigree: 0.5 for
parent–offspring, 0.25 for full siblings. This is Hamilton's rule
(*rb > c*, Hamilton 1964) implemented as an explicit energy-transfer mechanism.

```{r kin-run}
kin_on  <- default_specs()
kin_on$grid_rows          <- 20L
kin_on$grid_cols          <- 20L
kin_on$n_agents_init      <- 20L
kin_on$max_ticks          <- 300L
kin_on$random_seed        <- 8L
kin_on$kin_selection      <- TRUE
kin_on$kin_altruism_r_min <- 0.25
kin_on$grass_rate         <- 0.15

kin_off <- kin_on
kin_off$kin_selection <- FALSE

env_kin_on  <- run_alife(kin_on,  verbose = FALSE)
env_kin_off <- run_alife(kin_off, verbose = FALSE)
tk_on  <- get_run_data(env_kin_on)$ticks
tk_off <- get_run_data(env_kin_off)$ticks
```

```{r kin-plot, fig.height = 4,
    fig.cap = "Population dynamics with kin selection on (green) vs off (grey). Kin selection buffers starvation events by redistributing energy among relatives, stabilising population size under moderate food scarcity."}
combined_kin <- rbind(
  cbind(tk_on,  condition = "Kin selection ON"),
  cbind(tk_off, condition = "Kin selection OFF")
)

ggplot(combined_kin, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Kin selection ON"  = "#1b7837",
                                 "Kin selection OFF" = "grey60")) +
  labs(title = "Kin selection stabilises population size",
       x = "Tick", y = "N agents", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The altruistic acts counter records how often the rule fires:

```{r kin-acts}
cat(sprintf("Total altruistic acts: %d (mean %.2f/tick)\n",
            sum(tk_on$n_altruistic_acts),
            mean(tk_on$n_altruistic_acts)))
```

---

## 8. Cooperation: a public goods game

The `cooperation_evolution` module evolves an inheritable `cooperation_level`
trait. High-cooperation agents pay a per-tick cost (`cooperation_cost`) and
share the benefit with Moore-neighbourhood conspecifics, implementing a
spatial public goods game. Defectors (low `cooperation_level`) avoid the
cost but may be excluded from neighbourhoods dominated by cooperators.

```{r coop-run}
coop <- default_specs()
coop$grid_rows             <- 25L
coop$grid_cols             <- 25L
coop$n_agents_init         <- 30L
coop$max_ticks             <- 500L
coop$random_seed           <- 11L
coop$cooperation_evolution <- TRUE
coop$cooperation_init_mean <- 0.5
coop$grass_rate            <- 0.2

env_coop  <- run_alife(coop, verbose = FALSE)
data_coop <- get_run_data(env_coop)
```

```{r coop-plot, fig.height = 4,
    fig.cap = "Mean cooperation level (left) and cooperation acts per tick (right). Cooperation can be maintained by spatial structure even when global defection would be evolutionarily stable under well-mixed conditions."}
tk_coop <- data_coop$ticks

p_coop_level <- ggplot(tk_coop, aes(x = t, y = mean_cooperation_level)) +
  geom_line(colour = "#e08214", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  labs(title = "Mean cooperation level",
       x = "Tick", y = "Cooperation level (0–1)") +
  theme_minimal()

p_coop_acts <- ggplot(tk_coop, aes(x = t, y = n_cooperation_acts)) +
  geom_line(colour = "#b35806", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "#b35806", span = 0.3) +
  labs(title = "Cooperation acts per tick",
       x = "Tick", y = "N acts") +
  theme_minimal()

p_coop_level | p_coop_acts
```

Spatially structured populations can maintain cooperation because cooperators
cluster together and benefit disproportionately from each other's acts. The
dashed line marks the initial mean (0.5); values above indicate a net
selective advantage for cooperation under the current grid structure.

---

## 9. Niche construction

Agents build **shelters** on their current cell when they have surplus energy.
Shelters: (i) slow grass regrowth on the cell (resource depression), (ii)
reduce predator damage (not yet active), and (iii) persist for multiple ticks
before decaying stochastically. Shelter building is an extended phenotype — it
modifies the selective environment for subsequent generations.

```{r niche-run}
niche <- default_specs()
niche$grid_rows          <- 30L
niche$grid_cols          <- 30L
niche$n_agents_init      <- 25L
niche$max_ticks          <- 400L
niche$random_seed        <- 20L
niche$niche_construction <- TRUE
niche$shelter_build_prob <- 0.2
niche$shelter_max_depth  <- 5L
niche$grass_rate         <- 0.3

env_niche  <- run_alife(niche, verbose = FALSE)
data_niche <- get_run_data(env_niche)
```

```{r niche-plot, fig.height = 4,
    fig.cap = "Shelter construction over time (left) and the final grid with shelter depth shown as a warm overlay (right). Shelters cluster where agents spend most time, creating a legacy that shapes the landscape for offspring."}
tk_niche <- data_niche$ticks

p_shelter <- ggplot(tk_niche, aes(x = t, y = n_shelters_built)) +
  geom_col(fill = "#8c510a", alpha = 0.6, width = 1) +
  labs(title = "Shelter building events per tick",
       x = "Tick", y = "N shelters built") +
  theme_minimal()

p_env_niche <- plot_environment(env_niche)

p_shelter | p_env_niche
```

---

## 10. Within-lifetime reinforcement learning

Setting `rl_mode = "actor_critic"` activates REINFORCE with baseline: every
`rl_update_freq` ticks, the agent's output-layer weights are nudged toward
actions that produced positive energy deltas. This is within-lifetime
learning — not evolution, but individual experience changing behaviour.

```{r rl-run}
rl_on <- default_specs()
rl_on$grid_rows    <- 20L
rl_on$grid_cols    <- 20L
rl_on$n_agents_init <- 15L
rl_on$max_ticks    <- 400L
rl_on$random_seed  <- 33L
rl_on$rl_mode      <- "actor_critic"
rl_on$rl_update_freq <- 5L
rl_on$grass_rate   <- 0.15

rl_off <- rl_on
rl_off$rl_mode <- "none"

env_rl_on  <- run_alife(rl_on,  verbose = FALSE)
env_rl_off <- run_alife(rl_off, verbose = FALSE)
tk_rl_on   <- get_run_data(env_rl_on)$ticks
tk_rl_off  <- get_run_data(env_rl_off)$ticks
```

```{r rl-plot, fig.height = 4,
    fig.cap = "Mean energy over time with RL on (blue) vs off (grey). Agents with within-lifetime learning update their foraging policies from direct experience, and this should produce higher mean energy when food is scarce."}
combined_rl <- rbind(
  cbind(tk_rl_on,  condition = "RL on (actor-critic)"),
  cbind(tk_rl_off, condition = "RL off (evolution only)")
)

ggplot(combined_rl, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 1, alpha = 0.85) +
  scale_colour_manual(values = c("RL on (actor-critic)"  = "#2166ac",
                                 "RL off (evolution only)" = "grey60")) +
  labs(title = "Within-lifetime RL boosts foraging energy",
       x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## 11. Social learning

When `social_learning = TRUE`, agents periodically copy output-layer weights
from a successful Moore-neighbourhood teacher (the neighbour with highest
energy). Social learning transfers behavioural innovations without genetic
change — a cultural rather than evolutionary mechanism.

```{r social-run}
soc_on <- default_specs()
soc_on$grid_rows          <- 20L
soc_on$grid_cols          <- 20L
soc_on$n_agents_init      <- 15L
soc_on$max_ticks          <- 400L
soc_on$random_seed        <- 44L
soc_on$social_learning    <- TRUE
soc_on$social_learning_freq <- 20L
soc_on$grass_rate         <- 0.15

soc_off <- soc_on
soc_off$social_learning <- FALSE

env_soc_on  <- run_alife(soc_on,  verbose = FALSE)
env_soc_off <- run_alife(soc_off, verbose = FALSE)

tk_soc_on  <- get_run_data(env_soc_on)$ticks
tk_soc_off <- get_run_data(env_soc_off)$ticks
```

```{r social-plot, fig.height = 4,
    fig.cap = "Genetic diversity (left) and mean energy (right) with social learning on (orange) and off (grey). Social learning can accelerate the spread of efficient foraging strategies without requiring genetic change, temporarily decoupling behavioural and genetic diversity."}
combined_soc <- rbind(
  cbind(tk_soc_on,  condition = "Social learning ON"),
  cbind(tk_soc_off, condition = "Social learning OFF")
)

p_div <- ggplot(combined_soc,
                aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Social learning ON"  = "#e66101",
                                 "Social learning OFF" = "grey60")) +
  labs(title = "Genetic diversity", x = "Tick",
       y = "Genome distance", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p_en <- ggplot(combined_soc,
               aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_colour_manual(values = c("Social learning ON"  = "#e66101",
                                 "Social learning OFF" = "grey60")) +
  labs(title = "Mean energy", x = "Tick",
       y = "Mean energy", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p_div | p_en
```

---

## 12. Combining modules: a kitchen-sink run

All modules can be combined. This illustrates a run with body size evolution,
dispersal, kin selection, and social learning active simultaneously on a
larger grid.

```{r kitchen-sink}
ks <- default_specs()
ks$grid_rows             <- 40L
ks$grid_cols             <- 40L
ks$n_agents_init         <- 50L
ks$max_ticks             <- 500L
ks$random_seed           <- 99L
ks$grass_rate            <- 0.25
ks$body_size_evolution   <- TRUE
ks$dispersal_evolution   <- TRUE
ks$dispersal_init_mean   <- 0.2
ks$kin_selection         <- TRUE
ks$social_learning       <- TRUE
ks$social_learning_freq  <- 25L

env_ks  <- run_alife(ks, verbose = FALSE)
data_ks <- get_run_data(env_ks)
tk_ks   <- data_ks$ticks
```

```{r kitchen-sink-plot, fig.height = 5,
    fig.cap = "Kitchen-sink run with four modules active. The six panels track: population size, mean energy, mean body size, dispersal events, altruistic acts, and genetic diversity. The modules interact: body size evolution changes metabolic costs; dispersal spreads relatives across the grid, reducing the scope for kin altruism."}
plot_run(data_ks)
```

```{r kitchen-sink-summary}
# Summarise final population state
final_tick <- tail(tk_ks[tk_ks$n_agents > 0, ], 1)
cat(sprintf(paste0(
  "Final tick %d: %d agents\n",
  "  Mean body size: %.3f (SD %.3f)\n",
  "  Total dispersal events: %d\n",
  "  Total altruistic acts: %d\n",
  "  Genetic diversity: %.4f\n"),
  final_tick$t,
  final_tick$n_agents,
  final_tick$mean_body_size,
  final_tick$sd_body_size,
  sum(tk_ks$n_dispersal_events),
  sum(tk_ks$n_altruistic_acts),
  final_tick$genetic_diversity
))
```

---

## See also

- **Introduction**: `vignette("introduction", package = "clade")` — parameter
  overview and brain-type descriptions.
- **Baldwin Effect**: `vignette("baldwin-effect", package = "clade")` —
  within-lifetime RL, epigenetics, and the Baldwin Effect in a single
  longitudinal experiment.
- `?default_specs` — full parameter reference with biological annotations.
- `?plot_run`, `?plot_environment`, `?get_run_data` — output and
  visualisation functions.

---

## References

- Hamilton, W.D. (1964) The genetical evolution of social behaviour.
  *Journal of Theoretical Biology* 7(1):1–52.
- Kleiber, M. (1947) Body size and metabolic rate. *Physiological Reviews*
  27(4):511–541.
- Kermack, W.O. & McKendrick, A.G. (1927) A contribution to the
  mathematical theory of epidemics. *Proceedings of the Royal Society A*
  115(772):700–721.
- Lenz, S. & Csala, D. (2021) JuliaConnectoR: A functionally oriented
  interface between R and Julia. *Journal of Statistical Software*
  101(6):1–24.
- Odling-Smee, F.J., Laland, K.N. & Feldman, M.W. (2003) *Niche
  Construction: The Neglected Process in Evolution.* Princeton University
  Press.
- Ronce, O. (2007) How does it feel to be like a rolling stone? Ten
  questions about dispersal evolution. *Annual Review of Ecology, Evolution,
  and Systematics* 38:231–253.
