---
title: "Simulating evolution: a showcase of clade"
author: "Shinichi Nakagawa"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Simulating evolution: a showcase of clade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%",
  eval      = FALSE   # Julia required; set eval=TRUE to reproduce
)
# Figures are pre-built by vignettes/generate_figures.R and committed to
# man/figures/. Each section shows: (1) the code to produce the run
# (eval=FALSE, visible) followed by (2) the pre-built image (eval=TRUE,
# hidden code).
FIG <- function(name, ...) {
  knitr::include_graphics(
    system.file("figures", paste0("showcase_", name, ".png"),
                package = "clade"),
    ...
  )
}
```

This vignette walks through twelve biological scenarios in `clade`, from a
bare foraging world to kitchen-sink multi-module runs. Each section shows the
code needed to reproduce the result and a pre-built figure so that the
vignette renders without a Julia session.

To reproduce any figure interactively, copy the code chunk and run it in an R
session with Julia available. To regenerate all figures, run:

```r
source(system.file("generate_figures.R", package = "clade"))
```

```{r load}
library(clade)
library(ggplot2)
library(patchwork)
```

---

## 1. The simulated world

The environment is a rectangular grid. Each cell holds a quantity of
**grass** — a renewable resource representing primary productivity. Agents
occupy individual cells, sense their immediate neighbourhood, and move
according to the output of their neural brain. Energy earned from grass minus
energy spent on maintenance determines survival and reproduction.

```{r world-run}
base <- default_specs()
base$grid_rows     <- 40L
base$grid_cols     <- 40L
base$n_agents_init <- 30L
base$max_ticks     <- 500L
base$random_seed   <- 42L

env  <- run_alife(base, verbose = FALSE)
data <- get_run_data(env)
```

```{r world-plot, eval=FALSE}
plot_environment(env)
```

```{r fig-world-grid, eval=TRUE, echo=FALSE, fig.cap="Final grid state after 500 ticks (40 × 40, seed 42). Bright cells have abundant grass; dark cells have been depleted. Agents (white dots, sized by energy) cluster near high-productivity patches."}
FIG("01_world_grid")
```

```{r run-plot, eval=FALSE}
plot_run(data)
```

```{r fig-run-dashboard, eval=TRUE, echo=FALSE, fig.cap="Six-panel run summary: population size, mean energy (±SD ribbon), genetic diversity, births and deaths per tick, grass coverage, and BNN prior sigma (Baldwin Effect panel).", fig.height=6}
FIG("01_run_dashboard")
```

`plot_run()` reveals the characteristic logistic growth curve: the population
grows until resource competition limits further expansion, then stabilises near
the carrying capacity set by `max_agents`. The BNN prior sigma panel tracks the
Baldwin Effect: as agents' genomes encode the foraging solution, the Bayesian
prior narrows.

---

## 2. Natural selection: evidence from the time series

Natural selection increases foraging efficiency across generations. The
clearest signature is a rise in **mean age** over time: agents that locate
food consistently survive longer and leave more offspring.

```{r selection-evidence, eval=FALSE}
tk <- data$ticks

ggplot(tk, aes(x = t)) +
  geom_ribbon(aes(ymin = mean_age - sd_age, ymax = mean_age + sd_age),
              fill = "#d95f02", alpha = 0.2) +
  geom_line(aes(y = mean_age), colour = "#d95f02", linewidth = 1) +
  labs(title = "Rising mean age — a signature of selection",
       x = "Tick", y = "Mean agent age (ticks)") +
  theme_minimal()

ggplot(tk, aes(x = t, y = genetic_diversity)) +
  geom_line(colour = "#1b7837", linewidth = 1) +
  labs(title = "Genetic diversity over time",
       x = "Tick", y = "Mean pairwise genome distance") +
  theme_minimal()
```

```{r fig-selection, eval=TRUE, echo=FALSE, fig.cap="Left: mean age (±SD) rises as efficient foragers replace inefficient ones. Right: genetic diversity declines as the fit genotype sweeps, then partially recovers as spatial structure opens niches."}
FIG("02_selection")
```

---

## 3. Food availability as a selection pressure

We compare three `grass_rate` conditions: abundant (0.5), default (0.1), and
scarce (0.02).

```{r grass-compare, eval=FALSE}
run_grass <- function(rate) {
  s <- default_specs()
  s$grid_rows     <- 20L; s$grid_cols     <- 20L
  s$n_agents_init <- 20L; s$max_ticks     <- 300L
  s$max_agents    <- 400L; s$grass_rate   <- rate
  s$random_seed   <- 1L
  cbind(get_run_data(run_alife(s, verbose = FALSE))$ticks,
        grass_rate = rate)
}

grass_results <- rbind(
  run_grass(0.5), run_grass(0.1), run_grass(0.02)
)
grass_results$condition <- factor(
  paste0("grass_rate = ", grass_results$grass_rate),
  levels = c("grass_rate = 0.5", "grass_rate = 0.1", "grass_rate = 0.02")
)

p1 <- ggplot(grass_results, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#1b7837", "#4dac26", "#d6604d")) +
  labs(title = "Population size", x = "Tick", y = "N agents",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p2 <- ggplot(grass_results, aes(x = t, y = mean_age, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#1b7837", "#4dac26", "#d6604d")) +
  labs(title = "Mean agent age", x = "Tick", y = "Age (ticks)",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p1 | p2
```

```{r fig-food, eval=TRUE, echo=FALSE, fig.cap="Population size (left) and mean age (right) across three food-availability conditions. Scarce food imposes stronger selection (older mean age) but sustains fewer individuals."}
FIG("03_food_scarcity")
```

Scarce food (`grass_rate = 0.02`) sustains a small but highly adapted
population; abundant food (`grass_rate = 0.5`) allows rapid growth but reduces
selective pressure. The package default (`grass_rate = 0.1`) balances viability
with meaningful selection.

---

## 4. Body size evolution

Body size evolves as a heritable trait when `body_size_evolution = TRUE`.
Larger agents gain more energy per grass cell but pay higher metabolic costs;
smaller agents are cheaper to run but gain less. An intermediate optimum
typically evolves (Kleiber 1947).

```{r body-size-run, eval=FALSE}
bs <- default_specs()
bs$grid_rows           <- 30L; bs$grid_cols    <- 30L
bs$n_agents_init       <- 30L; bs$max_ticks    <- 500L
bs$random_seed         <- 7L;  bs$grass_rate   <- 0.2
bs$body_size_evolution <- TRUE
bs$body_size_init_mean <- 1.0

env_bs  <- run_alife(bs, verbose = FALSE)
data_bs <- get_run_data(env_bs)
tk_bs   <- data_bs$ticks
```

```{r body-size-plot, eval=FALSE}
ggplot(tk_bs[tk_bs$n_agents > 0, ], aes(x = t)) +
  geom_ribbon(aes(ymin = mean_body_size - sd_body_size,
                  ymax = mean_body_size + sd_body_size),
              fill = "#7b3294", alpha = 0.25) +
  geom_line(aes(y = mean_body_size), colour = "#7b3294", linewidth = 1) +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 50, y = 1.03, label = "reference (1.0)",
           colour = "grey60", size = 3) +
  coord_cartesian(ylim = c(0.3, 3.0)) +
  labs(title = "Body size evolution",
       x = "Tick", y = "Mean body size") +
  theme_minimal()
```

```{r fig-bodysize, eval=TRUE, echo=FALSE, fig.cap="Mean body size (±1 SD ribbon) over 500 ticks. The dashed line marks the reference size (1.0 = no metabolic correction). Selection typically drives mean body size toward a value balancing foraging gain against metabolic cost."}
FIG("04_body_size")
```

---

## 5. Natal dispersal

When `dispersal_evolution = TRUE`, each agent carries a heritable
`dispersal_tendency` trait (0 to 0.5). Each tick, with this probability, the
agent moves away from its birthplace. Dispersal reduces inbreeding and kin
competition but costs `dispersal_cost` energy (Ronce 2007).

```{r dispersal-run, eval=FALSE}
disp <- default_specs()
disp$grid_rows          <- 40L; disp$grid_cols    <- 40L
disp$n_agents_init      <- 30L; disp$max_ticks    <- 300L
disp$random_seed        <- 15L; disp$grass_rate   <- 0.3
disp$dispersal_evolution <- TRUE
disp$dispersal_init_mean <- 0.3

env_disp  <- run_alife(disp, verbose = FALSE)
data_disp <- get_run_data(env_disp)
```

```{r dispersal-plot, eval=FALSE}
tk_d <- data_disp$ticks

p_nd <- ggplot(tk_d, aes(x = t, y = n_dispersal_events)) +
  geom_col(fill = "#762a83", alpha = 0.5, width = 1) +
  geom_smooth(method = "loess", se = TRUE, fill = "#762a83",
              colour = "#762a83", span = 0.3, alpha = 0.3) +
  labs(title = "Dispersal events per tick",
       x = "Tick", y = "N events") +
  theme_minimal()

p_nd | plot_environment(env_disp)
```

```{r fig-dispersal, eval=TRUE, echo=FALSE, fig.cap="Left: dispersal events per tick (bars) with smoothed trend. Dispersal spikes during high-density periods when kin competition is strongest. Right: final grid snapshot — agents are more evenly distributed than without dispersal."}
FIG("05_dispersal")
```

---

## 6. Disease dynamics (SIR)

The `disease` module implements susceptible–infected–recovered (SIR) epidemic
dynamics. Infected agents pay an energy surcharge and transmit to susceptible
Moore-neighbourhood agents with probability `transmission_prob` (Kermack &
McKendrick 1927).

```{r disease-run, eval=FALSE}
dis <- default_specs()
dis$grid_rows           <- 30L; dis$grid_cols          <- 30L
dis$n_agents_init       <- 40L; dis$max_ticks          <- 400L
dis$random_seed         <- 3L;  dis$disease            <- TRUE
dis$disease_seed_prob   <- 0.05; dis$transmission_prob <- 0.3
dis$disease_duration    <- 20L; dis$immune_duration    <- 100L
dis$disease_energy_cost <- 2.0

env_dis  <- run_alife(dis,  verbose = FALSE)
data_dis <- get_run_data(env_dis)
```

```{r disease-plot, eval=FALSE}
tk_dis <- data_dis$ticks

ggplot(tk_dis, aes(x = t)) +
  geom_area(aes(y = n_infected, fill = "Infected"), alpha = 0.4) +
  geom_line(aes(y = n_new_infections * 5,
                colour = "New infections × 5"), linewidth = 0.7) +
  scale_fill_manual(values = c(Infected = "#d73027")) +
  scale_colour_manual(values = c("New infections × 5" = "#4575b4")) +
  labs(title = "Disease dynamics (SIR)", x = "Tick",
       y = "Agent count", fill = NULL, colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r fig-disease, eval=TRUE, echo=FALSE, fig.cap="SIR epidemic waves over 400 ticks. Infected counts (red area) rise and fall as herd immunity builds. New infections per tick × 5 (blue line) marks the epidemic's leading edge. Waves recur as immune agents die and naive offspring are born."}
FIG("06_disease")
```

---

## 7. Kin selection

When `kin_selection = TRUE`, agents with surplus energy donate energy to their
closest relative in the Moore neighbourhood (if relatedness ≥
`kin_altruism_r_min`). Relatedness is tracked via the pedigree. This
implements Hamilton's rule (*rB > C*; Hamilton 1964).

```{r kin-run, eval=FALSE}
kin_on <- default_specs()
kin_on$grid_rows      <- 20L; kin_on$grid_cols    <- 20L
kin_on$n_agents_init  <- 20L; kin_on$max_ticks    <- 300L
kin_on$random_seed    <- 8L;  kin_on$grass_rate   <- 0.15
kin_on$kin_selection  <- TRUE
kin_on$kin_altruism_r_min <- 0.25

kin_off <- kin_on
kin_off$kin_selection <- FALSE

tk_on  <- get_run_data(run_alife(kin_on,  verbose = FALSE))$ticks
tk_off <- get_run_data(run_alife(kin_off, verbose = FALSE))$ticks

combined_kin <- rbind(cbind(tk_on,  condition = "Kin selection ON"),
                      cbind(tk_off, condition = "Kin selection OFF"))

ggplot(combined_kin, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Kin selection ON"  = "#1b7837",
                                 "Kin selection OFF" = "grey60")) +
  labs(title = "Kin selection stabilises population size",
       x = "Tick", y = "N agents", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r fig-kin, eval=TRUE, echo=FALSE, fig.cap="Population dynamics with kin selection on (green) vs off (grey). Kin altruism buffers starvation events by redistributing energy among relatives, raising the population floor under moderate food scarcity."}
FIG("07_kin_selection")
```

---

## 8. Cooperation: a public goods game

The `cooperation_evolution` module evolves an inheritable `cooperation_level`
trait. Cooperators pay a cost and share the benefit with Moore-neighbourhood
conspecifics. Spatial structure allows cooperators to cluster and benefit
disproportionately from each other.

```{r coop-run, eval=FALSE}
coop <- default_specs()
coop$grid_rows             <- 25L; coop$grid_cols           <- 25L
coop$n_agents_init         <- 30L; coop$max_ticks           <- 500L
coop$random_seed           <- 11L; coop$grass_rate          <- 0.2
coop$cooperation_evolution <- TRUE
coop$cooperation_init_mean <- 0.5

env_coop  <- run_alife(coop, verbose = FALSE)
data_coop <- get_run_data(env_coop)
tk_coop   <- data_coop$ticks

p1 <- ggplot(tk_coop, aes(x = t, y = mean_cooperation_level)) +
  geom_line(colour = "#e08214", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  labs(title = "Mean cooperation level", x = "Tick",
       y = "Cooperation level (0–1)") +
  theme_minimal()

p2 <- ggplot(tk_coop, aes(x = t, y = n_cooperation_acts)) +
  geom_line(colour = "#b35806", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "#b35806", span = 0.3) +
  labs(title = "Cooperation acts per tick", x = "Tick", y = "N acts") +
  theme_minimal()

p1 | p2
```

```{r fig-coop, eval=TRUE, echo=FALSE, fig.cap="Left: mean cooperation level over time (dashed = initial mean 0.5). Values above 0.5 indicate net selective advantage for cooperation under this grid structure. Right: cooperation acts per tick track population growth."}
FIG("08_cooperation")
```

---

## 9. Niche construction

Agents build **shelters** on their current cell when they have surplus energy.
Shelters slow grass regrowth on the cell (resource depression) and persist for
several ticks before decaying stochastically. Shelter building is an extended
phenotype that modifies the selective environment for subsequent generations
(Odling-Smee, Laland & Feldman 2003).

```{r niche-run, eval=FALSE}
niche <- default_specs()
niche$grid_rows          <- 30L; niche$grid_cols     <- 30L
niche$n_agents_init      <- 25L; niche$max_ticks     <- 400L
niche$random_seed        <- 20L; niche$grass_rate    <- 0.3
niche$niche_construction <- TRUE
niche$shelter_build_prob <- 0.2
niche$shelter_max_depth  <- 5L

env_niche  <- run_alife(niche, verbose = FALSE)
data_niche <- get_run_data(env_niche)
tk_niche   <- data_niche$ticks

ggplot(tk_niche, aes(x = t, y = n_shelters_built)) +
  geom_col(fill = "#8c510a", alpha = 0.6, width = 1) +
  labs(title = "Shelter building events per tick",
       x = "Tick", y = "N shelters built") +
  theme_minimal()
```

```{r fig-niche, eval=TRUE, echo=FALSE, fig.cap="Left: shelter building events per tick — construction rates track population density. Right: final grid snapshot. Shelters cluster where agents spend most time, creating a spatial legacy that shapes the landscape for their offspring."}
FIG("09_niche")
```

---

## 10. Within-lifetime reinforcement learning

Setting `rl_mode = "actor_critic"` activates REINFORCE with baseline: every
`rl_update_freq` ticks, the agent's output-layer weights are nudged toward
actions that produced positive energy deltas. This is within-lifetime learning
— not evolution, but individual experience changing behaviour.

```{r rl-run, eval=FALSE}
rl_on <- default_specs()
rl_on$grid_rows     <- 20L; rl_on$grid_cols    <- 20L
rl_on$n_agents_init <- 15L; rl_on$max_ticks    <- 400L
rl_on$random_seed   <- 33L; rl_on$grass_rate   <- 0.15
rl_on$rl_mode       <- "actor_critic"
rl_on$rl_update_freq <- 5L

rl_off <- rl_on
rl_off$rl_mode <- "none"

combined_rl <- rbind(
  cbind(get_run_data(run_alife(rl_on,  verbose = FALSE))$ticks,
        condition = "RL on (actor-critic)"),
  cbind(get_run_data(run_alife(rl_off, verbose = FALSE))$ticks,
        condition = "RL off (evolution only)")
)

ggplot(combined_rl, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 1, alpha = 0.85) +
  scale_colour_manual(
    values = c("RL on (actor-critic)"    = "#2166ac",
               "RL off (evolution only)" = "grey60")) +
  labs(title = "Within-lifetime RL boosts foraging energy",
       x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r fig-rl, eval=TRUE, echo=FALSE, fig.cap="Mean energy over time with within-lifetime RL on (blue) vs off (grey). Agents with actor-critic learning update their foraging policies from direct experience, typically yielding higher mean energy than purely evolutionary agents under the same food regime."}
FIG("10_rl")
```

---

## 11. Social learning

When `social_learning = TRUE`, agents periodically copy output-layer weights
from a successful Moore-neighbourhood teacher (the neighbour with highest
energy). Social learning transfers behavioural innovations without genetic
change — a cultural rather than evolutionary mechanism.

```{r social-run, eval=FALSE}
soc_on <- default_specs()
soc_on$grid_rows          <- 20L; soc_on$grid_cols    <- 20L
soc_on$n_agents_init      <- 15L; soc_on$max_ticks    <- 400L
soc_on$random_seed        <- 44L; soc_on$grass_rate   <- 0.15
soc_on$social_learning    <- TRUE
soc_on$social_learning_freq <- 20L

soc_off <- soc_on
soc_off$social_learning <- FALSE

combined_soc <- rbind(
  cbind(get_run_data(run_alife(soc_on,  verbose = FALSE))$ticks,
        condition = "Social learning ON"),
  cbind(get_run_data(run_alife(soc_off, verbose = FALSE))$ticks,
        condition = "Social learning OFF")
)
soc_cols <- c("Social learning ON" = "#e66101",
              "Social learning OFF" = "grey60")

p_div2 <- ggplot(combined_soc,
                 aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 1) + scale_colour_manual(values = soc_cols) +
  labs(title = "Genetic diversity", x = "Tick",
       y = "Genome distance", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p_en2 <- ggplot(combined_soc,
                aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 1, alpha = 0.8) + scale_colour_manual(values = soc_cols) +
  labs(title = "Mean energy", x = "Tick",
       y = "Mean energy", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p_div2 | p_en2
```

```{r fig-social, eval=TRUE, echo=FALSE, fig.cap="Social learning (orange) vs no learning (grey). Left: social learning spreads efficient strategies, temporarily reducing genetic diversity below the purely evolutionary baseline. Right: mean energy is higher when learning accelerates the cultural spread of good foraging policies."}
FIG("11_social_learning")
```

---

## 12. Combining modules: a kitchen-sink run

All modules can be combined. This run activates body size evolution,
dispersal, kin selection, and social learning simultaneously on a larger grid.

```{r kitchen-sink, eval=FALSE}
ks <- default_specs()
ks$grid_rows            <- 40L; ks$grid_cols          <- 40L
ks$n_agents_init        <- 50L; ks$max_ticks          <- 500L
ks$random_seed          <- 99L; ks$grass_rate         <- 0.25
ks$body_size_evolution  <- TRUE; ks$dispersal_evolution <- TRUE
ks$dispersal_init_mean  <- 0.2;  ks$kin_selection      <- TRUE
ks$social_learning      <- TRUE; ks$social_learning_freq <- 25L

env_ks  <- run_alife(ks, verbose = FALSE)
data_ks <- get_run_data(env_ks)

plot_run(data_ks)
```

```{r fig-ks-run, eval=TRUE, echo=FALSE, fig.cap="Six-panel run summary for the kitchen-sink configuration (body size + dispersal + kin selection + social learning). The body-size panel (bottom-right) replaces the BNN sigma panel from Section 1.", fig.height=6}
FIG("12_kitchen_sink")
```

```{r kitchen-sink-dashboard, eval=FALSE}
visualize_progress(env_ks, data_ks)
```

```{r fig-ks-dash, eval=TRUE, echo=FALSE, fig.cap="Full dashboard for the kitchen-sink run: grid snapshot, population and energy dynamics, diversity trajectory, death scatter (age vs energy by cause), lifespan histogram, and body-size evolution.", fig.height=7}
FIG("12_kitchen_dashboard")
```

---

## Generating the figures yourself

If Julia is available, you can regenerate all figures from source:

```r
source(system.file("generate_figures.R", package = "clade"))
```

or, from the package source directory:

```r
source("vignettes/generate_figures.R")
```

This creates `man/figures/showcase_*.png`. Re-build the vignette with
`devtools::build_vignettes()` to embed the updated images.

---

## See also

- **Introduction**: `vignette("introduction", package = "clade")`
- `?default_specs` — full parameter reference with biological annotations.
- `?plot_run`, `?visualize_progress`, `?plot_environment` — output functions.

---

## References

- Hamilton, W.D. (1964) The genetical evolution of social behaviour.
  *Journal of Theoretical Biology* 7(1):1–52.
- Kermack, W.O. & McKendrick, A.G. (1927) A contribution to the mathematical
  theory of epidemics. *Proceedings of the Royal Society A* 115(772):700–721.
- Kleiber, M. (1947) Body size and metabolic rate.
  *Physiological Reviews* 27(4):511–541.
- Odling-Smee, F.J., Laland, K.N. & Feldman, M.W. (2003) *Niche Construction:
  The Neglected Process in Evolution.* Princeton University Press.
- Ronce, O. (2007) How does it feel to be like a rolling stone? Ten questions
  about dispersal evolution. *Annual Review of Ecology, Evolution, and
  Systematics* 38:231–253.
