---
title: "Simulating evolution: a showcase of clade"
author: "Shinichi Nakagawa"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Simulating evolution: a showcase of clade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%",
  eval      = FALSE   # Julia required; set eval=TRUE to reproduce
)
# Figures are pre-built by vignettes/generate_figures.R and committed to
# man/figures/. Each section shows: (1) the code to produce the run
# (eval=FALSE, visible) followed by (2) the pre-built image (eval=TRUE,
# hidden code).
FIG <- function(name, ...) {
  knitr::include_graphics(
    system.file("figures", paste0("showcase_", name, ".png"),
                package = "clade"),
    ...
  )
}
```

This vignette walks through twelve biological scenarios in `clade`, from a
bare foraging world to kitchen-sink multi-module runs. Each section shows the
code needed to reproduce the result and a pre-built figure so that the
vignette renders without a Julia session.

To reproduce any figure interactively, copy the code chunk and run it in an R
session with Julia available. To regenerate all figures, run:

```r
source(system.file("generate_figures.R", package = "clade"))
```

```{r load}
library(clade)
library(ggplot2)
library(patchwork)
```

---

## 1. The simulated world

The environment is a rectangular grid. Each cell holds a quantity of
**grass** — a renewable resource representing primary productivity. Agents
occupy individual cells, sense their immediate neighbourhood, and move
according to the output of their neural brain. Energy earned from grass minus
energy spent on maintenance determines survival and reproduction.

```{r world-run}
base <- default_specs()
base$grid_rows     <- 40L
base$grid_cols     <- 40L
base$n_agents_init <- 30L
base$max_ticks     <- 500L
base$random_seed   <- 42L

env  <- run_alife(base, verbose = FALSE)
data <- get_run_data(env)
```

```{r world-plot, eval=FALSE}
plot_environment(env)
```

```{r fig-world-grid, eval=TRUE, echo=FALSE, fig.cap="Final grid state after 500 ticks (40 × 40, seed 42). Bright cells have abundant grass; dark cells have been depleted. Agents (white dots, sized by energy) cluster near high-productivity patches."}
FIG("01_world_grid")
```

```{r run-plot, eval=FALSE}
plot_run(data)
```

```{r fig-run-dashboard, eval=TRUE, echo=FALSE, fig.cap="Six-panel run summary: population size, mean energy (±SD ribbon), genetic diversity, births and deaths per tick, grass coverage, and BNN prior sigma (Baldwin Effect panel).", fig.height=6}
FIG("01_run_dashboard")
```

`plot_run()` reveals the characteristic logistic growth curve: the population
grows until resource competition limits further expansion, then stabilises near
the carrying capacity set by `max_agents`. The BNN prior sigma panel tracks the
Baldwin Effect: as agents' genomes encode the foraging solution, the Bayesian
prior narrows.

---

## 2. Natural selection: evidence from the time series

Natural selection increases foraging efficiency across generations. The
clearest signature is a rise in **mean age** over time: agents that locate
food consistently survive longer and leave more offspring.

```{r selection-evidence, eval=FALSE}
tk <- data$ticks

ggplot(tk, aes(x = t)) +
  geom_ribbon(aes(ymin = mean_age - sd_age, ymax = mean_age + sd_age),
              fill = "#d95f02", alpha = 0.2) +
  geom_line(aes(y = mean_age), colour = "#d95f02", linewidth = 1) +
  labs(title = "Rising mean age — a signature of selection",
       x = "Tick", y = "Mean agent age (ticks)") +
  theme_minimal()

ggplot(tk, aes(x = t, y = genetic_diversity)) +
  geom_line(colour = "#1b7837", linewidth = 1) +
  labs(title = "Genetic diversity over time",
       x = "Tick", y = "Mean pairwise genome distance") +
  theme_minimal()
```

```{r fig-selection, eval=TRUE, echo=FALSE, fig.cap="Left: mean age (±SD) rises as efficient foragers replace inefficient ones. Right: genetic diversity declines as the fit genotype sweeps, then partially recovers as spatial structure opens niches."}
FIG("02_selection")
```

---

## 3. Food availability as a selection pressure

We compare three `grass_rate` conditions: abundant (0.5), default (0.1), and
scarce (0.02).

```{r grass-compare, eval=FALSE}
run_grass <- function(rate) {
  s <- default_specs()
  s$grid_rows     <- 20L; s$grid_cols     <- 20L
  s$n_agents_init <- 20L; s$max_ticks     <- 300L
  s$max_agents    <- 400L; s$grass_rate   <- rate
  s$random_seed   <- 1L
  cbind(get_run_data(run_alife(s, verbose = FALSE))$ticks,
        grass_rate = rate)
}

grass_results <- rbind(
  run_grass(0.5), run_grass(0.1), run_grass(0.02)
)
grass_results$condition <- factor(
  paste0("grass_rate = ", grass_results$grass_rate),
  levels = c("grass_rate = 0.5", "grass_rate = 0.1", "grass_rate = 0.02")
)

p1 <- ggplot(grass_results, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#1b7837", "#4dac26", "#d6604d")) +
  labs(title = "Population size", x = "Tick", y = "N agents",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p2 <- ggplot(grass_results, aes(x = t, y = mean_age, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#1b7837", "#4dac26", "#d6604d")) +
  labs(title = "Mean agent age", x = "Tick", y = "Age (ticks)",
       colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p1 | p2
```

```{r fig-food, eval=TRUE, echo=FALSE, fig.cap="Population size (left) and mean age (right) across three food-availability conditions. Scarce food imposes stronger selection (older mean age) but sustains fewer individuals."}
FIG("03_food_scarcity")
```

Scarce food (`grass_rate = 0.02`) sustains a small but highly adapted
population; abundant food (`grass_rate = 0.5`) allows rapid growth but reduces
selective pressure. The package default (`grass_rate = 0.1`) balances viability
with meaningful selection.

---

## 4. Body size evolution

Body size evolves as a heritable trait when `body_size_evolution = TRUE`.
Larger agents gain more energy per grass cell but pay higher metabolic costs;
smaller agents are cheaper to run but gain less. An intermediate optimum
typically evolves (Kleiber 1947).

```{r body-size-run, eval=FALSE}
bs <- default_specs()
bs$grid_rows           <- 30L; bs$grid_cols    <- 30L
bs$n_agents_init       <- 30L; bs$max_ticks    <- 500L
bs$random_seed         <- 7L;  bs$grass_rate   <- 0.2
bs$body_size_evolution <- TRUE
bs$body_size_init_mean <- 1.0

env_bs  <- run_alife(bs, verbose = FALSE)
data_bs <- get_run_data(env_bs)
tk_bs   <- data_bs$ticks
```

```{r body-size-plot, eval=FALSE}
ggplot(tk_bs[tk_bs$n_agents > 0, ], aes(x = t)) +
  geom_ribbon(aes(ymin = mean_body_size - sd_body_size,
                  ymax = mean_body_size + sd_body_size),
              fill = "#7b3294", alpha = 0.25) +
  geom_line(aes(y = mean_body_size), colour = "#7b3294", linewidth = 1) +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 50, y = 1.03, label = "reference (1.0)",
           colour = "grey60", size = 3) +
  coord_cartesian(ylim = c(0.3, 3.0)) +
  labs(title = "Body size evolution",
       x = "Tick", y = "Mean body size") +
  theme_minimal()
```

```{r fig-bodysize, eval=TRUE, echo=FALSE, fig.cap="Mean body size (±1 SD ribbon) over 500 ticks. The dashed line marks the reference size (1.0 = no metabolic correction). Selection typically drives mean body size toward a value balancing foraging gain against metabolic cost."}
FIG("04_body_size")
```

---

## 5. Natal dispersal

When `dispersal_evolution = TRUE`, each agent carries a heritable
`dispersal_tendency` trait (0 to 0.5). Each tick, with this probability, the
agent moves away from its birthplace. Dispersal reduces inbreeding and kin
competition but costs `dispersal_cost` energy (Ronce 2007).

```{r dispersal-run, eval=FALSE}
disp <- default_specs()
disp$grid_rows          <- 40L; disp$grid_cols    <- 40L
disp$n_agents_init      <- 30L; disp$max_ticks    <- 300L
disp$random_seed        <- 15L; disp$grass_rate   <- 0.3
disp$dispersal_evolution <- TRUE
disp$dispersal_init_mean <- 0.3

env_disp  <- run_alife(disp, verbose = FALSE)
data_disp <- get_run_data(env_disp)
```

```{r dispersal-plot, eval=FALSE}
tk_d <- data_disp$ticks

p_nd <- ggplot(tk_d, aes(x = t, y = n_dispersal_events)) +
  geom_col(fill = "#762a83", alpha = 0.5, width = 1) +
  geom_smooth(method = "loess", se = TRUE, fill = "#762a83",
              colour = "#762a83", span = 0.3, alpha = 0.3) +
  labs(title = "Dispersal events per tick",
       x = "Tick", y = "N events") +
  theme_minimal()

p_nd | plot_environment(env_disp)
```

```{r fig-dispersal, eval=TRUE, echo=FALSE, fig.cap="Left: dispersal events per tick (bars) with smoothed trend. Dispersal spikes during high-density periods when kin competition is strongest. Right: final grid snapshot — agents are more evenly distributed than without dispersal."}
FIG("05_dispersal")
```

---

## 6. Disease dynamics (SIR)

The `disease` module implements susceptible–infected–recovered (SIR) epidemic
dynamics. Infected agents pay an energy surcharge and transmit to susceptible
Moore-neighbourhood agents with probability `transmission_prob` (Kermack &
McKendrick 1927).

```{r disease-run, eval=FALSE}
dis <- default_specs()
dis$grid_rows           <- 30L; dis$grid_cols          <- 30L
dis$n_agents_init       <- 40L; dis$max_ticks          <- 400L
dis$random_seed         <- 3L;  dis$disease            <- TRUE
dis$disease_seed_prob   <- 0.05; dis$transmission_prob <- 0.3
dis$disease_duration    <- 20L; dis$immune_duration    <- 100L
dis$disease_energy_cost <- 2.0

env_dis  <- run_alife(dis,  verbose = FALSE)
data_dis <- get_run_data(env_dis)
```

```{r disease-plot, eval=FALSE}
tk_dis <- data_dis$ticks

ggplot(tk_dis, aes(x = t)) +
  geom_area(aes(y = n_infected, fill = "Infected"), alpha = 0.4) +
  geom_line(aes(y = n_new_infections * 5,
                colour = "New infections × 5"), linewidth = 0.7) +
  scale_fill_manual(values = c(Infected = "#d73027")) +
  scale_colour_manual(values = c("New infections × 5" = "#4575b4")) +
  labs(title = "Disease dynamics (SIR)", x = "Tick",
       y = "Agent count", fill = NULL, colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r fig-disease, eval=TRUE, echo=FALSE, fig.cap="SIR epidemic waves over 400 ticks. Infected counts (red area) rise and fall as herd immunity builds. New infections per tick × 5 (blue line) marks the epidemic's leading edge. Waves recur as immune agents die and naive offspring are born."}
FIG("06_disease")
```

---

## 7. Kin selection

When `kin_selection = TRUE`, agents with surplus energy donate energy to their
closest relative in the Moore neighbourhood (if relatedness ≥
`kin_altruism_r_min`). Relatedness is tracked via the pedigree. This
implements Hamilton's rule (*rB > C*; Hamilton 1964).

```{r kin-run, eval=FALSE}
kin_on <- default_specs()
kin_on$grid_rows      <- 20L; kin_on$grid_cols    <- 20L
kin_on$n_agents_init  <- 20L; kin_on$max_ticks    <- 300L
kin_on$random_seed    <- 8L;  kin_on$grass_rate   <- 0.15
kin_on$kin_selection  <- TRUE
kin_on$kin_altruism_r_min <- 0.25

kin_off <- kin_on
kin_off$kin_selection <- FALSE

tk_on  <- get_run_data(run_alife(kin_on,  verbose = FALSE))$ticks
tk_off <- get_run_data(run_alife(kin_off, verbose = FALSE))$ticks

combined_kin <- rbind(cbind(tk_on,  condition = "Kin selection ON"),
                      cbind(tk_off, condition = "Kin selection OFF"))

ggplot(combined_kin, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("Kin selection ON"  = "#1b7837",
                                 "Kin selection OFF" = "grey60")) +
  labs(title = "Kin selection stabilises population size",
       x = "Tick", y = "N agents", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r fig-kin, eval=TRUE, echo=FALSE, fig.cap="Population dynamics with kin selection on (green) vs off (grey). Kin altruism buffers starvation events by redistributing energy among relatives, raising the population floor under moderate food scarcity."}
FIG("07_kin_selection")
```

---

## 8. Cooperation: a public goods game

The `cooperation_evolution` module evolves an inheritable `cooperation_level`
trait. Cooperators pay a cost and share the benefit with Moore-neighbourhood
conspecifics. Spatial structure allows cooperators to cluster and benefit
disproportionately from each other.

```{r coop-run, eval=FALSE}
coop <- default_specs()
coop$grid_rows             <- 25L; coop$grid_cols           <- 25L
coop$n_agents_init         <- 30L; coop$max_ticks           <- 500L
coop$random_seed           <- 11L; coop$grass_rate          <- 0.2
coop$cooperation_evolution <- TRUE
coop$cooperation_init_mean <- 0.5

env_coop  <- run_alife(coop, verbose = FALSE)
data_coop <- get_run_data(env_coop)
tk_coop   <- data_coop$ticks

p1 <- ggplot(tk_coop, aes(x = t, y = mean_cooperation_level)) +
  geom_line(colour = "#e08214", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  labs(title = "Mean cooperation level", x = "Tick",
       y = "Cooperation level (0–1)") +
  theme_minimal()

p2 <- ggplot(tk_coop, aes(x = t, y = n_cooperation_acts)) +
  geom_line(colour = "#b35806", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "#b35806", span = 0.3) +
  labs(title = "Cooperation acts per tick", x = "Tick", y = "N acts") +
  theme_minimal()

p1 | p2
```

```{r fig-coop, eval=TRUE, echo=FALSE, fig.cap="Left: mean cooperation level over time (dashed = initial mean 0.5). Values above 0.5 indicate net selective advantage for cooperation under this grid structure. Right: cooperation acts per tick track population growth."}
FIG("08_cooperation")
```

---

## 9. Niche construction

Agents build **shelters** on their current cell when they have surplus energy.
Shelters slow grass regrowth on the cell (resource depression) and persist for
several ticks before decaying stochastically. Shelter building is an extended
phenotype that modifies the selective environment for subsequent generations
(Odling-Smee, Laland & Feldman 2003).

```{r niche-run, eval=FALSE}
niche <- default_specs()
niche$grid_rows          <- 30L; niche$grid_cols     <- 30L
niche$n_agents_init      <- 25L; niche$max_ticks     <- 400L
niche$random_seed        <- 20L; niche$grass_rate    <- 0.3
niche$niche_construction <- TRUE
niche$shelter_build_prob <- 0.2
niche$shelter_max_depth  <- 5L

env_niche  <- run_alife(niche, verbose = FALSE)
data_niche <- get_run_data(env_niche)
tk_niche   <- data_niche$ticks

ggplot(tk_niche, aes(x = t, y = n_shelters_built)) +
  geom_col(fill = "#8c510a", alpha = 0.6, width = 1) +
  labs(title = "Shelter building events per tick",
       x = "Tick", y = "N shelters built") +
  theme_minimal()
```

```{r fig-niche, eval=TRUE, echo=FALSE, fig.cap="Left: shelter building events per tick — construction rates track population density. Right: final grid snapshot. Shelters cluster where agents spend most time, creating a spatial legacy that shapes the landscape for their offspring."}
FIG("09_niche")
```

---

## 10. Within-lifetime reinforcement learning

Setting `rl_mode = "actor_critic"` activates REINFORCE with baseline: every
`rl_update_freq` ticks, the agent's output-layer weights are nudged toward
actions that produced positive energy deltas. This is within-lifetime learning
— not evolution, but individual experience changing behaviour.

```{r rl-run, eval=FALSE}
rl_on <- default_specs()
rl_on$grid_rows     <- 20L; rl_on$grid_cols    <- 20L
rl_on$n_agents_init <- 15L; rl_on$max_ticks    <- 400L
rl_on$random_seed   <- 33L; rl_on$grass_rate   <- 0.15
rl_on$rl_mode       <- "actor_critic"
rl_on$rl_update_freq <- 5L

rl_off <- rl_on
rl_off$rl_mode <- "none"

combined_rl <- rbind(
  cbind(get_run_data(run_alife(rl_on,  verbose = FALSE))$ticks,
        condition = "RL on (actor-critic)"),
  cbind(get_run_data(run_alife(rl_off, verbose = FALSE))$ticks,
        condition = "RL off (evolution only)")
)

ggplot(combined_rl, aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 1, alpha = 0.85) +
  scale_colour_manual(
    values = c("RL on (actor-critic)"    = "#2166ac",
               "RL off (evolution only)" = "grey60")) +
  labs(title = "Within-lifetime RL boosts foraging energy",
       x = "Tick", y = "Mean energy", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r fig-rl, eval=TRUE, echo=FALSE, fig.cap="Mean energy over time with within-lifetime RL on (blue) vs off (grey). Agents with actor-critic learning update their foraging policies from direct experience, typically yielding higher mean energy than purely evolutionary agents under the same food regime."}
FIG("10_rl")
```

---

## 11. Social learning

When `social_learning = TRUE`, agents periodically copy output-layer weights
from a successful Moore-neighbourhood teacher (the neighbour with highest
energy). Social learning transfers behavioural innovations without genetic
change — a cultural rather than evolutionary mechanism.

```{r social-run, eval=FALSE}
soc_on <- default_specs()
soc_on$grid_rows          <- 20L; soc_on$grid_cols    <- 20L
soc_on$n_agents_init      <- 15L; soc_on$max_ticks    <- 400L
soc_on$random_seed        <- 44L; soc_on$grass_rate   <- 0.15
soc_on$social_learning    <- TRUE
soc_on$social_learning_freq <- 20L

soc_off <- soc_on
soc_off$social_learning <- FALSE

combined_soc <- rbind(
  cbind(get_run_data(run_alife(soc_on,  verbose = FALSE))$ticks,
        condition = "Social learning ON"),
  cbind(get_run_data(run_alife(soc_off, verbose = FALSE))$ticks,
        condition = "Social learning OFF")
)
soc_cols <- c("Social learning ON" = "#e66101",
              "Social learning OFF" = "grey60")

p_div2 <- ggplot(combined_soc,
                 aes(x = t, y = genetic_diversity, colour = condition)) +
  geom_line(linewidth = 1) + scale_colour_manual(values = soc_cols) +
  labs(title = "Genetic diversity", x = "Tick",
       y = "Genome distance", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p_en2 <- ggplot(combined_soc,
                aes(x = t, y = mean_energy, colour = condition)) +
  geom_line(linewidth = 1, alpha = 0.8) + scale_colour_manual(values = soc_cols) +
  labs(title = "Mean energy", x = "Tick",
       y = "Mean energy", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p_div2 | p_en2
```

```{r fig-social, eval=TRUE, echo=FALSE, fig.cap="Social learning (orange) vs no learning (grey). Left: social learning spreads efficient strategies, temporarily reducing genetic diversity below the purely evolutionary baseline. Right: mean energy is higher when learning accelerates the cultural spread of good foraging policies."}
FIG("11_social_learning")
```

---

## 12. Combining modules: a kitchen-sink run

All modules can be combined. This run activates body size evolution,
dispersal, kin selection, and social learning simultaneously on a larger grid.

```{r kitchen-sink, eval=FALSE}
ks <- default_specs()
ks$grid_rows            <- 40L; ks$grid_cols          <- 40L
ks$n_agents_init        <- 50L; ks$max_ticks          <- 500L
ks$random_seed          <- 99L; ks$grass_rate         <- 0.25
ks$body_size_evolution  <- TRUE; ks$dispersal_evolution <- TRUE
ks$dispersal_init_mean  <- 0.2;  ks$kin_selection      <- TRUE
ks$social_learning      <- TRUE; ks$social_learning_freq <- 25L

env_ks  <- run_alife(ks, verbose = FALSE)
data_ks <- get_run_data(env_ks)

plot_run(data_ks)
```

```{r fig-ks-run, eval=TRUE, echo=FALSE, fig.cap="Six-panel run summary for the kitchen-sink configuration (body size + dispersal + kin selection + social learning). The body-size panel (bottom-right) replaces the BNN sigma panel from Section 1.", fig.height=6}
FIG("12_kitchen_sink")
```

```{r kitchen-sink-dashboard, eval=FALSE}
visualize_progress(env_ks, data_ks)
```

```{r fig-ks-dash, eval=TRUE, echo=FALSE, fig.cap="Full dashboard for the kitchen-sink run: grid snapshot, population and energy dynamics, diversity trajectory, death scatter (age vs energy by cause), lifespan histogram, and body-size evolution.", fig.height=7}
FIG("12_kitchen_dashboard")
```

---

## 13. Disease and Immunity Dynamics {#section-13}

Pathogens impose powerful selection on host populations. When `disease = TRUE`
the simulation runs a susceptible–infected–recovered (SIR) model; when
`immune_evolution = TRUE` each agent carries a heritable `immune_strength`
trait that scales recovery speed and transmission resistance. Over time,
selection should drive immune investment upward in populations under sustained
pathogen pressure (Boots & Bowers 2004).

```{r sec-13-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$disease           <- TRUE
specs$immune_evolution  <- TRUE
specs$n_agents_init     <- 100L
specs$max_ticks         <- 300L
specs$disease_seed_prob <- 0.05
specs$transmission_prob <- 0.3
specs$disease_duration  <- 20L
specs$immune_duration   <- 80L
specs$random_seed       <- 13L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)
plot_run(rd)
```

```{r sec-13-fig, eval=TRUE, echo=FALSE, fig.cap="Population and immune dynamics under evolving immunity. The n_infected panel shows SIR epidemic waves declining in amplitude as mean immune_strength rises across generations under continued pathogen pressure."}
FIG("13_disease")
```

---

## 14. Predator-Prey Dynamics {#section-14}

Predators are mobile heterotrophs that consume agents. The classic
Lotka-Volterra expectation — asynchronous oscillations in prey and predator
density — emerges from spatial agent interactions without any deliberately
engineered cycle (Lotka 1925). Predator populations collapse when prey density
falls below the level needed to sustain maintenance costs.

```{r sec-14-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$n_predators_init  <- 5L
specs$n_agents_init     <- 100L
specs$max_ticks         <- 300L
specs$grid_rows         <- 30L
specs$grid_cols         <- 30L
specs$grass_rate        <- 0.3
specs$random_seed       <- 14L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)
plot_run(rd)
```

```{r sec-14-fig, eval=TRUE, echo=FALSE, fig.cap="Dual time series of prey (n_agents) and predator (n_predators) population size over 300 ticks. Asynchronous boom-bust cycles emerge from local predator-prey interactions on the spatial grid."}
FIG("14_predators")
```

---

## 15. Group Defense (Dilution of Risk) {#section-15}

When `group_defense = TRUE`, the per-agent attack probability from a predator
scales inversely with local group size — the dilution-of-risk effect (Hamilton
1971). Agents in dense clusters are individually safer, creating positive
fitness feedback for aggregation even when aggregation competes for grass.

```{r sec-15-code, eval=FALSE}
library(clade)

run_defense <- function(defense) {
  s <- default_specs()
  s$n_predators_init <- 5L
  s$group_defense    <- defense
  s$n_agents_init    <- 100L
  s$max_ticks        <- 300L
  s$grid_rows        <- 30L
  s$grid_cols        <- 30L
  s$grass_rate       <- 0.3
  s$random_seed      <- 15L
  cbind(get_run_data(run_alife(s, verbose = FALSE))$ticks,
        group_defense = defense)
}

results <- rbind(run_defense(TRUE), run_defense(FALSE))
results$condition <- ifelse(results$group_defense,
                            "Group defense ON", "Group defense OFF")

ggplot(results, aes(x = t, y = n_agents, colour = condition)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    values = c("Group defense ON" = "#2166ac", "Group defense OFF" = "#d73027")
  ) +
  labs(title = "Dilution of risk: group defense vs no defense",
       x = "Tick", y = "N agents", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r sec-15-fig, eval=TRUE, echo=FALSE, fig.cap="Prey population size over time with group defense on (blue) and off (red) under identical predator pressure. Dilution of risk reduces per-capita predation and typically sustains higher population densities."}
FIG("15_group_defense")
```

---

## 16. Habitat Preference and Ideal Free Distribution {#section-16}

When `habitat_preference_evolution = TRUE`, agents carry a heritable
`habitat_preference` trait that biases movement toward high-quality cells. The
ideal free distribution (IFD) predicts that, at evolutionary equilibrium, all
habitat patches yield equal fitness (Fretwell & Lucas 1970). Deviations from
IFD appear in early generations before selection has had time to tune preference.

```{r sec-16-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$habitat_preference_evolution <- TRUE
specs$n_agents_init                <- 80L
specs$max_ticks                    <- 300L
specs$grid_rows                    <- 30L
specs$grid_cols                    <- 30L
specs$grass_rate                   <- 0.2
specs$random_seed                  <- 16L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)

plot_dispersal_events(rd)
```

```{r sec-16-fig, eval=TRUE, echo=FALSE, fig.cap="Habitat preference evolution and dispersal events over 300 ticks. As habitat preference traits evolve, agents increasingly concentrate on productive patches, and dispersal event rates reflect the changing spatial distribution of resources and competitors."}
FIG("16_habitat_preference")
```

---

## 17. Seasonal Dynamics {#section-17}

Many real environments oscillate between productive and lean seasons.
`seasonal_amplitude` scales the within-year grass growth variation; agents
that survive winter (`winter_death_prob`) must persist on stored energy until
spring. Population size therefore tracks the seasonal calendar, and life-history
traits are expected to align with the seasonal cycle (Dingle 1996).

```{r sec-17-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$seasonal_amplitude <- 0.6
specs$season_length      <- 100L
specs$winter_death_prob  <- 0.02
specs$n_agents_init      <- 80L
specs$max_ticks          <- 400L
specs$grid_rows          <- 30L
specs$grid_cols          <- 30L
specs$random_seed        <- 17L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)
plot_run(rd)
```

```{r sec-17-fig, eval=TRUE, echo=FALSE, fig.cap="Population and resource dynamics under seasonal forcing (amplitude = 0.6, season_length = 100). Population size oscillates in phase with the seasonal grass cycle, with winter mortality troughs visible every 100 ticks."}
FIG("17_seasons")
```

---

## 18. Speciation and Genetic Divergence {#section-18}

When `speciation = TRUE`, agents that share fewer than `isolation_threshold`
neural-weight similarity with the general population form a reproductively
isolated lineage. Over time, spatially segregated clusters may diverge
sufficiently to constitute distinct species, increasing `n_species` (Gavrilets
2004).

```{r sec-18-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$speciation         <- TRUE
specs$isolation_threshold <- 0.4
specs$n_agents_init      <- 100L
specs$max_ticks          <- 500L
specs$grid_rows          <- 40L
specs$grid_cols          <- 40L
specs$grass_rate         <- 0.25
specs$random_seed        <- 18L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)

tk <- rd$ticks

p1 <- ggplot(tk, aes(x = t, y = n_species)) +
  geom_step(colour = "#5e3c99", linewidth = 1) +
  labs(title = "Number of species over time",
       x = "Tick", y = "N species") +
  theme_minimal()

p2 <- ggplot(tk, aes(x = t, y = genetic_diversity)) +
  geom_line(colour = "#e66101", linewidth = 1) +
  labs(title = "Genetic diversity",
       x = "Tick", y = "Mean pairwise genome distance") +
  theme_minimal()

p1 | p2
```

```{r sec-18-fig, eval=TRUE, echo=FALSE, fig.cap="Left: cumulative number of species over 500 ticks — each step marks a new lineage passing the isolation threshold. Right: genetic diversity rises as diverging lineages accumulate sequence differences."}
FIG("18_speciation")
```

---

## 19. Parental Care and Life History {#section-19}

Parental care (`parental_care = TRUE`) allows parents to shelter and feed
juveniles. This imposes a direct energetic cost on the parent (`care_cost_per_tick`)
while increasing juvenile survival. The result is a shift toward slower life
histories — longer gestation, higher offspring quality, lower turnover — a
trade-off well documented across vertebrates (Clutton-Brock 1991).

```{r sec-19-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$parental_care      <- TRUE
specs$care_cost_per_tick <- 1.5
specs$feeding_rate       <- 8.0
specs$n_agents_init      <- 60L
specs$max_ticks          <- 300L
specs$grid_rows          <- 30L
specs$grid_cols          <- 30L
specs$grass_rate         <- 0.3
specs$random_seed        <- 19L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)

tk <- rd$ticks

p1 <- ggplot(tk, aes(x = t)) +
  geom_line(aes(y = n_agents,    colour = "Adults"),   linewidth = 1) +
  geom_line(aes(y = n_juveniles, colour = "Juveniles"), linewidth = 1) +
  scale_colour_manual(
    values = c(Adults = "#1b7837", Juveniles = "#d95f02")) +
  labs(title = "Adults and juveniles over time",
       x = "Tick", y = "Count", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p2 <- plot_dead_agents(rd)

p1 | p2
```

```{r sec-19-fig, eval=TRUE, echo=FALSE, fig.cap="Left: adult and juvenile population counts over time — juveniles track adult breeding pulses. Right: lifespan distribution from plot_dead_agents(), showing that care-reared offspring reach later ages compared to no-care baselines."}
FIG("19_parental_care")
```

---

## 20. Cooperative Breeding {#section-20}

Cooperative breeding (`cooperative_breeding = TRUE`) allows non-breeding helpers
to assist at the nest of kin. Helpers pay an energetic cost but increase the
breeding success of relatives, gaining inclusive fitness when relatedness
exceeds `helper_kin_threshold`. The evolution of helping is therefore an
extension of Hamilton's rule operating at the family group level (Cockburn 1998).

```{r sec-20-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$cooperative_breeding <- TRUE
specs$parental_care        <- TRUE
specs$helper_kin_threshold <- 0.25
specs$n_agents_init        <- 60L
specs$max_ticks            <- 300L
specs$grid_rows            <- 30L
specs$grid_cols            <- 30L
specs$grass_rate           <- 0.3
specs$random_seed          <- 20L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)
tk  <- rd$ticks

p1 <- ggplot(tk, aes(x = t, y = mean_helper_tendency)) +
  geom_line(colour = "#762a83", linewidth = 1) +
  labs(title = "Evolution of helping tendency",
       x = "Tick", y = "Mean helper tendency (0-1)") +
  theme_minimal()

p2 <- ggplot(tk, aes(x = t, y = n_helpers)) +
  geom_col(fill = "#9970ab", alpha = 0.6, width = 1) +
  labs(title = "Active helpers per tick",
       x = "Tick", y = "N helpers") +
  theme_minimal()

p1 | p2
```

```{r sec-20-fig, eval=TRUE, echo=FALSE, fig.cap="Left: mean helper tendency (0-1) rises over time as alleles favouring kin-directed helping spread. Right: active helpers per tick track the rise of the helping trait and population density."}
FIG("20_cooperative_breeding")
```

---

## 21. Mimicry and Toxicity {#section-21}

The `mimicry` module evolves a heritable `toxicity` trait (0–1). Toxic prey
damage predators that attack them; predators learn to associate warning signals
with toxicity via Rescorla-Wagner updating of `signal_memory`. Over time,
selection drives toxicity upward while predator avoidance co-evolves, producing
the classic Mullerian dynamics between defended and undefended forms (Mallet &
Joron 1999).

```{r sec-21-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$mimicry              <- TRUE
specs$n_predators_init     <- 5L
specs$toxicity_init_mean   <- 0.2
specs$n_agents_init        <- 100L
specs$max_ticks            <- 300L
specs$grid_rows            <- 30L
specs$grid_cols            <- 30L
specs$grass_rate           <- 0.3
specs$random_seed          <- 21L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)
tk  <- rd$ticks

p1 <- ggplot(tk, aes(x = t, y = mean_toxicity)) +
  geom_line(colour = "#d73027", linewidth = 1) +
  geom_hline(yintercept = 0.2, linetype = "dashed", colour = "grey60") +
  labs(title = "Evolution of toxicity",
       x = "Tick", y = "Mean toxicity (0-1)") +
  theme_minimal()

p2 <- ggplot(tk, aes(x = t)) +
  geom_line(aes(y = n_toxic_attacks,   colour = "Attacks on toxic prey"),
            linewidth = 0.9) +
  geom_line(aes(y = n_avoided_attacks, colour = "Avoided attacks"),
            linewidth = 0.9, linetype = "dashed") +
  scale_colour_manual(
    values = c("Attacks on toxic prey" = "#d73027",
               "Avoided attacks"       = "#4575b4")) +
  labs(title = "Predator attack behaviour",
       x = "Tick", y = "N events per tick", colour = NULL) +
  theme_minimal() + theme(legend.position = "bottom")

p1 | p2
```

```{r sec-21-fig, eval=TRUE, echo=FALSE, fig.cap="Left: mean toxicity rises from its initial value (0.2, dashed) as costly undefended genotypes are purged. Right: attacks on toxic prey decline and avoided attacks rise as predator signal memory accumulates, illustrating the coevolution of defence and avoidance."}
FIG("21_mimicry")
```

---

## 22. Phenotypic Plasticity {#section-22}

Phenotypic plasticity (`phenotypic_plasticity = TRUE`) allows agents to adjust
expressed phenotype in response to local environmental conditions without genetic
change. A heritable `plasticity` trait (0–1) scales the sensitivity of the
expressed phenotype to sensory input. High plasticity is advantageous in
variable environments but may be costly when the environment is stable enough
for a fixed genotype to suffice (DeWitt, Scheiner & Wolpert 1998).

```{r sec-22-code, eval=FALSE}
library(clade)
specs <- default_specs()
specs$phenotypic_plasticity <- TRUE
specs$plasticity_init_mean  <- 0.3
specs$n_agents_init         <- 80L
specs$max_ticks             <- 300L
specs$grid_rows             <- 30L
specs$grid_cols             <- 30L
specs$grass_rate            <- 0.2
specs$seasonal_amplitude    <- 0.4
specs$season_length         <- 60L
specs$random_seed           <- 22L

env <- run_alife(specs, verbose = FALSE)
rd  <- get_run_data(env)
tk  <- rd$ticks

ggplot(tk, aes(x = t)) +
  geom_ribbon(aes(ymin = mean_plasticity - sd_plasticity,
                  ymax = mean_plasticity + sd_plasticity),
              fill = "#4dac26", alpha = 0.25) +
  geom_line(aes(y = mean_plasticity), colour = "#4dac26", linewidth = 1) +
  geom_hline(yintercept = 0.3, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 20, y = 0.32,
           label = "initial mean (0.3)", colour = "grey50", size = 3) +
  labs(title = "Evolution of phenotypic plasticity",
       subtitle = "Seasonal environment (amplitude = 0.4, period = 60 ticks)",
       x = "Tick", y = "Mean plasticity (0-1)") +
  theme_minimal()
```

```{r sec-22-fig, eval=TRUE, echo=FALSE, fig.cap="Mean plasticity (±1 SD ribbon) over 300 ticks under moderate seasonal forcing. The dashed line marks the initial mean (0.3). Selection in a variable environment is expected to drive plasticity upward until the cost of maintaining plastic machinery equals the benefit of phenotypic adjustment."}
FIG("22_plasticity")
```

---

## Generating the figures yourself

If Julia is available, you can regenerate all figures from source:

```r
source(system.file("generate_figures.R", package = "clade"))
```

or, from the package source directory:

```r
source("vignettes/generate_figures.R")
```

This creates `man/figures/showcase_*.png`. Re-build the vignette with
`devtools::build_vignettes()` to embed the updated images.

---

## See also

- **Introduction**: `vignette("introduction", package = "clade")`
- `?default_specs` — full parameter reference with biological annotations.
- `?plot_run`, `?visualize_progress`, `?plot_environment` — output functions.

---

## References

- Hamilton, W.D. (1964) The genetical evolution of social behaviour.
  *Journal of Theoretical Biology* 7(1):1–52.
- Kermack, W.O. & McKendrick, A.G. (1927) A contribution to the mathematical
  theory of epidemics. *Proceedings of the Royal Society A* 115(772):700–721.
- Kleiber, M. (1947) Body size and metabolic rate.
  *Physiological Reviews* 27(4):511–541.
- Odling-Smee, F.J., Laland, K.N. & Feldman, M.W. (2003) *Niche Construction:
  The Neglected Process in Evolution.* Princeton University Press.
- Ronce, O. (2007) How does it feel to be like a rolling stone? Ten questions
  about dispersal evolution. *Annual Review of Ecology, Evolution, and
  Systematics* 38:231–253.
- Boots, M. & Bowers, R.G. (2004) The evolution of resistance through costly
  acquired immunity. *Proceedings of the Royal Society B* 271:715–723.
- Clutton-Brock, T.H. (1991) *The Evolution of Parental Care.* Princeton
  University Press.
- Cockburn, A. (1998) Evolution of helping behavior in cooperatively breeding
  birds. *Annual Review of Ecology and Systematics* 29:141–177.
- DeWitt, T.J., Scheiner, S.M. & Wolpert, D. (1998) Costs and limits of
  phenotypic plasticity. *Trends in Ecology and Evolution* 13(2):77–81.
- Dingle, H. (1996) *Migration: The Biology of Life on the Move.* Oxford
  University Press.
- Fretwell, S.D. & Lucas, H.L. (1970) On territorial behavior and other factors
  influencing habitat distribution in birds. *Acta Biotheoretica* 19:16–36.
- Gavrilets, S. (2004) *Fitness Landscapes and the Origin of Species.* Princeton
  University Press.
- Hamilton, W.D. (1971) Geometry for the selfish herd.
  *Journal of Theoretical Biology* 31(2):295–311.
- Lotka, A.J. (1925) *Elements of Physical Biology.* Williams & Wilkins.
- Mallet, J. & Joron, M. (1999) Evolution of diversity in warning color and
  mimicry: polymorphisms, shifting balance, and speciation.
  *Annual Review of Ecology and Systematics* 30:201–233.
