---
title: "Troubleshooting and FAQ"
author: "Shinichi Nakagawa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting and FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

*Common problems and their fixes. Organised by the symptom you're
likely to see. If your issue isn't covered here, please [open an
issue](https://github.com/itchyshin/clade/issues).*

---

## Julia and installation

### `julia_is_ready()` hangs or returns `FALSE`

The first call to `julia_is_ready()` compiles the Julia kernel. This
takes **60–90 seconds** on a typical machine. Subsequent calls in the
same R session are ~instant.

If it still returns `FALSE` after 2+ minutes:

1. Verify Julia is installed: `system("julia --version")` in R.
2. Verify `JuliaConnectoR` can find it: `JuliaConnectoR::juliaSetupOk()`.
3. On macOS, Julia installed via Homebrew can have PATH issues. Install
   via [juliaup](https://github.com/JuliaLang/juliaup) instead —
   `curl -fsSL https://install.julialang.org | sh`.
4. If `JULIA_HOME` is set to an old installation, unset it:
   `Sys.unsetenv("JULIA_HOME")` then restart R.

### "Package Clade not found" or similar Julia-side errors

The Julia environment lives in `inst/julia/`. If clade was recently
updated (e.g. you pulled new kernel changes), rebuild the Julia
environment:

```bash
cd ~/.julia/packages/  # or wherever your R library has clade installed
cd $(R -e 'cat(system.file("julia", package = "clade"))' -s)
julia --project=. -e 'using Pkg; Pkg.instantiate()'
```

### First-compile is extremely slow (>5 minutes)

On first compile, Julia pre-compiles all kernel modules. This is
cached per-Julia-environment. If you see >5 minute compiles every
time, something is invalidating the cache — often a package-manager
update or a stale `Manifest.toml`. Clean with:

```julia
using Pkg
Pkg.resolve()
Pkg.instantiate()
```

---

## "My population crashed" — viability failures

### Symptom: final `n_agents` near zero

Your simulation ran but the population went extinct. Almost always
one of:

- **Mortality too high**. Check `death_prob`, `senescence_rate`,
  `winter_death_prob`. If multiple mortality sources are on, they
  compound.
- **Food too scarce**. Check `grass_rate` (per-tick regrowth) and
  `grass_max`. If `grass_rate` is below ~0.1 the world can starve.
- **Too few initial agents** combined with high stochasticity. Try
  `n_agents_init = 40L` minimum for a 30×30 grid.
- **Reproduction threshold too high** relative to foraging rate.
  Check `min_repro_energy` vs. what an agent can accumulate in its
  lifetime.

**Always use `viability_report()` before interpreting a result.**
Extinct populations have no signal to report:

```{r}
rd  <- get_run_data(env)
via <- viability_report(rd)
print(via)   # populations survived? heritability non-zero?
```

### Symptom: populations survive but show `n_agents` oscillating wildly

Usually **Allee effects + small grid**. Two fixes:

1. Use `realistic_specs()` or `ultra_realistic_specs()` — these
   scale the grid and initial population to reduce Allee stochasticity.
2. Relax `min_repro_energy` slightly, or increase `max_ticks` so the
   population has time to recover between crashes.

---

## Parallel runs and `batch_alife()`

### "All connections are in use" or hanging mclapply

**Never use `parallel::mclapply()` with clade.** `mclapply` forks the
R process; JuliaConnectoR doesn't survive a fork and deadlocks.
`batch_alife()` uses PSOCK by default — that's the only safe backend.

### PSOCK workers don't pick up kernel changes

PSOCK workers use the *installed* clade package, not your
`devtools::load_all()` development tree. If you changed Julia kernel
files in `inst/julia/` and tried a PSOCK batch, you'll see stale
behaviour.

Fix: reinstall before launching:

```r
devtools::install(quick = TRUE)   # then restart R and the batch job
```

### "All 64 connections are in use" on a big machine

PSOCK has per-machine limits on the number of parallel R connections.
On this project's machine we cap at `n_cores = 64` even when more
physical cores exist. If you need more, either:

1. Split the job across multiple R sessions manually.
2. Use the `future` package with a multisession plan (more robust
   than raw PSOCK, but slightly slower startup).

---

## "Direction right, magnitude wrong"

### What it means

The audit ledger uses three tiers:

- ✅ **PASS** — direction correct, **|t| ≥ 2** at the tested number of seeds
- 🟠 **Passed-consistent** — direction correct but **|t| < 2**
- 🔴 **Contradicts** — direction opposite to theory

If your result is 🟠, usually one of these is going on:

1. **Noise dominates**: more seeds reveal a real signal. Scale from
   8 → 16 → 32 seeds before concluding anything.
2. **Kernel-limited**: the mechanism you need isn't in clade at
   sufficient strength. Check whether the module has a "strength"
   parameter that can be cranked up (e.g. `habitat_preference_strength`,
   `parasite_pressure`).
3. **Honest null**: theory is wrong at this parameter regime. Document
   it. The audit ledger has several honest nulls — they're not
   failures, they're findings.

### "Signal vanishes at realistic scale"

This is a real phenomenon, not a bug. Some signals (e.g. Hamilton
1971 selfish-herd) depend on population size via 1/√N and literally
become smaller at larger scales. Check the audit notes for the
relevant scenario — they flag where scale matters.

---

## Spec conflicts

### "`kin_selection` does nothing" / "`cooperation` does nothing"

Many modules require other modules to be on. Common dependencies:

- **`kin_selection`** needs `ploidy = 2` (diploid). It's meaningless
  under haploid reproduction.
- **`mate_choice`** needs `signal_mating` or `signal_evolution_drift`
  to have something to choose on.
- **`parental_care`** needs `neonatal_foraging_deficit > 0` to have
  a demographic reason to care.
- **`coevolving_parasites`** needs `signal_dims > 0` for continuous-
  trait matching, or `n_parasite_loci > 0` for Hamming-haplotype.
- **`cooperative_breeding`** requires `cooperation_evolution = TRUE`
  plus a delayed-dispersal mechanism (e.g. saturated habitat).

The [Parameter reference](parameter-reference.html) documents these
dependencies per flag.

### "`stress_hypermutation` produces no signal"

`stress_threshold` must be **above** `min_repro_energy` or the module
is structurally silent — agents never reach a stress state before
reproducing. The default is set conservatively. For stress-scarcity
studies, try `stress_threshold = 150` with default `min_repro_energy
= 120`.

---

## RNG / reproducibility

### "My runs don't reproduce even with `random_seed`"

clade's kernel uses `random_seed` but does NOT fully reset its RNG
between `run_alife()` calls within one R session. Consecutive runs
share state. If you need bit-exact reproducibility:

1. Restart R between runs, or
2. Use `batch_alife()` with explicit `seeds = ...` — each worker gets
   a fresh Julia environment.

### Figure scripts don't reproduce the audit

The figure-generation scripts in `vignettes/` and the audit scripts
in `dev/audit/fidelity/` must call `run_alife()` in the **same order**
to reproduce the exact same population. If you refactor call order,
expect seed-level mismatch even with identical `random_seed`.

---

## Getting help

If your problem isn't here:

1. Check the [Scenarios](scenarios.html) index — your question may
   match an audited scenario.
2. Check the [Kernel as biology](k-README.html) articles for
   mechanism-level documentation.
3. Skim [open issues](https://github.com/itchyshin/clade/issues) —
   someone may have hit the same wall.
4. [Open a new issue](https://github.com/itchyshin/clade/issues/new)
   with:
   - `sessionInfo()` output
   - The specs list you ran (redact anything sensitive)
   - The error message / unexpected behaviour
   - What you expected vs. what you got

---

## Anti-patterns to avoid

- **Single-seed conclusions.** Never report a result from one seed.
  8 minimum for pilots, 16+ for claims.
- **Deleting the `viability_report()` check** because "the result
  looks reasonable". A dead population produces plausible-looking
  but meaningless final-state numbers.
- **Citing clade without citing the underlying theory**. clade is
  the instrument; the theory (Hamilton, Williams, DeWitt, etc.) is
  the question. Always cite both.
- **Running `pkgdown::build_site()` on CI** before running
  `devtools::check()` locally — the build site rebuilds vignettes
  and will fail in non-obvious ways if your specs have typos.
- **`git add .`** in a clade working directory. There are often large
  cached `.rds` files that you don't want to commit. Always add
  specific files.
